### 2024-08-10 main.tex
Initial Overleaf Import

\documentclass[a4paper]{article}

\usepackage[pages=all, color=black, position={current page.south}, placement=bottom, scale=1, opacity=1, vshift=5mm]{background}
\SetBgContents{This is a basic template to start working on, will be modified to the appropriate journal later on}      % copyright

\usepackage[margin=1in]{geometry} % full-width

% AMS Packages
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}

% Unicode
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\hypersetup{
	unicode,
%	colorlinks,
%	breaklinks,
%	urlcolor=cyan, 
%	linkcolor=blue, 
	pdfauthor={Benjamin Post},
	pdftitle={A simple article template},
	pdfsubject={A simple article template},
	pdfkeywords={article, template, simple},
	pdfproducer={LaTeX},
	pdfcreator={pdflatex}
}

% Vietnamese
%\usepackage{vntex}

% Natbib
\usepackage[sort&compress,numbers,square]{natbib}
\bibliographystyle{mplainnat}

\usepackage{graphicx, color}
\graphicspath{{fig/}}

%\usepackage[linesnumbered,ruled,vlined,commentsnumbered]{algorithm2e} % use algorithm2e for typesetting algorithms
\usepackage{algorithm, algpseudocode} % use algorithm and algorithmicx for typesetting algorithms
\usepackage{mathrsfs} % for \mathscr command

% Author info
\title{Drivers of Phytoplankton Community Variability in the Cariaco Basin, Venezuela}
\author{Benjamin Post$^{1,2}$ \and Esteban Acevedo-Trejos$^3$ \and Subhendu Chakraborty$^1$ \and Andrew Barton$^3$ \and Agostino Merico$^1$}

\date{
	$^1$Systems Ecology Group, Leibniz Centre for Tropical Marine Research (ZMT), Bremen, Germany \\ \\%
        $^2$School of Science, Constructor University, Bremen, Germany \\ \\%
	$^3$Earth Surface Process Modelling, GFZ German Research Centre for Geosciences, Potsdam, Germany\\ \\%
	$^4$Scripps Institution of Oceanography and Department of Ecology, Behavior and Evolution, University of California San Diego, La Jolla, CA, United States \\ \\[2ex]%
%	\today
corresponding author: \texttt{benjaminpost@aoop.de}
}

\begin{document}
    \maketitle

    \begin{abstract}
        this is the abstract

        \noindent\textbf{Keywords:} Phytoplankton, Diversity, Gradient Forest
    \end{abstract}

    \tableofcontents

    \section{Introduction}
    \label{sec:intro}

        \begin{itemize}

            %% Overarching theme: Stability of Ecosystem faunctions, Drivers of community change

            \item "The Cariaco Basin, located ...." - general hydrography/depth, etc.

            \item "The CARIACO Time Series ...." - talk about the Time Series, what they did, why

            \item "The Cariaco Basin has undergone marked shifts..." Talk about ecosystem, what has changed, reference prev literature

            \item "Here's where we come in..." Talk about how most of the analysis stopped after 2012, time series died out, but there is more data there and particularly looking at community data, we can see a return.

            \item "In this paper, we..."

        \end{itemize}

    \section{Methods}
    \label{sec:methods}

        \begin{itemize}

            %%

            \item Talk about the Basin shortly, then about the time series

            %% DATA SOURCES
            \item Niskin Bottle Dataset

            \item CTD Dataset

            \item Phytoplankton Microscopy

            \item HPLC (?)

            \item sediment traps (?)

            \item ERA 5 climate data for wind and precipitation

            \item AMO and MEI v.2 Climate indices

            %% METHODS
            \item Z-Scores calculation

            \item Clustering

            \item NMDS

            \item Gradient Forest

        \end{itemize}

    \section{Results}
    \label{sec:results}

            First of all, we have a map:

        \begin{figure}[ht]
            \centering
            \includegraphics[width=0.7\textwidth]{fig/Artboard 1.pdf}
            \caption{Map of Location of Time Series in the Cariaco Basin, Venezuela.}
            \label{fig:example}
        \end{figure}

            Then we want to show some overall dynamics of the time series:

            \begin{figure}[ht]
            \centering
            \includegraphics[width=0.7\textwidth]{fig/PLOTZScores.pdf}
            \caption{Z Scores over yearly time series..}
            \label{fig:example}
        \end{figure}

             Then we compare the overall dynamics with the community clustering:

            \begin{figure}[ht]
            \centering
            \includegraphics[width=0.7\textwidth]{fig/Screenshot 2024-08-09 at 11.10.58.png}
            \caption{Community Clustering.}
            \label{fig:example}
        \end{figure}

    \section{Discussion}
    \label{sec:discussion}

        \section{Conclusion}
    \label{sec:conclusion}

    \paragraph{Acknowledgements} Thank you to all who have helped, communicated, provided data. Claudia Benitez-Nelson, Digna Rueda-Roa, James Pinckney.

    %	\newpage
    \bibliography{refs}

    \appendix

    \section{Appendix 1}
    \label{app:1}

\end{document}

\textbf
	Unordered List (taken from Overleaf)
	\begin{itemize}
		\item The individual entries are indicated with a black dot, a so-called bullet.
		\item The text in the entries may be of any length.
	\end{itemize}

	Ordered List (taken from Overleaf)
	\begin{enumerate}
		\item The labels consists of sequential numbers.
		\item The numbers starts at 1 with every call to the enumerate environment.
	\end{enumerate}

	\begin{table}[ht]
		\centering
		\begin{tabular}{|c|c|}
			\hline
			\textbf{Odd} & \textbf{Even} \\
			\hline\hline
			One & Two \\
			\hline
			Three & Four \\
			\hline
		\end{tabular}
		\caption{This is a table}
		\label{tbl:1}
	\end{table}

	Table~\ref*{tbl:1} is an example of a table.

	\section{Discussion}
	\label{sec:examples}

	Now we include a figure.
	(See Figure~\ref{fig:example}.)
	\begin{figure}[ht]
		\centering
		\includegraphics[width=0.3\textwidth]{example}
		\caption{An example of a figure}
		\label{fig:example}
	\end{figure}

### 2024-09-26 agujournaltemplate.tex
Updated to AGU Template

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% AGUJournalTemplate.tex: this template file is for articles formatted with LaTeX
%
% This file includes commands and instructions
% given in the order necessary to produce a final output that will
% satisfy AGU requirements, including customized APA reference formatting.
%
% You may copy this file and give it your
% article name, and enter your text.
%
% guidelines and troubleshooting are here:

%% To submit your paper:
\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
%%%%%%%
% As of 2018 we recommend use of the TrackChanges package to mark revisions.
% The trackchanges package adds five new LaTeX commands:
%
%  \note[editor]{The note}
%  \annote[editor]{Text to annotate}{The note}
%  \add[editor]{Text to add}
%  \remove[editor]{Text to remove}
%  \change[editor]{Text to remove}{Text to add}
%
% complete documentation is here: http://trackchanges.sourceforge.net/
%%%%%%%

\draftfalse

%% Enter journal name below.
%% Choose from this list of Journals:
%
% JGR: Atmospheres
% JGR: Biogeosciences
% JGR: Earth Surface
% JGR: Oceans
% JGR: Planets
% JGR: Solid Earth
% JGR: Space Physics
% Global Biogeochemical Cycles
% Geophysical Research Letters
% Paleoceanography and Paleoclimatology
% Radio Science
% Reviews of Geophysics
% Tectonics
% Space Weather
% Water Resources Research
% Geochemistry, Geophysics, Geosystems
% Journal of Advances in Modeling Earth Systems (JAMES)
% Earth's Future
% Earth and Space Science
% Geohealth
%
% ie, \journalname{Water Resources Research}

\journalname{Enter journal name here}

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  TITLE
%
% (A title should be specific, informative, and brief. Use
% abbreviations only if they are defined in the abstract. Titles that
% start with general keywords then specific terms are optimized in
% searches)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Example: \title{This is a test title}

\title{Drivers of Phytoplankton Community Variability in the Cariaco Basin, Venezuela}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  AUTHORS AND AFFILIATIONS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Authors are individuals who have significantly contributed to the
% research and preparation of the article. Group authors are allowed, if
% each author in the group is separately identified in an appendix.)

% List authors by first name or initial followed by last name and
% separated by commas. Use \affil{} to number affiliations, and
% \thanks{} for author notes.
% Additional author notes should be indicated with \thanks{} (for
% example, for current addresses).

% Example: \authors{A. B. Author\affil{1}\thanks{Current address, Antartica}, B. C. Author\affil{2,3}, and D. E.
% Author\affil{3,4}\thanks{Also funded by Monsanto.}}

\authors{B. Post\affil{1,2}, E. Acevedo-Trejos\affil{3}, S. Chakraborty\affil{1}, A.D. Barton\affil{4}, A. Merico\affil{1}}

% \affiliation{1}{First Affiliation}
% \affiliation{2}{Second Affiliation}
% \affiliation{3}{Third Affiliation}
% \affiliation{4}{Fourth Affiliation}

\affiliation{1}{Systems Ecology Group, Leibniz Centre for Tropical Marine Research (ZMT), Bremen, Germany}
\affiliation{2}{School of Science, Constructor University, Bremen, Germany}
\affiliation{3}{Earth Surface Process Modelling, GFZ German Research Centre for Geosciences, Potsdam, Germany}
\affiliation{4}{Scripps Institution of Oceanography and Department of Ecology, Behavior and Evolution, University of California San Diego, La Jolla, CA, United States}
%(repeat as many times as is necessary)

% Corresponding author mailing address and e-mail address:

% (include name and email addresses of the corresponding author.  More
% than one corresponding author is allowed in this LaTeX file and for
% publication; but only one corresponding author is allowed in our
% editorial system.)

% Example: \correspondingauthor{First and Last Name}{email@address.edu}

\correspondingauthor{Benjamin Post}{benjaminpost@aoop.de}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% KEY POINTS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  List up to three key points (at least one is required)
%  Key Points summarize the main points and conclusions of the article
%  Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences

% Example:
% \begin{keypoints}
% \item	List up to three key points (at least one is required)
% \item	Key Points summarize the main points and conclusions of the article
% \item	Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences
% \end{keypoints}

\begin{keypoints}
\item The phytoplankton community data identified by microscopy from the CARIACO Time Series between 1996 and 2016 shows two separate clusters characterized by high and low upwelling regimes respectively.
\item Despite the differing physical regimes and large shifts between functional groups, there is no significant difference in diversity metrics between the two clusters. Overall the data shows a positive relationship between Richness and Biomass, Productivity as well as Resource Use Efficiency.
\item A gradient forest analysis of the community data reveals that the strongest predictor of shifts within the community is the AMO index, which points to a strong impact of large scale climatic oscillations in the tropical coastal ecosystem.
\end{keypoints}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  ABSTRACT and PLAIN LANGUAGE SUMMARY
%
% A good Abstract will begin with a short description of the problem
% being addressed, briefly describe the new data or analyses, then
% briefly states the main conclusion(s) and how they are supported and
% uncertainties.

% The Plain Language Summary should be written for a broad audience,
% including journalists and the science-interested public, that will not have 
% a background in your field.
%
% A Plain Language Summary is required in GRL, JGR: Planets, JGR: Biogeosciences,
% JGR: Oceans, G-Cubed, Reviews of Geophysics, and JAMES.
% see http://sharingscience.agu.org/creating-plain-language-summary/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% \begin{abstract} starts the second page

\begin{abstract}
[ enter your Abstract here ]
\end{abstract}

\section*{Plain Language Summary}
Enter your Plain Language Summary here or delete this section.
Here are instructions on writing a Plain Language Summary: 
https://www.agu.org/Share-and-Advocate/Share/Community/Plain-language-summary

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  BODY TEXT
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Suggested section heads:
% \section{Introduction}
%
% The main text should start with an introduction. Except for short
% manuscripts (such as comments and replies), the text should be divided
% into sections, each with its own heading.

% Headings should be sentence fragments and do not begin with a
% lowercase letter or number. Examples of good headings are:

% \section{Materials and Methods}
% Here is text on Materials and Methods.
%
% \subsection{A descriptive heading about methods}
% More about Methods.
%
% \section{Data} (Or section title might be a descriptive heading about data)
%
% \section{Results} (Or section title might be a descriptive heading about the
% results)
%
% \section{Conclusions}

\section{Draft Results}
%Text here ===>>>
Below here I will put the draft plots to give an overview over the current state of the results.

\begin{figure}
\includegraphics{example.png}
\caption{caption}
\end{figure}

%%

%  Numbered lines in equations:
%  To add line numbers to lines in equations,
%  \begin{linenomath*}
%  \begin{equation}
%  \end{equation}
%  \end{linenomath*}

%% Enter Figures and Tables near as possible to where they are first mentioned:
%
% DO NOT USE \psfrag or \subfigure commands.
%
% Figure captions go below the figure.
% Acronyms used in figure captions will be spelled out in the final, published version.

% Table titles go above tables;  other caption information
%  should be placed in last line of the table, using
% \multicolumn2l{$^a$ This is a table note.}
% NOTE that there is no difference between table caption and table heading in the final, published version
%
%----------------
% EXAMPLE FIGURES
%
% \begin{figure}
% \includegraphics{example.png}
% \caption{caption}
% \end{figure}
%
% Giving latex a width will help it to scale the figure properly. A simple trick is to use \textwidth. Try this if large figures run off the side of the page.
% \begin{figure}
% \noindent\includegraphics[width=\textwidth]{anothersample.png}
%\caption{caption}
%\label{pngfiguresample}
%\end{figure}
%
%
% If you get an error about an unknown bounding box, try specifying the width and height of the figure with the natwidth and natheight options. This is common when trying to add a PDF figure without pdflatex.
% \begin{figure}
% \noindent\includegraphics[natwidth=800px,natheight=600px]{samplefigure.pdf}
%\caption{caption}
%\label{pdffiguresample}
%\end{figure}
%
%
% PDFLatex does not seem to be able to process EPS figures. You may want to try the epstopdf package.
%

%
% ---------------
% EXAMPLE TABLE
%
% \begin{table}
% \caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
% \centering
% \begin{tabular}{l c}
% \hline
%  Run  & Time (min)  \\
% \hline
%   $l1$  & 260   \\
%   $l2$  & 300   \\
%   $l3$  & 340   \\
%   $h1$  & 270   \\
%   $h2$  & 250   \\
%   $h3$  & 380   \\
%   $r1$  & 370   \\
%   $r2$  & 390   \\
% \hline
% \multicolumn{2}{l}{$^{a}$Footnote text here.}
% \end{tabular}
% \end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% SIDEWAYS FIGURES and TABLES
% AGU prefers the use of {sidewaystable} over {landscapetable} as it causes fewer problems.
%
% \begin{sidewaysfigure}
% \includegraphics[width=20pc]{figsamp}
% \caption{caption here}
% \label{newfig}
% \end{sidewaysfigure}
%
%  \begin{sidewaystable}
%  \caption{Caption here}
% \label{tab:signif_gap_clos}
%  \begin{tabular}{ccc}
% one&two&three\\
% four&five&six
%  \end{tabular}
%  \end{sidewaystable}

%% If using numbered lines, please surround equations with \begin{linenomath*}...\end{linenomath*}
%\begin{linenomath*}
%\begin{equation}
%y|{f} \sim g(m, \sigma),
%\end{equation}
%\end{linenomath*}

%%% End of body of article

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Optional Appendices go here
%
% The \appendix command resets counters and redefines section heads
%
% After typing \appendix
%
%\section{Here Is Appendix Title}
% will show
% A: Here Is Appendix Title
%
%\appendix
%\section{Here is a sample appendix}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Optional Glossary, Notation or Acronym section goes here:
%
% Glossary is only allowed in Reviews of Geophysics
%  \begin{glossary}
%  \term{Term}
%   Term Definition here
%  \term{Term}
%   Term Definition here
%  \term{Term}
%   Term Definition here
%  \end{glossary}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Acronyms
%% NOTE that acronyms in the final published version will be spelled out when used in figure captions.
%   \begin{acronyms}
%   \acro{Acronym}
%   Definition here
%   \acro{EMOS}
%   Ensemble model output statistics
%   \acro{ECMWF}
%   Centre for Medium-Range Weather Forecasts
%   \end{acronyms}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Notation
%   \begin{notation}
%   \notation{$a+b$} Notation Definition here
%   \notation{$e=mc^2$}
%   Equation in German-born physicist Albert Einstein's theory of special
%  relativity that showed that the increased relativistic mass ($m$) of a
%  body comes from the energy of motion of the body—that is, its kinetic
%  energy ($E$)—divided by the speed of light squared ($c^2$).
%   \end{notation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% DATA SECTION and ACKNOWLEDGMENTS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Open Research Section}
This section MUST contain a statement that describes where the data supporting the conclusions can be obtained. Data cannot be listed as ''Available from authors'' or stored solely in supporting information. Citations to archived data should be included in your reference list. Wiley will publish it as a separate section on the paper’s page. Examples and complete information are here:
https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors

\section*{As Applicable – Inclusion in Global Research Statement}
The Authorship: Inclusion in Global Research policy aims to promote greater equity and transparency in research collaborations. AGU Publications encourage research collaborations between regions, countries, and communities and expect authors to include their local collaborators as co-authors when they meet the AGU Publications authorship criteria (described here: https://www.agu.org/publications/authors/policies\#authorship). Those who do not meet the criteria should be included in the Acknowledgement section. We encourage researchers to consider recommendations from The TRUST CODE - A Global Code of Conduct for Equitable Research Partnerships (https://www.globalcodeofconduct.org/) when conducting and reporting their research, as applicable, and encourage authors to include a disclosure statement pertaining to the ethical and scientific considerations of their research collaborations in an “Inclusion in Global Research Statement’ as a standalone section in the manuscript following the Conclusions section. This can include disclosure of permits, authorizations, permissions and/or any formal agreements with local communities or other authorities, additional acknowledgements of local help received, and/or description of end-users of the research. You can learn more about the policy in this editorial. Example statements can be found in the following published papers: 
Holt et al. (https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2022JG007188), 
Sánchez-Gutiérrez et al. (https://agupubs.onlinelibrary.wiley.com/doi/abs/10.1029/2023JG007554), 
Tully et al. (https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1029/2022JG007128) 
Please note that these statements are titled as “Global Research Collaboration Statements” from a previous pilot requirement in JGR Biogeosciences. The pilot has ended and statements should now be titled “Inclusion in Global Research Statement”.

\acknowledgments
Enter acknowledgments here. This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% REFERENCES and BIBLIOGRAPHY
%
% \bibliography{<name of your .bib file>} don't specify the file extension
% don't specify bibliographystyle
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\bibliography{ enter your bibtex bibliography filename here }

%Reference citation instructions and examples:
%
% Please use ONLY \cite and \citeA for reference citations.
% \cite for parenthetical references
% ...as shown in recent studies (Simpson et al., 2019)
% \citeA for in-text citations
% ...Simpson et al. (2019) have shown...
%
%
%...as shown by \citeA{jskilby}.
%...as shown by \citeA{lewin76}, \citeA{carson86}, \citeA{bartoldy02}, and \citeA{rinaldi03}.
%...has been shown \cite{jskilbye}.
%...has been shown \cite{lewin76,carson86,bartoldy02,rinaldi03}.
%... \cite <i.e.>[]{lewin76,carson86,bartoldy02,rinaldi03}.
%...has been shown by \cite <e.g.,>[and others]{lewin76}.
%
% apacite uses < > for prenotes and [ ] for postnotes
% DO NOT use other cite commands (e.g., \citet, \citep, \citeyear, \nocite, \citealp, etc.).
%

\end{document}

More Information and Advice:

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  SECTION HEADS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Capitalize the first letter of each word (except for
% prepositions, conjunctions, and articles that are
% three or fewer letters).

% AGU follows standard outline style; therefore, there cannot be a section 1 without
% a section 2, or a section 2.3.1 without a section 2.3.2.
% Please make sure your section numbers are balanced.
% ---------------
% Level 1 head
%
% Use the \section{} command to identify level 1 heads;
% type the appropriate head wording between the curly
% brackets, as shown below.
%
%An example:
%\section{Level 1 Head: Introduction}
%
% ---------------
% Level 2 head
%
% Use the \subsection{} command to identify level 2 heads.
%An example:
%\subsection{Level 2 Head}
%
% ---------------
% Level 3 head
%
% Use the \subsubsection{} command to identify level 3 heads
%An example:
%\subsubsection{Level 3 Head}
%
%---------------
% Level 4 head
%
% Use the \subsubsubsection{} command to identify level 3 heads
% An example:
%\subsubsubsection{Level 4 Head} An example.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  IN-TEXT LISTS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Do not use bulleted lists; enumerated lists are okay.
% \begin{enumerate}
% \item
% \item
% \item
% \end{enumerate}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  EQUATIONS
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Single-line equations are centered.
% Equation arrays will appear left-aligned.

Math coded inside display math mode \[ ...\]
 will not be numbered, e.g.,:
 \[ x^2=y^2 + z^2\]

 Math coded inside \begin{equation} and \end{equation} will
 be automatically numbered, e.g.,:
 \begin{equation}
 x^2=y^2 + z^2
 \end{equation}

% To create multiline equations, use the
% \begin{eqnarray} and \end{eqnarray} environment
% as demonstrated below.
\begin{eqnarray}
  x_{1} & = & (x - x_{0}) \cos \Theta \nonumber \\
        && + (y - y_{0}) \sin \Theta  \nonumber \\
  y_{1} & = & -(x - x_{0}) \sin \Theta \nonumber \\
        && + (y - y_{0}) \cos \Theta.
\end{eqnarray}

%If you don't want an equation number, use the star form:
%\begin{eqnarray*}...\end{eqnarray*}

% Break each line at a sign of operation
% (+, -, etc.) if possible, with the sign of operation
% on the new line.

% Indent second and subsequent lines to align with
% the first character following the equal sign on the
% first line.

% Use an \hspace{} command to insert horizontal space
% into your equation if necessary. Place an appropriate
% unit of measure between the curly braces, e.g.
% \hspace{1in}; you may have to experiment to achieve
% the correct amount of space.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  EQUATION NUMBERING: COUNTER
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% You may change equation numbering by resetting
% the equation counter or by explicitly numbering
% an equation.

% To explicitly number an equation, type \eqnum{}
% (with the desired number between the brackets)
% after the \begin{equation} or \begin{eqnarray}
% command.  The \eqnum{} command will affect only
% the equation it appears with; LaTeX will number
% any equations appearing later in the manuscript
% according to the equation counter.
%

% If you have a multiline equation that needs only
% one equation number, use a \nonumber command in
% front of the double backslashes (\\) as shown in
% the multiline equation above.

% If you are using line numbers, remember to surround
% equations with \begin{linenomath*}...\end{linenomath*}

%  To add line numbers to lines in equations:
%  \begin{linenomath*}
%  \begin{equation}
%  \end{equation}
%  \end{linenomath*}

### 2024-09-26 main.tex
Updated to AGU Template

@@
 
+            Understanding the link between ocean surface and the sinking flux of particulate carbon in the Cariaco Basin.
+                a large (\∼160 km long, 70 km wide) and deep (∼1,400 m) basin, located on the Venezuelan continental shelf
+                bound to the north by a sill at a mean depth of about 100 m
+                productive area (\~400 gC/m²y)
+                Primary production and vertical particulate organic matter fluxes are higher than BATS and HOT locations → anoxic bottom waters!
+
             %% DATA SOURCES
@@
             \item HPLC (?)
-
-            \item sediment traps (?)
 

### 2024-09-26 agujournaltemplate.tex
Added Figures to Draft

@@
 % guidelines and troubleshooting are here:
+
+%% user added for convenience, can be removed before submission:
+\usepackage{comment}
 
@@
 \begin{figure}
-\includegraphics{example.png}
+\noindent\includegraphics[width=\textwidth]{fig/Artboard 1.pdf}
+\caption{caption}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/PLOTZScores.pdf}
+\caption{caption}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/TimeSeriesOverview2.pdf}
+\caption{caption}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/AllCorrelationsPlot.pdf}
+\caption{caption}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot.pdf}
+\caption{caption}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot2.pdf}
 \caption{caption}
@@
 
-\section*{Open Research Section}
-This section MUST contain a statement that describes where the data supporting the conclusions can be obtained.
+% \section*{Open Research Section}
+%This section MUST contain a statement that describes where the data supporting the conclusions can be obtained.
 Data cannot be listed as ''Available from authors'' or stored solely in supporting information.
@@
 Examples and complete information are here:
-https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors
+%https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors
 
-\section*{As Applicable – Inclusion in Global Research Statement}
-The Authorship: Inclusion in Global Research policy aims to promote greater equity and transparency in research collaborations.
+%\section*{As Applicable – Inclusion in Global Research Statement}
+%The Authorship: Inclusion in Global Research policy aims to promote greater equity and transparency in research collaborations.
 AGU Publications encourage research collaborations between regions, countries, and communities and expect authors to include their local collaborators as co-authors when they meet the AGU Publications authorship criteria (described here: https://www.agu.org/publications/authors/policies\#authorship).
@@
 Example statements can be found in the following published papers:
-Holt et al. (https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2022JG007188), 
-Sánchez-Gutiérrez et al. (https://agupubs.onlinelibrary.wiley.com/doi/abs/10.1029/2023JG007554), 
-Tully et al. (https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1029/2022JG007128) 
-Please note that these statements are titled as “Global Research Collaboration Statements” from a previous pilot requirement in JGR Biogeosciences.
+%Holt et al. (https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2022JG007188), 
+%Sánchez-Gutiérrez et al. (https://agupubs.onlinelibrary.wiley.com/doi/abs/10.1029/2023JG007554), 
+%Tully et al. (https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1029/2022JG007128) 
+%Please note that these statements are titled as “Global Research Collaboration Statements” from a previous pilot requirement in JGR Biogeosciences.
 The pilot has ended and statements should now be titled “Inclusion in Global Research Statement”.

### 2024-09-26 agujournaltemplate.tex

@@
 Overall the data shows a positive relationship between Richness and Biomass, Productivity as well as Resource Use Efficiency.
-\item A gradient forest analysis of the community data reveals that the strongest predictor of shifts within the community is the AMO index, which points to a strong impact of large scale climatic oscillations in the tropical coastal ecosystem.
+\item A gradient forest analysis of the community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 4 months, which points to a strong impact of large scale climatic oscillations in the tropical coastal ecosystem.
 \end{keypoints}
@@
 
-\section*{Plain Language Summary}
-Enter your Plain Language Summary here or delete this section.
-Here are instructions on writing a Plain Language Summary: 
-https://www.agu.org/Share-and-Advocate/Share/Community/Plain-language-summary
+%1
+%\section*{Plain Language Summary}
+%Enter your Plain Language Summary here or delete this section.
+%Here are instructions on writing a Plain Language Summary: 
+%https://www.agu.org/Share-and-Advocate/Share/Community/Plain-language-summary
 
@@
 \begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot2.pdf}
+\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3.pdf}
 \caption{caption}
@@
 
+%!
 % \section*{Open Research Section}
@@
 
-\acknowledgments
-Enter acknowledgments here. This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.
+%!
+%\acknowledgments
+%Enter acknowledgments here. This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.
 

### 2024-11-14 agujournaltemplate.tex

--- 2024-09-26 agujournaltemplate.tex
@@
 % guidelines and troubleshooting are here:
+
+%% user added for convenience, can be removed before submission:
 
@@
 \item Despite the differing physical regimes and large shifts between functional groups, there is no significant difference in diversity metrics between the two clusters.
-Overall the data shows a positive relationship between Richness and Biomass, Productivity as well as Resource Use Efficiency.
-\item A gradient forest analysis of the community data reveals that the strongest predictor of shifts within the community is the AMO index, which points to a strong impact of large scale climatic oscillations in the tropical coastal ecosystem.
+Overall, diversity (i.e.
+Genus Richness) shows a positive linear relationship to Biomass, Productivity as well as Resource Use Efficiency.
+\item A gradient forest analysis of the community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 5 months, which points to a strong impact of large scale climatic oscillations in the tropical coastal ecosystem.
+The next strongest predictor is nitrate concentration, with the MEI v2 index (lag of 4 months) and salinity following behind.
 \end{keypoints}
@@
 %%% Suggested section heads:
-% \section{Introduction}
+\section{Introduction}
 %
@@
 
-% \section{Materials and Methods}
+    \begin{itemize}
+
+        %% Overarching theme: Stability of Ecosystem faunctions, Drivers of community change
+
+        \item "The Cariaco Basin, located ...." - general hydrography/depth, etc.
+
+        \item "The CARIACO Time Series ...." - talk about the Time Series, what they did, why
+
+        \item "The Cariaco Basin has undergone marked shifts..." Talk about ecosystem, what has changed, reference prev literature
+
+        \item "Here's where we come in..." Talk about how most of the analysis stopped after 2012, time series died out, but there is more data there and particularly looking at community data, we can see a return.
+
+        \item "In this paper, we..."
+
+    \end{itemize}
+
+\section{Materials and Methods}
 % Here is text on Materials and Methods.
 %
-% \subsection{A descriptive heading about methods}
-% More about Methods.
+\subsection{The CARIACO Time Series}
+  %Talk about the Basin shortly, then about the time series
+    The Cariaco Ocean Time-Series Program was established to investigate the link between ocean surface processes and the sinking flux of particulate carbon.
+The Cariaco Basin is a large ($\sim$ 160 km long, 70 km wide) and deep ($\sim$ 1,400 m) basin, located on the north-eastern Venezuelan continental shelf, bound to the north by a sill at a mean depth of about 100 m.
+It is a relatively productive area ($\sim$ 400 $g C m^{-2} y^{-1}$), with primary production and vertical particulate organic matter fluxes that are higher than BATS and HOT locations, leading to the accumulation of anoxic bottom waters below 200 m depth.
+
+    \subsubsection{Niskin Bottle Dataset}
+        Samples were collected as part of the Cariaco Ocean Time-Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (10.50° N, 64.67° W).
+
+    \subsubsection{CTD Dataset}
+
+    \subsubsection{Phytoplankton Microscopy}
+
+    \subsubsection{ERA 5 climate data for wind}
+
+    \subsubsection{AMO and MEI v.2 Climate indices}
+
+        %% METHODS
+    \subsubsection{Z-Scores calculation}
+
+    \subsubsection{Clustering and NMDS}
+
+        \subsubsection{Gradient Forest}
+
 %
-% \section{Data} (Or section title might be a descriptive heading about data)
+\section{Data} (Or section title might be a descriptive heading about data)
 %
@@
 %
-% \section{Conclusions}
 
@@
 Below here I will put the draft plots to give an overview over the current state of the results.
+As of now all plots are in a draft state, sometimes missing appropriate units.
+Please excuse this, as I would like to first trim down the plots and order them for the final manuscript, before committing further time into finalizing the details.
 
 \begin{figure}
-\includegraphics{example.png}
-\caption{caption}
+\noindent\includegraphics[width=\textwidth]{fig/Artboard 1.pdf}
+\caption{Map of CARIACO Time Series location in the Cariaco basin in Venezuela in the Caribbean Ocean, located at 10.50° N, 64.67° W}
 \end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/PLOTZScores.pdf}
+\caption{Heatmap of yearly average values, normalized for comparison as z-scores.
+All data is available at monthly resolution (see next draft figure for raw data).
+Red indicates a larger value, blue indicates a smaller value for that particular year.
+The years 1995 (n=2) and 2017 (n=1) were removed due to lack of data.
+Diversity Indices are calculated based on microscopy counts at the Genus level.}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\noindent\includegraphics[width=480]{fig/TimeSeriesOverview2.pdf}
+\end{center}
+\caption{Overview over all variables used in this manuscript.
+Red and blue bars are yearly averages above or below mean over the entire time series.
+Black dots are raw or interpolated monthly data.
+The microscopy data is simplified to counts per functional group, to show the shifts within the community.\newline
+\textit{Honestly, I did not find another reasonable way to show a clean and concise overview with matching columns and rows, that shows the raw data.
+This figure could be put into the supplementary or split into separate plots.
+The previous Z-Scores plot shows the same dynamics in a much more concise way.}}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/DivEcoFunctionPlot.pdf}
+\caption{Linear regression of diversity indices to log-transformed Chlorophyll Biomass, Primary Production, Resource Use Efficiency for nitrate ($RUE_N$) and for phosphate ($RUE_P$).
+$RUE$ is calculated as the ratio of Chlorophyll to the available nutrient ($NO_3$ and $PO_4$ respectively).
+All variables show a significant increasing linear trend with Genus Richness, but none for the Shannon Index or Pielou\'s Evenness Index.
+\newline
+\textit{I included this plot to show something about the Diversity to Ecosystem Function relationship at CARIACO.
+This can be left out, if it does not make sense to discuss in the context of this manuscript, or much further simplified, by removing the correlations to Shannon and Pielou, since none of them are significant.
+All correlations to Genus Richness are below p=0.05.
+A more advanced way to quantify these relationships would be mixed linear models, which I could do, if that makes sense in the context of the other results.
+For now, I wanted to keep it simple.}}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot.pdf}
+\caption{(A) Clustering of microscopy data at the Genus level, using the binary Jaccard distance.
+The community for each year is binned together, to smooth out seasonal aspects.
+(B) NMDS plot of monthly community data (cell counts) that is cube root transformed.
+Stress: 0.261, 2 dimensions, converging solution.
+Color of dots based on which yearly cluster the monthly data is in.
+Vector overlay of ENVFIT shows underlying environmental variables that are co-varying with the community data.
+\newline
+Plots (A-L) below show density distributions of the environmental variables as well as diversity indices between the clustered years. Dashed line shows the mean. \newline
+\textit{
+The clustering above is pretty straightforward and shows that we have a return to a previous community, as was noted in the 2019 Review of CARIACO.
+NMDS shows that the yearly clustering seems to relate at least somewhat to the monthly data.
+Density plots below ideally showcase why these "community clusters" exist, since the physical environment is different.
+I have not yet included ANOVA comparisons between the clusters into the plot, however quite a few of these are non-significant.
+Nevertheless, there are obvious differences, particularly towards the extremes.
+Diversity does not seem to significantly differ in the mean, although the variability (i.e.
+deviation from mean) seems to differ.
+}}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/GF_simplified_weightedImp.pdf}
+\caption{Now we move on to the final portion: The Gradient Forest (GF) analysis.
+This is relatively complex, so I tried to show two plots first, that are relatively simplified and where the analysis does not include time lags, and a final plot that includes almost all the available output/data and where time lags were included.
+\newline
+Above here, you can see the result of running the GF algorithm with the environmental and climatic predictor variables at a monthly resolution on the community data at a Genus level.
+Of roughly 100 genus counts put in, the model can predict 43 genus abundances using the predictor variables above.
+GF builds many random forest models for each variable (i.e.
+genus abundance) to predict, then subsets the predictors and variables, checks for correlation, and notes where in the forest of random trees "splits" occur, i.e.
+at what value of a predictor the abundance of a genus changes.
+Accumulating all the predictor accuracies between variables (i.e.
+genus abundances) yields the plot above.}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/GF_simplified_overallCumImp.pdf}
+\caption{This second plot is output of the same run, which shows the cumulative importance along the predictor variables.
+A steep increase means that at this range there were many "splits" in the random forest models, so this is where shifts in the community occur.
+For nitrate for example, the range between 5 and 7.5 \units{µM} seems to have the greatest effect on the community.
+For temperature, lower ranges affect the community most, all of which points to the importance of upwelling in structuring the community.}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3.pdf}
+\caption{\textit{Since I have spent much time playing with the GF analysis and data, this is the result of optimizing the model predictions, by including and testing multiple time lags of variables.} Interestingly only time lags of the climatic indices (AMO and MEIv2) actually showed better predictive capabilities, there were none for lags of wind, nutrients or the isotherm.
+What I also included in the plot above is the underlying mechanism of fitting a random forest model for each genus.
+The plot on the top right shows the weighted importance for each genus, ranked according to the predictive capability of the model, and color coded by their respective group.
+Similarly, the cumulative importance plots below show the overall importance, as well as the individual color lines that show the cumulative importance per genus predicted.
+This is not necessarily useful to show in the final manuscript, unless we want to discuss the model output down to the genus level.
+Not surprisingly, it is mostly diatoms that are predictable, but also other groups are present here.\newline
+\textit{What I find particularly interesting for this set of results, is that the AMO index with a time lab of 5 months is by far the strongest predictor of abundances of genus present.
+The MEIv2 also scores just slightly below nitrate.
+I am not entirely certain how to discuss this, except that the system and community seems to be driven by multi-decadal climatic oscillations.}}
+\end{figure}
+
+\section{Conclusions}
 
@@
 
+%!
 \section*{Open Research Section}
@@
 Example statements can be found in the following published papers:
-Holt et al. (https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2022JG007188), 
-Sánchez-Gutiérrez et al. (https://agupubs.onlinelibrary.wiley.com/doi/abs/10.1029/2023JG007554), 
-Tully et al. (https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1029/2022JG007128) 
-Please note that these statements are titled as “Global Research Collaboration Statements” from a previous pilot requirement in JGR Biogeosciences.
+%Holt et al. (https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2022JG007188), 
+%Sánchez-Gutiérrez et al. (https://agupubs.onlinelibrary.wiley.com/doi/abs/10.1029/2023JG007554), 
+%Tully et al. (https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1029/2022JG007128) 
+%Please note that these statements are titled as “Global Research Collaboration Statements” from a previous pilot requirement in JGR Biogeosciences.
 The pilot has ended and statements should now be titled “Inclusion in Global Research Statement”.
 
+%!
 \acknowledgments
-Enter acknowledgments here. This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.
+Thank you to all who have helped, communicated, provided data. Claudia Benitez-Nelson, Digna Rueda-Roa, James Pinckney.
+
+%Enter acknowledgments here. This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.
 
@@
 
-%\bibliography{ enter your bibtex bibliography filename here }
+\bibliography{references}
 

### 2024-11-14 agujournaltemplate.tex
Updates from Overleaf

@@
 Below here I will put the draft plots to give an overview over the current state of the results.
-As of now all plots are in a draft state, sometimes missing appropriate units.
+As of now, all plots are in a draft state, sometimes missing appropriate units.
 Please excuse this, as I would like to first trim down the plots and order them for the final manuscript, before committing further time into finalizing the details.
@@
 \begin{center}
-\noindent\includegraphics[width=480]{fig/TimeSeriesOverview2.pdf}
+\noindent\includegraphics[width=480]{fig/DepthDivTimeSeries.pdf}
 \end{center}
-\caption{Overview over all variables used in this manuscript.
-Red and blue bars are yearly averages above or below mean over the entire time series.
-Black dots are raw or interpolated monthly data.
-The microscopy data is simplified to counts per functional group, to show the shifts within the community.\newline
-\textit{Honestly, I did not find another reasonable way to show a clean and concise overview with matching columns and rows, that shows the raw data.
-This figure could be put into the supplementary or split into separate plots.
-The previous Z-Scores plot shows the same dynamics in a much more concise way.}}
+\caption{Overview over Chlorophyll and Diversity Variables over Depth}
 \end{figure}
@@
 \begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/DivEcoFunctionPlot.pdf}
+\noindent\includegraphics[width=\textwidth]{fig/BDEF_depth_PLOT.pdf}
 \caption{Linear regression of diversity indices to log-transformed Chlorophyll Biomass, Primary Production, Resource Use Efficiency for nitrate ($RUE_N$) and for phosphate ($RUE_P$).
@@
 \begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot.pdf}
+\noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot_NEW.pdf}
 \caption{(A) Clustering of microscopy data at the Genus level, using the binary Jaccard distance.
@@
 \newline
-Plots (A-L) below show density distributions of the environmental variables as well as diversity indices between the clustered years. Dashed line shows the mean. \newline
-\textit{
-The clustering above is pretty straightforward and shows that we have a return to a previous community, as was noted in the 2019 Review of CARIACO.
-NMDS shows that the yearly clustering seems to relate at least somewhat to the monthly data.
-Density plots below ideally showcase why these "community clusters" exist, since the physical environment is different.
-I have not yet included ANOVA comparisons between the clusters into the plot, however quite a few of these are non-significant.
-Nevertheless, there are obvious differences, particularly towards the extremes.
-Diversity does not seem to significantly differ in the mean, although the variability (i.e.
-deviation from mean) seems to differ.
-}}
-\end{figure}
-
-\begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/GF_simplified_weightedImp.pdf}
-\caption{Now we move on to the final portion: The Gradient Forest (GF) analysis.
-This is relatively complex, so I tried to show two plots first, that are relatively simplified and where the analysis does not include time lags, and a final plot that includes almost all the available output/data and where time lags were included.
-\newline
-Above here, you can see the result of running the GF algorithm with the environmental and climatic predictor variables at a monthly resolution on the community data at a Genus level.
-Of roughly 100 genus counts put in, the model can predict 43 genus abundances using the predictor variables above.
-GF builds many random forest models for each variable (i.e.
-genus abundance) to predict, then subsets the predictors and variables, checks for correlation, and notes where in the forest of random trees "splits" occur, i.e.
-at what value of a predictor the abundance of a genus changes.
-Accumulating all the predictor accuracies between variables (i.e.
-genus abundances) yields the plot above.}
-\end{figure}
-
-\begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/GF_simplified_overallCumImp.pdf}
-\caption{This second plot is output of the same run, which shows the cumulative importance along the predictor variables.
-A steep increase means that at this range there were many "splits" in the random forest models, so this is where shifts in the community occur.
-For nitrate for example, the range between 5 and 7.5 \units{µM} seems to have the greatest effect on the community.
-For temperature, lower ranges affect the community most, all of which points to the importance of upwelling in structuring the community.}
+Plots (A-L) below show density distributions of the environmental variables as well as diversity indices between the clustered years. Dashed line shows the mean.}
 \end{figure}
@@
 \noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3.pdf}
-\caption{\textit{Since I have spent much time playing with the GF analysis and data, this is the result of optimizing the model predictions, by including and testing multiple time lags of variables.} Interestingly only time lags of the climatic indices (AMO and MEIv2) actually showed better predictive capabilities, there were none for lags of wind, nutrients or the isotherm.
+\caption{Interestingly only time lags of the climatic indices (AMO and MEIv2) actually showed better predictive capabilities, there were none for lags of wind, nutrients or the isotherm.
 What I also included in the plot above is the underlying mechanism of fitting a random forest model for each genus.
@@
 This is not necessarily useful to show in the final manuscript, unless we want to discuss the model output down to the genus level.
-Not surprisingly, it is mostly diatoms that are predictable, but also other groups are present here.\newline
-\textit{What I find particularly interesting for this set of results, is that the AMO index with a time lab of 5 months is by far the strongest predictor of abundances of genus present.
-The MEIv2 also scores just slightly below nitrate.
-I am not entirely certain how to discuss this, except that the system and community seems to be driven by multi-decadal climatic oscillations.}}
+Not surprisingly, it is mostly diatoms that are predictable, but also other groups are present here.}
 \end{figure}

### 2024-11-19 agujournaltemplate.tex
Start working on Methods

@@
 
-\title{Drivers of Phytoplankton Community Variability in the Cariaco Basin, Venezuela}
+\title{Bottum-up drivers of phytoplankton community diversity in the Cariaco Basin, Venezuela}
 
@@
 
+        % below is from 2019 REVIEW
+        Hydrographic observations (Figure 2) show strong seasonality and interannual variability in the upper 400 m in the Cariaco Basin.
+This is controlled by seasonal and interannual changes in the wind and in the intensity of the geostrophic Caribbean Current (Muller-Karger et al.
+1989).
+Primary upwelling was strongest between November and May.
+In general, during upwelling, sea surface temperatures ranged between approximately 23◦Cand25◦C.
+The secondary upwelling that occurs between June and August is shorter (∼5 weeks) and is approximately 1.5\°C warmer than the primary upwelling.
+It is driven by Ekman transport and an intensification of the curl of the wind close to the southern coast of the Caribbean (Rueda-Roa \& Muller-Karger 2013, Rueda-Roa et al.
+2018).
+Upwelled waters move offshore in plumes that extended as far as 50 km to more than 250 km from the coast (Muller-Karger et al.
+1989, 2010).
+From July to October, there is strong stratification in the upper 50 m of the water column after the wind weakens and the rain and rivers deliver less saline waters (Figure 2).
+        Sea surface temperatures ranged from approximately 28◦C to 30◦C during the period of stratification (Figure 2).
+The mixed-layer depth varied from less than 10 m to 35 m during the 21-year time series (Figure 3).
+Satellite sea surface temperature and ocean color observations show that upwelling pulsates throughout the year at the scale of days to weeks, leading to marked variation in primary production (Figure 3).
+CARIACO rigorously characterized the annual seasonal variation in primary productivity and stratification.
+The length of the time series was key in capturing interannual variations that spanned decades.
+Satellite sea surface temperature and field data show that upwelling was stronger between 1996 and 2002.
+After 2002, trade wind intensity became weaker (Taylor et al.
+2012).
+        The field observations, particularly those in the upper 50 m, showed what appeared to be a return to stronger upwelling conditions starting in 2014 (Figure 2a).
+
         \item "Here's where we come in..." Talk about how most of the analysis stopped after 2012, time series died out, but there is more data there and particularly looking at community data, we can see a return.
@@
 %
-\subsection{The CARIACO Time Series}
+\subsection{Study Site}
   %Talk about the Basin shortly, then about the time series
@@
 
-    \subsubsection{Niskin Bottle Dataset}
+    \subsubsection{In-Situ Data}
     Samples were collected as part of the Cariaco Ocean Time-Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (10.50° N, 64.67° W).
+% https://imars.usf.edu/?page_id=1027 (10°30’N, 64°40’W) ???
 
-    \subsubsection{CTD Dataset}
+    % from 2019 Review:
+    The CARIACO field program consisted of a monthly core oceanographic cruise, two sediment trap deployment/recovery cruises per year, and two microbiogeochemical cruises targeting microbial processes across the oxic–anoxic interface per year.
+Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela.
+Details of the history of sampling procedures in CARIACO as well as sampling schemes, analytical methods, and data quality control and intercalibration procedures have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.
 
-    \subsubsection{Phytoplankton Microscopy}
+    % below paragraph is from Taylor et al. 2012
+    % I should similarly reference the other more comprehensive data protocols like they do
+    % but also explain the methodology quickly, like they did
+    Sampling and processing protocols for hydrography, NPP, Chla, phaeo, nutrient concentrations, and sediment trap deployment and processing appear in Muller-Karger et al.
+(8).
+These data are available from www.imars.
+usf.edu/CAR, the National Oceanographic Data Center (NODC) (www.nodc.
+noaa.gov), and the Biological and Chemical Oceanography Data Management Office (BCO-DMO) (http://bco-dmo.org/data).
+From late 1995 to early 1998, nutrient data were provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and by Dr.
+K.
+Fanning (University of South Florida, St.
+Petersburg, FL) from mid-1997 to the present.
+Samples from eight cruises were analyzed by both laboratories; PO43− and Si[OH]4 data agreed very well ($±$5\%), but NOx and NH4+ did not.
+Therefore, the entire PO43− and Si[OH]4 dataset and only the USF NOx and NH4+ dataset were used here.
 
-    \subsubsection{ERA 5 climate data for wind}
+    Phytoplankton enumeration was conducted on water samples (500 mL) collected by Niskin bottles at depths of 1, 7, 15, 25, 35, 55 m and fixed with 5\% sodium tetraborate-buffered formaldehyde (final concentration).
+Quantitative analyses were performed using settling chambers and an inverted microscope according the Utermöhl method (38).
 
-    \subsubsection{AMO and MEI v.2 Climate indices}
+    % The below is from Pinckney et al. 2015
+    % They have more detail about Phytoplankton, either use this, or reference them
+    Samples were collected as part of the Cariaco Ocean Time-Series Program at a station located in the eastern Cariaco Basin (101300N and 641400W) between November 1995 and January 2017.
+All samples were collected using a 12 bottle rosette system equipped with a CTD (SeaBird, model SBE 25) (Astor and Varela, 2013).
+Water samples for chl a and phaeopigment analysis were obtained on a monthly basis from the first cast of each of cruise.
+Samples were collected at 8 standard depths in the upper 100 m (1, 7, 15, 25, 35, 55, 75, and 100 m) following Varela (2013).
+For phytoplankton species determination by microscopy, 500 ml of seawater were collected at the same depths into HDPE bottles and fixed with a 4\% formalin solution neutralized with sodium tetraborate.
+Samples were analyzed at the Universidad de Oriente, Venezuela, using the Utermöhl technique (Hasle, 1978) with 100 ml sedimentation chambers and a settling period of 48 h.
+Identifications and enumerations by microscopy were limited to microphytoplankton (cells $>$20 mm) (Mutshinda et al., 2013).
 
-        %% METHODS
-    \subsubsection{Z-Scores calculation}
+    % The below is from Taylor et al. 2012 again
+    % I can use as template to introduce ERA5, MEIv2 etc.
+    To gauge ITCZ changes relevant to Cariaco Basin data, we developed ITCZ indices over South America (20°S to 5°N and 290°W to 320°W) and the Tropical Atlantic (15°S to 20°N and 300°W to 357.5°W) as area-weighted precipitation and convergence, using the precipitation intensity and negative divergence of horizontal winds, respectively.
+We calculate ITCZ indices using the previously described center of action (COA) approach (39).
+The NCEP Reanalysis (40) was the source of data for NAO (www.cgd.ucar.edu/cas/ jhurrell/ indices.html), Azores High, and ITCZ analyses.
+Precipitation intensities were obtained from Global Precipitation Climatology Project (41).
+In all cases except the NAO, derived data were obtained from the National Oceanic and Atmospheric Administration/Office of Oceanic and Atmospheric Research/Earth System Research Laboratory Physical Sciences Division (NOAA/ OAR/ESRL PSD) (www.esrl.noaa.gov/psd).
+
+    \subsubsection{Climate data}
+    %from Taylor et al. 2012
+    Zonal wind stress (−u, easterly Trade Winds) across the southern Caribbean Sea drives near-shore upwelling of nutrient-rich waters by Ekman transport.
+We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis at the grid cell closest to the Cariaco Time Series sampling coordinates (10.5°N,64.8°W ?double check?), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024)
+
+    The monthly AMO index data (10 year lowpass), defined as an area average of detrended low-pass filtered North Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} (accessed in June, 2024).
+
+    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of the surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (30°S-30°N and 100°E-70°W) were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024).
+In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).
+
+    \subsubsection{Data processing}
+    % Niskin, CTD and Phytoplankton data processing
+    - quality control, removed outliers, interpolation, averaging
+    All done in R
+
+    \subsection{Data analysis}
+
+    \subsubsection{Z-Scores}
+    % from Fraker et al. 2022 - Lake Erie paper
+    To visualize how ecosystem components changed through time and to identify similarities in temporal trends among variables, each standardized data point (i.e., annual value) was color-coded based on whether its standardized score was greater or less than the long-term mean (negative anomaly: red; positive anomaly: blue; deeper shades indicating greater values).
+The variables were then ordered based on their loading onto PCA axis 1 (R package ‘ade4’; Dray and Dufour, 2007; R version 3.6.3; RCoreTeam, 2020) to indicate their degree of change over the time interval and their similarity with one another.
+Such visualizations have been shown to provide a useful portrayal of ecosystem change (e.g., Choi et al., 2005; Vasilakopoulos et al., 2017; Harvey et al., 2020).
+
+    \subsubsection{BDEF and Diversity}
+    % from Ptacnik et al. 2008
+    Diversity was defined as the number of genera present in a given sample.
+Reducing the taxonomic resolution to genus level made the dataset more homogeneous and more robust to differences in skill and effort among the individual phytoplankton analysts.
+
+    % Regression analysis of RUE
+    To analyze the variability of phytoplankton diversity across temporal and depth ranges, we interpolated the discrete depth counts and divided the resulting depth profile by 4 equal depth intervals (0-25, 25-50, 50-75 and 75-100 m).
+
+    RUE: We excluded observations with TP concentrations 2 gliter1 (n  15) because they are close to the detection limit, resulting in very imprecise measurements.
+    Phytoplankton biomass divided by the limiting resource (total phosphorus) gave the yield or resource use efficiency (RUE) for the community.
+We estimated RUE both in terms of chlorophyll-a (RUEchl = chlorophyll-a:TP) and carbon biomass (RUEcarb  PPC:TP)
 
     \subsubsection{Clustering and NMDS}
+    % from Winder and Hunter 2008
+    Non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering analysis were used to extract gradients in phytoplankton community composition over time based on the Sørenson (Bray–Curtis) distance matrix.
+NMDS is a powerful tool for ecological community data that include a large portion of zero values because it is not biased by the ‘‘joint absences’’ effect (McCune et al.
+2002) and outperformed linear models (such as principal components analysis or redundancy analysis).
+NMDS is an iterative search for the positions of species and time on few dimensions (axes) that minimizes the departure from monotonicity in the association between distance (dissimilarity) in the original data and ordination space.
+It plots dissimilar objects far apart in the ordination space and similar objects close to one another.
+The temporal dynamics of the phytoplankton community was characterized based on biovolume averages to reflect the true contribution of individual species to phytoplankton community composition (Interlandi and Kilham 2001).
+Temporal change of phytoplankton genera is conducted on a matrix of annual averages (year) by genera biovolume to extract years with similar phytoplankton structure.
+Data were transformed by square root transformation and double standardized (species by maximum species biovolume and dates to unit totals).
+NMDS is started from random starting configurations and the solution with the lowest ‘‘stress’’ was selected from 500 runs.
+Stress measures departure of fit between the ordination and the data and is calculated as a departure from monotonicity between the two series of distances.
+A stable three-dimensional solution was identified for all ordinations.
+Orthogonal rotation of the NMDS solution was used to maximize correlations of axes so that the variance of points is maximized on first dimension.
+Environmental variables were related to the strongest gradients (axes) in species composition and, using overlays, in the ordination space towards which the environmental vectors change most rapidly and to which they have maximal correlations with the ordination configuration.
+Linear regression was used to relate the environmental data of nutrient concentration, optical and physical (mixing) properties, as well as zooplankton (crustacean) density to the NMDS axis scores (Jovan and McCune 2005).
+Multivariate analysis was conducted using the vegan package in R.
+    % could add indicator species per cluster, as they have done
 
     \subsubsection{Gradient Forest}
+    % from Fraker et al. 2022 - Lake Erie paper
+    gradient forest analysis was used to identify and compare among the drivers that predicted the temporal trends within each of the indicator datasets and time intervals examined.
+Gradient forest analysis is a community-level extension of random forest analysis that aggregates regression trees to quantify turnover in indicator variables across driver (or pressure) variables (Pitcher et al., 2012; Large et al., 2015; Tam et al., 2017).
+It partitions a goodness-of-fit statistic among each driver variable for each indicator variable, then averages over indicators to produce an overall importance for that driver (R2P).
+The split values of the regression trees also identify thresholds for the driver variables at which the chosen indicator variables exhibit large changes in value (e.g., community turnover in the fish, bird, and zooplankton abundance indicator datasets).
+Plots of the cumulative predictive importance of each driver variable can suggest target values for management of key drivers of ecosystem condition (i.e., where thresholds occur; Tam et al., 2017).
+These analyses were conducted for abundances of each taxa and the fish-derived ecosystem indicators within each time interval (the long, intermediate, and short-term intervals, as well as each 20-year interval), except for zooplankton during the 1969–2018, 1969–1988, and 1979–1998 intervals due to lack of monitoring data during these times.
+These analyses were conducted using R package ‘gradientforest’ with 500 trees, a maximum of three splits, and correlation threshold of 0.5; \cite{ellis_gradient_2012}.
 
@@
 %
+
+- Phytoplankton.csv - https://www.bco-dmo.org/dataset/3095
+	last accessed Jun 2024 (2.6.2024)
+	Troccoli, L., Diaz-Ramos, R., Subero-Pino, S., Muller-Karger, F., Astor, Y., Varela, R., Rueda-Roa, D., Klein, E.
+(2019) Time-Series of Phytoplankton Taxonomy and Density collected by the CARIACO Ocean Time-Series Project from November 1995 to January 2017.
+Biological and Chemical Oceanography Data Management Office (BCO-DMO).
+(Version 1) Version Date 2019-08-14.
+doi:10.1575/1912/bco-dmo.3095.1 [last accessed June 2024]
+
+- Niskin.csv - https://www.bco-dmo.org/dataset/3093
+	last accessed Jun 2024 (2.6.2024)
+	Muller-Karger, F., Astor, Y., Scranton, M., Taylor, G., Thunell, R., Varela, R., Benitez-Nelson, C., Buck, K., Fanning, K., Capelo, J., Gutierrez, J., Guzman, L., Lorenzoni, L., Montes, E., Rojas, J., Rondon, A., Rueda-Roa, D., Tappa, E.
+(2019) Time-series Niskin-bottle sample data from R/V Hermano Gines cruises in the Cariaco Basin from 1995 through 2017 (CARIACO Ocean Time-Series Program).
+Biological and Chemical Oceanography Data Management Office (BCO-DMO).
+(Version 1) Version Date 2019-06-06.
+doi:10.1575/1912/bco-dmo.3093.1 [last accessed June 2024]
+
+- CTD - https://www.bco-dmo.org/dataset/3092
+	Muller-Karger, F., Astor, Y., Benitez-Nelson, C., Scranton, M., Taylor, G., Thunell, R., Varela, R., Capelo, J., Guzman, L., Lorenzoni, L., Montes, E., Rojas, J., Rueda-Roa, D.
+(2019) Time series composite CTD profiles from R/V Hermano Ginés cruises in the Cariaco Basin from 1995 through 2017 (CARIACO Ocean Time-Series Program).
+Biological and Chemical Oceanography Data Management Office (BCO-DMO).
+(Version 1) Version Date 2019-06-04 [if applicable, indicate subset used].
+doi:10.1575/1912/bco-dmo.3092.1 [access date]
+
+- ERA5 - https://cds.climate.copernicus.eu/cdsapp#!/dataset/10.24381/cds.f17050d7
+	Hersbach, H., Bell, B., Berrisford, P., Biavati, G., Horányi, A., Muñoz Sabater, J., Nicolas, J., Peubey, C., Radu, R., Rozum, I., Schepers, D., Simmons, A., Soci, C., Dee, D., Thépaut, J-N.
+(2023): ERA5 monthly averaged data on single levels from 1940 to present.
+Copernicus Climate Change Service (C3S) Climate Data Store (CDS), DOI: 10.24381/cds.f17050d7 (Accessed on 02-06-2024)
+
+- AMO - https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo 
+	Trenberth, Kevin &, Zhang, Rong & National Center for Atmospheric Research Staff (Eds).
+Last modified 2023-09-01 "The Climate Data Guide: Atlantic Multi-decadal Oscillation (AMO) and Atlantic Multidecadal Variability (AMV).” Retrieved from [https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo](https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo) on 2024-06-06.
+- MEI v.2 - https://psl.noaa.gov/enso/mei/
 
@@
 $RUE$ is calculated as the ratio of Chlorophyll to the available nutrient ($NO_3$ and $PO_4$ respectively).
-All variables show a significant increasing linear trend with Genus Richness, but none for the Shannon Index or Pielou\'s Evenness Index.
-\newline
-\textit{I included this plot to show something about the Diversity to Ecosystem Function relationship at CARIACO.
-This can be left out, if it does not make sense to discuss in the context of this manuscript, or much further simplified, by removing the correlations to Shannon and Pielou, since none of them are significant.
-All correlations to Genus Richness are below p=0.05.
-A more advanced way to quantify these relationships would be mixed linear models, which I could do, if that makes sense in the context of the other results.
-For now, I wanted to keep it simple.}}
+All variables show a significant increasing linear trend with Genus Richness, but none for the Shannon Index or Pielou\'s Evenness Index.}
 \end{figure}
@@
 \end{figure}
+
+\section{Discussion}
+This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem, that is driven by local environmental variability, which is in turn affected by large scale climatic shifts.
+In the Cariaco basin, long-term climatic oscillations have affected the upwelling regime and ...
+In the Cariaco basin, wind-driven seasonal upwelling dictates the nutrient availability in the upper mixed layer, which in turn affects the phytoplankton dynamics \cite{ptacnik_diversity_2008, bopp_response_2005}.
+In particular, Diatoms are responsive to changes in the mixing regime \cite{huisman_reduced_2006}.
+Alterations in physical mixing affected phytoplankton dynamics,
+% MEI index relevant (from 2019 REVIEW)
+The Caribbean experiences the most marked of the El Niño–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Niño–Southern Oscillation (Enfield & Mayer 1997).
+
+% REALLY ONLY LOOK AT BOTTOM UP HERE; No Top Down effects quantified
 
@@
 %
-% \bibliography{<name of your .bib file>} don't specify the file extension
+%\bibliography{<name of your .bib file>} don't specify the file extension
 % don't specify bibliographystyle

### 2024-11-21 agujournaltemplate.tex
first draft of methods section

--- 2024-11-14 agujournaltemplate.tex
@@
         \item "In this paper, we..."
+
+        %Highlight that I am only looking at Phytoplankton Microscopy data, and a bit of Niskin, but nothing else. Only bottom-up drivers.
 
@@
 %
-\subsection{The CARIACO Time Series}
+\subsection{Study Site}
     %Talk about the Basin shortly, then about the time series
-    The Cariaco Ocean Time-Series Program was established to investigate the link between ocean surface processes and the sinking flux of particulate carbon.
-The Cariaco Basin is a large ($\sim$ 160 km long, 70 km wide) and deep ($\sim$ 1,400 m) basin, located on the north-eastern Venezuelan continental shelf, bound to the north by a sill at a mean depth of about 100 m.
-It is a relatively productive area ($\sim$ 400 $g C m^{-2} y^{-1}$), with primary production and vertical particulate organic matter fluxes that are higher than BATS and HOT locations, leading to the accumulation of anoxic bottom waters below 200 m depth.
+    The Cariaco Ocean Time-Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon.
+The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (about 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around 100 m.
+This area is relatively productive ($\sim$ 400 $g C m^{-2} y^{-1}$), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below 200 m depth.
 
-    \subsubsection{Niskin Bottle Dataset}
-        Samples were collected as part of the Cariaco Ocean Time-Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (10.50° N, 64.67° W).
+\subsection{In-Situ Data}
+    Samples were collected as part of the Cariaco Ocean Time-Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (10.50° N, 64.66° W) \cite{muller-karger_scientific_2019}.
+    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this manuscript.
+Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela.
+Details of the history of sampling procedures in CARIACO as well as sampling schemes, analytical methods, and data quality control and intercalibration procedures have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.
 
-    \subsubsection{CTD Dataset}
+    All samples were collected using a 12-bottle rosette system equipped with a CTD (SeaBird, model SBE 25) \cite{astor_yrene_m_handbook_2013}.
+Water samples for chlorophyll a and nutrient analysis were taken monthly during the first cast of each cruise.
+These samples were collected at 8 standard depths within the upper 100 meters (1, 7, 15, 25, 35, 55, 75, and 100 m).
 
-    \subsubsection{Phytoplankton Microscopy}
+    Between late 1995 and early 1998, nutrient data were supplied by William Senior (Universidad de Oriente, Cumana, Venezuela), and from mid-1997 to the present, by Dr.
+K.
+Fanning (University of South Florida, St.
+Petersburg, FL).
+Samples from eight cruises were analyzed by both laboratories.
+The data for $PO_4$ and $SiO_4$ were in strong agreement (within ±5\%), whereas the $NO_x$ and $NH_4$ data showed discrepancies.
+\cite{taylor_ecosystem_2012}.
+For the overview of the entire time-series, we used a merged dataset for all nutrients, where coincidental monthly measurements were averaged.
+For the analysis of Resource Use Efficiency, only the USF $NO_3$ measurements were used.
 
-    \subsubsection{ERA 5 climate data for wind}
+    For phytoplankton species identification by microscopy, 500 mL seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate.
+Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using 100 mL sedimentation chambers with a 48-hour settling period.
+Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$20 μm) \cite{mutshinda_environmental_2013}.
+    %from Mutshinda et al. 2013
+    To quality control the microscopy data, we assembled the taxonomic identifications made across all cruises, removed observations that were not identified to a species or genus level, corrected spelling variations, and grouped synonyms using taxonomic information in the World Register of Marine Species (\url{marinespecies.org}).
+After processing the data, there were 482 phytoplankton species observed over the course of the time series, which group into 223 phytoplankton genera.
 
-    \subsubsection{AMO and MEI v.2 Climate indices}
+    The datasets for Niskin bottle samples, CTD and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024).
 
-        %% METHODS
-    \subsubsection{Z-Scores calculation}
+    From the CARIACO Time Series data, we used only those variables, that cover the entire length of the time series without large gaps in coverage.
+Therefore, we did not include Zooplankton net tow and HPLC pigment data in our analysis.
+
+\subsection{Climate data}
+    %from Taylor et al. 2012
+    Zonal wind stress (−u, easterly Trade Winds) across the southern Caribbean Sea drives near-shore upwelling of nutrient-rich waters by Ekman transport \cite{taylor_ecosystem_2012}.
+We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis at the grid cell closest to the Cariaco Time Series sampling coordinates (10.4°N,64.8°W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024)
+
+    The monthly AMO index data (10 year lowpass), which is as an area average of detrended low-pass filtered North-Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} (accessed in June, 2024).
+
+    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of the surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (30°S-30°N and 100°E-70°W), were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024).
+In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).
+
+    \subsubsection{Data processing}
+    % Niskin, CTD and Phytoplankton data processing
+    Data was processed and analyzed using the R programming langauge (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with the tidyverse packages \cite{wickham_welcome_2019}.
+All data processing and analysis code is publically available on Github \url{https://github.com/ben1post/cariaco-ms2}.
+%Note: Repo currently still set to private
+    For all depth discrete measurements, the data was interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm used by the U.S.
+National Oceanographic Data Center \cite{johnson2006world}.
+The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis.
+For the phytoplankton microscopy data, the individual identified taxonomic units (Species or Genus) were interpolated, integrated and then grouped by Genus via the sum of interpolated counts.
+
+    \subsection{Data analysis}    
+    To give a general overview of the data and trends over the time series, we calculated yearly means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$).
+The cruises from 1995 and 2017 were excluded, as there were too few measurements to provide a reliable yearly mean estimate.
+
+    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to make the dataset more homogeneous and more robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008}.
+Additionally, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\'s evenness $J' = -\sum p_i ln( p_i )/ln(H')$ based on the genera present for each interpolated and integrated monthly sampling.
+
+    %RUE/BDEF ANALYSIS
+    To investigates the effect of diversity on ecosystem functioning, we compared measurements of chlorophyll a, primary productivity and Resource Use Efficiency for Nitrate ($RUE_N$) against Genus Richness, Shannon Index and Pielou\'s Index, each integrated at the four depth intervals across each monthly sampling.
+$RUE_N$ is calculated as the ration between the limiting resource ($NO_3$) and phytoplankton biomass (chlorophyll a).
+    We excluded observations with $NO_3$ concentrations below 0.01 µM (n = 36) because they are close to the detection limit, resulting in imprecise measurements \cite{ptacnik_diversity_2008}.
+    To quantify the trend in the relationships, we use linear regression models using the "lm" function \cite{r_core_team_r_2024}.
+    %here I want to extend to glmm that correct for depth and seasonality for the final results
 
     \subsubsection{Clustering and NMDS}
+    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of yearly aggregates of genera observed integrated over 100 m depth.
+Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}.
+    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}.
+%Here I want to add "significant diferences calculation" using e.g.
+ANOVA.
+    % detailed explanation of NMDS in Winder and Hunter 2008
 
     \subsubsection{Gradient Forest}
-
-%
-\section{Data} (Or section title might be a descriptive heading about data)
-%
-% \section{Results} (Or section title might be a descriptive heading about the
-% results)
-%
+    To quantify the effects of bottom-up drivers in the Cariaco basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis.
+Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables\cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}.
+It calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor.
+The regression tree split values also identify ranges of the predictors where significant changes occur in indicator variables, which signal a threshold for community turnover in the phytoplankton community.
+Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}.
+These analyses were conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data.
+The drivers/predictor variables considered in the analysis were the AMO index, MEI v.2 index, -u 10m wind component, the 21°C Isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$).
+The environmental variables from ocean time series tend to be strongly correlated, which Gradient Forest methodology takes into account by implementing a conditional permutation of correlated predictors \cite{ellis_gradient_2012}.
+To take into account the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all variables in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run.
+% see figure in supplementary with initial runs TODO
 
 \section{Draft Results}
+
+%from Mutshinda et al. 2013
+Phytoplankton species richness varied from 4 to 100 species per cruise, with a mean of 40 over the CARIACO time series (167 cruises from November 1995 to March 2011).
+Over a third of the species (187 of 525) were only observed during one or two cruises in the entire study.
+
 %Text here ===>>>

### 2024-11-28 agujournaltemplate.tex
pre update plots

@@
     As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to make the dataset more homogeneous and more robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008}.
-Additionally, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\'s evenness $J' = -\sum p_i ln( p_i )/ln(H')$ based on the genera present for each interpolated and integrated monthly sampling.
+Additionally, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
 
     %RUE/BDEF ANALYSIS
-    To investigates the effect of diversity on ecosystem functioning, we compared measurements of chlorophyll a, primary productivity and Resource Use Efficiency for Nitrate ($RUE_N$) against Genus Richness, Shannon Index and Pielou\'s Index, each integrated at the four depth intervals across each monthly sampling.
+    To investigates the effect of diversity on ecosystem functioning, we compared measurements of chlorophyll a, primary productivity and Resource Use Efficiency for Nitrate ($RUE_N$) against Genus Richness, Shannon Index and Pielou\textquotesingle s Index, each integrated at the four depth intervals across each monthly sampling.
 $RUE_N$ is calculated as the ration between the limiting resource ($NO_3$) and phytoplankton biomass (chlorophyll a).
@@
 Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}.
+We used the Bray distance matrix in a 2-dimensional ordination, that reached a converging solution within 30 steps.
+To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations.
+The arrows indicate the gradient of a 2-dimensional linear trend surface.
     To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}.
@@
 \section{Draft Results}
+Our study focused on the phytoplankton microscopy data from the CARIACO Time Series, which consists of mostly monthly samplings taken at multiple depths between November 1995 and January 2017.
+We compared this data with nutrient, temperature and salinity data from Niskin bottle samples, as well as climate data and indices, focusing on the bottom-up drivers of variability and diversity.
+In total, we used a complete set of 254 monthly data points, on which all variables were available.
+% Figure 1
+Normalized yearly means for all used environmental variables, as well as chlorophyll a and functional group counts, are shown in Figure \ref{fig:zscore}.
+The patterns emerging from -u 10m wind component, Temperature and the directly related 21°C Isotherm show clearly the years of positive anomalies of wind (red),  upwelling (blue, i.e.
+shallower depth of isotherm) and negative for temperature (blue) within the top 100 m.
+These patterns are reflected in the mode of the Atlantic Multidecadal Oscillation (AMO), where a positive anomaly seems to correlate with weak upwelling regimes.
+The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing oscillating patterns.
+Nitrate, and to some extent phosphate, seem to follow the upwelling anomalies, but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration.
+Silicate does not follow a clear pattern, but shows an elevated mean concentration towards the later part of the time series (between 2008 and 2013), which could be driven by surface runoff and reduced depletion via diminished Diatom abundances.
+
+% Figure 2
+
+% Figure 3
+
+% Figure 4
+
+% Figure 5
 
@@
 \begin{figure}
+\label{fig:map}
 \noindent\includegraphics[width=\textwidth]{fig/Artboard 1.pdf}
@@
 \begin{figure}
+\label{fig:zscore}
 \noindent\includegraphics[width=\textwidth]{fig/PLOTZScores.pdf}
@@
 \begin{figure}
+\label{fig:divts}
 \begin{center}
@@
 \begin{figure}
+\label{fig:bdef}
 \noindent\includegraphics[width=\textwidth]{fig/BDEF_depth_PLOT.pdf}
@@
 $RUE$ is calculated as the ratio of Chlorophyll to the available nutrient ($NO_3$ and $PO_4$ respectively).
-All variables show a significant increasing linear trend with Genus Richness, but none for the Shannon Index or Pielou\'s Evenness Index.}
+All variables show a significant increasing linear trend with Genus Richness, but none for the Shannon Index or Pielou\textquotesingle s Evenness Index.}
 \end{figure}
@@
 \begin{figure}
+\label{fig:clustering}
 \noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot_NEW.pdf}
@@
 \begin{figure}
+\label{fig:GF}
 \noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3.pdf}
@@
 % MEI index relevant (from 2019 REVIEW)
-The Caribbean experiences the most marked of the El Niño–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Niño–Southern Oscillation (Enfield & Mayer 1997).
+The Caribbean experiences the most marked of the El Niño–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Niño–Southern Oscillation \cite{enfield_tropical_1997}.
 

### 2024-12-02 agujournaltemplate.tex
monday afternoon corrections and results writing

@@
     %Talk about the Basin shortly, then about the time series
-    The Cariaco Ocean Time-Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon.
-The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (about 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around 100 m.
+    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon.
+The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around 100 m.
 This area is relatively productive ($\sim$ 400 $g C m^{-2} y^{-1}$), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below 200 m depth.
@@
 \subsection{In-Situ Data}
-    Samples were collected as part of the Cariaco Ocean Time-Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (10.50° N, 64.66° W) \cite{muller-karger_scientific_2019}.
+    Samples were collected as part of the Cariaco Ocean Time Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (10.50° N, 64.66° W) \cite{muller-karger_scientific_2019}.
     The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this manuscript.
 Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela.
-Details of the history of sampling procedures in CARIACO as well as sampling schemes, analytical methods, and data quality control and intercalibration procedures have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.
+Details of the history of sampling procedures in CARIACO, as well as sampling schemes, analytical methods, and data quality control and inter-calibration procedures, have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.
 
@@
 Water samples for chlorophyll a and nutrient analysis were taken monthly during the first cast of each cruise.
-These samples were collected at 8 standard depths within the upper 100 meters (1, 7, 15, 25, 35, 55, 75, and 100 m).
+These samples were collected at eight standard depths within the upper 100 meters (1, 7, 15, 25, 35, 55, 75, and 100 m).
 
-    Between late 1995 and early 1998, nutrient data were supplied by William Senior (Universidad de Oriente, Cumana, Venezuela), and from mid-1997 to the present, by Dr.
+    Between late 1995 and early 1998, nutrient data were supplied by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to the present by Dr.
 K.
@@
 Samples from eight cruises were analyzed by both laboratories.
-The data for $PO_4$ and $SiO_4$ were in strong agreement (within ±5\%), whereas the $NO_x$ and $NH_4$ data showed discrepancies.
+Data for $PO_4$ and $SiO_4$ were in strong agreement (within ±5\%), whereas the $NO_x$ and $NH_4$ measurements showed discrepancies.
 \cite{taylor_ecosystem_2012}.
@@
 
-    For phytoplankton species identification by microscopy, 500 mL seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate.
+    For the identification of phytoplankton species by microscopy, 500 mL seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate.
 Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using 100 mL sedimentation chambers with a 48-hour settling period.
@@
 %Note: Repo currently still set to private
-    For all depth discrete measurements, the data was interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm used by the U.S.
+    For all depth discrete measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm used by the U.S.
 National Oceanographic Data Center \cite{johnson2006world}.
 The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis.
-For the phytoplankton microscopy data, the individual identified taxonomic units (Species or Genus) were interpolated, integrated and then grouped by Genus via the sum of interpolated counts.
+For the phytoplankton microscopy data, the individual identified taxonomic units (Species or Genus) were interpolated, integrated, and then grouped by Genus via the sum of interpolated counts.
 
     \subsection{Data analysis}    
-    To give a general overview of the data and trends over the time series, we calculated yearly means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$).
-The cruises from 1995 and 2017 were excluded, as there were too few measurements to provide a reliable yearly mean estimate.
+    To give a general overview of the data and trends in the time series, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$).
+Cruises from 1995 and 2017 were excluded, as there were too few measurements to provide a reliable yearly mean estimate.
 
-    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to make the dataset more homogeneous and more robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008}.
-Additionally, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
+    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to make the data set more homogeneous and robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008}.
+Furthermore, we calculate the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
 
     %RUE/BDEF ANALYSIS
-    To investigates the effect of diversity on ecosystem functioning, we compared measurements of chlorophyll a, primary productivity and Resource Use Efficiency for Nitrate ($RUE_N$) against Genus Richness, Shannon Index and Pielou\textquotesingle s Index, each integrated at the four depth intervals across each monthly sampling.
+    To investigate the effect of diversity on ecosystem functioning, we compared measurements of chlorophyll a, primary productivity, and resource use efficiency for Nitrate ($RUE_N$) against Genus Richness, Shannon Index, and Pielou\textquotesingle s Index, each integrated at the four depth intervals across each monthly sampling.
 $RUE_N$ is calculated as the ration between the limiting resource ($NO_3$) and phytoplankton biomass (chlorophyll a).
@@
     \subsubsection{Clustering and NMDS}
-    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of yearly aggregates of genera observed integrated over 100 m depth.
+    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth.
 Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}.
-We used the Bray distance matrix in a 2-dimensional ordination, that reached a converging solution within 30 steps.
+We used the Bray distance matrix in a 2-dimensional ordination that reached a converging solution within 30 steps.
 To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations.
@@
 It calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor.
-The regression tree split values also identify ranges of the predictors where significant changes occur in indicator variables, which signal a threshold for community turnover in the phytoplankton community.
+The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for community turnover in the phytoplankton community.
 Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}.
-These analyses were conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data.
-The drivers/predictor variables considered in the analysis were the AMO index, MEI v.2 index, -u 10m wind component, the 21°C Isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$).
-The environmental variables from ocean time series tend to be strongly correlated, which Gradient Forest methodology takes into account by implementing a conditional permutation of correlated predictors \cite{ellis_gradient_2012}.
+This analysis was conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data.
+The drivers/predictor variables considered in the analysis were the AMO index, MEI v.2 index, the -u 10m wind component, the 21 ° C isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$).
+The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology takes into account by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}.
 To take into account the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all variables in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run.
@@
 \section{Draft Results}
-Our study focused on the phytoplankton microscopy data from the CARIACO Time Series, which consists of mostly monthly samplings taken at multiple depths between November 1995 and January 2017.
-We compared this data with nutrient, temperature and salinity data from Niskin bottle samples, as well as climate data and indices, focusing on the bottom-up drivers of variability and diversity.
+Our study focused on phytoplankton microscopy data from the CARIACO time series, which consists of monthly samplings taken at multiple depths between November 1995 and January 2017.
+We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity.
 In total, we used a complete set of 254 monthly data points, on which all variables were available.
 % Figure 1
-Normalized yearly means for all used environmental variables, as well as chlorophyll a and functional group counts, are shown in Figure \ref{fig:zscore}.
-The patterns emerging from -u 10m wind component, Temperature and the directly related 21°C Isotherm show clearly the years of positive anomalies of wind (red),  upwelling (blue, i.e.
-shallower depth of isotherm) and negative for temperature (blue) within the top 100 m.
+The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll a and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}.
+The patterns emerging from the -u 10m wind component, the temperature and the directly related 21 ° C isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m.
 These patterns are reflected in the mode of the Atlantic Multidecadal Oscillation (AMO), where a positive anomaly seems to correlate with weak upwelling regimes.
-The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing oscillating patterns.
-Nitrate, and to some extent phosphate, seem to follow the upwelling anomalies, but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration.
-Silicate does not follow a clear pattern, but shows an elevated mean concentration towards the later part of the time series (between 2008 and 2013), which could be driven by surface runoff and reduced depletion via diminished Diatom abundances.
+The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing interannual oscillations.
+Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration.
+Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.
 
 % Figure 2
+To further investigate the depth distribution and dynamics of chlorophyll biomass and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as across seasons.
+The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters.
+Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$).
+The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters, which during some years is below the concentration between 25 and 50 meters during the years 2000, 2005, 2007, 2008 and 2013 (see Figure \ref{fig:divts} $a$).
+In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012.
+The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value.
+Nevertheless, the monthly measurements for the remainder of the year 2013 show a marked reduction in biomass (65\% compared to the mean over the entire time series for these months), which recovers in 2014.
 
@@
 \label{fig:zscore}
-\noindent\includegraphics[width=\textwidth]{fig/PLOTZScores.pdf}
+\noindent\includegraphics[width=\textwidth]{fig/PLOTZScores_updated.pdf}
 \caption{Heatmap of yearly average values, normalized for comparison as z-scores.
@@
 \begin{center}
-\noindent\includegraphics[width=480]{fig/DepthDivTimeSeries.pdf}
+\noindent\includegraphics[width=480]{fig/DepthDivTimeSeries_updated.pdf}
 \end{center}

### 2024-12-05 agujournaltemplate.tex
First rough draft of Results

@@
 Primary upwelling was strongest between November and May.
-In general, during upwelling, sea surface temperatures ranged between approximately 23◦Cand25◦C.
-The secondary upwelling that occurs between June and August is shorter (∼5 weeks) and is approximately 1.5\°C warmer than the primary upwelling.
+In general, during upwelling, sea surface temperatures ranged between approximately 23\°C and 25\°C.
+The secondary upwelling that occurs between June and August is shorter ($\sim$ 5 weeks) and is approximately 1.5\°C warmer than the primary upwelling.
 It is driven by Ekman transport and an intensification of the curl of the wind close to the southern coast of the Caribbean (Rueda-Roa \& Muller-Karger 2013, Rueda-Roa et al.
@@
 From July to October, there is strong stratification in the upper 50 m of the water column after the wind weakens and the rain and rivers deliver less saline waters (Figure 2).
-        Sea surface temperatures ranged from approximately 28◦C to 30◦C during the period of stratification (Figure 2).
+        Sea surface temperatures ranged from approximately 28\°C to 30\°C during the period of stratification (Figure 2).
 The mixed-layer depth varied from less than 10 m to 35 m during the 21-year time series (Figure 3).
@@
 The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value.
-Nevertheless, the monthly measurements for the remainder of the year 2013 show a marked reduction in biomass (65\% compared to the mean over the entire time series for these months), which recovers in 2014.
+Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% compared to the mean over the entire time series for these months), which recovers in 2014.
+Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series.
+In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
+The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter.
+Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011.
+Scaling this genus diversity by the cell count abundances, i.e.
+the Shannon Index, shows a similar but less pronounced dynamic.
+Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016.
+The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$), yet there is a consistent pattern of higher Shannon diversity at lower depths for all seasons except for Fall.
+Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$).
+The interannual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value.
+There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016.
+Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.
 
 % Figure 3
+The effects of diversity on ecosystem functions can be described by the relationship between measured diversity indicators and corresponding measurements of biomass, productivity, and resource use efficiency (RUE).
+In order to take into account effects of depth on the resulting response, we used the same interpolated depth intervals (0-25, 25-50, 50-75, and 75-100 meters) and plotted coincidental measurements against each other.
+The considered diversity indices are the genus richness, Shannon index, and Pielou index, which are plotted against chlorophyll a biomass, primary productivity, as well as RUE for Nitrate ($RUE_N$).
+Immediately apparent, is that depth has a strong effect on the measured relationship, in particular for genus richness, where most instances of high biomass, productivity and $RUE_N$ at high diversity were observed in the uppwer water column (see Figure \ref{fig:bdef} $a, d, g$).
+For the Shannon index, diversity at the different depths shows a much wider variability and overlap, so the pattern is less clear.
+Similarly, community evenness shows no clear depth structure, except that very high biomass and productivity measurements in the upper 25 meters tend to be clustered at intermediate evenness.
+To quantify the relationships, we performed simple linear regression... %TODO perhaps check the glmm method to account for depth and seasonality
 
 % Figure 4
+Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment.
+For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements.
+Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix.
+The community is separated into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013.
+This confirms previously observed patterns in the Cariaco ecosystem, where a strong shift in the community happened in 2004, but the community returned to the previous regime past 2014 \cite{taylor_ecosystem_2012, muller-karger_scientific_2019}.
+Interannual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, turnover is higher again.
+This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years.
+To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$).
+This NMDS clustering is overlayed with environmental data and climate indices to help explain some of the underlying trends.
+The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards postive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index.
+The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed.
+Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering.
+Figure \ref{fig:clustering} $d$ shows density distributions for environmental variables, climate and diversity indices between the two clusters.
+The patterns for upwelling related variables show clear differences between the two clusters, for example for cluster 1 there was a higher occurrence of high wind, shallower 21 \°C Isotherm, as well as lower temperature and higher salinity and nutrient concentrations, in particular for nitrate.
+The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies.
+Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences.
+For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.
 
 % Figure 5
-
-%from Mutshinda et al. 2013
-Phytoplankton species richness varied from 4 to 100 species per cruise, with a mean of 40 over the CARIACO time series (167 cruises from November 1995 to March 2011).
-Over a third of the species (187 of 525) were only observed during one or two cruises in the entire study.
-
-%Text here ===>>>
-Below here I will put the draft plots to give an overview over the current state of the results.
-As of now, all plots are in a draft state, sometimes missing appropriate units.
-Please excuse this, as I would like to first trim down the plots and order them for the final manuscript, before committing further time into finalizing the details.
+To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
+Preliminary model runs were used to identify if time lags of the predictor variables could better explain the data (see Appendix), and this resulted in choosing the AMO with a lag of 5 months and MEI v.2 with a lag of 4 months.
+In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the highest predictive capability for the climate indices with time lags, and upwelling related variables such as nitrate in the upper 100 meters and wind speed scoring slightly lower.
+The 21 \°C Isotherm and silicate and phosphate concentration scored lowest in predicting shifts within the community.
+The way Gradient Forest predicts shifts in the community, is by building a random forest model for each observed genus across the time series.
+The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$.
+The microscopy data is biased towards larger cell sizes, and therefore most of the predicted genus cell counts are from diatoms.
+However, two Haptophytes and four Dinoflagellates, as well as the Cyanobacteria \textit{Synechococcus} do rank highly.
+Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the paramater range.
+The black line indicates the cumulative importance across the entire community, and the colored dashed lines indicate the importance per phytoplankton group.
+For the AMO index, we see a clear pattern that negative anomalies highly affect the community.
+This contrasts with a higher importance at positive anomalies in the MEI v.2 index, in particular for Haptophytes.
+Nitrate concentrations between 6 and 8 $µM$ is where the community changed most, which indicates these values as a threshold for the composition of the community.
+For wind speed, temperature and salinity, changes occurred most at parameter ranges related to high upwelling, i.e.
+high wind (between -4 and -6 $m s^{-1}$), low temperature (below 23 \°C) and high salinity (above 36.8 PSU).
+One interesting aspect is that for both climate indices, Haptophytes showed the strongest cumulative importance, but for Diatoms the nitrate concentration seem to be a stronger predictor.
 

### 2024-12-18 agujournaltemplate.tex
Savepoint before removing BDEF

@@
 \usepackage{soul}
+
+% MY OWN ADDED PACKAGES TO RENDER UNITS
+\usepackage{gensymb}
+\usepackage{booktabs}
+
 \linenumbers
@@
 Primary upwelling was strongest between November and May.
-In general, during upwelling, sea surface temperatures ranged between approximately 23\°C and 25\°C.
-The secondary upwelling that occurs between June and August is shorter ($\sim$ 5 weeks) and is approximately 1.5\°C warmer than the primary upwelling.
+In general, during upwelling, sea surface temperatures ranged between approximately 23\degree C and 25\degree C.
+The secondary upwelling that occurs between June and August is shorter ($\sim$ 5 weeks) and is approximately 1.5\degree C warmer than the primary upwelling.
 It is driven by Ekman transport and an intensification of the curl of the wind close to the southern coast of the Caribbean (Rueda-Roa \& Muller-Karger 2013, Rueda-Roa et al.
@@
 From July to October, there is strong stratification in the upper 50 m of the water column after the wind weakens and the rain and rivers deliver less saline waters (Figure 2).
-        Sea surface temperatures ranged from approximately 28\°C to 30\°C during the period of stratification (Figure 2).
+        Sea surface temperatures ranged from approximately 28\degree C to 30\degree C during the period of stratification (Figure 2).
 The mixed-layer depth varied from less than 10 m to 35 m during the 21-year time series (Figure 3).
@@
 Figure \ref{fig:clustering} $d$ shows density distributions for environmental variables, climate and diversity indices between the two clusters.
-The patterns for upwelling related variables show clear differences between the two clusters, for example for cluster 1 there was a higher occurrence of high wind, shallower 21 \°C Isotherm, as well as lower temperature and higher salinity and nutrient concentrations, in particular for nitrate.
+The patterns for upwelling related variables show clear differences between the two clusters, for example for cluster 1 there was a higher occurrence of high wind, shallower 21 \degree C Isotherm, as well as lower temperature and higher salinity and nutrient concentrations, in particular for nitrate.
 The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies.
@@
 In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the highest predictive capability for the climate indices with time lags, and upwelling related variables such as nitrate in the upper 100 meters and wind speed scoring slightly lower.
-The 21 \°C Isotherm and silicate and phosphate concentration scored lowest in predicting shifts within the community.
+The 21 \degree C Isotherm and silicate and phosphate concentration scored lowest in predicting shifts within the community.
 The way Gradient Forest predicts shifts in the community, is by building a random forest model for each observed genus across the time series.
@@
 For wind speed, temperature and salinity, changes occurred most at parameter ranges related to high upwelling, i.e.
-high wind (between -4 and -6 $m s^{-1}$), low temperature (below 23 \°C) and high salinity (above 36.8 PSU).
+high wind (between -4 and -6 $m s^{-1}$), low temperature (below 23 \degree C) and high salinity (above 36.8 PSU).
 One interesting aspect is that for both climate indices, Haptophytes showed the strongest cumulative importance, but for Diatoms the nitrate concentration seem to be a stronger predictor.
@@
 Vector overlay of ENVFIT shows underlying environmental variables that are co-varying with the community data.
-\newline
 Plots (A-L) below show density distributions of the environmental variables as well as diversity indices between the clustered years. Dashed line shows the mean.}
 \end{figure}
+
+\begin{table}
+\label{table:clustering}
+\caption{Wilcoxon sum rank test with continuity correction for variables between the two clusters.
+Values for in-situ data are interpolated across the top 100 meters for each individual monthly cruise sampling.}
+\centering
+\begin{tabular}[t]{lrrl}
+\toprule
+Variable & W & p.value & Significance\\
+\midrule
+AMO & 3494.0 & 1.86e-14 & ***\\
+MEI v.2 & 9523.5 & 0.006 & ***\\
+u-component 10 m wind speed & 7074.0 & 0.143 & n.s.\\
+Isotherm 21 \degree C & 5293.0 & 0.034 & **\\
+\addlinespace
+Temperature & 4763.0 & 0.003 & ***\\
+$NO_3$ & 6378.0 & 0.018 & **\\
+$PO_4$ & 7359.0 & 0.012 & **\\
+$SiO_4$ & 5998.0 & 0.736 & n.s.\\
+Salinity & 7636.0 & 0.002 & ***\\
+\addlinespace
+Genus Richness & 5913.0 & 0.954 & n.s.\\
+Shannon Index & 5142.0 & 0.108 & n.s.\\
+Pielou Index & 4874.0 & 0.029 & **\\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 %
-%\appendix
-%\section{Here is a sample appendix}
+
+\appendix
+
+\section{Gradient Forest Analysis supplement}
+
+\begin{table}
+\ref{app:table1}
+
+\caption{Individual Gradient Forest model runs for each variable and the corresponding time lags. 
+        For in-situ variables, for which no coverage extends over the time series, 
+        we tested all measurements up to a lag of 3 months. 
+        For climate variables, we tested up to a lag of 6 months. 
+        Time lag with maximum importance per variable are highlighted in bold and chosen for the final model run.}
+\centering
+\begin{tabular}[t]{rrrrrrrrrr}
+\toprule
+lag & NO3 & PO4 & SiO4 & Temperature & Isotherm\_21 & Salinity\_bottles & u10 & AMO & MEIv2\\
+\midrule
+0 & \textbf{0.038} & \textbf{0.038} & \textbf{0.034} & \textbf{0.071} & \textbf{0.028} & 0.007 & 0.004 & 0.016 & 0.015\\
+1 & 0.017 & 0.011 & 0.015 & 0.002 & 0.002 & 0.004 & \textbf{0.010} & 0.014 & 0.018\\
+2 & 0.022 & 0.008 & 0.013 & 0.006 & 0.008 & \textbf{0.011} & 0.009 & \textbf{0.030} & 0.019\\
+3 & 0.016 & 0.007 & 0.004 & 0.005 & 0.007 & 0.010 & 0.007 & 0.018 & 0.018\\
+4 & - & - & - & - & - & - & 0.008 & 0.019 & \textbf{0.029}\\
+5 & - & - & - & - & - & - & 0.008 & 0.018 & 0.022\\
+6 & - & - & - & - & - & - & 0.010 & 0.018 & 0.013\\
+\bottomrule
+\end{tabular}
+\end{table}
 

### 2024-12-18 agujournaltemplate.tex
presync

@@
 %Note: Repo currently still set to private
-    For all depth discrete measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm used by the U.S.
+    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm used by the U.S.
 National Oceanographic Data Center \cite{johnson2006world}.
@@
 Furthermore, we calculate the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
-
-    %RUE/BDEF ANALYSIS
-    To investigate the effect of diversity on ecosystem functioning, we compared measurements of chlorophyll a, primary productivity, and resource use efficiency for Nitrate ($RUE_N$) against Genus Richness, Shannon Index, and Pielou\textquotesingle s Index, each integrated at the four depth intervals across each monthly sampling.
-$RUE_N$ is calculated as the ration between the limiting resource ($NO_3$) and phytoplankton biomass (chlorophyll a).
-    We excluded observations with $NO_3$ concentrations below 0.01 µM (n = 36) because they are close to the detection limit, resulting in imprecise measurements \cite{ptacnik_diversity_2008}.
-    To quantify the trend in the relationships, we use linear regression models using the "lm" function \cite{r_core_team_r_2024}.
-    %here I want to extend to glmm that correct for depth and seasonality for the final results
 
@@
 % Figure 3
-The effects of diversity on ecosystem functions can be described by the relationship between measured diversity indicators and corresponding measurements of biomass, productivity, and resource use efficiency (RUE).
-In order to take into account effects of depth on the resulting response, we used the same interpolated depth intervals (0-25, 25-50, 50-75, and 75-100 meters) and plotted coincidental measurements against each other.
-The considered diversity indices are the genus richness, Shannon index, and Pielou index, which are plotted against chlorophyll a biomass, primary productivity, as well as RUE for Nitrate ($RUE_N$).
-Immediately apparent, is that depth has a strong effect on the measured relationship, in particular for genus richness, where most instances of high biomass, productivity and $RUE_N$ at high diversity were observed in the uppwer water column (see Figure \ref{fig:bdef} $a, d, g$).
-For the Shannon index, diversity at the different depths shows a much wider variability and overlap, so the pattern is less clear.
-Similarly, community evenness shows no clear depth structure, except that very high biomass and productivity measurements in the upper 25 meters tend to be clustered at intermediate evenness.
-To quantify the relationships, we performed simple linear regression... %TODO perhaps check the glmm method to account for depth and seasonality
-
-% Figure 4
 Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment.
@@
 
-% Figure 5
+% Figure 4
 To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
@@
 \caption{Overview over Chlorophyll and Diversity Variables over Depth}
-\end{figure}
-
-\begin{figure}
-\label{fig:bdef}
-\noindent\includegraphics[width=\textwidth]{fig/BDEF_depth_PLOT.pdf}
-\caption{Linear regression of diversity indices to log-transformed Chlorophyll Biomass, Primary Production, Resource Use Efficiency for nitrate ($RUE_N$) and for phosphate ($RUE_P$).
-$RUE$ is calculated as the ratio of Chlorophyll to the available nutrient ($NO_3$ and $PO_4$ respectively).
-All variables show a significant increasing linear trend with Genus Richness, but none for the Shannon Index or Pielou\textquotesingle s Evenness Index.}
 \end{figure}
@@
 \label{fig:GF}
-\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3.pdf}
+\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3_new.pdf}
 \caption{Interestingly only time lags of the climatic indices (AMO and MEIv2) actually showed better predictive capabilities, there were none for lags of wind, nutrients or the isotherm.

### 2024-12-19 agujournaltemplate.tex

@@
 
-\title{Bottum-up drivers of phytoplankton community diversity in the Cariaco Basin, Venezuela}
+\title{Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela}
 
@@
 \begin{keypoints}
-\item The phytoplankton community data identified by microscopy from the CARIACO Time Series between 1996 and 2016 shows two separate clusters characterized by high and low upwelling regimes respectively.
-\item Despite the differing physical regimes and large shifts between functional groups, there is no significant difference in diversity metrics between the two clusters.
-Overall, diversity (i.e.
-Genus Richness) shows a positive linear relationship to Biomass, Productivity as well as Resource Use Efficiency.
-\item A gradient forest analysis of the community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 5 months, which points to a strong impact of large scale climatic oscillations in the tropical coastal ecosystem.
-The next strongest predictor is nitrate concentration, with the MEI v2 index (lag of 4 months) and salinity following behind.
+\item The CARIACO time series showed a highly variable phytoplankton community over the course of sampling between 1995 and 2017.
+Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the AMO index.
+\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, although community evenness was higher in low upwelling conditions.
+\item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem.
+The next strongest predictor is the MEI v2 index (lag of 4 months), with in-situ temperature and nitrate concentration following behind.
 \end{keypoints}
@@
 \begin{abstract}
-[ enter your Abstract here ]
+The structure of the phytoplankton community was identified using microscopy in the Cariaco basin, in Venezuela, between 1995 and 2017.
+Microscopy data are limited to larger cell sizes (> 20 µm), where the community showed a large change in the abundances of functional groups and shifts in diversity patterns.
 \end{abstract}
@@
 For the overview of the entire time-series, we used a merged dataset for all nutrients, where coincidental monthly measurements were averaged.
-For the analysis of Resource Use Efficiency, only the USF $NO_3$ measurements were used.
 
@@
     %from Mutshinda et al. 2013
-    To quality control the microscopy data, we assembled the taxonomic identifications made across all cruises, removed observations that were not identified to a species or genus level, corrected spelling variations, and grouped synonyms using taxonomic information in the World Register of Marine Species (\url{marinespecies.org}).
-After processing the data, there were 482 phytoplankton species observed over the course of the time series, which group into 223 phytoplankton genera.
+    Quality control the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified to species or genus level, correcting spelling variations, and grouping synonyms using taxonomic information in the World Register of Marine Species (\url{marinespecies.org}).
+After processing the data, 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.
 
     The datasets for Niskin bottle samples, CTD and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024).
-
-    From the CARIACO Time Series data, we used only those variables, that cover the entire length of the time series without large gaps in coverage.
+    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps in coverage.
 Therefore, we did not include Zooplankton net tow and HPLC pigment data in our analysis.
@@
     % Niskin, CTD and Phytoplankton data processing
-    Data was processed and analyzed using the R programming langauge (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with the tidyverse packages \cite{wickham_welcome_2019}.
-All data processing and analysis code is publically available on Github \url{https://github.com/ben1post/cariaco-ms2}.
+    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with tidyverse packages \cite{wickham_welcome_2019}.
+All data processing and analysis code is publically available on GitHub \url{https://github.com/ben1post/cariaco-ms2}.
 %Note: Repo currently still set to private
-    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm used by the U.S.
+    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S.
 National Oceanographic Data Center \cite{johnson2006world}.
 The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis.
-For the phytoplankton microscopy data, the individual identified taxonomic units (Species or Genus) were interpolated, integrated, and then grouped by Genus via the sum of interpolated counts.
+For phytoplankton microscopy data, individual identified taxonomic units (Species or Genus) were interpolated, integrated, and then grouped by Genus via the sum of interpolated counts.
 
@@
     To quantify the effects of bottom-up drivers in the Cariaco basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis.
-Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables\cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}.
-It calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor.
-The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for community turnover in the phytoplankton community.
+Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}.
+Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor.
+The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover.
 Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}.
 This analysis was conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data.
-The drivers/predictor variables considered in the analysis were the AMO index, MEI v.2 index, the -u 10m wind component, the 21 ° C isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$).
+The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, the -u 10m wind component, the 21 ° C isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$).
 The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology takes into account by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}.
-To take into account the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all variables in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run.
-% see figure in supplementary with initial runs TODO
+To consider the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all climate variables, as well as 3 months lags for in-situ measurements in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run (see Table \ref{app:table1} in the Appendix).
 
@@
 These patterns are reflected in the mode of the Atlantic Multidecadal Oscillation (AMO), where a positive anomaly seems to correlate with weak upwelling regimes.
-The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing interannual oscillations.
+The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing inter-annual oscillations.
 Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration.
@@
 The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value.
-Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% compared to the mean over the entire time series for these months), which recovers in 2014.
+Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014.
 Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series.
@@
 Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$).
-The interannual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value.
+The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value.
 There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016.

### 2024-12-19 agujournaltemplate.tex

@@
 
-    Between late 1995 and early 1998, nutrient data were supplied by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to the present by Dr.
+    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to the present by Dr.
 K.
@@
 \cite{taylor_ecosystem_2012}.
-For the overview of the entire time-series, we used a merged dataset for all nutrients, where coincidental monthly measurements were averaged.
+For our analysis, we used a merged dataset for all nutrients, where coincidental monthly measurements were averaged.
 
@@
     %from Mutshinda et al. 2013
-    Quality control the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified to species or genus level, correcting spelling variations, and grouping synonyms using taxonomic information in the World Register of Marine Species (\url{marinespecies.org}).
+    Quality control the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level, correcting spelling variations and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}).
 After processing the data, 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.
 
-    The datasets for Niskin bottle samples, CTD and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024).
+    Data sets for Niskin bottle samples, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024).
     From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps in coverage.
@@
 To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$).
-This NMDS clustering is overlayed with environmental data and climate indices to help explain some of the underlying trends.
-The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards postive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index.
+This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends.
+The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index.
 The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed.
 Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering.
-Figure \ref{fig:clustering} $d$ shows density distributions for environmental variables, climate and diversity indices between the two clusters.
+Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters.
 The patterns for upwelling related variables show clear differences between the two clusters, for example for cluster 1 there was a higher occurrence of high wind, shallower 21 \degree C Isotherm, as well as lower temperature and higher salinity and nutrient concentrations, in particular for nitrate.
@@
 \caption{Heatmap of yearly average values, normalized for comparison as z-scores.
-All data is available at monthly resolution (see next draft figure for raw data).
 Red indicates a larger value, blue indicates a smaller value for that particular year.
-The years 1995 (n=2) and 2017 (n=1) were removed due to lack of data.
+The years 1995 (n=2) and 2017 (n=1) were removed due to lack of data, additionally we marked years with more than .
 Diversity Indices are calculated based on microscopy counts at the Genus level.}
@@
 \label{fig:clustering}
-\noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot_NEW.pdf}
+\noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot_NEW 2.pdf}
 \caption{(A) Clustering of microscopy data at the Genus level, using the binary Jaccard distance.

### 2025-01-30 agujournaltemplate.tex
Post Subhi-Comments and fixing units

@@
 \usepackage{booktabs}
+\usepackage{siunitx}
 
@@
 \item The CARIACO time series showed a highly variable phytoplankton community over the course of sampling between 1995 and 2017.
-Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the AMO index.
-\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, although community evenness was higher in low upwelling conditions.
+Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the anomalies in the AMO index.
+\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, although community evenness was higher in low-upwelling conditions.
 \item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem.
@@
 
+    \item - Phytoplankton is important
+
+    %“Time series of biological observations provide insights into the responses of marine ecosystems to cyclic processes, transient events, and, when maintained over many decades, changes in climate (Carstensen 2014, Henson et al.
+2016).
+Such time series are needed for assessments on the dynamic state of the ocean.” ([Muller-Karger et al., 2019, p.
+2](zotero://select/library/items/7YW7MFTU)) ([pdf](zotero://open-pdf/library/items/TTZQBAXW?page=2&annotation=R4DSZPK6))
+
+    - Phytoplankton is important
+    - Why community structure is important, taxonomic diversity (aspects of diversity) - why look at taxonomic diversity?
+
+    - some general intro to DIVERSITY.. what is it? Why important?
+    %"Whereas the positive biomass response to high nutrient and light availability is well known (Tilman 1982; Leibold 1999; Siegel et al.
+2013), the biodiversity response is much more complex.
+First, biodiversity itself is a multifaceted construct, comprising the number of different taxa (e.g.
+species richness) and the equality of species contribution to the assemblage as aspects of local alpha diversity (evenness).
+But the change in richness is only the net effect of immigration and local extinction, while the turnover in species composition with time is an additional aspect of biodiversity.
+Extending to larger spatial scales, heterogeneity in space (beta diversity) and composition of the regional species pool (gamma diversity) are further biodiversity dimensions.
+Second, for each of these facets, all metrics are effort- and scaledependent.
+Thus, in contrast to other aspects of ecosystem assessments, biodiversity metrics do not qualify to set absolute target values for good or bad status and effective ecosystem management.
+But, the relevance of the metrics arises from analysing their temporal trends over a long time period.
+Third, no single variable captures even the most important aspects of community composition and change.
+Therefore, the analysis of phytoplankton biodiversity requires a multimetric approach." from Di Cavalho et al.
+2023 (with Hillebrand)
+
+    - why are we looking at Functional Groups?
+    %"The complexity of biodiversity assessment has led to the development of other approaches addressing compositional change in plankton communities, most prominently the creation of functional groups.
+Functional groups are meant to comprise species that show similar responses to environmental drivers and share morphological and/or physiological characteristics.
+Functional group indicators have been shown to be relevant for describing community structure and biodiversity and are more comparable with other studies than species-based indicators (Mouillot et al.
+2006)." from Di Cavalho et al.
+2023 (with Hillebrand)
+
     %% Overarching theme: Stability of Ecosystem faunctions, Drivers of community change
+    \item "Regime shifts + Diversity" - general intro on current findings related to "climate driven ecosystem shifts"
+        - %"We find that the rate of turnover in the phytoplankton community becomes faster during this century, that is, the community structure becomes increasingly unstable in response to climate change.
+Combined with alterations to phytoplankton diversity, our results imply a loss of ecological resilience, with likely knock-on effects on the productivity and functioning of the marine environment." \cite{henson_future_2021}
+        -
 
     \item "The Cariaco Basin, located ...." - general hydrography/depth, etc.
+        The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}.
+        A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May, and a smaller secondary peak in July and August \cite{astor_seasonal_2003}.
+        The basin is effectively separated from the Caribbean Sea by a sill with a mean depth of about 100 m.
+The water below the sill is poorly ventilated, which, combined with the high surface productivity, results in permanent anoxic conditions at depths below 275 m \cite{thunell_organic_2000}.
+
+        Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!
 
@@
         % below is from 2019 REVIEW
-        Hydrographic observations (Figure 2) show strong seasonality and interannual variability in the upper 400 m in the Cariaco Basin.
+        %Hydrographic observations (Figure 2) show strong seasonality and interannual variability in the upper 400 m in the Cariaco Basin.
 This is controlled by seasonal and interannual changes in the wind and in the intensity of the geostrophic Caribbean Current (Muller-Karger et al.
@@
 From July to October, there is strong stratification in the upper 50 m of the water column after the wind weakens and the rain and rivers deliver less saline waters (Figure 2).
-        Sea surface temperatures ranged from approximately 28\degree C to 30\degree C during the period of stratification (Figure 2).
+        %Sea surface temperatures ranged from approximately 28\degree C to 30\degree C during the period of stratification (Figure 2).
 The mixed-layer depth varied from less than 10 m to 35 m during the 21-year time series (Figure 3).
@@
 
+        %“The Caribbean experiences the most marked of the El Nino–Southern Oscillation telecon- ˜ nections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Nino–Southern Oscillation (Enfield & Mayer 1997).
+˜” ([Muller-Karger et al., 2019, p.
+4](zotero://select/library/items/7YW7MFTU)) ([pdf](zotero://open-pdf/library/items/TTZQBAXW?page=4&annotation=D3H7YG7Y)) link to El Nino of Cariaco/Caribbean sea!
+
+        %“When the CARIACO time series was implemented in 1995, major changes were happening in the Atlantic.
+We now know that at that time, the Atlantic Multidecadal Oscillation (Kerr 2000, Enfield et al.
+2001) started shifting to a positive phase, which then persisted for more than 20 years and led to oceanographic anomalies in the Atlantic that were larger than any recorded over the previous 100 years (McCarthy et al.
+2015).
+The changes had profound biological impacts throughout the Atlantic (Alexander et al.
+2014, Nye et al.
+2014), including in the Cariaco Basin (Taylor et al.
+2012, Pinckney et al.
+2015).” (Muller-Karger et al., 2019, p.
+3) Interesting aspect that the AMO changed, as apparently the main climate indices to affect the Atlantic/CARIACO
+
     \item "Here's where we come in..." Talk about how most of the analysis stopped after 2012, time series died out, but there is more data there and particularly looking at community data, we can see a return.
+    - what has been done (Phytoplankton - Statistics)?
+    - where are the gaps?
+
+    So we would hypothesize, that if wind is driving the ecosystem, then wind would be best predictor of shits.
+    Or nutrients, but if it is a climate indices, then the indices capture some variablity that is not related to wind or nutrients....
 
@@
 The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around 100 m.
-This area is relatively productive ($\sim$ 400 $g C m^{-2} y^{-1}$), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below 200 m depth.
+This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below 200 m depth.
 
 \subsection{In-Situ Data}
-    Samples were collected as part of the Cariaco Ocean Time Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (10.50° N, 64.66° W) \cite{muller-karger_scientific_2019}.
+    Samples were collected as part of the Cariaco Ocean Time Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
     The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this manuscript.
@@
 
-    For the identification of phytoplankton species by microscopy, 500 mL seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate.
-Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using 100 mL sedimentation chambers with a 48-hour settling period.
-Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$20 μm) \cite{mutshinda_environmental_2013}.
+    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate.
+Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period.
+Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
     %from Mutshinda et al. 2013
@@
     Data sets for Niskin bottle samples, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024).
-    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps in coverage.
+    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (more than three months) in coverage.
 Therefore, we did not include Zooplankton net tow and HPLC pigment data in our analysis.
+Our study instead focused on the phytoplankton microscopy data.
+We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity.
+In total, we used a complete set of 254 monthly data points, on which all variables were available.
 
@@
     %from Taylor et al. 2012
-    Zonal wind stress (−u, easterly Trade Winds) across the southern Caribbean Sea drives near-shore upwelling of nutrient-rich waters by Ekman transport \cite{taylor_ecosystem_2012}.
-We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis at the grid cell closest to the Cariaco Time Series sampling coordinates (10.4°N,64.8°W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024)
+    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives near-shore upwelling of nutrient-rich waters by Ekman transport \cite{taylor_ecosystem_2012}.
+We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling coordinates (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024).
 
@@
 
-    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of the surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (30°S-30°N and 100°E-70°W), were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024).
+    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of the surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024).
 In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).
@@
     Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with tidyverse packages \cite{wickham_welcome_2019}.
-All data processing and analysis code is publically available on GitHub \url{https://github.com/ben1post/cariaco-ms2}.
+All data processing and analysis code is publicly available on GitHub \url{https://github.com/ben1post/cariaco-ms2}.
 %Note: Repo currently still set to private
@@
 
-    \subsubsection{Clustering and NMDS}
+    \subsubsection{Phytoplankton Community Structure Analysis}
     To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth.
@@
 
-    \subsubsection{Gradient Forest}
+    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
     To quantify the effects of bottom-up drivers in the Cariaco basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis.
@@
 This analysis was conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data.
-The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, the -u 10m wind component, the 21 ° C isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$).
+The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, the -u 10m wind component, the \qty{21}{\celsius} isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$).
 The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology takes into account by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}.
@@
 \section{Draft Results}
-Our study focused on phytoplankton microscopy data from the CARIACO time series, which consists of monthly samplings taken at multiple depths between November 1995 and January 2017.
-We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity.
-In total, we used a complete set of 254 monthly data points, on which all variables were available.
+
 % Figure 1
-The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll a and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}.
-The patterns emerging from the -u 10m wind component, the temperature and the directly related 21 ° C isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m.
-These patterns are reflected in the mode of the Atlantic Multidecadal Oscillation (AMO), where a positive anomaly seems to correlate with weak upwelling regimes.
+The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}.
+The patterns emerging from the -u 10m wind component, the temperature and the directly related \qty{21}{\celcius} isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m.
+These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes.
 The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing inter-annual oscillations.
@@
 % Figure 2
-To further investigate the depth distribution and dynamics of chlorophyll biomass and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as across seasons.
+To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between seasons.
 The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters.
 Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$).
-The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters, which during some years is below the concentration between 25 and 50 meters during the years 2000, 2005, 2007, 2008 and 2013 (see Figure \ref{fig:divts} $a$).
+The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters.
+In some years, particularly during 2000, 2005, 2007, 2008 and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$).
 In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012.
@@
 Scaling this genus diversity by the cell count abundances, i.e.
-the Shannon Index, shows a similar but less pronounced dynamic.
+the Shannon index, shows a similar but less pronounced dynamic.
 Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016.
@@
 The community is separated into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013.
-This confirms previously observed patterns in the Cariaco ecosystem, where a strong shift in the community happened in 2004, but the community returned to the previous regime past 2014 \cite{taylor_ecosystem_2012, muller-karger_scientific_2019}.
-Interannual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, turnover is higher again.
+Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, turnover is higher again.
 This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years.
@@
 Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters.
-The patterns for upwelling related variables show clear differences between the two clusters, for example for cluster 1 there was a higher occurrence of high wind, shallower 21 \degree C Isotherm, as well as lower temperature and higher salinity and nutrient concentrations, in particular for nitrate.
+Upwelling related variables show distinct patterns between the two clusters.
+For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celcius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate.
 The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies.
@@
 In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the highest predictive capability for the climate indices with time lags, and upwelling related variables such as nitrate in the upper 100 meters and wind speed scoring slightly lower.
-The 21 \degree C Isotherm and silicate and phosphate concentration scored lowest in predicting shifts within the community.
+The \qty{21}{\celsius} Isotherm and silicate and phosphate concentrations scored lowest in predicting shifts within the community.
 The way Gradient Forest predicts shifts in the community, is by building a random forest model for each observed genus across the time series.
@@
 However, two Haptophytes and four Dinoflagellates, as well as the Cyanobacteria \textit{Synechococcus} do rank highly.
-Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the paramater range.
+Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range.
 The black line indicates the cumulative importance across the entire community, and the colored dashed lines indicate the importance per phytoplankton group.
@@
 This contrasts with a higher importance at positive anomalies in the MEI v.2 index, in particular for Haptophytes.
-Nitrate concentrations between 6 and 8 $µM$ is where the community changed most, which indicates these values as a threshold for the composition of the community.
+Nitrate concentrations between 6 and 8 \unit{\micro \mole} is where the community changed most, which indicates these values as a threshold for the composition of the community.
 For wind speed, temperature and salinity, changes occurred most at parameter ranges related to high upwelling, i.e.
-high wind (between -4 and -6 $m s^{-1}$), low temperature (below 23 \degree C) and high salinity (above 36.8 PSU).
+high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}) and high salinity (above \qty{36.8}{PSU}).
 One interesting aspect is that for both climate indices, Haptophytes showed the strongest cumulative importance, but for Diatoms the nitrate concentration seem to be a stronger predictor.
@@
 \begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/Artboard 1.pdf}
+\caption{Map of CARIACO Time Series location in the Cariaco basin in Venezuela in the Caribbean Ocean, located at \ang{10.50}N, \ang{64.67}W}
 \label{fig:map}
-\noindent\includegraphics[width=\textwidth]{fig/Artboard 1.pdf}
-\caption{Map of CARIACO Time Series location in the Cariaco basin in Venezuela in the Caribbean Ocean, located at 10.50° N, 64.67° W}
 \end{figure}
@@
 \begin{figure}
-\label{fig:zscore}
 \noindent\includegraphics[width=\textwidth]{fig/PLOTZScores_updated.pdf}
@@
 Red indicates a larger value, blue indicates a smaller value for that particular year.
-The years 1995 (n=2) and 2017 (n=1) were removed due to lack of data, additionally we marked years with more than .
+The years 1995 (n=2) and 2017 (n=1) were removed due to lack of data, additionally we marked years with two or less and three or more missing values.
 Diversity Indices are calculated based on microscopy counts at the Genus level.}
+\label{fig:zscore}
 \end{figure}
@@
 \begin{figure}
-\label{fig:divts}
 \begin{center}
@@
 \end{center}
-\caption{Overview over Chlorophyll and Diversity Variables over Depth}
+\caption{Time series and seasonality of chlorophyll and diversity indices across four depth intervals.
+(a,c,e,g) Annual mean across time series for years with more than 5 monthly sampling.
+Transparent dots show monthly raw values.
+(b,d,f,h) Boxplots of monthly measurements across four depth intervals grouped by season.}
+\label{fig:divts}
 \end{figure}
@@
 \begin{figure}
-\label{fig:clustering}
 \noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot_NEW 2.pdf}
-\caption{(A) Clustering of microscopy data at the Genus level, using the binary Jaccard distance.
+\caption{(a) Clustering of microscopy data at the Genus level, using the binary Jaccard distance.
 The community for each year is binned together, to smooth out seasonal aspects.
-(B) NMDS plot of monthly community data (cell counts) that is cube root transformed.
+(b) Species turnover between years for annual aggregate of identified phytoplankton to a genus level.
+(c) NMDS plot of monthly community data (cell counts) that is cube-root transformed.
 Stress: 0.261, 2 dimensions, converging solution.
@@
 Vector overlay of ENVFIT shows underlying environmental variables that are co-varying with the community data.
-Plots (A-L) below show density distributions of the environmental variables as well as diversity indices between the clustered years. Dashed line shows the mean.}
+Plots (d-o) below show density distributions of the environmental variables as well as diversity indices between the clustered years.
+Dashed line shows the median and significance levels of Wilcoxon sum rank tests between clusters are shown above plots.}
+\label{fig:clustering}
 \end{figure}
@@
 \begin{table}
-\label{table:clustering}
 \caption{Wilcoxon sum rank test with continuity correction for variables between the two clusters.
 Values for in-situ data are interpolated across the top 100 meters for each individual monthly cruise sampling.}
+\label{table:clustering}
 \centering
@@
 u-component 10 m wind speed & 7074.0 & 0.143 & n.s.\\
-Isotherm 21 \degree C & 5293.0 & 0.034 & **\\
+Isotherm \qty{21}{\celcius} & 5293.0 & 0.034 & **\\
 \addlinespace
@@
 \begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3_new.pdf}
+\caption{Gradient Forest analysis of monthly community data at a genus level.
+(a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances.
+(b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance.
+(c) Cumulative importance curves for the predictor variables.
+Black line shows cumulative importance accross predictor range for the entire model and all genera.
+Dashed colored lines represent the responses of individual functional groups.}
 \label{fig:GF}
-\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3_new.pdf}
-\caption{Interestingly only time lags of the climatic indices (AMO and MEIv2) actually showed better predictive capabilities, there were none for lags of wind, nutrients or the isotherm.
-What I also included in the plot above is the underlying mechanism of fitting a random forest model for each genus.
-The plot on the top right shows the weighted importance for each genus, ranked according to the predictive capability of the model, and color coded by their respective group.
-Similarly, the cumulative importance plots below show the overall importance, as well as the individual color lines that show the cumulative importance per genus predicted.
-This is not necessarily useful to show in the final manuscript, unless we want to discuss the model output down to the genus level.
-Not surprisingly, it is mostly diatoms that are predictable, but also other groups are present here.}
 \end{figure}
@@
 In the Cariaco basin, long-term climatic oscillations have affected the upwelling regime and ...
+
 In the Cariaco basin, wind-driven seasonal upwelling dictates the nutrient availability in the upper mixed layer, which in turn affects the phytoplankton dynamics \cite{ptacnik_diversity_2008, bopp_response_2005}.
@@
 % REALLY ONLY LOOK AT BOTTOM UP HERE; No Top Down effects quantified
+
+% AMO
+The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
+
+% "Shifts in global and basin-specific primary production and phytoplankton abundances are related to large-scale climatic variability in the physical environment.
+Despite the lack of a definitive period associated with the AMO, the relationship between the atmospheric variability associated with the NAO and thermohaline circulation characterized by the AMO suggests a coupling of ocean and atmospheric processes on time scales that will facilitate understanding and prediction of inter-annual weather, precipitation, fluvial output and estuarine and marine productivity.
+Several studies have documented shifts related to the AMO in oceanic (Martinez et al., 2009), coastal and continental shelf (Rodrigues et al., 2009; Schofield et al., 2008) and estuarine (Hubeny et al., 2006) productivity and responses in forage fish populations to this increase in planktonic productivity (Tourre et al., 2007).
+Correlations between the AMO indices and marine productivity have been documented at immediate (Schofield et al., 2008), historical (Martinez et al., 2009; Tourre et al., 2007), and pre-historical (Hubeny et al., 2006; Rodrigues et al., 2009) scales.
+There are several mechanisms through which this might occur.
+Stratification within the water column separates nutrients at lower depths from entering the euphotic zone, thus limiting primary productivity.
+Long-term climate oscillations may augment or suppress seasonal conditions that promote or erode stratification and regulate mixing.
+Schofield et al.
+(2008) have demonstrated links between the AMO and phytoplankton blooms in the Mid-Atlantic Bight, such that increased nutrient availability during the annual fall transition were augmented by increased winter winds associated with a positive phase in the AMO.
+Martinez et al.
+(2009) demonstrated that global phytoplankton productivity is linked to multi-decadal oscillations in both the Atlantic and Pacific, suggesting that links between chlorophyll and sea surface temperature are driven by interactions between the pycnocline and mixed surface layers.
+The influence of large-scale climate fluctuations on surface stratification has particular influence on phytoplankton productivity in the tropics and mid-latitudes, waters which are often nutrient-limited.
+Alternatively, warmer sea surface temperatures themselves may influence productivity.
+Shifts in fluvial inputs associated with altered precipitation patterns will also alter river-induced fertilization of marine productivity." from Nye et al.
+2014
+
+%"The AMO is also associated with changes in wind and current regimes (Delworth et al., 2007; Häkkinen et al., 2011).
+Thus, passive dispersal of marine organisms, many of which are planktonic or have planktonic larvae, will be affected resulting in changes in spatial distribution, survival, and population abundance." from Nye et al 2014
+
+FOR THE CLUSTERING (moved from Results): This confirms previously observed patterns in the Cariaco ecosystem, where a strong shift in the community happened in 2004, but the community returned to the previous regime past 2014 \cite{taylor_ecosystem_2012, muller-karger_scientific_2019}.
 
@@
 \begin{table}
-\ref{app:table1}
 
@@
         Time lag with maximum importance per variable are highlighted in bold and chosen for the final model run.}
+
+\ref{app:table1}
 \centering

### 2025-02-18 agujournaltemplate.tex
Intro draft Day 2 of Focus

@@
 Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the anomalies in the AMO index.
-\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, although community evenness was higher in low-upwelling conditions.
+% \index Previous studies proposed a shift in the community, but with the full dataset we see a shift and a return.
+The community shows a departure form cluster one around 2004, but a return to that community in 2014.
+\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, although community evenness was higher and community turnover was lower in the second cluster (low-upwelling conditions).
 \item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem.
@@
 
-\begin{itemize}
+    Phytoplankton communities provide the foundation of marine food webs and play a crucial role in global biogeochemical cycles \cite{falkowski_biogeochemical_1998}.
+These communities consist of a diverse set of species that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}.
+The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them a highly suitable indicator of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
+    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term climate changes, especially when data collection spans several decades \cite{carstensen_need_2014,henson_observing_2016}.
+    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin off the coast of Venezuela between 1995 and 2017 \cite{muller-karger_scientific_2019}.
+The Cariaco basin is one of the most productive ecosystems in the Caribbean Sea \cite{bringue_physico-chemical_2018}, with seasonal upwelling driven by trade winds during December through May and a secondary upwelling observed in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
+The basin is anoxic at depths below \qty{\sim 275}{m}, due to microbial respiration of sinking organic matter and restricted bottom water circulation, leaving a rich sedimentary record for the study of past climate \cite{hughen1996nature}.
 
-    \item - Phytoplankton is important
+    % In recent years, significant changes have been observed in phytoplankton populations within the Cariaco Basin.
+These changes are believed to be influenced by various factors, including climate variability, nutrient fluxes, and ocean circulation patterns.
+Understanding the drivers and impacts of these changes in the phytoplankton community is crucial to understand the broader implications for marine food webs and carbon cycling in this and similar ecosystems.
 
-    %“Time series of biological observations provide insights into the responses of marine ecosystems to cyclic processes, transient events, and, when maintained over many decades, changes in climate (Carstensen 2014, Henson et al.
-2016).
-Such time series are needed for assessments on the dynamic state of the ocean.” ([Muller-Karger et al., 2019, p.
-2](zotero://select/library/items/7YW7MFTU)) ([pdf](zotero://open-pdf/library/items/TTZQBAXW?page=2&annotation=R4DSZPK6))
-
-    - Phytoplankton is important
-    - Why community structure is important, taxonomic diversity (aspects of diversity) - why look at taxonomic diversity?
-
-    - some general intro to DIVERSITY.. what is it? Why important?
-    %"Whereas the positive biomass response to high nutrient and light availability is well known (Tilman 1982; Leibold 1999; Siegel et al.
-2013), the biodiversity response is much more complex.
-First, biodiversity itself is a multifaceted construct, comprising the number of different taxa (e.g.
-species richness) and the equality of species contribution to the assemblage as aspects of local alpha diversity (evenness).
-But the change in richness is only the net effect of immigration and local extinction, while the turnover in species composition with time is an additional aspect of biodiversity.
-Extending to larger spatial scales, heterogeneity in space (beta diversity) and composition of the regional species pool (gamma diversity) are further biodiversity dimensions.
-Second, for each of these facets, all metrics are effort- and scaledependent.
-Thus, in contrast to other aspects of ecosystem assessments, biodiversity metrics do not qualify to set absolute target values for good or bad status and effective ecosystem management.
-But, the relevance of the metrics arises from analysing their temporal trends over a long time period.
-Third, no single variable captures even the most important aspects of community composition and change.
-Therefore, the analysis of phytoplankton biodiversity requires a multimetric approach." from Di Cavalho et al.
-2023 (with Hillebrand)
-
-    - why are we looking at Functional Groups?
-    %"The complexity of biodiversity assessment has led to the development of other approaches addressing compositional change in plankton communities, most prominently the creation of functional groups.
-Functional groups are meant to comprise species that show similar responses to environmental drivers and share morphological and/or physiological characteristics.
-Functional group indicators have been shown to be relevant for describing community structure and biodiversity and are more comparable with other studies than species-based indicators (Mouillot et al.
-2006)." from Di Cavalho et al.
-2023 (with Hillebrand)
-
-    %% Overarching theme: Stability of Ecosystem faunctions, Drivers of community change
-    \item "Regime shifts + Diversity" - general intro on current findings related to "climate driven ecosystem shifts"
-        - %"We find that the rate of turnover in the phytoplankton community becomes faster during this century, that is, the community structure becomes increasingly unstable in response to climate change.
-Combined with alterations to phytoplankton diversity, our results imply a loss of ecological resilience, with likely knock-on effects on the productivity and functioning of the marine environment." \cite{henson_future_2021}
-        -
-
-    \item "The Cariaco Basin, located ...." - general hydrography/depth, etc.
+    "The Cariaco Basin, located ...." - general hydrography/depth, etc.
         The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}.
@@
 
-    \item "The CARIACO Time Series ...." - talk about the Time Series, what they did, why
-
-    \item "The Cariaco Basin has undergone marked shifts..." Talk about ecosystem, what has changed, reference prev literature
+    "The Cariaco Basin has undergone marked shifts..." Talk about ecosystem, what has changed, reference prev literature
 
@@
 
-    \item "Here's where we come in..." Talk about how most of the analysis stopped after 2012, time series died out, but there is more data there and particularly looking at community data, we can see a return.
+    - Why community structure is important, taxonomic diversity (aspects of diversity) - why look at taxonomic diversity?
+
+    - some general intro to DIVERSITY.. what is it? Why important?
+
+    % "Phytoplankton is the major primary producer group in marine ecosystems and at the same time a highly suitable indicator for ecosystem change.
+Due to its high diversity and short generation time, phytoplankton responds quickly to changes in the environment (Reynolds 1998).
+Especially changes in temperature and nutrient supply are known to change phytoplankton composition and biodiversity, with higher temperatures and lower nutrients favouring small species (Litchman et al.
+2007; Winder and Sommer 2012), whereas Si:N and N:P supply ratios change the relative importance of diatoms and other groups such as dino(flagellates) and cyanobacteria (Sommer et al.
+2004; Makareviciute-Fichtner et al.
+2020)." from Di Cavalho et al.
+2023
+
+    %"Whereas the positive biomass response to high nutrient and light availability is well known (Tilman 1982; Leibold 1999; Siegel et al.
+2013), the biodiversity response is much more complex.
+First, biodiversity itself is a multifaceted construct, comprising the number of different taxa (e.g.
+species richness) and the equality of species contribution to the assemblage as aspects of local alpha diversity (evenness).
+But the change in richness is only the net effect of immigration and local extinction, while the turnover in species composition with time is an additional aspect of biodiversity.
+Extending to larger spatial scales, heterogeneity in space (beta diversity) and composition of the regional species pool (gamma diversity) are further biodiversity dimensions.
+Second, for each of these facets, all metrics are effort- and scaledependent.
+Thus, in contrast to other aspects of ecosystem assessments, biodiversity metrics do not qualify to set absolute target values for good or bad status and effective ecosystem management.
+But, the relevance of the metrics arises from analysing their temporal trends over a long time period.
+Third, no single variable captures even the most important aspects of community composition and change.
+Therefore, the analysis of phytoplankton biodiversity requires a multimetric approach." from Di Cavalho et al.
+2023 (with Hillebrand)
+
+    - why are we looking at Functional Groups?
+    %"The complexity of biodiversity assessment has led to the development of other approaches addressing compositional change in plankton communities, most prominently the creation of functional groups.
+Functional groups are meant to comprise species that show similar responses to environmental drivers and share morphological and/or physiological characteristics.
+Functional group indicators have been shown to be relevant for describing community structure and biodiversity and are more comparable with other studies than species-based indicators (Mouillot et al.
+2006)." from Di Cavalho et al.
+2023 (with Hillebrand)
+
+    %% Overarching theme: Stability of Ecosystem faunctions, Drivers of community change
+    "Regime shifts + Diversity" - general intro on current findings related to "climate driven ecosystem shifts"
+        - %"We find that the rate of turnover in the phytoplankton community becomes faster during this century, that is, the community structure becomes increasingly unstable in response to climate change.
+Combined with alterations to phytoplankton diversity, our results imply a loss of ecological resilience, with likely knock-on effects on the productivity and functioning of the marine environment." \cite{henson_future_2021}
+        -
+
+        "Here's where we come in..." Talk about how most of the analysis stopped after 2012, time series died out, but there is more data there and particularly looking at community data, we can see a return.
     - what has been done (Phytoplankton - Statistics)?
     - where are the gaps?
+    - less studied aspects: Diversity + Phytoplankton Microscopy
 
     So we would hypothesize, that if wind is driving the ecosystem, then wind would be best predictor of shits.
-    Or nutrients, but if it is a climate indices, then the indices capture some variablity that is not related to wind or nutrients....
+    Or nutrients, but if it is a climate indices, then the indices capture some variability that is not related to wind or nutrients....
 
-    \item "In this paper, we..."
+    "In this paper, we..."
 
         %Highlight that I am only looking at Phytoplankton Microscopy data, and a bit of Niskin, but nothing else. Only bottom-up drivers.
-
-    \end{itemize}
+        %What are the questions I ask myself?? 1.
 
@@
 (c) Cumulative importance curves for the predictor variables.
-Black line shows cumulative importance accross predictor range for the entire model and all genera.
+Black line shows cumulative importance across predictor range for the entire model and all genera.
 Dashed colored lines represent the responses of individual functional groups.}
@@
 
+- The 20 year record of CARIACO time series is a great opportunity to better understand the connection between pelagic proccesses, climate and sedimentation.
+
+- Our main results show that...
+
+- We are not able to fully resolve....
+
+- We are analyzing long term time series of data, where probably different analysts have counted the phytoplankton samples...
+%It is well known that the personal component has an influence on how well species are identified.
+Even the same person counting the samples can evolve over time, gaining more experience, which can also influence the results (Löder et al.
+2012; Nohe et al.
+2018).
+A recommendation for further improvement of the assessment is a case wise cross-validation of some samples by the monitoring agencies involved.
+Another important issue to be considered are the differences in sampling frequency between countries and also between stations (see ESM Fig. S1).
+In highly dynamic and turbid waters such as the Wadden Sea, low frequency and infrequent sampling is not sufficient to properly capture temporal signals (Fettweis et al.
+2023).
+% from Di Cavalho et al.
+2023
+
+- The decrease in observed phytoplankton cells between the two clusters indicates...
+% talk about the general trends and what they can mean, changes in size distribution, changes in functional groups
+
+- What could have driven these shifts?
+% talk about link to bottom up drivers, but also mention top down aspects...
+then go deep on Gradient Forest analysis, what that could mean.
+Relate this to other papers about community changes, climate effects and gradient forest applications.
+
 In the Cariaco basin, wind-driven seasonal upwelling dictates the nutrient availability in the upper mixed layer, which in turn affects the phytoplankton dynamics \cite{ptacnik_diversity_2008, bopp_response_2005}.
@@
 The Caribbean experiences the most marked of the El Niño–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Niño–Southern Oscillation \cite{enfield_tropical_1997}.
+
+% Pinckney et al. 2015:
+- major thing is that previous paper only looked at microscopy to 55m (Mutshinda and Pinckney), but now we look at full 100 m!
+%The Cariaco Basin has undergone marked shifts in biogeochemistry and food web structure over the past two decades (Taylor et al., 2012; Montes et al., 2013; Bates et al., 2014; Scranton et al., 2014).
+Since the initiation of the CARIACO Ocean Time-Series, near-surface (0–55 m) chl a concentrations (measured by fluorescence of pigments extracted in methanol) and net primary production rates (NPP, measured using 14C incubations) have declined, particularly between 2004 and 2010 (Taylor et al., 2012; Scranton et al., 2014).
+This is hypothesized to be the result of a measured decrease in seasonal trade wind intensity, which has reduced coastal upwelling and led to shallower mixed layer depths (Taylor et al., 2012).
+Microscopy observations indicate that microphytoplankton assemblages (420 mm) within the euphotic zone have shifted from a community comprised mainly of diatoms, dinoflagellates, and coccolithophorids to one with increasing numbers of smaller taxa including cyanobacteria (Montes et al., 2013).
+During this period, a collapse in the sardine industry also occurred.
+The causes are not clear, but the phenomenon was accompanied by an increase in zooplankton biomass and an increase in particulate organic carbon export below the euphotic zone (Taylor et al., 2012).
+These results suggest that changes in zooplankton abundance and therefore grazing may have contributed to the decline in diatoms and to more efficient carbon export to depth.
+% Consistent with the results of Taylor et al.
+(2012), the microscopy data show large scale reductions in overall microphytoplankton (4 20 mm) cell densities and a large decrease in both diatoms and coccolithophores over the upper 55 m.
+However, the observed photopigment decline in diatom abundance is not nearly to the same extent and coccolithophores had a significant concentration increase of 60.1% over this same depth interval (Table 3).
+We argue that the reduction in cell numbers is likely only true of larger cells.
+Microscopic cell counts are inherently selective and biased toward easily identifiable species and cells that remain intact after preservation (Lund et al., 1958; Rott, 1981; Wilhelm et al., 1991; Schlüter et al., 2000).
+Furthermore, Mousing et al.
+(2014) found that small increases in temperature (as observed in the Cariaco Basin) lead to a decrease in the relative contribution of large cells in the community, regardless of ambient nutrient availability.
+Finally, depth limited microscopy data misses the downward movement of diatoms and coccolithophores deeper in the water column (i.e.
+below 56 m).
+In fact, there was also a dramatic shift in diatom speciation, with centric diatoms (Guinardia, Leptocylindrus, Skeletonema) dominating during the first half of the time-series and a mix of centric and pennate diatoms (Pseudonitzschia, Leptocylindrus, Guinardia) dominating in more recent years.
+In microscopy, smaller sample volumes may lead to underestimations of species richness or diversity (Cermeño et al., 2014).
+Nonetheless, both photopigment and microscopy measurements suggest that phytoplankton communities became more homogeneously distributed over the years, with the majority of the difference due to declining diatom populations that occur during upwelling events.
 
@@
 \section{Conclusions}
+
+- Our analysis shows that...
+
+- We also show that...
+
+- But we cannot confirm or test. so we would recommend X for the future.
 

### 2025-02-20 agujournaltemplate.tex
Focus Phase Day 4 - Intro Progress

@@
 The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them a highly suitable indicator of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
-    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term climate changes, especially when data collection spans several decades \cite{carstensen_need_2014,henson_observing_2016}.
-    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin off the coast of Venezuela between 1995 and 2017 \cite{muller-karger_scientific_2019}.
-The Cariaco basin is one of the most productive ecosystems in the Caribbean Sea \cite{bringue_physico-chemical_2018}, with seasonal upwelling driven by trade winds during December through May and a secondary upwelling observed in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
-The basin is anoxic at depths below \qty{\sim 275}{m}, due to microbial respiration of sinking organic matter and restricted bottom water circulation, leaving a rich sedimentary record for the study of past climate \cite{hughen1996nature}.
+
+    %DIVERSITY PARAGRAPH HERE? yes...
+
+    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term climate changes, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.
+    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin between 1995 and 2017 \cite{muller-karger_scientific_2019}.
+    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}.
+A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May, and a smaller secondary peak in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
+The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}.
+The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}.
+Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test-bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
+
+    % Instead of going paper by paper below, I should go topic by topic (i.e. all points made on "DIVERSITY" should be in a paragraph, and all on changes in FGs or Size Classes in another etc...)
+
+    Since the inception of the time series, the Cariaco basin has undergone marked shifts...
+
+    The time series includes multiple measurements of the phytoplankton community, which are the bulk fluorometric chlorophyll, microscopy cell counts and high performance liquid chromatography (HPLC) pigment analysis.
+Of these, all cover the entire time series except the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.
+
+    - Taylor 2012 ()
+    \citeA{taylor_ecosystem_2012} analyzed data up until 2010 and found significant decadal trends of sea surface temperature (SST) increase, upwelling intensity decrease and diminished phytoplankton bloom intensities, which the authors related to the impact of global warming.
+The phytoplankton community identified via microscopy showed a shift towards smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass.
+The authors explained the change in physical regime in the Cariaco basin with a north-easterly shift of the Inter-Tropical Convergence Zone (ITCZ) and a resulting weakening trend in Trade Winds.
+
+    - Mutshinda 2013 ()
+    Using a hierarchical Bayesian modeling approach, changes in dominant species quantified via microscopy cell counts between 1995 and 2011 were found to be best predicted by changes in temperature, pH and irradiance, and less so by salinity or macronutrient concentrations \cite{mutshinda_environmental_2013}.
+In particular, higher temperatures were associated with a lower abundance of diatoms.
+
+    - Irwin 2015 ()
+    With the same dataset of microscopy cell counts and a MaxEnt statistical modeling approach, \citeA{irwin_phytoplankton_2015} calculated realized niches for the dominant species for SST, irradiance and nitrate concentration.
+As trends in the realized niches were observed over the course of the time series, this seemed to show the adaption of species to climate change effects.
+
+    - Pinckney 2015 ()
+    The phytoplankton community was investigated in detail by \citeA{pinckney_phytoplankton_2015}, using the microscopy counts and HPLC data from the time ranges 1996 to 2000 and 2006 to 2010.
+Between these two time periods, they observed that the drastic reduction in diatoms counted using microscopy did not correspond to a similar reduction in diatom abundance calculated from the pigment composition, which is most likely explained by a shift to diatom species with smaller cell sizes.
+Additionally, the pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity and a reduction in seasonal variability in the community composition.
+These changes are most likely a response to the reduced upwelling  and a deepening of the euphotic zone associated with a decrease in the strength of seasonal trade winds.
+Interestingly, the shifts in the phytoplankton community did not result in a significant change to the export of particulate organic carbon (POC) to depth as measured by sediment traps deployed during the time series.
+
+    - Lorenzoni 2015 ()
+    \citeA{lorenzoni_characterization_2015} focused their analysis on HPLC data retrieved between 2006 and 2012 and related seasonal changes in the pigment composition to spectral absorption from remote sensing data and concentrations of organic carbon.
+They observed a high seasonal variability, with microphytoplankton (mostly diatoms) dominating upwelling periods and picophytoplankton (mostly cyanobacteria) dominating during the rainy season.
+These community compositions were also reflected in the POC:TChla ratio, which was higher for diatom dominated assemblages and lower for those dominated by cyanobacteria.
+    % "The difference between the taxonomic data and the pigment data suggests that the microscopy taxonomy may not capture the variability in the pico-phytoplankton.
+Since the beginning of the time-series in 1995, there has been a change in the abundance of phytoplankton size classes based on diagnostic pigments: micro-phytoplankton has decreased year-round, while picophytoplankton has increased, particularly during the rainy season.
+This is consistent with the decrease in larger phytoplankton previously reported for the region." Lorenzoni et al.
+2015
+
+    - AMO (ref to 2019 Review)
+    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al.
+2015).
+These climatic changes had a profound impact throughout the Atlantic (Alexander et al.
+2014, Nye et al.
+2014) and in the Cariaco basin \cite{taylor_ecosystem_2012,pinckney_phytoplankton_2015}.
+
+    - Finally what it says in Review about Phytoplankton community
+    - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!
+
+    % Hone in on Diversity Aspects, and what has been done there?
+    - Highlight diversity aspect of Pinckney
+    - Highlight Mutshinda etc. How they used Niches/Phyto data
+    - What are the questions in Diversity TS data? -> Esteban's papers
+
+    % Finally say what WE are going to do! What are the open questions, what is new?
 
@@
 Understanding the drivers and impacts of these changes in the phytoplankton community is crucial to understand the broader implications for marine food webs and carbon cycling in this and similar ecosystems.
-
-    "The Cariaco Basin, located ...." - general hydrography/depth, etc.
-        The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}.
-        A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May, and a smaller secondary peak in July and August \cite{astor_seasonal_2003}.
-        The basin is effectively separated from the Caribbean Sea by a sill with a mean depth of about 100 m.
-The water below the sill is poorly ventilated, which, combined with the high surface productivity, results in permanent anoxic conditions at depths below 275 m \cite{thunell_organic_2000}.
-
-        Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!
 
@@
     The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon.
-The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around 100 m.
-This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below 200 m depth.
+The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around \qty{100}{m}.
+This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below \qty{\sim 200}{m} depth.
 
@@
 - We are not able to fully resolve....
+% "Phytoplankton communities exhibit a great deal of variation due to a combination of regulation of growth by bottom-up factors such as resource availability (Grover 1997), variation in mortality due to grazing, viral and parasitoid attack (Alpine & Cloern 1992; Mu ̈ hling et al.
+2005), and demographic drift among species (neutral variation).
+Bottom-up factors can be complex, depending on nutrient supply rates and ratios (Tilman 1982), light availability and photoinhibition (Six et al.
+2007; Alderkamp et al.
+2010), species-specific competitive ability for resources (Litchman 2007) and other factors influencing physiological responses such as temperature (Eppley 1972).
+Variation in mortality is much less frequently measured than the factors promoting growth due to the technical challenges in making measurements.
+None of our environmental variables are good proxies of grazing rates or whether the abundance of a particular population is determined by grazer control or resource supply at a particular time.
+We have ident" from Mutshinda et al.
+2013
+
+%"Taylor et al.
+2012 observed that the decline in cell abundances was disproportionate compared with the smaller decrease in total chlorophyll a, which implies that phytoplankton of smaller size were not detected in the surveys..
+This was supported by high-performance liquid chromatography analyses, which indicated a reduction in diatom concentrations and an increase in phytoplankton diversity.
+A return to a phytoplankton community structure similar to that observed in the late 1990s was observed in 2014 and 2015 (Figure 5a), as upwelling intensity increased to levels similar to those seen before 2003." from 2019 Review Legacy of CARIACO
 
@@
 % Pinckney et al. 2015:
-- major thing is that previous paper only looked at microscopy to 55m (Mutshinda and Pinckney), but now we look at full 100 m!
+- major thing is that previous paper only looked at microscopy to 55m (Mutshinda to 75m and Pinckney), but now we look at full 100 m!
 %The Cariaco Basin has undergone marked shifts in biogeochemistry and food web structure over the past two decades (Taylor et al., 2012; Montes et al., 2013; Bates et al., 2014; Scranton et al., 2014).

### 2025-02-23 agujournaltemplate.tex
Focus Phase Day 7 Commit - restructure to final draft of Intro

@@
 
+    % Phytoplankton is important
     Phytoplankton communities provide the foundation of marine food webs and play a crucial role in global biogeochemical cycles \cite{falkowski_biogeochemical_1998}.
-These communities consist of a diverse set of species that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}.
+    Phytoplankton consists of a diverse set of photoautotrophic microorganisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}.
 The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them a highly suitable indicator of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
 
-    %DIVERSITY PARAGRAPH HERE? yes...
-
+    - Phytoplankton communities can be described from taxonomic/functional group aspects, but also from biodiversity aspects (or trait-based, but this I am not doing here)
+
+    % Phytoplankton is under threat
+    On a global scale, human activities are driving a decline in biodiversity \cite{tittensor_mid-term_2014}.
+Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}.
+
+    %%% Phytoplankton functional group shifts through climate change
+    %%% Phytoplankton diversity shifts through climate change
+
+    The loss of biodiversity is predicted to have a lasting impact on the functioning of an ecosystem, since links have been found between the richness of a phytoplankton community and the biomass and resource use efficiency \cite{ptacnik_diversity_2008, lehtinen_phytoplankton_2017}.
+
+    - mention BDEF in Tropics paper, or some other paper, to highlight that the tropics are notoriously understudied in relation to phytoplankton diversity.
+
+    %%% This paper is clearly not BDEF, but environmental impacts on diversity and community!section
+
+    % We need time series in order to understand effects and changes (threats) 
     Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term climate changes, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.
+
+    % CARIACO is one such time series
     One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin between 1995 and 2017 \cite{muller-karger_scientific_2019}.
+
+    % Describe CARIACO location
     The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}.
@@
 The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}.
-Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test-bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
-
-    % Instead of going paper by paper below, I should go topic by topic (i.e. all points made on "DIVERSITY" should be in a paragraph, and all on changes in FGs or Size Classes in another etc...)
+Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
+
+    % Describe CARIACO changes in ecosystem
+
+    %% Changes in Functional Groups
+
+    %% Changes in Diversity
+
+    % Now we come into the picture: Which gap are we adressing?
+
+    %% in depth look at biodiversity aspect, in particular microscopy data
+
+    %% try to identify drivers of shift in community with Gradient Forest
+
+    %%% Instead of going paper by paper below, I should go topic by topic (i.e. all points made on "DIVERSITY" should be in a paragraph, and all on changes in FGs or Size Classes in another etc...)
 
@@
 
-    The time series includes multiple measurements of the phytoplankton community, which are the bulk fluorometric chlorophyll, microscopy cell counts and high performance liquid chromatography (HPLC) pigment analysis.
-Of these, all cover the entire time series except the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.
-
+    The time series includes multiple measurements of the phytoplankton community, which are the bulk fluorometric chlorophyll, microscopy cell counts, and high-performance liquid chromatography (HPLC) pigment analysis.
+Of these, all cover the entire time series except for the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.
+
+    The most often measured indicator of bulk phytoplankton biomass is fluorometric Chlorophyll \textit{a} concentration, which gives a general measure of biomass standing stock and distribution in the water column.
+A more advanced methodology, high-performance liquid chromatography (HPLC), allows individual pigment types to be quantified.
+These pigment distributions can be used to estimate the groups of phytoplankton that contribute to biomass.
+The most specific methodology to quantify phytoplankton communities is the counting and identification of individual phytoplankton cells, either manually under a microscope or automated in flow-through systems.
+The major drawbacks are the size limit of resolving smaller cells and, if done manually, the complexity of identifying phytoplankton species.
+
+    % PAPER TIMELINE:
     - Taylor 2012 ()
-    \citeA{taylor_ecosystem_2012} analyzed data up until 2010 and found significant decadal trends of sea surface temperature (SST) increase, upwelling intensity decrease and diminished phytoplankton bloom intensities, which the authors related to the impact of global warming.
+    \citeA{taylor_ecosystem_2012} analyzed data up to 2010 and found significant decadal trends of increase in sea surface temperature (SST), coinciding with a decrease in upwelling intensity and phytoplankton bloom intensities, which the authors related to the impact of global warming.
 The phytoplankton community identified via microscopy showed a shift towards smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass.
@@
 
-    - Finally what it says in Review about Phytoplankton community
+    - Finally, says in Review about Phytoplankton community
     - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!
@@
         %What are the questions I ask myself?? 1.
+
+    % INTRO TEXT DUMP:
+
+    Biodiversity is a complex concept composed of genetic, taxonomic, and ecological aspects that can additionally be quantified in temporal and spatial scales.
+In its basic sense, a measure of diversity should quantify two components: the number of items and their relative frequencies \cite{borics_freshwater_2021}.
+    Richness, the number of species or genera present at a particular time in a sample, has been established as the most basic descriptor of community diversity, but it is not the best indicator of ecosystem change \cite{hillebrand_biodiversity_2018}.
+    - Other metrics, evenness, and dominance
 
@@
 2023
+- Discuss limitation of taxonomic diversity, i.e. not direct translation to functional diversity, but here no Data foundation to look at functional diversity.. (inspired by Borics et al. 2021)
 

### 2025-02-28 agujournaltemplate.tex
Focus Phase Day 12 - Intro almost done, start Discussion

@@
     % Phytoplankton is important
-    Phytoplankton communities provide the foundation of marine food webs and play a crucial role in global biogeochemical cycles \cite{falkowski_biogeochemical_1998}.
-    Phytoplankton consists of a diverse set of photoautotrophic microorganisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}.
+    Phytoplankton provides the foundation for marine food webs and plays a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
+    Phytoplankton consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}.
 The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them a highly suitable indicator of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
-
-    - Phytoplankton communities can be described from taxonomic/functional group aspects, but also from biodiversity aspects (or trait-based, but this I am not doing here)
+    The structure of the community can be considered an integrator of environmental factors, as it is strongly determined by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}.
+    %%% Mention Top-Down and Bottom-Up forces structuring communities and affection biodiversity.
 
@@
 
-    The loss of biodiversity is predicted to have a lasting impact on the functioning of an ecosystem, since links have been found between the richness of a phytoplankton community and the biomass and resource use efficiency \cite{ptacnik_diversity_2008, lehtinen_phytoplankton_2017}.
-
-    - mention BDEF in Tropics paper, or some other paper, to highlight that the tropics are notoriously understudied in relation to phytoplankton diversity.
-
-    %%% This paper is clearly not BDEF, but environmental impacts on diversity and community!section
+    The loss of biodiversity is predicted to have a lasting impact on the functioning of an ecosystem, since links have been found between the richness of a phytoplankton community and ecosystem functions, such as biomass and resource use efficiency \cite{ptacnik_diversity_2008, lehtinen_phytoplankton_2017}.
+Such links need to be further investigated, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014} and under greater threat of extinction, when compared to temperate regions \cite{finnegan_paleontological_2015}.
+    % DON'T FORGET TO MENTION CLIMATE IMPACTS (as I investigate AMO linkage and ENSO MEI etc.)
+    % this would actually make the most sense here!
 
@@
     The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}.
-A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May, and a smaller secondary peak in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
 The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}.
@@
 Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
+    %%% mention climate impacts (AMO + ENSO) here!
+
+    % Describe general seasonal cycle with functional groups etc.
+    A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward.
+A smaller wind-driven secondary upwelling often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
+During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids and nanoplankton also contributing.
+In the low-wind rainy season, the Cariaco basin exhibits oligotrophic conditions, where the portion of large phytoplankton cells is diminished, but nano- and picophytoplankton are present in considerable numbers throughout the year \cite{lorenzoni_characterization_2015}.
 
     % Describe CARIACO changes in ecosystem
-
+    Since the inception of the time series in 1995, the ecosystem of the Cariaco basin has undergone marked changes.
+\citeA{taylor_ecosystem_2012} analyzed time series data up to 2010 and found significant decadal trends of increase in sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors related to the impact of global warming.
+The phytoplankton community identified by microscopy showed a drastic reduction in Diatom abundance and a shift toward smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass.
+    The trends in microscopy community data were confirmed and expanded by \citeA{pinckney_phytoplankton_2015}.
+They observed that the drastic reduction in diatoms counted using microscopy corresponded, but to a lesser degree, to diatom abundance calculated from the pigment composition measured by high-performance liquid chromatography (HPLC).
+This is most likely explained by a shift to diatom species with smaller cell sizes, which are more difficult to detect and identify using light microscopy \cite{lorenzoni_characterization_2015}.
+In addition, the pigment data indicate a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability in the community composition.
+These changes are most likely a response to the reduced upwelling and a deepening of the euphotic zone associated with a decrease in the strength of seasonal trade winds.
+Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.
+    %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
+    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift.
+Data past 2012 until the end of the CARIACO time series in 2017, which was not included in previous studies of the phytoplankton community, indicates a return to stronger upwelling intensity and similar productivity levels to the late 1990s \cite{muller-karger_scientific_2019}.
+
+    % NOw we come in!
+    - The question arises, what were the drivers of change in phytoplankton community?
+    - Bottom-up or top-down processes? response to regional changes in oceanographic conditions, in turn driven by climate change or oscillations?
+    - Highlight, that there is a large mismatch in data-coverage particularly for Top-Down aspects, fisheries is landing data (hard to quantify anthropogenic impact on this data and ecosystem) and Zooplankton data is only from after 2002.
+Also HPLC data is lacking in coverage.
+    - We focus on Bottom-up, given the available data, and try to identify changes in biodiversity and community and relate these to both local effects and climatic drivers.
+
+    - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!
+
+    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+    % "Changes in plankton biodiversity in response to regional changes in oceangraphic conditions" driven by climate oscillaitons (from REVIEW 2019).
     %% Changes in Functional Groups
@@
 
-    %% in depth look at biodiversity aspect, in particular microscopy data
+    % FOCUS ON Community Changes + BOTTOM UP
+
+    These changes are believed to be influenced by various factors, including climate variability, nutrient fluxes, and ocean circulation patterns.
+    % WHAT ARE THE DRIVERS OF COMMUNITY CHANGE? Bottom up OR top down, we focus on bottom up...
+
+    Describe physical changes!
+
+    - AMO (ref to 2019 Review)
+    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al.
+2015).
+These climatic changes had a profound impact throughout the Atlantic (Alexander et al.
+2014, Nye et al.
+2014) and in the Cariaco basin \cite{taylor_ecosystem_2012,pinckney_phytoplankton_2015}.
+    - ENSO teleconnection (ref to 2019 Review)
+    The Caribbean experiences the most marked of the El Ni  ̃no–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Ni  ̃no–Southern Oscillation (Enfield & Mayer 1997).
+
+     - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!
+
+    % Finally say what WE are going to do! What are the open questions, what is new?
 
@@
 
-    - AMO (ref to 2019 Review)
-    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al.
-2015).
-These climatic changes had a profound impact throughout the Atlantic (Alexander et al.
-2014, Nye et al.
-2014) and in the Cariaco basin \cite{taylor_ecosystem_2012,pinckney_phytoplankton_2015}.
-
-    - Finally, says in Review about Phytoplankton community
+    - Finally, add the info from Review about Phytoplankton community, about community and upwelling returning in 2014 to previous levels/state.
     - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!
@@
     - What are the questions in Diversity TS data? -> Esteban's papers
-
-    % Finally say what WE are going to do! What are the open questions, what is new?
 
@@
 \section{Discussion}
-This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem, that is driven by local environmental variability, which is in turn affected by large scale climatic shifts.
-In the Cariaco basin, long-term climatic oscillations have affected the upwelling regime and ...
-
-- The 20 year record of CARIACO time series is a great opportunity to better understand the connection between pelagic proccesses, climate and sedimentation.
+This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes.
+In the Cariaco basin, long-term climatic oscillations have affected the upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
 
 - Our main results show that...
+1. we see Overlap in yearly mean anomalies between certain variables
+
+2. mismatch between Diversity variables, and cell count abundances + community clustering / ecosystem shift
+
+3. Clustering shows return of community to previous cluster, confirmed by monthly NMDS, but EnvFit is interesting, as no variable clearly goes along gradient separating the two cluster (NMDS2)
+
+3.1 Turnover is different between two clusters, also corresponding to evennness between clusters
+
+3.2 Discuss which variables actually vary between clusters and how they could mechanistically explain difference between communities
+
+4. Gradient Forest clearly identifies climate indices as best predictors of community change, then nutrients etc.
+This is really interesting, as it would be expected, that in situ-measurements are best predictor of community.
+BUT either the community + top down processes are capturing more of that variance, or we simply are missing a bottom-up variable that can explain it.
+but that most likely requires mechanistic modeling to investigate further...
+
+5. Compare to what happened at other long term time series in the ocean...
 

### 2025-02-28 text-backup.tex
Focus Phase Day 12 - Intro almost done, start Discussion

--- 2025-02-23 agujournaltemplate.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% AGUJournalTemplate.tex: this template file is for articles formatted with LaTeX
-%
-% This file includes commands and instructions
-% given in the order necessary to produce a final output that will
-% satisfy AGU requirements, including customized APA reference formatting.
-%
-% You may copy this file and give it your
-% article name, and enter your text.
-%
-% guidelines and troubleshooting are here:
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BELOW HERE ARE INTRO DRAFT TEXTS! %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%% user added for convenience, can be removed before submission:
-
-%% To submit your paper:
-\documentclass[draft]{agujournal2019}
-\usepackage{url} %this package should fix any errors with URLs in refs.
-\usepackage{lineno}
-\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
-\usepackage{soul}
-
-% MY OWN ADDED PACKAGES TO RENDER UNITS
-\usepackage{gensymb}
-\usepackage{booktabs}
-\usepackage{siunitx}
-
-\linenumbers
-%%%%%%%
-% As of 2018 we recommend use of the TrackChanges package to mark revisions.
-% The trackchanges package adds five new LaTeX commands:
-%
-%  \note[editor]{The note}
-%  \annote[editor]{Text to annotate}{The note}
-%  \add[editor]{Text to add}
-%  \remove[editor]{Text to remove}
-%  \change[editor]{Text to remove}{Text to add}
-%
-% complete documentation is here: http://trackchanges.sourceforge.net/
-%%%%%%%
-
-\draftfalse
-
-%% Enter journal name below.
-%% Choose from this list of Journals:
-%
-% JGR: Atmospheres
-% JGR: Biogeosciences
-% JGR: Earth Surface
-% JGR: Oceans
-% JGR: Planets
-% JGR: Solid Earth
-% JGR: Space Physics
-% Global Biogeochemical Cycles
-% Geophysical Research Letters
-% Paleoceanography and Paleoclimatology
-% Radio Science
-% Reviews of Geophysics
-% Tectonics
-% Space Weather
-% Water Resources Research
-% Geochemistry, Geophysics, Geosystems
-% Journal of Advances in Modeling Earth Systems (JAMES)
-% Earth's Future
-% Earth and Space Science
-% Geohealth
-%
-% ie, \journalname{Water Resources Research}
-
-\journalname{Enter journal name here}
-
-\begin{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%  TITLE
-%
-% (A title should be specific, informative, and brief. Use
-% abbreviations only if they are defined in the abstract. Titles that
-% start with general keywords then specific terms are optimized in
-% searches)
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Example: \title{This is a test title}
-
-\title{Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  AUTHORS AND AFFILIATIONS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Authors are individuals who have significantly contributed to the
-% research and preparation of the article. Group authors are allowed, if
-% each author in the group is separately identified in an appendix.)
-
-% List authors by first name or initial followed by last name and
-% separated by commas. Use \affil{} to number affiliations, and
-% \thanks{} for author notes.
-% Additional author notes should be indicated with \thanks{} (for
-% example, for current addresses).
-
-% Example: \authors{A. B. Author\affil{1}\thanks{Current address, Antartica}, B. C. Author\affil{2,3}, and D. E.
-% Author\affil{3,4}\thanks{Also funded by Monsanto.}}
-
-\authors{B. Post\affil{1,2}, E. Acevedo-Trejos\affil{3}, S. Chakraborty\affil{1}, A.D. Barton\affil{4}, A. Merico\affil{1}}
-
-% \affiliation{1}{First Affiliation}
-% \affiliation{2}{Second Affiliation}
-% \affiliation{3}{Third Affiliation}
-% \affiliation{4}{Fourth Affiliation}
-
-\affiliation{1}{Systems Ecology Group, Leibniz Centre for Tropical Marine Research (ZMT), Bremen, Germany}
-\affiliation{2}{School of Science, Constructor University, Bremen, Germany}
-\affiliation{3}{Earth Surface Process Modelling, GFZ German Research Centre for Geosciences, Potsdam, Germany}
-\affiliation{4}{Scripps Institution of Oceanography and Department of Ecology, Behavior and Evolution, University of California San Diego, La Jolla, CA, United States}
-%(repeat as many times as is necessary)
-
-% Corresponding author mailing address and e-mail address:
-
-% (include name and email addresses of the corresponding author.  More
-% than one corresponding author is allowed in this LaTeX file and for
-% publication; but only one corresponding author is allowed in our
-% editorial system.)
-
-% Example: \correspondingauthor{First and Last Name}{email@address.edu}
-
-\correspondingauthor{Benjamin Post}{benjaminpost@aoop.de}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% KEY POINTS
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%  List up to three key points (at least one is required)
-%  Key Points summarize the main points and conclusions of the article
-%  Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences
-
-% Example:
-% \begin{keypoints}
-% \item	List up to three key points (at least one is required)
-% \item	Key Points summarize the main points and conclusions of the article
-% \item	Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences
-% \end{keypoints}
-
-\begin{keypoints}
-\item The CARIACO time series showed a highly variable phytoplankton community over the course of sampling between 1995 and 2017.
-Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the anomalies in the AMO index.
-% \index Previous studies proposed a shift in the community, but with the full dataset we see a shift and a return.
-The community shows a departure form cluster one around 2004, but a return to that community in 2014.
-\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, although community evenness was higher and community turnover was lower in the second cluster (low-upwelling conditions).
-\item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem.
-The next strongest predictor is the MEI v2 index (lag of 4 months), with in-situ temperature and nitrate concentration following behind.
-\end{keypoints}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  ABSTRACT and PLAIN LANGUAGE SUMMARY
-%
-% A good Abstract will begin with a short description of the problem
-% being addressed, briefly describe the new data or analyses, then
-% briefly states the main conclusion(s) and how they are supported and
-% uncertainties.
-
-% The Plain Language Summary should be written for a broad audience,
-% including journalists and the science-interested public, that will not have 
-% a background in your field.
-%
-% A Plain Language Summary is required in GRL, JGR: Planets, JGR: Biogeosciences,
-% JGR: Oceans, G-Cubed, Reviews of Geophysics, and JAMES.
-% see http://sharingscience.agu.org/creating-plain-language-summary/)
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%% \begin{abstract} starts the second page
-
-\begin{abstract}
-The structure of the phytoplankton community was identified using microscopy in the Cariaco basin, in Venezuela, between 1995 and 2017.
-Microscopy data are limited to larger cell sizes (> 20 µm), where the community showed a large change in the abundances of functional groups and shifts in diversity patterns.
-\end{abstract}
-
-%1
-%\section*{Plain Language Summary}
-%Enter your Plain Language Summary here or delete this section.
-%Here are instructions on writing a Plain Language Summary: 
-%https://www.agu.org/Share-and-Advocate/Share/Community/Plain-language-summary
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  BODY TEXT
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%% Suggested section heads:
-\section{Introduction}
-%
-% The main text should start with an introduction. Except for short
-% manuscripts (such as comments and replies), the text should be divided
-% into sections, each with its own heading.
-
-% Headings should be sentence fragments and do not begin with a
-% lowercase letter or number. Examples of good headings are:
-
-    % Phytoplankton is important
-    Phytoplankton communities provide the foundation of marine food webs and play a crucial role in global biogeochemical cycles \cite{falkowski_biogeochemical_1998}.
-    Phytoplankton consists of a diverse set of photoautotrophic microorganisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}.
-The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them a highly suitable indicator of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
-
-    - Phytoplankton communities can be described from taxonomic/functional group aspects, but also from biodiversity aspects (or trait-based, but this I am not doing here)
-
-    % Phytoplankton is under threat
-    On a global scale, human activities are driving a decline in biodiversity \cite{tittensor_mid-term_2014}.
-Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}.
-
-    %%% Phytoplankton functional group shifts through climate change
-    %%% Phytoplankton diversity shifts through climate change
-
-    The loss of biodiversity is predicted to have a lasting impact on the functioning of an ecosystem, since links have been found between the richness of a phytoplankton community and the biomass and resource use efficiency \cite{ptacnik_diversity_2008, lehtinen_phytoplankton_2017}.
-
-    - mention BDEF in Tropics paper, or some other paper, to highlight that the tropics are notoriously understudied in relation to phytoplankton diversity.
-
-    %%% This paper is clearly not BDEF, but environmental impacts on diversity and community!section
-
-    % We need time series in order to understand effects and changes (threats) 
-    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term climate changes, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.
-
-    % CARIACO is one such time series
-    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin between 1995 and 2017 \cite{muller-karger_scientific_2019}.
-
-    % Describe CARIACO location
-    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}.
-A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May, and a smaller secondary peak in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
-The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}.
-The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}.
-Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
-
-    % Describe CARIACO changes in ecosystem
-
+    % "Changes in plankton biodiversity in response to regional changes in oceangraphic conditions" driven by climate oscillaitons (from REVIEW 2019).
     %% Changes in Functional Groups
@@
 
-    %% in depth look at biodiversity aspect, in particular microscopy data
+    % FOCUS ON Community Changes + BOTTOM UP
+
+    These changes are believed to be influenced by various factors, including climate variability, nutrient fluxes, and ocean circulation patterns.
+    % WHAT ARE THE DRIVERS OF COMMUNITY CHANGE? Bottom up OR top down, we focus on bottom up...
+
+    Describe physical changes!
+
+    - AMO (ref to 2019 Review)
+    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al.
+2015).
+These climatic changes had a profound impact throughout the Atlantic (Alexander et al.
+2014, Nye et al.
+2014) and in the Cariaco basin \cite{taylor_ecosystem_2012,pinckney_phytoplankton_2015}.
+    - ENSO teleconnection (ref to 2019 Review)
+    The Caribbean experiences the most marked of the El Ni  ̃no–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Ni  ̃no–Southern Oscillation (Enfield & Mayer 1997).
+
+     - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!
+
+    % Finally say what WE are going to do! What are the open questions, what is new?
 
     %% try to identify drivers of shift in community with Gradient Forest
+
+    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
     - Other metrics, evenness, and dominance
-
-\section{Materials and Methods}
-% Here is text on Materials and Methods.
-%
-\subsection{Study Site}
-    %Talk about the Basin shortly, then about the time series
-    The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon.
-The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around \qty{100}{m}.
-This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below \qty{\sim 200}{m} depth.
-
-\subsection{In-Situ Data}
-    Samples were collected as part of the Cariaco Ocean Time Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
-    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this manuscript.
-Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela.
-Details of the history of sampling procedures in CARIACO, as well as sampling schemes, analytical methods, and data quality control and inter-calibration procedures, have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.
-
-    All samples were collected using a 12-bottle rosette system equipped with a CTD (SeaBird, model SBE 25) \cite{astor_yrene_m_handbook_2013}.
-Water samples for chlorophyll a and nutrient analysis were taken monthly during the first cast of each cruise.
-These samples were collected at eight standard depths within the upper 100 meters (1, 7, 15, 25, 35, 55, 75, and 100 m).
-
-    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to the present by Dr.
-K.
-Fanning (University of South Florida, St.
-Petersburg, FL).
-Samples from eight cruises were analyzed by both laboratories.
-Data for $PO_4$ and $SiO_4$ were in strong agreement (within ±5\%), whereas the $NO_x$ and $NH_4$ measurements showed discrepancies.
-\cite{taylor_ecosystem_2012}.
-For our analysis, we used a merged dataset for all nutrients, where coincidental monthly measurements were averaged.
-
-    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate.
-Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period.
-Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
-    %from Mutshinda et al. 2013
-    Quality control the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level, correcting spelling variations and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}).
-After processing the data, 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.
-
-    Data sets for Niskin bottle samples, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024).
-    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (more than three months) in coverage.
-Therefore, we did not include Zooplankton net tow and HPLC pigment data in our analysis.
-Our study instead focused on the phytoplankton microscopy data.
-We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity.
-In total, we used a complete set of 254 monthly data points, on which all variables were available.
-
-\subsection{Climate data}
-    %from Taylor et al. 2012
-    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives near-shore upwelling of nutrient-rich waters by Ekman transport \cite{taylor_ecosystem_2012}.
-We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling coordinates (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024).
-
-    The monthly AMO index data (10 year lowpass), which is as an area average of detrended low-pass filtered North-Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} (accessed in June, 2024).
-
-    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of the surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024).
-In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).
-
-    \subsubsection{Data processing}
-    % Niskin, CTD and Phytoplankton data processing
-    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with tidyverse packages \cite{wickham_welcome_2019}.
-All data processing and analysis code is publicly available on GitHub \url{https://github.com/ben1post/cariaco-ms2}.
-%Note: Repo currently still set to private
-    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S.
-National Oceanographic Data Center \cite{johnson2006world}.
-The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis.
-For phytoplankton microscopy data, individual identified taxonomic units (Species or Genus) were interpolated, integrated, and then grouped by Genus via the sum of interpolated counts.
-
-\subsection{Data analysis}    
-    To give a general overview of the data and trends in the time series, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$).
-Cruises from 1995 and 2017 were excluded, as there were too few measurements to provide a reliable yearly mean estimate.
-
-    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to make the data set more homogeneous and robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008}.
-Furthermore, we calculate the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
-
-    \subsubsection{Phytoplankton Community Structure Analysis}
-    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth.
-Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}.
-We used the Bray distance matrix in a 2-dimensional ordination that reached a converging solution within 30 steps.
-To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations.
-The arrows indicate the gradient of a 2-dimensional linear trend surface.
-    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}.
-%Here I want to add "significant diferences calculation" using e.g.
-ANOVA.
-    % detailed explanation of NMDS in Winder and Hunter 2008
-
-    \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
-    To quantify the effects of bottom-up drivers in the Cariaco basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis.
-Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}.
-Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor.
-The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover.
-Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}.
-This analysis was conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data.
-The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, the -u 10m wind component, the \qty{21}{\celsius} isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$).
-The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology takes into account by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}.
-To consider the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all climate variables, as well as 3 months lags for in-situ measurements in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run (see Table \ref{app:table1} in the Appendix).
-
-\section{Draft Results}
-
-% Figure 1
-The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}.
-The patterns emerging from the -u 10m wind component, the temperature and the directly related \qty{21}{\celcius} isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m.
-These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes.
-The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing inter-annual oscillations.
-Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration.
-Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.
-
-% Figure 2
-To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between seasons.
-The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters.
-Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$).
-The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters.
-In some years, particularly during 2000, 2005, 2007, 2008 and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$).
-In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012.
-The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value.
-Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014.
-Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series.
-In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
-The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter.
-Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011.
-Scaling this genus diversity by the cell count abundances, i.e.
-the Shannon index, shows a similar but less pronounced dynamic.
-Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016.
-The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$), yet there is a consistent pattern of higher Shannon diversity at lower depths for all seasons except for Fall.
-Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$).
-The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value.
-There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016.
-Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.
-
-% Figure 3
-Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment.
-For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements.
-Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix.
-The community is separated into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013.
-Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, turnover is higher again.
-This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years.
-To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$).
-This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends.
-The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index.
-The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed.
-Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering.
-Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters.
-Upwelling related variables show distinct patterns between the two clusters.
-For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celcius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate.
-The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies.
-Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences.
-For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.
-
-% Figure 4
-To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
-Preliminary model runs were used to identify if time lags of the predictor variables could better explain the data (see Appendix), and this resulted in choosing the AMO with a lag of 5 months and MEI v.2 with a lag of 4 months.
-In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the highest predictive capability for the climate indices with time lags, and upwelling related variables such as nitrate in the upper 100 meters and wind speed scoring slightly lower.
-The \qty{21}{\celsius} Isotherm and silicate and phosphate concentrations scored lowest in predicting shifts within the community.
-The way Gradient Forest predicts shifts in the community, is by building a random forest model for each observed genus across the time series.
-The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$.
-The microscopy data is biased towards larger cell sizes, and therefore most of the predicted genus cell counts are from diatoms.
-However, two Haptophytes and four Dinoflagellates, as well as the Cyanobacteria \textit{Synechococcus} do rank highly.
-Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range.
-The black line indicates the cumulative importance across the entire community, and the colored dashed lines indicate the importance per phytoplankton group.
-For the AMO index, we see a clear pattern that negative anomalies highly affect the community.
-This contrasts with a higher importance at positive anomalies in the MEI v.2 index, in particular for Haptophytes.
-Nitrate concentrations between 6 and 8 \unit{\micro \mole} is where the community changed most, which indicates these values as a threshold for the composition of the community.
-For wind speed, temperature and salinity, changes occurred most at parameter ranges related to high upwelling, i.e.
-high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}) and high salinity (above \qty{36.8}{PSU}).
-One interesting aspect is that for both climate indices, Haptophytes showed the strongest cumulative importance, but for Diatoms the nitrate concentration seem to be a stronger predictor.
-
-\begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/Artboard 1.pdf}
-\caption{Map of CARIACO Time Series location in the Cariaco basin in Venezuela in the Caribbean Ocean, located at \ang{10.50}N, \ang{64.67}W}
-\label{fig:map}
-\end{figure}
-
-\begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/PLOTZScores_updated.pdf}
-\caption{Heatmap of yearly average values, normalized for comparison as z-scores.
-Red indicates a larger value, blue indicates a smaller value for that particular year.
-The years 1995 (n=2) and 2017 (n=1) were removed due to lack of data, additionally we marked years with two or less and three or more missing values.
-Diversity Indices are calculated based on microscopy counts at the Genus level.}
-\label{fig:zscore}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\noindent\includegraphics[width=480]{fig/DepthDivTimeSeries_updated.pdf}
-\end{center}
-\caption{Time series and seasonality of chlorophyll and diversity indices across four depth intervals.
-(a,c,e,g) Annual mean across time series for years with more than 5 monthly sampling.
-Transparent dots show monthly raw values.
-(b,d,f,h) Boxplots of monthly measurements across four depth intervals grouped by season.}
-\label{fig:divts}
-\end{figure}
-
-\begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot_NEW 2.pdf}
-\caption{(a) Clustering of microscopy data at the Genus level, using the binary Jaccard distance.
-The community for each year is binned together, to smooth out seasonal aspects.
-(b) Species turnover between years for annual aggregate of identified phytoplankton to a genus level.
-(c) NMDS plot of monthly community data (cell counts) that is cube-root transformed.
-Stress: 0.261, 2 dimensions, converging solution.
-Color of dots based on which yearly cluster the monthly data is in.
-Vector overlay of ENVFIT shows underlying environmental variables that are co-varying with the community data.
-Plots (d-o) below show density distributions of the environmental variables as well as diversity indices between the clustered years.
-Dashed line shows the median and significance levels of Wilcoxon sum rank tests between clusters are shown above plots.}
-\label{fig:clustering}
-\end{figure}
-
-\begin{table}
-\caption{Wilcoxon sum rank test with continuity correction for variables between the two clusters.
-Values for in-situ data are interpolated across the top 100 meters for each individual monthly cruise sampling.}
-\label{table:clustering}
-\centering
-\begin{tabular}[t]{lrrl}
-\toprule
-Variable & W & p.value & Significance\\
-\midrule
-AMO & 3494.0 & 1.86e-14 & ***\\
-MEI v.2 & 9523.5 & 0.006 & ***\\
-u-component 10 m wind speed & 7074.0 & 0.143 & n.s.\\
-Isotherm \qty{21}{\celcius} & 5293.0 & 0.034 & **\\
-\addlinespace
-Temperature & 4763.0 & 0.003 & ***\\
-$NO_3$ & 6378.0 & 0.018 & **\\
-$PO_4$ & 7359.0 & 0.012 & **\\
-$SiO_4$ & 5998.0 & 0.736 & n.s.\\
-Salinity & 7636.0 & 0.002 & ***\\
-\addlinespace
-Genus Richness & 5913.0 & 0.954 & n.s.\\
-Shannon Index & 5142.0 & 0.108 & n.s.\\
-Pielou Index & 4874.0 & 0.029 & **\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3_new.pdf}
-\caption{Gradient Forest analysis of monthly community data at a genus level.
-(a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances.
-(b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance.
-(c) Cumulative importance curves for the predictor variables.
-Black line shows cumulative importance across predictor range for the entire model and all genera.
-Dashed colored lines represent the responses of individual functional groups.}
-\label{fig:GF}
-\end{figure}
-
-\section{Discussion}
-This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem, that is driven by local environmental variability, which is in turn affected by large scale climatic shifts.
-In the Cariaco basin, long-term climatic oscillations have affected the upwelling regime and ...
-
-- The 20 year record of CARIACO time series is a great opportunity to better understand the connection between pelagic proccesses, climate and sedimentation.
-
-- Our main results show that...
-
-- We are not able to fully resolve....
-% "Phytoplankton communities exhibit a great deal of variation due to a combination of regulation of growth by bottom-up factors such as resource availability (Grover 1997), variation in mortality due to grazing, viral and parasitoid attack (Alpine & Cloern 1992; Mu ̈ hling et al.
-2005), and demographic drift among species (neutral variation).
-Bottom-up factors can be complex, depending on nutrient supply rates and ratios (Tilman 1982), light availability and photoinhibition (Six et al.
-2007; Alderkamp et al.
-2010), species-specific competitive ability for resources (Litchman 2007) and other factors influencing physiological responses such as temperature (Eppley 1972).
-Variation in mortality is much less frequently measured than the factors promoting growth due to the technical challenges in making measurements.
-None of our environmental variables are good proxies of grazing rates or whether the abundance of a particular population is determined by grazer control or resource supply at a particular time.
-We have ident" from Mutshinda et al.
-2013
-
-%"Taylor et al.
-2012 observed that the decline in cell abundances was disproportionate compared with the smaller decrease in total chlorophyll a, which implies that phytoplankton of smaller size were not detected in the surveys..
-This was supported by high-performance liquid chromatography analyses, which indicated a reduction in diatom concentrations and an increase in phytoplankton diversity.
-A return to a phytoplankton community structure similar to that observed in the late 1990s was observed in 2014 and 2015 (Figure 5a), as upwelling intensity increased to levels similar to those seen before 2003." from 2019 Review Legacy of CARIACO
-
-- We are analyzing long term time series of data, where probably different analysts have counted the phytoplankton samples...
-%It is well known that the personal component has an influence on how well species are identified.
-Even the same person counting the samples can evolve over time, gaining more experience, which can also influence the results (Löder et al.
-2012; Nohe et al.
-2018).
-A recommendation for further improvement of the assessment is a case wise cross-validation of some samples by the monitoring agencies involved.
-Another important issue to be considered are the differences in sampling frequency between countries and also between stations (see ESM Fig. S1).
-In highly dynamic and turbid waters such as the Wadden Sea, low frequency and infrequent sampling is not sufficient to properly capture temporal signals (Fettweis et al.
-2023).
-% from Di Cavalho et al.
-2023
-- Discuss limitation of taxonomic diversity, i.e. not direct translation to functional diversity, but here no Data foundation to look at functional diversity.. (inspired by Borics et al. 2021)
-
-- The decrease in observed phytoplankton cells between the two clusters indicates...
-% talk about the general trends and what they can mean, changes in size distribution, changes in functional groups
-
-- What could have driven these shifts?
-% talk about link to bottom up drivers, but also mention top down aspects...
-then go deep on Gradient Forest analysis, what that could mean.
-Relate this to other papers about community changes, climate effects and gradient forest applications.
-
-In the Cariaco basin, wind-driven seasonal upwelling dictates the nutrient availability in the upper mixed layer, which in turn affects the phytoplankton dynamics \cite{ptacnik_diversity_2008, bopp_response_2005}.
-In particular, Diatoms are responsive to changes in the mixing regime \cite{huisman_reduced_2006}.
-Alterations in physical mixing affected phytoplankton dynamics,
-% MEI index relevant (from 2019 REVIEW)
-The Caribbean experiences the most marked of the El Niño–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Niño–Southern Oscillation \cite{enfield_tropical_1997}.
-
-% Pinckney et al. 2015:
-- major thing is that previous paper only looked at microscopy to 55m (Mutshinda to 75m and Pinckney), but now we look at full 100 m!
-%The Cariaco Basin has undergone marked shifts in biogeochemistry and food web structure over the past two decades (Taylor et al., 2012; Montes et al., 2013; Bates et al., 2014; Scranton et al., 2014).
-Since the initiation of the CARIACO Ocean Time-Series, near-surface (0–55 m) chl a concentrations (measured by fluorescence of pigments extracted in methanol) and net primary production rates (NPP, measured using 14C incubations) have declined, particularly between 2004 and 2010 (Taylor et al., 2012; Scranton et al., 2014).
-This is hypothesized to be the result of a measured decrease in seasonal trade wind intensity, which has reduced coastal upwelling and led to shallower mixed layer depths (Taylor et al., 2012).
-Microscopy observations indicate that microphytoplankton assemblages (420 mm) within the euphotic zone have shifted from a community comprised mainly of diatoms, dinoflagellates, and coccolithophorids to one with increasing numbers of smaller taxa including cyanobacteria (Montes et al., 2013).
-During this period, a collapse in the sardine industry also occurred.
-The causes are not clear, but the phenomenon was accompanied by an increase in zooplankton biomass and an increase in particulate organic carbon export below the euphotic zone (Taylor et al., 2012).
-These results suggest that changes in zooplankton abundance and therefore grazing may have contributed to the decline in diatoms and to more efficient carbon export to depth.
-% Consistent with the results of Taylor et al.
-(2012), the microscopy data show large scale reductions in overall microphytoplankton (4 20 mm) cell densities and a large decrease in both diatoms and coccolithophores over the upper 55 m.
-However, the observed photopigment decline in diatom abundance is not nearly to the same extent and coccolithophores had a significant concentration increase of 60.1% over this same depth interval (Table 3).
-We argue that the reduction in cell numbers is likely only true of larger cells.
-Microscopic cell counts are inherently selective and biased toward easily identifiable species and cells that remain intact after preservation (Lund et al., 1958; Rott, 1981; Wilhelm et al., 1991; Schlüter et al., 2000).
-Furthermore, Mousing et al.
-(2014) found that small increases in temperature (as observed in the Cariaco Basin) lead to a decrease in the relative contribution of large cells in the community, regardless of ambient nutrient availability.
-Finally, depth limited microscopy data misses the downward movement of diatoms and coccolithophores deeper in the water column (i.e.
-below 56 m).
-In fact, there was also a dramatic shift in diatom speciation, with centric diatoms (Guinardia, Leptocylindrus, Skeletonema) dominating during the first half of the time-series and a mix of centric and pennate diatoms (Pseudonitzschia, Leptocylindrus, Guinardia) dominating in more recent years.
-In microscopy, smaller sample volumes may lead to underestimations of species richness or diversity (Cermeño et al., 2014).
-Nonetheless, both photopigment and microscopy measurements suggest that phytoplankton communities became more homogeneously distributed over the years, with the majority of the difference due to declining diatom populations that occur during upwelling events.
-
-% REALLY ONLY LOOK AT BOTTOM UP HERE; No Top Down effects quantified
-
-% AMO
-The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
-
-% "Shifts in global and basin-specific primary production and phytoplankton abundances are related to large-scale climatic variability in the physical environment.
-Despite the lack of a definitive period associated with the AMO, the relationship between the atmospheric variability associated with the NAO and thermohaline circulation characterized by the AMO suggests a coupling of ocean and atmospheric processes on time scales that will facilitate understanding and prediction of inter-annual weather, precipitation, fluvial output and estuarine and marine productivity.
-Several studies have documented shifts related to the AMO in oceanic (Martinez et al., 2009), coastal and continental shelf (Rodrigues et al., 2009; Schofield et al., 2008) and estuarine (Hubeny et al., 2006) productivity and responses in forage fish populations to this increase in planktonic productivity (Tourre et al., 2007).
-Correlations between the AMO indices and marine productivity have been documented at immediate (Schofield et al., 2008), historical (Martinez et al., 2009; Tourre et al., 2007), and pre-historical (Hubeny et al., 2006; Rodrigues et al., 2009) scales.
-There are several mechanisms through which this might occur.
-Stratification within the water column separates nutrients at lower depths from entering the euphotic zone, thus limiting primary productivity.
-Long-term climate oscillations may augment or suppress seasonal conditions that promote or erode stratification and regulate mixing.
-Schofield et al.
-(2008) have demonstrated links between the AMO and phytoplankton blooms in the Mid-Atlantic Bight, such that increased nutrient availability during the annual fall transition were augmented by increased winter winds associated with a positive phase in the AMO.
-Martinez et al.
-(2009) demonstrated that global phytoplankton productivity is linked to multi-decadal oscillations in both the Atlantic and Pacific, suggesting that links between chlorophyll and sea surface temperature are driven by interactions between the pycnocline and mixed surface layers.
-The influence of large-scale climate fluctuations on surface stratification has particular influence on phytoplankton productivity in the tropics and mid-latitudes, waters which are often nutrient-limited.
-Alternatively, warmer sea surface temperatures themselves may influence productivity.
-Shifts in fluvial inputs associated with altered precipitation patterns will also alter river-induced fertilization of marine productivity." from Nye et al.
-2014
-
-%"The AMO is also associated with changes in wind and current regimes (Delworth et al., 2007; Häkkinen et al., 2011).
-Thus, passive dispersal of marine organisms, many of which are planktonic or have planktonic larvae, will be affected resulting in changes in spatial distribution, survival, and population abundance." from Nye et al 2014
-
-FOR THE CLUSTERING (moved from Results): This confirms previously observed patterns in the Cariaco ecosystem, where a strong shift in the community happened in 2004, but the community returned to the previous regime past 2014 \cite{taylor_ecosystem_2012, muller-karger_scientific_2019}.
-
-\section{Conclusions}
-
-- Our analysis shows that...
-
-- We also show that...
-
-- But we cannot confirm or test. so we would recommend X for the future.
-
-%%
-
-%  Numbered lines in equations:
-%  To add line numbers to lines in equations,
-%  \begin{linenomath*}
-%  \begin{equation}
-%  \end{equation}
-%  \end{linenomath*}
-
-%% Enter Figures and Tables near as possible to where they are first mentioned:
-%
-% DO NOT USE \psfrag or \subfigure commands.
-%
-% Figure captions go below the figure.
-% Acronyms used in figure captions will be spelled out in the final, published version.
-
-% Table titles go above tables;  other caption information
-%  should be placed in last line of the table, using
-% \multicolumn2l{$^a$ This is a table note.}
-% NOTE that there is no difference between table caption and table heading in the final, published version
-%
-%----------------
-% EXAMPLE FIGURES
-%
-% \begin{figure}
-% \includegraphics{example.png}
-% \caption{caption}
-% \end{figure}
-%
-% Giving latex a width will help it to scale the figure properly. A simple trick is to use \textwidth. Try this if large figures run off the side of the page.
-% \begin{figure}
-% \noindent\includegraphics[width=\textwidth]{anothersample.png}
-%\caption{caption}
-%\label{pngfiguresample}
-%\end{figure}
-%
-%
-% If you get an error about an unknown bounding box, try specifying the width and height of the figure with the natwidth and natheight options.
-This is common when trying to add a PDF figure without pdflatex.
-% \begin{figure}
-% \noindent\includegraphics[natwidth=800px,natheight=600px]{samplefigure.pdf}
-%\caption{caption}
-%\label{pdffiguresample}
-%\end{figure}
-%
-%
-% PDFLatex does not seem to be able to process EPS figures. You may want to try the epstopdf package.
-%
-
-%
-% ---------------
-% EXAMPLE TABLE
-%
-% \begin{table}
-% \caption{Time of the Transition Between Phase 1 and Phase 2$^{a}$}
-% \centering
-% \begin{tabular}{l c}
-% \hline
-%  Run  & Time (min)  \\
-% \hline
-%   $l1$  & 260   \\
-%   $l2$  & 300   \\
-%   $l3$  & 340   \\
-%   $h1$  & 270   \\
-%   $h2$  & 250   \\
-%   $h3$  & 380   \\
-%   $r1$  & 370   \\
-%   $r2$  & 390   \\
-% \hline
-% \multicolumn{2}{l}{$^{a}$Footnote text here.}
-% \end{tabular}
-% \end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% SIDEWAYS FIGURES and TABLES
-% AGU prefers the use of {sidewaystable} over {landscapetable} as it causes fewer problems.
-%
-% \begin{sidewaysfigure}
-% \includegraphics[width=20pc]{figsamp}
-% \caption{caption here}
-% \label{newfig}
-% \end{sidewaysfigure}
-%
-%  \begin{sidewaystable}
-%  \caption{Caption here}
-% \label{tab:signif_gap_clos}
-%  \begin{tabular}{ccc}
-% one&two&three\\
-% four&five&six
-%  \end{tabular}
-%  \end{sidewaystable}
-
-%% If using numbered lines, please surround equations with \begin{linenomath*}...\end{linenomath*}
-%\begin{linenomath*}
-%\begin{equation}
-%y|{f} \sim g(m, \sigma),
-%\end{equation}
-%\end{linenomath*}
-
-%%% End of body of article
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Optional Appendices go here
-%
-% The \appendix command resets counters and redefines section heads
-%
-% After typing \appendix
-%
-%\section{Here Is Appendix Title}
-% will show
-% A: Here Is Appendix Title
-%
-
-\appendix
-
-\section{Gradient Forest Analysis supplement}
-
-\begin{table}
-
-\caption{Individual Gradient Forest model runs for each variable and the corresponding time lags. 
-        For in-situ variables, for which no coverage extends over the time series, 
-        we tested all measurements up to a lag of 3 months. 
-        For climate variables, we tested up to a lag of 6 months. 
-        Time lag with maximum importance per variable are highlighted in bold and chosen for the final model run.}
-
-\ref{app:table1}
-\centering
-\begin{tabular}[t]{rrrrrrrrrr}
-\toprule
-lag & NO3 & PO4 & SiO4 & Temperature & Isotherm\_21 & Salinity\_bottles & u10 & AMO & MEIv2\\
-\midrule
-0 & \textbf{0.038} & \textbf{0.038} & \textbf{0.034} & \textbf{0.071} & \textbf{0.028} & 0.007 & 0.004 & 0.016 & 0.015\\
-1 & 0.017 & 0.011 & 0.015 & 0.002 & 0.002 & 0.004 & \textbf{0.010} & 0.014 & 0.018\\
-2 & 0.022 & 0.008 & 0.013 & 0.006 & 0.008 & \textbf{0.011} & 0.009 & \textbf{0.030} & 0.019\\
-3 & 0.016 & 0.007 & 0.004 & 0.005 & 0.007 & 0.010 & 0.007 & 0.018 & 0.018\\
-4 & - & - & - & - & - & - & 0.008 & 0.019 & \textbf{0.029}\\
-5 & - & - & - & - & - & - & 0.008 & 0.018 & 0.022\\
-6 & - & - & - & - & - & - & 0.010 & 0.018 & 0.013\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Optional Glossary, Notation or Acronym section goes here:
-%
-% Glossary is only allowed in Reviews of Geophysics
-%  \begin{glossary}
-%  \term{Term}
-%   Term Definition here
-%  \term{Term}
-%   Term Definition here
-%  \term{Term}
-%   Term Definition here
-%  \end{glossary}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Acronyms
-%% NOTE that acronyms in the final published version will be spelled out when used in figure captions.
-%   \begin{acronyms}
-%   \acro{Acronym}
-%   Definition here
-%   \acro{EMOS}
-%   Ensemble model output statistics
-%   \acro{ECMWF}
-%   Centre for Medium-Range Weather Forecasts
-%   \end{acronyms}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Notation
-%   \begin{notation}
-%   \notation{$a+b$} Notation Definition here
-%   \notation{$e=mc^2$}
-%   Equation in German-born physicist Albert Einstein's theory of special
-%  relativity that showed that the increased relativistic mass ($m$) of a
-%  body comes from the energy of motion of the body—that is, its kinetic
-%  energy ($E$)—divided by the speed of light squared ($c^2$).
-%   \end{notation}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% DATA SECTION and ACKNOWLEDGMENTS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%!
-\section*{Open Research Section}
-This section MUST contain a statement that describes where the data supporting the conclusions can be obtained.
-Data cannot be listed as ''Available from authors'' or stored solely in supporting information.
-Citations to archived data should be included in your reference list.
-Wiley will publish it as a separate section on the paper’s page.
-Examples and complete information are here:
-https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors
-
-\section*{As Applicable – Inclusion in Global Research Statement}
-The Authorship: Inclusion in Global Research policy aims to promote greater equity and transparency in research collaborations.
-AGU Publications encourage research collaborations between regions, countries, and communities and expect authors to include their local collaborators as co-authors when they meet the AGU Publications authorship criteria (described here: https://www.agu.org/publications/authors/policies\#authorship).
-Those who do not meet the criteria should be included in the Acknowledgement section.
-We encourage researchers to consider recommendations from The TRUST CODE - A Global Code of Conduct for Equitable Research Partnerships (https://www.globalcodeofconduct.org/) when conducting and reporting their research, as applicable, and encourage authors to include a disclosure statement pertaining to the ethical and scientific considerations of their research collaborations in an “Inclusion in Global Research Statement’ as a standalone section in the manuscript following the Conclusions section.
-This can include disclosure of permits, authorizations, permissions and/or any formal agreements with local communities or other authorities, additional acknowledgements of local help received, and/or description of end-users of the research.
-You can learn more about the policy in this editorial.
-Example statements can be found in the following published papers:
-%Holt et al. (https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2022JG007188), 
-%Sánchez-Gutiérrez et al. (https://agupubs.onlinelibrary.wiley.com/doi/abs/10.1029/2023JG007554), 
-%Tully et al. (https://agupubs.onlinelibrary.wiley.com/doi/epdf/10.1029/2022JG007128) 
-%Please note that these statements are titled as “Global Research Collaboration Statements” from a previous pilot requirement in JGR Biogeosciences.
-The pilot has ended and statements should now be titled “Inclusion in Global Research Statement”.
-
-%!
-\acknowledgments
-Thank you to all who have helped, communicated, provided data. Claudia Benitez-Nelson, Digna Rueda-Roa, James Pinckney.
-
-%Enter acknowledgments here. This section is to acknowledge funding, thank colleagues, enter any secondary affiliations, and so on.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% REFERENCES and BIBLIOGRAPHY
-%
-%\bibliography{<name of your .bib file>} don't specify the file extension
-% don't specify bibliographystyle
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\bibliography{references}
-
-%Reference citation instructions and examples:
-%
-% Please use ONLY \cite and \citeA for reference citations.
-% \cite for parenthetical references
-% ...as shown in recent studies (Simpson et al., 2019)
-% \citeA for in-text citations
-% ...Simpson et al. (2019) have shown...
-%
-%
-%...as shown by \citeA{jskilby}.
-%...as shown by \citeA{lewin76}, \citeA{carson86}, \citeA{bartoldy02}, and \citeA{rinaldi03}.
-%...has been shown \cite{jskilbye}.
-%...has been shown \cite{lewin76,carson86,bartoldy02,rinaldi03}.
-%... \cite <i.e.>[]{lewin76,carson86,bartoldy02,rinaldi03}.
-%...has been shown by \cite <e.g.,>[and others]{lewin76}.
-%
-% apacite uses < > for prenotes and [ ] for postnotes
-% DO NOT use other cite commands (e.g., \citet, \citep, \citeyear, \nocite, \citealp, etc.).
-%
-
-\end{document}
-
-More Information and Advice:
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  SECTION HEADS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Capitalize the first letter of each word (except for
-% prepositions, conjunctions, and articles that are
-% three or fewer letters).
-
-% AGU follows standard outline style; therefore, there cannot be a section 1 without
-% a section 2, or a section 2.3.1 without a section 2.3.2.
-% Please make sure your section numbers are balanced.
-% ---------------
-% Level 1 head
-%
-% Use the \section{} command to identify level 1 heads;
-% type the appropriate head wording between the curly
-% brackets, as shown below.
-%
-%An example:
-%\section{Level 1 Head: Introduction}
-%
-% ---------------
-% Level 2 head
-%
-% Use the \subsection{} command to identify level 2 heads.
-%An example:
-%\subsection{Level 2 Head}
-%
-% ---------------
-% Level 3 head
-%
-% Use the \subsubsection{} command to identify level 3 heads
-%An example:
-%\subsubsection{Level 3 Head}
-%
-%---------------
-% Level 4 head
-%
-% Use the \subsubsubsection{} command to identify level 3 heads
-% An example:
-%\subsubsubsection{Level 4 Head} An example.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  IN-TEXT LISTS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-% Do not use bulleted lists; enumerated lists are okay.
-% \begin{enumerate}
-% \item
-% \item
-% \item
-% \end{enumerate}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  EQUATIONS
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Single-line equations are centered.
-% Equation arrays will appear left-aligned.
-
-Math coded inside display math mode \[ ...\]
- will not be numbered, e.g.,:
- \[ x^2=y^2 + z^2\]
-
- Math coded inside \begin{equation} and \end{equation} will
- be automatically numbered, e.g.,:
- \begin{equation}
- x^2=y^2 + z^2
- \end{equation}
-
-% To create multiline equations, use the
-% \begin{eqnarray} and \end{eqnarray} environment
-% as demonstrated below.
-\begin{eqnarray}
-  x_{1} & = & (x - x_{0}) \cos \Theta \nonumber \\
-        && + (y - y_{0}) \sin \Theta  \nonumber \\
-  y_{1} & = & -(x - x_{0}) \sin \Theta \nonumber \\
-        && + (y - y_{0}) \cos \Theta.
-\end{eqnarray}
-
-%If you don't want an equation number, use the star form:
-%\begin{eqnarray*}...\end{eqnarray*}
-
-% Break each line at a sign of operation
-% (+, -, etc.) if possible, with the sign of operation
-% on the new line.
-
-% Indent second and subsequent lines to align with
-% the first character following the equal sign on the
-% first line.
-
-% Use an \hspace{} command to insert horizontal space
-% into your equation if necessary. Place an appropriate
-% unit of measure between the curly braces, e.g.
-% \hspace{1in}; you may have to experiment to achieve
-% the correct amount of space.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  EQUATION NUMBERING: COUNTER
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% You may change equation numbering by resetting
-% the equation counter or by explicitly numbering
-% an equation.
-
-% To explicitly number an equation, type \eqnum{}
-% (with the desired number between the brackets)
-% after the \begin{equation} or \begin{eqnarray}
-% command.  The \eqnum{} command will affect only
-% the equation it appears with; LaTeX will number
-% any equations appearing later in the manuscript
-% according to the equation counter.
-%
-
-% If you have a multiline equation that needs only
-% one equation number, use a \nonumber command in
-% front of the double backslashes (\\) as shown in
-% the multiline equation above.
-
-% If you are using line numbers, remember to surround
-% equations with \begin{linenomath*}...\end{linenomath*}
-
-%  To add line numbers to lines in equations:
-%  \begin{linenomath*}
-%  \begin{equation}
-%  \end{equation}
-%  \end{linenomath*}

### 2025-03-05 agujournaltemplate.tex
Focus Phase Day 16 - Intro Draft done, Discussion 40%

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%  BODY TEXT
-%
+%                                             %
+%  BODY TEXT                                  %
+%                                             %
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
     Phytoplankton consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}.
-The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them a highly suitable indicator of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
+The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them highly suitable indicators of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
     The structure of the community can be considered an integrator of environmental factors, as it is strongly determined by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}.
-    %%% Mention Top-Down and Bottom-Up forces structuring communities and affection biodiversity.
+In addition to such bottom-up effects, there are also forces of higher trophic levels, such as zooplankton, that structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}.
+Disentangling the effects driving changes in communities and biodiversity of phytoplankton is a difficult but crucial task to better understand our oceans.
 
@@
 Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}.
-
-    %%% Phytoplankton functional group shifts through climate change
-    %%% Phytoplankton diversity shifts through climate change
-
-    The loss of biodiversity is predicted to have a lasting impact on the functioning of an ecosystem, since links have been found between the richness of a phytoplankton community and ecosystem functions, such as biomass and resource use efficiency \cite{ptacnik_diversity_2008, lehtinen_phytoplankton_2017}.
-Such links need to be further investigated, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014} and under greater threat of extinction, when compared to temperate regions \cite{finnegan_paleontological_2015}.
-    % DON'T FORGET TO MENTION CLIMATE IMPACTS (as I investigate AMO linkage and ENSO MEI etc.)
-    % this would actually make the most sense here!
+The drivers and effects of changes in biodiversity need to be further investigated, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014} and is under greater threat of extinction, compared to temperate regions \cite{finnegan_paleontological_2015}.
 
@@
     Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term climate changes, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.
-
     % CARIACO is one such time series
     One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin between 1995 and 2017 \cite{muller-karger_scientific_2019}.
-
     % Describe CARIACO location
@@
     % NOw we come in!
-    - The question arises, what were the drivers of change in phytoplankton community?
-    - Bottom-up or top-down processes? response to regional changes in oceanographic conditions, in turn driven by climate change or oscillations?
-    - Highlight, that there is a large mismatch in data-coverage particularly for Top-Down aspects, fisheries is landing data (hard to quantify anthropogenic impact on this data and ecosystem) and Zooplankton data is only from after 2002.
-Also HPLC data is lacking in coverage.
-    - We focus on Bottom-up, given the available data, and try to identify changes in biodiversity and community and relate these to both local effects and climatic drivers.
-
-    - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!
-
-    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-    % "Changes in plankton biodiversity in response to regional changes in oceangraphic conditions" driven by climate oscillaitons (from REVIEW 2019).
-    %% Changes in Functional Groups
-
-    %% Changes in Diversity
-
-    % Now we come into the picture: Which gap are we adressing?
-
-    % FOCUS ON Community Changes + BOTTOM UP
-
-    These changes are believed to be influenced by various factors, including climate variability, nutrient fluxes, and ocean circulation patterns.
-    % WHAT ARE THE DRIVERS OF COMMUNITY CHANGE? Bottom up OR top down, we focus on bottom up...
-
-    Describe physical changes!
-
-    - AMO (ref to 2019 Review)
-    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al.
-2015).
-These climatic changes had a profound impact throughout the Atlantic (Alexander et al.
-2014, Nye et al.
-2014) and in the Cariaco basin \cite{taylor_ecosystem_2012,pinckney_phytoplankton_2015}.
-    - ENSO teleconnection (ref to 2019 Review)
-    The Caribbean experiences the most marked of the El Ni  ̃no–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Ni  ̃no–Southern Oscillation (Enfield & Mayer 1997).
-
-     - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!
-
-    % Finally say what WE are going to do! What are the open questions, what is new?
-
-    %% try to identify drivers of shift in community with Gradient Forest
-
-    %%% Instead of going paper by paper below, I should go topic by topic (i.e. all points made on "DIVERSITY" should be in a paragraph, and all on changes in FGs or Size Classes in another etc...)
-
-    Since the inception of the time series, the Cariaco basin has undergone marked shifts...
-
-    The time series includes multiple measurements of the phytoplankton community, which are the bulk fluorometric chlorophyll, microscopy cell counts, and high-performance liquid chromatography (HPLC) pigment analysis.
-Of these, all cover the entire time series except for the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.
-
-    The most often measured indicator of bulk phytoplankton biomass is fluorometric Chlorophyll \textit{a} concentration, which gives a general measure of biomass standing stock and distribution in the water column.
-A more advanced methodology, high-performance liquid chromatography (HPLC), allows individual pigment types to be quantified.
-These pigment distributions can be used to estimate the groups of phytoplankton that contribute to biomass.
-The most specific methodology to quantify phytoplankton communities is the counting and identification of individual phytoplankton cells, either manually under a microscope or automated in flow-through systems.
-The major drawbacks are the size limit of resolving smaller cells and, if done manually, the complexity of identifying phytoplankton species.
-
-    % PAPER TIMELINE:
-    - Taylor 2012 ()
-    \citeA{taylor_ecosystem_2012} analyzed data up to 2010 and found significant decadal trends of increase in sea surface temperature (SST), coinciding with a decrease in upwelling intensity and phytoplankton bloom intensities, which the authors related to the impact of global warming.
-The phytoplankton community identified via microscopy showed a shift towards smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass.
-The authors explained the change in physical regime in the Cariaco basin with a north-easterly shift of the Inter-Tropical Convergence Zone (ITCZ) and a resulting weakening trend in Trade Winds.
-
-    - Mutshinda 2013 ()
-    Using a hierarchical Bayesian modeling approach, changes in dominant species quantified via microscopy cell counts between 1995 and 2011 were found to be best predicted by changes in temperature, pH and irradiance, and less so by salinity or macronutrient concentrations \cite{mutshinda_environmental_2013}.
-In particular, higher temperatures were associated with a lower abundance of diatoms.
-
-    - Irwin 2015 ()
-    With the same dataset of microscopy cell counts and a MaxEnt statistical modeling approach, \citeA{irwin_phytoplankton_2015} calculated realized niches for the dominant species for SST, irradiance and nitrate concentration.
-As trends in the realized niches were observed over the course of the time series, this seemed to show the adaption of species to climate change effects.
-
-    - Pinckney 2015 ()
-    The phytoplankton community was investigated in detail by \citeA{pinckney_phytoplankton_2015}, using the microscopy counts and HPLC data from the time ranges 1996 to 2000 and 2006 to 2010.
-Between these two time periods, they observed that the drastic reduction in diatoms counted using microscopy did not correspond to a similar reduction in diatom abundance calculated from the pigment composition, which is most likely explained by a shift to diatom species with smaller cell sizes.
-Additionally, the pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity and a reduction in seasonal variability in the community composition.
-These changes are most likely a response to the reduced upwelling  and a deepening of the euphotic zone associated with a decrease in the strength of seasonal trade winds.
-Interestingly, the shifts in the phytoplankton community did not result in a significant change to the export of particulate organic carbon (POC) to depth as measured by sediment traps deployed during the time series.
-
-    - Lorenzoni 2015 ()
-    \citeA{lorenzoni_characterization_2015} focused their analysis on HPLC data retrieved between 2006 and 2012 and related seasonal changes in the pigment composition to spectral absorption from remote sensing data and concentrations of organic carbon.
-They observed a high seasonal variability, with microphytoplankton (mostly diatoms) dominating upwelling periods and picophytoplankton (mostly cyanobacteria) dominating during the rainy season.
-These community compositions were also reflected in the POC:TChla ratio, which was higher for diatom dominated assemblages and lower for those dominated by cyanobacteria.
-    % "The difference between the taxonomic data and the pigment data suggests that the microscopy taxonomy may not capture the variability in the pico-phytoplankton.
-Since the beginning of the time-series in 1995, there has been a change in the abundance of phytoplankton size classes based on diagnostic pigments: micro-phytoplankton has decreased year-round, while picophytoplankton has increased, particularly during the rainy season.
-This is consistent with the decrease in larger phytoplankton previously reported for the region." Lorenzoni et al.
-2015
-
-    - Finally, add the info from Review about Phytoplankton community, about community and upwelling returning in 2014 to previous levels/state.
-    - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!
-
-    % Hone in on Diversity Aspects, and what has been done there?
-    - Highlight diversity aspect of Pinckney
-    - Highlight Mutshinda etc. How they used Niches/Phyto data
-    - What are the questions in Diversity TS data? -> Esteban's papers
-
-    % In recent years, significant changes have been observed in phytoplankton populations within the Cariaco Basin.
-These changes are believed to be influenced by various factors, including climate variability, nutrient fluxes, and ocean circulation patterns.
-Understanding the drivers and impacts of these changes in the phytoplankton community is crucial to understand the broader implications for marine food webs and carbon cycling in this and similar ecosystems.
-
-    "The Cariaco Basin has undergone marked shifts..." Talk about ecosystem, what has changed, reference prev literature
-
-        % below is from 2019 REVIEW
-        %Hydrographic observations (Figure 2) show strong seasonality and interannual variability in the upper 400 m in the Cariaco Basin.
-This is controlled by seasonal and interannual changes in the wind and in the intensity of the geostrophic Caribbean Current (Muller-Karger et al.
-1989).
-Primary upwelling was strongest between November and May.
-In general, during upwelling, sea surface temperatures ranged between approximately 23\degree C and 25\degree C.
-The secondary upwelling that occurs between June and August is shorter ($\sim$ 5 weeks) and is approximately 1.5\degree C warmer than the primary upwelling.
-It is driven by Ekman transport and an intensification of the curl of the wind close to the southern coast of the Caribbean (Rueda-Roa \& Muller-Karger 2013, Rueda-Roa et al.
-2018).
-Upwelled waters move offshore in plumes that extended as far as 50 km to more than 250 km from the coast (Muller-Karger et al.
-1989, 2010).
-From July to October, there is strong stratification in the upper 50 m of the water column after the wind weakens and the rain and rivers deliver less saline waters (Figure 2).
-        %Sea surface temperatures ranged from approximately 28\degree C to 30\degree C during the period of stratification (Figure 2).
-The mixed-layer depth varied from less than 10 m to 35 m during the 21-year time series (Figure 3).
-Satellite sea surface temperature and ocean color observations show that upwelling pulsates throughout the year at the scale of days to weeks, leading to marked variation in primary production (Figure 3).
-CARIACO rigorously characterized the annual seasonal variation in primary productivity and stratification.
-The length of the time series was key in capturing interannual variations that spanned decades.
-Satellite sea surface temperature and field data show that upwelling was stronger between 1996 and 2002.
-After 2002, trade wind intensity became weaker (Taylor et al.
-2012).
-        The field observations, particularly those in the upper 50 m, showed what appeared to be a return to stronger upwelling conditions starting in 2014 (Figure 2a).
-
-        %“The Caribbean experiences the most marked of the El Nino–Southern Oscillation telecon- ˜ nections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Nino–Southern Oscillation (Enfield & Mayer 1997).
-˜” ([Muller-Karger et al., 2019, p.
-4](zotero://select/library/items/7YW7MFTU)) ([pdf](zotero://open-pdf/library/items/TTZQBAXW?page=4&annotation=D3H7YG7Y)) link to El Nino of Cariaco/Caribbean sea!
-
-        %“When the CARIACO time series was implemented in 1995, major changes were happening in the Atlantic.
-We now know that at that time, the Atlantic Multidecadal Oscillation (Kerr 2000, Enfield et al.
-2001) started shifting to a positive phase, which then persisted for more than 20 years and led to oceanographic anomalies in the Atlantic that were larger than any recorded over the previous 100 years (McCarthy et al.
-2015).
-The changes had profound biological impacts throughout the Atlantic (Alexander et al.
-2014, Nye et al.
-2014), including in the Cariaco Basin (Taylor et al.
-2012, Pinckney et al.
-2015).” (Muller-Karger et al., 2019, p.
-3) Interesting aspect that the AMO changed, as apparently the main climate indices to affect the Atlantic/CARIACO
-
-    - Why community structure is important, taxonomic diversity (aspects of diversity) - why look at taxonomic diversity?
-
-    - some general intro to DIVERSITY.. what is it? Why important?
-
-    % "Phytoplankton is the major primary producer group in marine ecosystems and at the same time a highly suitable indicator for ecosystem change.
-Due to its high diversity and short generation time, phytoplankton responds quickly to changes in the environment (Reynolds 1998).
-Especially changes in temperature and nutrient supply are known to change phytoplankton composition and biodiversity, with higher temperatures and lower nutrients favouring small species (Litchman et al.
-2007; Winder and Sommer 2012), whereas Si:N and N:P supply ratios change the relative importance of diatoms and other groups such as dino(flagellates) and cyanobacteria (Sommer et al.
-2004; Makareviciute-Fichtner et al.
-2020)." from Di Cavalho et al.
-2023
-
-    %"Whereas the positive biomass response to high nutrient and light availability is well known (Tilman 1982; Leibold 1999; Siegel et al.
-2013), the biodiversity response is much more complex.
-First, biodiversity itself is a multifaceted construct, comprising the number of different taxa (e.g.
-species richness) and the equality of species contribution to the assemblage as aspects of local alpha diversity (evenness).
-But the change in richness is only the net effect of immigration and local extinction, while the turnover in species composition with time is an additional aspect of biodiversity.
-Extending to larger spatial scales, heterogeneity in space (beta diversity) and composition of the regional species pool (gamma diversity) are further biodiversity dimensions.
-Second, for each of these facets, all metrics are effort- and scaledependent.
-Thus, in contrast to other aspects of ecosystem assessments, biodiversity metrics do not qualify to set absolute target values for good or bad status and effective ecosystem management.
-But, the relevance of the metrics arises from analysing their temporal trends over a long time period.
-Third, no single variable captures even the most important aspects of community composition and change.
-Therefore, the analysis of phytoplankton biodiversity requires a multimetric approach." from Di Cavalho et al.
-2023 (with Hillebrand)
-
-    - why are we looking at Functional Groups?
-    %"The complexity of biodiversity assessment has led to the development of other approaches addressing compositional change in plankton communities, most prominently the creation of functional groups.
-Functional groups are meant to comprise species that show similar responses to environmental drivers and share morphological and/or physiological characteristics.
-Functional group indicators have been shown to be relevant for describing community structure and biodiversity and are more comparable with other studies than species-based indicators (Mouillot et al.
-2006)." from Di Cavalho et al.
-2023 (with Hillebrand)
-
-    %% Overarching theme: Stability of Ecosystem faunctions, Drivers of community change
-    "Regime shifts + Diversity" - general intro on current findings related to "climate driven ecosystem shifts"
-        - %"We find that the rate of turnover in the phytoplankton community becomes faster during this century, that is, the community structure becomes increasingly unstable in response to climate change.
-Combined with alterations to phytoplankton diversity, our results imply a loss of ecological resilience, with likely knock-on effects on the productivity and functioning of the marine environment." \cite{henson_future_2021}
-        -
-
-        "Here's where we come in..." Talk about how most of the analysis stopped after 2012, time series died out, but there is more data there and particularly looking at community data, we can see a return.
-    - what has been done (Phytoplankton - Statistics)?
-    - where are the gaps?
-    - less studied aspects: Diversity + Phytoplankton Microscopy
-
-    So we would hypothesize, that if wind is driving the ecosystem, then wind would be best predictor of shits.
-    Or nutrients, but if it is a climate indices, then the indices capture some variability that is not related to wind or nutrients....
-
-    "In this paper, we..."
-
-        %Highlight that I am only looking at Phytoplankton Microscopy data, and a bit of Niskin, but nothing else. Only bottom-up drivers.
-        %What are the questions I ask myself?? 1.
-
-    % INTRO TEXT DUMP:
-
-    Biodiversity is a complex concept composed of genetic, taxonomic, and ecological aspects that can additionally be quantified in temporal and spatial scales.
-In its basic sense, a measure of diversity should quantify two components: the number of items and their relative frequencies \cite{borics_freshwater_2021}.
-    Richness, the number of species or genera present at a particular time in a sample, has been established as the most basic descriptor of community diversity, but it is not the best indicator of ecosystem change \cite{hillebrand_biodiversity_2018}.
-    - Other metrics, evenness, and dominance
+    In this study, we document changes in biodiversity and community structure in the Cariaco basin between 1995 and 2017 and quantify the effect of bottom-up drivers and climate indices on changes in the phytoplankton community.
+We focus on the phytoplankton cell count data by microscopy, as these data have not been fully analyzed in previous studies, and we argue that despite the shortcomings of microscopy methods, they provide a consistent measure of changes in large phytoplankton.
+Our analysis focuses on bottom-up drivers of community changes, as these data are available throughout the time series and at monthly resolution.
+This study complements the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017.
+Understanding the regional dynamics of the Cariaco basin is crucial to understanding the broader ecological changes in tropical marine ecosystems driven by changes in climate.
 
@@
 Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix.
-The community is separated into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013.
-Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, turnover is higher again.
+The community is divided into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013.
+Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, the turnover is higher again.
 This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years.
@@
 To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
-Preliminary model runs were used to identify if time lags of the predictor variables could better explain the data (see Appendix), and this resulted in choosing the AMO with a lag of 5 months and MEI v.2 with a lag of 4 months.
-In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the highest predictive capability for the climate indices with time lags, and upwelling related variables such as nitrate in the upper 100 meters and wind speed scoring slightly lower.
-The \qty{21}{\celsius} Isotherm and silicate and phosphate concentrations scored lowest in predicting shifts within the community.
-The way Gradient Forest predicts shifts in the community, is by building a random forest model for each observed genus across the time series.
+Preliminary model runs were used to identify if time lags of the predictor variables could better explain the data (see Appendix), and this resulted in choosing the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months.
+In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the highest predictive capacity for climate indices with time lags and upwelling-related variables such as nitrate in the upper 100 meters and wind speed scoring slightly lower.
+The \qty{21}{\celsius} Isotherm and silicate and phosphate concentrations scored lowest in predicting changes within the community.
+The way Gradient Forest predicts shifts in the community is by building a random forest model for each observed genus across the time series.
 The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$.
-The microscopy data is biased towards larger cell sizes, and therefore most of the predicted genus cell counts are from diatoms.
+The microscopy data are biased towards larger cell sizes, and therefore most of the predicted genus cell counts are from diatoms.
 However, two Haptophytes and four Dinoflagellates, as well as the Cyanobacteria \textit{Synechococcus} do rank highly.
 Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range.
-The black line indicates the cumulative importance across the entire community, and the colored dashed lines indicate the importance per phytoplankton group.
-For the AMO index, we see a clear pattern that negative anomalies highly affect the community.
-This contrasts with a higher importance at positive anomalies in the MEI v.2 index, in particular for Haptophytes.
-Nitrate concentrations between 6 and 8 \unit{\micro \mole} is where the community changed most, which indicates these values as a threshold for the composition of the community.
-For wind speed, temperature and salinity, changes occurred most at parameter ranges related to high upwelling, i.e.
-high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}) and high salinity (above \qty{36.8}{PSU}).
-One interesting aspect is that for both climate indices, Haptophytes showed the strongest cumulative importance, but for Diatoms the nitrate concentration seem to be a stronger predictor.
+The black line indicates cumulative importance across the entire community and the colored dashed lines indicate importance per phytoplankton group.
+For the AMO index, we see a clear pattern in which negative anomalies highly affect the community.
+This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes.
+Nitrate concentrations between 6 and 8 \unit{\micro \mole} is where the community changed most, indicating these values as a threshold for the composition of the community.
+For wind speed, temperature, and salinity, changes occurred most at parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}) and high salinity (above \qty{36.8}{PSU}).
+An interesting aspect is that, for both climate indices, haptophytes showed the strongest cumulative importance, but for diatoms, the nitrate concentration seemed to be a stronger predictor.
 
@@
 \section{Discussion}
+
+% HERE I INTERPRET AND DISCUSS, whereas before I just described. Answer not the *what*, but the *how* and *why*.
+Our main results show that...
+1. we see Overlap in yearly mean anomalies between certain variables
+
+- The Cariaco basin is a system, that...
+    - Discuss cell count collapse, apparent shift, but not as strong! Limitations of microscopy.
+
 This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes.
-In the Cariaco basin, long-term climatic oscillations have affected the upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
-
-- Our main results show that...
-1. we see Overlap in yearly mean anomalies between certain variables
+In the Cariaco basin, long-term climatic oscillations have affected the wind-driven upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
+%need to rephrase below (from Pinckney)
+During the early years of the time series (1996–2004), the Cariaco basin experienced a regular seasonal upwelling cycle driven by strong Trade Winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
+This upwelling period represented \qty{\sim 70}{\%} of annual depth-integrated primary production with a phytoplankton community dominated by microphytoplankton (diatoms, dinoflagellates, and coccolithophorids) \cite{pinckney_phytoplankton_2015}.
+Following 2004, the physical and biogeochemical regime had changed to one with significantly reduced Trade Winds during the upwelling period, resulting in higher sea surface temperatures and less pronounced mixing of the surface and deep layers \cite{taylor_ecosystem_2012}.
+The microscopy cell counts indicate a collapse of large phytoplankton and diatoms in particular, which is confirmed to a lesser degree by pigment data analysis.
+In general, a shift in the phytoplankton community to smaller cell sizes and higher abundances of coccolithophores, cryptophytes, and other phytoflagellates, typical for stratified water columns was observed \cite{pinckney_phytoplankton_2015}.
+
+- Now mention how the variables overlap in anomalies in Figure 2, how that can be linked to the general changes already observed by other studies.
+
+- What we want to mention, but have not included in our analysis due to lack of data at appropriate coverage or resolution, is the indicators of top-down effects on the Cariaco basin ecosystem.
+Yearly landings of the Spanish sardine (\textit{Sardinella aurita}) fisheries off the NE coast of Venezuela collapsed after 2004 and measurements of zooplankton abundances since 2001 showed a steady increase in numbers up until 2010.
+These can be viewed as consequences of the changes apparent in the phytoplankton community, but can also be important drivers of that change, for example if the collapse in fisheries was mainly driven by overfishing.
+(MENTION THAT ZOO DATA IS TOO FEW DATAPOINTS TO INCLUDE IN GRADIENT FOREST MODEL RUN)
 
 2. mismatch between Diversity variables, and cell count abundances + community clustering / ecosystem shift
+
+   The time series includes multiple measurements of the phytoplankton community, which are the bulk fluorometric chlorophyll, microscopy cell counts, and high-performance liquid chromatography (HPLC) pigment analysis.
+Of these, all cover the entire time series except for the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.
+
+    The most often measured indicator of bulk phytoplankton biomass is fluorometric Chlorophyll \textit{a} concentration, which gives a general measure of biomass standing stock and distribution in the water column.
+A more advanced methodology, high-performance liquid chromatography (HPLC), allows individual pigment types to be quantified.
+These pigment distributions can be used to estimate the groups of phytoplankton that contribute to biomass.
+The most specific methodology to quantify phytoplankton communities is the counting and identification of individual phytoplankton cells, either manually under a microscope or automated in flow-through systems.
+The main drawbacks are the size limit for resolving smaller cells and, if done manually, the complexity of identifying phytoplankton species.
 
@@
     but that most likely requires mechanistic modeling to investigate further...
+
+    %- Discuss effect of AMO
+    % Which studies have already looked at effects of Climate on Ecosystem?
+    - AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
+The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
+    - Although regional effecs may vary, a positive phase in the AMO generally linked to periods of warmer temperatures and a shift of the ITCZ from north to south.
+
+    - Taylor 2012 looked at NAO and MEI, and mostly looked linear trends in variables (they missed the return of the community and regime).
+NAO is the north-south diplole in sea level pressure, that is primarily governed by atmospheric dynamics \cite{hurrell_north_2009}.
+NAO and AMO are l
+    - AMO is mentioned in 2019 review
+    "The observed large-scale multidecadal fluctuations in the Atlantic sea surface temperature (SST), has been referred to as the Atlantic Multidecadal Oscillation (AMO) \cite{kerr2000north} to emphasize the “multidecadal” character of this oceanic phenomenon and to distinguish it from the inter-annual variability associated with the atmospheric North Atlantic Oscillation (NAO) \cite{enfield2001atlantic}." from AMO website
+
+    %- Discuss effect of MEI
+    Since El Nino has strong teleconnection to Caribbean (see ref below from Review) we included this as variable to explain changes in community.
+From Results it is less pronounced than AMO, but still explains some variability (also ref to NMDS).
+
+    Interestingly, in Gradient Forest Model, Climate variables are the strongest predictors of changes in community (i.e. changes in genus abundances).
+    - Discuss what climate variables actually capture, and why they might be better predictors of community changes than local (directly measured) variables.
+
+    - AMO (ref to 2019 Review)
+    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al.
+2015).
+These climatic changes had a profound impact throughout the Atlantic (Alexander et al.
+2014, Nye et al.
+2014) and in the Cariaco basin \cite{taylor_ecosystem_2012,pinckney_phytoplankton_2015}.
+    - ENSO teleconnection (ref to 2019 Review)
+    The Caribbean experiences the most marked of the El Ni  ̃no–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Ni  ̃no–Southern Oscillation (Enfield & Mayer 1997).
 
@@
 
-In the Cariaco basin, wind-driven seasonal upwelling dictates the nutrient availability in the upper mixed layer, which in turn affects the phytoplankton dynamics \cite{ptacnik_diversity_2008, bopp_response_2005}.
-In particular, Diatoms are responsive to changes in the mixing regime \cite{huisman_reduced_2006}.
+In the Cariaco basin, wind-driven seasonal upwelling dictates nutrient availability in the upper mixed layer, which, in turn, affects phytoplankton dynamics \cite{ptacnik_diversity_2008, bopp_response_2005}.
+In particular, diatoms respond to changes in the mixing regime \cite{huisman_reduced_2006}.
 Alterations in physical mixing affected phytoplankton dynamics,
@@
 % AMO
-The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
 

### 2025-03-11 agujournaltemplate.tex
Discussion draft 40%

@@
 In addition to such bottom-up effects, there are also forces of higher trophic levels, such as zooplankton, that structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}.
-Disentangling the effects driving changes in communities and biodiversity of phytoplankton is a difficult but crucial task to better understand our oceans.
+Disentangling the effects driving changes in community composition and biodiversity of phytoplankton is a difficult but crucial task to better understand our oceans.
 
@@
 Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}.
-The drivers and effects of changes in biodiversity need to be further investigated, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014} and is under greater threat of extinction, compared to temperate regions \cite{finnegan_paleontological_2015}.
+The drivers and effects of changes in biodiversity need to be further investigated, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and is under greater threat of extinction, compared to temperate regions \cite{finnegan_paleontological_2015}.
 
@@
 The phytoplankton community identified by microscopy showed a drastic reduction in Diatom abundance and a shift toward smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass.
-    The trends in microscopy community data were confirmed and expanded by \citeA{pinckney_phytoplankton_2015}.
+    The trends in microscopy community data were partially confirmed and expanded by \citeA{pinckney_phytoplankton_2015}.
 They observed that the drastic reduction in diatoms counted using microscopy corresponded, but to a lesser degree, to diatom abundance calculated from the pigment composition measured by high-performance liquid chromatography (HPLC).
@@
 
-    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to make the data set more homogeneous and robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008}.
+    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to make the data set more homogeneous and robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}.
 Furthermore, we calculate the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
@@
 
-% Figure 1
+% Figure 2
 The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}.
@@
 
-% Figure 2
+% Figure 3
 To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between seasons.
@@
 The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter.
+Genus richness is consistently highest in the top \qty{25}{meters} depth interval, which also shows the most variability over the time series.
 Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011.
+The collapse in diversity following the start of the time series does not follow the signal shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}).
 Scaling this genus diversity by the cell count abundances, i.e.
@@
 
-% Figure 3
+% Figure 4
 Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment.
@@
 
-% Figure 4
+% Figure 5
 To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
@@
 % HERE I INTERPRET AND DISCUSS, whereas before I just described. Answer not the *what*, but the *how* and *why*.
-Our main results show that...
-1. we see Overlap in yearly mean anomalies between certain variables
-
-- The Cariaco basin is a system, that...
-    - Discuss cell count collapse, apparent shift, but not as strong! Limitations of microscopy.
-
 This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes.
-In the Cariaco basin, long-term climatic oscillations have affected the wind-driven upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
-%need to rephrase below (from Pinckney)
-During the early years of the time series (1996–2004), the Cariaco basin experienced a regular seasonal upwelling cycle driven by strong Trade Winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
-This upwelling period represented \qty{\sim 70}{\%} of annual depth-integrated primary production with a phytoplankton community dominated by microphytoplankton (diatoms, dinoflagellates, and coccolithophorids) \cite{pinckney_phytoplankton_2015}.
-Following 2004, the physical and biogeochemical regime had changed to one with significantly reduced Trade Winds during the upwelling period, resulting in higher sea surface temperatures and less pronounced mixing of the surface and deep layers \cite{taylor_ecosystem_2012}.
-The microscopy cell counts indicate a collapse of large phytoplankton and diatoms in particular, which is confirmed to a lesser degree by pigment data analysis.
-In general, a shift in the phytoplankton community to smaller cell sizes and higher abundances of coccolithophores, cryptophytes, and other phytoflagellates, typical for stratified water columns was observed \cite{pinckney_phytoplankton_2015}.
-
-- Now mention how the variables overlap in anomalies in Figure 2, how that can be linked to the general changes already observed by other studies.
-
-- What we want to mention, but have not included in our analysis due to lack of data at appropriate coverage or resolution, is the indicators of top-down effects on the Cariaco basin ecosystem.
-Yearly landings of the Spanish sardine (\textit{Sardinella aurita}) fisheries off the NE coast of Venezuela collapsed after 2004 and measurements of zooplankton abundances since 2001 showed a steady increase in numbers up until 2010.
-These can be viewed as consequences of the changes apparent in the phytoplankton community, but can also be important drivers of that change, for example if the collapse in fisheries was mainly driven by overfishing.
-(MENTION THAT ZOO DATA IS TOO FEW DATAPOINTS TO INCLUDE IN GRADIENT FOREST MODEL RUN)
-
-2. mismatch between Diversity variables, and cell count abundances + community clustering / ecosystem shift
-
-   The time series includes multiple measurements of the phytoplankton community, which are the bulk fluorometric chlorophyll, microscopy cell counts, and high-performance liquid chromatography (HPLC) pigment analysis.
-Of these, all cover the entire time series except for the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.
-
-    The most often measured indicator of bulk phytoplankton biomass is fluorometric Chlorophyll \textit{a} concentration, which gives a general measure of biomass standing stock and distribution in the water column.
-A more advanced methodology, high-performance liquid chromatography (HPLC), allows individual pigment types to be quantified.
-These pigment distributions can be used to estimate the groups of phytoplankton that contribute to biomass.
-The most specific methodology to quantify phytoplankton communities is the counting and identification of individual phytoplankton cells, either manually under a microscope or automated in flow-through systems.
-The main drawbacks are the size limit for resolving smaller cells and, if done manually, the complexity of identifying phytoplankton species.
-
-3. Clustering shows return of community to previous cluster, confirmed by monthly NMDS, but EnvFit is interesting, as no variable clearly goes along gradient separating the two cluster (NMDS2)
-
-3.1 Turnover is different between two clusters, also corresponding to evennness between clusters
-
-3.2 Discuss which variables actually vary between clusters and how they could mechanistically explain difference between communities
-
-4. Gradient Forest clearly identifies climate indices as best predictors of community change, then nutrients etc.
-    This is really interesting, as it would be expected, that in situ-measurements are best predictor of community.
-BUT either the community + top down processes are capturing more of that variance, or we simply are missing a bottom-up variable that can explain it.
-    but that most likely requires mechanistic modeling to investigate further...
-
-    %- Discuss effect of AMO
-    % Which studies have already looked at effects of Climate on Ecosystem?
-    - AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
-The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
-    - Although regional effecs may vary, a positive phase in the AMO generally linked to periods of warmer temperatures and a shift of the ITCZ from north to south.
-
-    - Taylor 2012 looked at NAO and MEI, and mostly looked linear trends in variables (they missed the return of the community and regime).
-NAO is the north-south diplole in sea level pressure, that is primarily governed by atmospheric dynamics \cite{hurrell_north_2009}.
-NAO and AMO are l
-    - AMO is mentioned in 2019 review
-    "The observed large-scale multidecadal fluctuations in the Atlantic sea surface temperature (SST), has been referred to as the Atlantic Multidecadal Oscillation (AMO) \cite{kerr2000north} to emphasize the “multidecadal” character of this oceanic phenomenon and to distinguish it from the inter-annual variability associated with the atmospheric North Atlantic Oscillation (NAO) \cite{enfield2001atlantic}." from AMO website
-
-    %- Discuss effect of MEI
-    Since El Nino has strong teleconnection to Caribbean (see ref below from Review) we included this as variable to explain changes in community.
-From Results it is less pronounced than AMO, but still explains some variability (also ref to NMDS).
-
-    Interestingly, in Gradient Forest Model, Climate variables are the strongest predictors of changes in community (i.e. changes in genus abundances).
-    - Discuss what climate variables actually capture, and why they might be better predictors of community changes than local (directly measured) variables.
-
-    - AMO (ref to 2019 Review)
-    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al.
-2015).
-These climatic changes had a profound impact throughout the Atlantic (Alexander et al.
-2014, Nye et al.
-2014) and in the Cariaco basin \cite{taylor_ecosystem_2012,pinckney_phytoplankton_2015}.
-    - ENSO teleconnection (ref to 2019 Review)
-    The Caribbean experiences the most marked of the El Ni  ̃no–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Ni  ̃no–Southern Oscillation (Enfield & Mayer 1997).
-
-5. Compare to what happened at other long term time series in the ocean...
+In the Cariaco basin, long-term climate oscillations have affected the wind-driven upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
+During the initial phase of the time series (1996–2004), the Cariaco basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
+In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed.
+During upwelling, the phytoplankton community was dominated by microphytoplankton, consisting primarily of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.
+Following 2004,  there was a transition in the physical and biogeochemical regime characterized by a significant reduction in Trade Winds during the upwelling period, which led to elevated sea surface temperatures and decreased mixing between the surface and deep ocean layers \cite{taylor_ecosystem_2012}.
+Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis.
+In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}.
+
+We found that including the time series data up to its end in 2017 in our analysis clearly shows that the observed shifts were reversible.
+With a return to higher upwelling conditions in 2014, we also see a corresponding increase in biomass (see Figure \ref{fig:divts} $a$) and a return to a community more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$).
+This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine capture observed in 2016 and 2017 in a location south-east of the Isla de Magarita \cite{gomez_gaspar_variacion_2025}.
+
+When comparing the community between the two clusters...
+- In microscopy counts we see a stark reduction in cell counts, which corresponds to a reduced concentration of chlorophyll \textit{a} biomass particularly in the top \qty{25}{meters} of the water column.
+When including the HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement (check Pinckneys discssion on this again) actually shows an increase...
+- The community shows a stark reduction in diatom cell counts, but also other functional groups --> shift towards smaller cell sizes, not observable in light microscopy
+- Discuss why there is a shift to smaller cell sizes at low upwelling -> increased nutrient efficiency, cite some papers here.
+- Discuss the dominant (most observed) genera per Cluster (TODO with DATA Analysis) and how that relates to the conditions
+- Discuss how similar Part 1 and 2 of Cluster 1 are.. why are they clustered together?
+- Discuss why we do not observe a clear distinction in Diversity betweent he two clusters, but mention the species turnover difference and the community evenness increase.
+
+When comparing the environmental conditions between the two clusters...
+
+- The decrease in observed phytoplankton cells between the two clusters indicates...
+% talk about the general trends and what they can mean, changes in size distribution, changes in functional groups
 
@@
 2012 observed that the decline in cell abundances was disproportionate compared with the smaller decrease in total chlorophyll a, which implies that phytoplankton of smaller size were not detected in the surveys..
-This was supported by high-performance liquid chromatography analyses, which indicated a reduction in diatom concentrations and an increase in phytoplankton diversity.
-A return to a phytoplankton community structure similar to that observed in the late 1990s was observed in 2014 and 2015 (Figure 5a), as upwelling intensity increased to levels similar to those seen before 2003." from 2019 Review Legacy of CARIACO
+This was supported by high-performance liquid chromatography analyses, which indicated a reduction in diatom concentrations and an increase in phytoplankton diversity." from 2019 Review
+
+%"A return to a phytoplankton community structure similar to that observed in the late 1990s was observed in 2014 and 2015 (Figure 5a), as upwelling intensity increased to levels similar to those seen before 2003." from 2019 Review Legacy of CARIACO
 
@@
 
-- The decrease in observed phytoplankton cells between the two clusters indicates...
-% talk about the general trends and what they can mean, changes in size distribution, changes in functional groups
-
 - What could have driven these shifts?
@@
 
+These changes are well reflected in the yearly mean anomalies shown in figure \ref{fig:zscore}.
+After 2004 we see higher SST, lower wind speed, reduced fluorometric chlorophyll \textit{a}, a collapse in functional group cell counts, all matching a shift in the AMO toward positive mean anomalies until 2014.
+However, the shift in yearly mean genus richness does not follow the abundances in the cell counts.
+Instead, we see a marked reduction following 1998 up until 2010, where richness recovers four years before the community seems to return to previous levels in 2014.
+The same trend is apparent, although to a less pronounced degree, in the other diversity metrics.
+Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance.
+The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy data to deduce changes in the phytoplankton community.
+This limits the size range of the observable phytoplankton and thus the measurable diversity.
+We could deduce that we see a marked reduction in the diversity of microphytoplankton ahead of the collapse in abundance, and similarly an increase in microphytoplankton diversity a few years ahead of the return in abundance to previous levels.
+
+- major thing is that previous paper only looked at microscopy to 55m (Mutshinda to 75m and Pinckney), but now we look at full 100 m!
+
+- Seasonality and Depth Distribution of biomass and diversity.
+The diversity of phytoplankton peaks in the surface ocean, similar to the patterns observed using genetic methods \cite{yeh_contrasting_2022}.
+
+2. mismatch between Diversity variables, and cell count abundances + community clustering / ecosystem shift
+
+   The time series includes multiple measurements of the phytoplankton community, which are the bulk fluorometric chlorophyll, microscopy cell counts, and high-performance liquid chromatography (HPLC) pigment analysis.
+Of these, all cover the entire time series except for the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.
+
+    The most often measured indicator of bulk phytoplankton biomass is fluorometric Chlorophyll \textit{a} concentration, which gives a general measure of biomass standing stock and distribution in the water column.
+A more advanced methodology, high-performance liquid chromatography (HPLC), allows individual pigment types to be quantified.
+These pigment distributions can be used to estimate the groups of phytoplankton that contribute to biomass.
+The most specific methodology to quantify phytoplankton communities is the counting and identification of individual phytoplankton cells, either manually under a microscope or automated in flow-through systems.
+The main drawbacks are the size limit for resolving smaller cells and, if done manually, the complexity of identifying phytoplankton species.
+
+3.
+Clustering shows the return of the community to the previous cluster, confirmed by monthly NMDS, but EnvFit is interesting, as no variable clearly follows the gradient separating the two clusters (NMDS2).
+
+3.1 Turnover is different between two clusters, also corresponding to evenness between clusters
+
+3.2 Discuss which variables actually vary between clusters and how they could mechanistically explain difference between communities
+
+4. Gradient Forest clearly identifies climate indices as best predictors of community change, then nutrients etc.
+    This is really interesting, as it would be expected, that in situ-measurements are best predictor of community.
+BUT either the community + top down processes are capturing more of that variance, or we simply are missing a bottom-up variable that can explain it.
+    but that most likely requires mechanistic modeling to investigate further...
+
+    %- Discuss effect of AMO
+    % Which studies have already looked at effects of Climate on Ecosystem?
+    - AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
+The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
+    - Although regional effecs may vary, a positive phase in the AMO generally linked to periods of warmer temperatures and a shift of the ITCZ from north to south.
+
+    - Taylor 2012 looked at NAO and MEI, and mostly looked linear trends in variables (they missed the return of the community and regime).
+NAO is the north-south diplole in sea level pressure, that is primarily governed by atmospheric dynamics \cite{hurrell_north_2009}.
+NAO and AMO are l
+    - AMO is mentioned in 2019 review
+    "The observed large-scale multidecadal fluctuations in the Atlantic sea surface temperature (SST), has been referred to as the Atlantic Multidecadal Oscillation (AMO) \cite{kerr2000north} to emphasize the “multidecadal” character of this oceanic phenomenon and to distinguish it from the inter-annual variability associated with the atmospheric North Atlantic Oscillation (NAO) \cite{enfield2001atlantic}." from AMO website
+
+    - We group species and genus counts, to genus level, and use genus counts as variables the model predicts.
+This assumes that there are not very specific niches etc.
+within genus, but that they are homogenous --> caveat!
+
+    %- Discuss effect of MEI
+    Since El Nino has strong teleconnection to Caribbean (see ref below from Review) we included this as variable to explain changes in community.
+From Results it is less pronounced than AMO, but still explains some variability (also ref to NMDS).
+
+    Interestingly, in Gradient Forest Model, Climate variables are the strongest predictors of changes in community (i.e. changes in genus abundances).
+    - Discuss what climate variables actually capture, and why they might be better predictors of community changes than local (directly measured) variables.
+
+    - AMO (ref to 2019 Review)
+    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al.
+2015).
+These climatic changes had a profound impact throughout the Atlantic (Alexander et al.
+2014, Nye et al.
+2014) and in the Cariaco basin \cite{taylor_ecosystem_2012, pinckney_phytoplankton_2015}.
+    - ENSO teleconnection (ref to 2019 Review)
+    The Caribbean experiences the most marked of the El Ni  ̃no–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Ni  ̃no–Southern Oscillation (Enfield & Mayer 1997).
+
+5. Compare to what happened at other long-term time series in the ocean...
+
 % Pinckney et al. 2015:
-- major thing is that previous paper only looked at microscopy to 55m (Mutshinda to 75m and Pinckney), but now we look at full 100 m!
 %The Cariaco Basin has undergone marked shifts in biogeochemistry and food web structure over the past two decades (Taylor et al., 2012; Montes et al., 2013; Bates et al., 2014; Scranton et al., 2014).
@@
 
+%%% DISCUSSION DUMP:
+
+- What we want to mention, but have not included in our analysis due to lack of data at appropriate coverage or resolution, is the indicators of top-down effects on the Cariaco basin ecosystem.
+Yearly landings of the Spanish sardine (\textit{Sardinella aurita}) fisheries off the NE coast of Venezuela collapsed after 2004 and measurements of zooplankton abundances since 2001 showed a steady increase in numbers up until 2010.
+These can be viewed as consequences of the changes apparent in the phytoplankton community, but can also be important drivers of that change, for example if the collapse in fisheries was mainly driven by overfishing.
+(MENTION THAT ZOO DATA IS TOO FEW DATAPOINTS TO INCLUDE IN GRADIENT FOREST MODEL RUN)
+
 \section{Conclusions}

### 2025-03-19 agujournaltemplate.tex
Discussion just missing GF 90%

@@
 
-% HERE I INTERPRET AND DISCUSS, whereas before I just described. Answer not the *what*, but the *how* and *why*.
 This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes.
-In the Cariaco basin, long-term climate oscillations have affected the wind-driven upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
+In the Cariaco basin, decadal-scale climate oscillations have affected the wind-driven upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
 During the initial phase of the time series (1996–2004), the Cariaco basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
@@
 During upwelling, the phytoplankton community was dominated by microphytoplankton, consisting primarily of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.
+
 Following 2004, there was a transition in the physical and biogeochemical regime characterized by a significant reduction in Trade Winds during the upwelling period, which led to elevated sea surface temperatures and decreased mixing between the surface and deep ocean layers \cite{taylor_ecosystem_2012}.
+These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}).
+After 2004 we see higher SST, lower wind speed, reduced fluorometric chlorophyll \textit{a}, a collapse in functional group cell counts, all matching a shift in the AMO toward positive mean anomalies until 2014.
 Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis.
 In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}.
-
-We found that including the time series data up to its end in 2017 in our analysis clearly shows that the observed shifts were reversible.
-With a return to higher upwelling conditions in 2014, we also see a corresponding increase in biomass (see Figure \ref{fig:divts} $a$) and a return to a community more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$).
+During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, which coincided with an increase in zooplankton biomass measured in the Cariaco basin \cite{taylor_ecosystem_2012}.
+
+Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes can not "be clearly identified as unidirectional trends driven by anthrophogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".
+We found that including the time series data up to its end in 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}.
+In 2014, physical conditions in the Cariaco basin returned to a regime of higher wind speeds, increased uwpelling, and higher nutrient concentrations (see Figure \ref{fig:zscore}).
+At the same time, there was an increase in chlorophyll \textit{a} biomass (see Figure \ref{fig:divts} $a$) and a return to a community more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$).
 This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine capture observed in 2016 and 2017 in a location south-east of the Isla de Magarita \cite{gomez_gaspar_variacion_2025}.
+
+When comparing the phytoplankton community between the two clusters, we see a marked reduction in cell counts, which corresponds to a reduced concentration of chlorophyll \textit{a} biomass measured by fluorometry particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}).
+When including HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement here actually shows an increase between the periods 1996-2000 and 2006-2010 \cite{pinckney_phytoplankton_2015}, particularly at depths below \qty{55}{meters}.
+It should be noted that the HPLC data were analyzed by multiple labs throughout the time series, with a gap in data coverage between 2000 and 2006.
+\citeA{pinckney_phytoplankton_2015} could not find a consistent method for quantifying total chlorophyll \textit{a} between the different laboratories.
+Here, we focus on the more consistently sampled fluorometric measurements.
+Independent spectophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Magarita Island follow the temporal trend of a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
+It should be noted that changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a}, but also other related forms of chlorophyll, which can vary depending on the composition of the community \cite{welschmeyer_fluorometric_1994}.
+
+Despite some of the difficulty in reconciling HPLC pigment data and microscopy cell counts, they agree in a reduction in diatom abundance after 2004.
+Cell counts of other functional groups decreased, where HPLC data indicate an increase and redistribution of biomass to depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}.
+The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community transitioned to smaller individuals that were not captured under the microscope \cite{muller-karger_scientific_2019}.
+This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}.
+Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but are particularly prevalent when (as observed in the Cariaco basin) temperatures increase and the nutrient supply decreases \cite{mousing_global_2014}.
+
+If we compare the two parts of the first cluster, the community did reorganize, particularly the diatoms shifted from a community dominated by centric diatoms (Guinardia, Leptocylindrus, Skeletonema) to one dominated by pennate diatoms (Pseudonitzschia) \cite{pinckney_phytoplankton_2015}.
+
+However, the shift in yearly mean genus richness does not follow the abundances in the cell counts.
+Instead, we see a marked reduction following 1998 up until 2010, where richness recovers four years before the community seems to return to previous levels in 2014.
+The same trend is apparent, although to a less pronounced degree, in the other diversity metrics.
+Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance.
+%(Here check if this is due to "synechococcus" counts, that were abandoned past 2003). 
+The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy counts to deduce changes in the phytoplankton community.
+This limits the size range of the observable phytoplankton and the measurable diversity through the limited sampling volume \cite{cermeno_sampling_2014}.
+We can see a marked reduction in the diversity of microphytoplankton ahead of the collapse in abundance, and similarly an increase in microphytoplankton diversity a few years ahead of the return in abundance to previous levels.
+This pattern results in the fact that no significant differences in diversity are observed from microscopy cell counts between two clusters, either by Shannon index or by genus richness.
+We do see a significant difference in the Pilou index between the two clusters.
+Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during the low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}.
+This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and thereby reduced disturbance of the community.
+The resulting reduced compositional turnover is also observed in a decrease in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$).
+As the productivity and biomass of phytoplankton communities are mainly driven by nutrients supplied through disturbance events, this confirms the observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
+
+% NOW let's jump to Gradient Forest and discussing CLIMATE impacts! for a final paragraph, and then we are done here!!
+
+--------------------------------------------------------------------------
+----------
+----------
+----------
+----------
+----------
 
@@
 When including the HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement (check Pinckneys discssion on this again) actually shows an increase...
+- Independent spectophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Magarita Island follow the temporal trend of the fluorometric chlorophyll reduction between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
 - The community shows a stark reduction in diatom cell counts, but also other functional groups --> shift towards smaller cell sizes, not observable in light microscopy
+
+In the Cariaco basin, wind-driven seasonal upwelling dictates nutrient availability in the upper mixed layer, which, in turn, affects phytoplankton dynamics \cite{ptacnik_diversity_2008, bopp_response_2005}.
+In particular, diatoms respond to changes in the mixing regime \cite{huisman_reduced_2006}.
+Alterations in physical mixing affected phytoplankton dynamics,
+
 - Discuss why there is a shift to smaller cell sizes at low upwelling -> increased nutrient efficiency, cite some papers here.
 - Discuss the dominant (most observed) genera per Cluster (TODO with DATA Analysis) and how that relates to the conditions
-- Discuss how similar Part 1 and 2 of Cluster 1 are.. why are they clustered together?
-- Discuss why we do not observe a clear distinction in Diversity betweent he two clusters, but mention the species turnover difference and the community evenness increase.
+- Discuss how similar Part 1 and 2 of Cluster 1 are.. Why are they clustered together?
+- Discuss why we do not observe a clear distinction in Diversity between the two clusters, but mention the species turnover difference and the community evenness increase.
+
+- What could have driven these shifts?
+% talk about link to bottom up drivers, but also mention top down aspects...
+then go deep on Gradient Forest analysis, what that could mean.
+Relate this to other papers about community changes, climate effects and gradient forest applications.
 
 When comparing the environmental conditions between the two clusters...
-
-- The decrease in observed phytoplankton cells between the two clusters indicates...
-% talk about the general trends and what they can mean, changes in size distribution, changes in functional groups
+- When looking at distributions of measured environmental variables, there is a clear trend signifying the reduction in upwelling conditions during the period from 2004 to 2014 (cluster 2).
+- We have lower wind speed, lower Isotherm, higher Temp, lower Nuts (except Silicate -> river runoff) lower salinity -> less mixing, higher precipitation.
+
+3.
+Clustering shows the return of the community to the previous cluster, confirmed by monthly NMDS, but EnvFit is interesting, as no variable clearly follows the gradient separating the two clusters (NMDS2).
+
+- Climate indices: AMO stark difference between clusters, cluster 1 is negative anomaly, cluster 2 is positive anomaly, corresponding to effects of AMO -> ITCZ etc.
+- MEIv2 less clear, but shows a preponderance of positive anomalies in cluster 1 ... why???
+% MEI index relevant (from 2019 REVIEW)
+The Caribbean experiences the most marked of the El Niño–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Niño–Southern Oscillation \cite{enfield_tropical_1997}.
+
+4. Gradient Forest clearly identifies climate indices as best predictors of community change, then nutrients etc.
+    This is really interesting, as it would be expected, that in situ-measurements are best predictor of community.
+BUT either the community + top down processes are capturing more of that variance, or we simply are missing a bottom-up variable that can explain it.
+    but that most likely requires mechanistic modeling to investigate further...
+
+    %- Discuss effect of AMO
+    % Which studies have already looked at effects of Climate on Ecosystem?
+    - AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
+The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
+    - Although regional effecs may vary, a positive phase in the AMO generally linked to periods of warmer temperatures and a shift of the ITCZ from north to south.
+
+    - Taylor 2012 looked at NAO and MEI, and mostly looked linear trends in variables (they missed the return of the community and regime).
+NAO is the north-south diplole in sea level pressure, that is primarily governed by atmospheric dynamics \cite{hurrell_north_2009}.
+NAO and AMO are l
+    - AMO is mentioned in 2019 review
+    "The observed large-scale multidecadal fluctuations in the Atlantic sea surface temperature (SST), has been referred to as the Atlantic Multidecadal Oscillation (AMO) \cite{kerr2000north} to emphasize the “multidecadal” character of this oceanic phenomenon and to distinguish it from the inter-annual variability associated with the atmospheric North Atlantic Oscillation (NAO) \cite{enfield2001atlantic}." from AMO website
+
+    %- Discuss effect of MEI
+    Since El Nino has strong teleconnection to Caribbean (see ref below from Review) we included this as variable to explain changes in community.
+From Results it is less pronounced than AMO, but still explains some variability (also ref to NMDS).
+
+    Interestingly, in Gradient Forest Model, Climate variables are the strongest predictors of changes in community (i.e. changes in genus abundances).
+    - Discuss what climate variables actually capture, and why they might be better predictors of community changes than local (directly measured) variables.
+
+    - AMO (ref to 2019 Review)
+    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al.
+2015).
+These climatic changes had a profound impact throughout the Atlantic (Alexander et al.
+2014, Nye et al.
+2014) and in the Cariaco basin \cite{taylor_ecosystem_2012, pinckney_phytoplankton_2015}.
+    - ENSO teleconnection (ref to 2019 Review)
+    The Caribbean experiences the most marked of the El Ni  ̃no–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Ni  ̃no–Southern Oscillation (Enfield & Mayer 1997).
 
@@
 2023
+- We group species and genus counts, to genus level, and use genus counts as variables the model predicts.
+This assumes that there are not very specific niches etc.
+within genus, but that they are homogenous --> caveat!
 - Discuss limitation of taxonomic diversity, i.e. not direct translation to functional diversity, but here no Data foundation to look at functional diversity.. (inspired by Borics et al. 2021)
-
-- What could have driven these shifts?
-% talk about link to bottom up drivers, but also mention top down aspects...
-then go deep on Gradient Forest analysis, what that could mean.
-Relate this to other papers about community changes, climate effects and gradient forest applications.
-
-In the Cariaco basin, wind-driven seasonal upwelling dictates nutrient availability in the upper mixed layer, which, in turn, affects phytoplankton dynamics \cite{ptacnik_diversity_2008, bopp_response_2005}.
-In particular, diatoms respond to changes in the mixing regime \cite{huisman_reduced_2006}.
-Alterations in physical mixing affected phytoplankton dynamics,
-% MEI index relevant (from 2019 REVIEW)
-The Caribbean experiences the most marked of the El Niño–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Niño–Southern Oscillation \cite{enfield_tropical_1997}.
 
@@
 Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance.
+%(Here check if this is due to "synechococcus" counts, that were abandoned past 2003). 
 The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy data to deduce changes in the phytoplankton community.
@@
 
-2. mismatch between Diversity variables, and cell count abundances + community clustering / ecosystem shift
-
-   The time series includes multiple measurements of the phytoplankton community, which are the bulk fluorometric chlorophyll, microscopy cell counts, and high-performance liquid chromatography (HPLC) pigment analysis.
+2. Mismatch between Diversity variables, and cell count abundances + community clustering / ecosystem shift
+
+   The time series includes multiple measurements of the phytoplankton community, which are bulk fluorometric chlorophyll, microscopy cell counts, and high-performance liquid chromatography (HPLC) pigment analysis.
 Of these, all cover the entire time series except for the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.
@@
 The main drawbacks are the size limit for resolving smaller cells and, if done manually, the complexity of identifying phytoplankton species.
-
-3.
-Clustering shows the return of the community to the previous cluster, confirmed by monthly NMDS, but EnvFit is interesting, as no variable clearly follows the gradient separating the two clusters (NMDS2).
-
-3.1 Turnover is different between two clusters, also corresponding to evenness between clusters
-
-3.2 Discuss which variables actually vary between clusters and how they could mechanistically explain difference between communities
-
-4. Gradient Forest clearly identifies climate indices as best predictors of community change, then nutrients etc.
-    This is really interesting, as it would be expected, that in situ-measurements are best predictor of community.
-BUT either the community + top down processes are capturing more of that variance, or we simply are missing a bottom-up variable that can explain it.
-    but that most likely requires mechanistic modeling to investigate further...
-
-    %- Discuss effect of AMO
-    % Which studies have already looked at effects of Climate on Ecosystem?
-    - AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
-The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
-    - Although regional effecs may vary, a positive phase in the AMO generally linked to periods of warmer temperatures and a shift of the ITCZ from north to south.
-
-    - Taylor 2012 looked at NAO and MEI, and mostly looked linear trends in variables (they missed the return of the community and regime).
-NAO is the north-south diplole in sea level pressure, that is primarily governed by atmospheric dynamics \cite{hurrell_north_2009}.
-NAO and AMO are l
-    - AMO is mentioned in 2019 review
-    "The observed large-scale multidecadal fluctuations in the Atlantic sea surface temperature (SST), has been referred to as the Atlantic Multidecadal Oscillation (AMO) \cite{kerr2000north} to emphasize the “multidecadal” character of this oceanic phenomenon and to distinguish it from the inter-annual variability associated with the atmospheric North Atlantic Oscillation (NAO) \cite{enfield2001atlantic}." from AMO website
-
-    - We group species and genus counts, to genus level, and use genus counts as variables the model predicts.
-This assumes that there are not very specific niches etc.
-within genus, but that they are homogenous --> caveat!
-
-    %- Discuss effect of MEI
-    Since El Nino has strong teleconnection to Caribbean (see ref below from Review) we included this as variable to explain changes in community.
-From Results it is less pronounced than AMO, but still explains some variability (also ref to NMDS).
-
-    Interestingly, in Gradient Forest Model, Climate variables are the strongest predictors of changes in community (i.e. changes in genus abundances).
-    - Discuss what climate variables actually capture, and why they might be better predictors of community changes than local (directly measured) variables.
-
-    - AMO (ref to 2019 Review)
-    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al.
-2015).
-These climatic changes had a profound impact throughout the Atlantic (Alexander et al.
-2014, Nye et al.
-2014) and in the Cariaco basin \cite{taylor_ecosystem_2012, pinckney_phytoplankton_2015}.
-    - ENSO teleconnection (ref to 2019 Review)
-    The Caribbean experiences the most marked of the El Ni  ̃no–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Ni  ̃no–Southern Oscillation (Enfield & Mayer 1997).
 
@@
 
-FOR THE CLUSTERING (moved from Results): This confirms previously observed patterns in the Cariaco ecosystem, where a strong shift in the community happened in 2004, but the community returned to the previous regime past 2014 \cite{taylor_ecosystem_2012, muller-karger_scientific_2019}.
-
 %%% DISCUSSION DUMP:
@@
 (MENTION THAT ZOO DATA IS TOO FEW DATAPOINTS TO INCLUDE IN GRADIENT FOREST MODEL RUN)
+
+FOR THE CLUSTERING (moved from Results): This confirms previously observed patterns in the Cariaco ecosystem, where a strong shift in the community happened in 2004, but the community returned to the previous regime past 2014 \cite{taylor_ecosystem_2012, muller-karger_scientific_2019}.
 

### 2025-03-24 agujournaltemplate.tex
Discussion 98% done - pre save

@@
     The monthly AMO index data (10 year lowpass), which is as an area average of detrended low-pass filtered North-Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} (accessed in June, 2024).
+    %The term "Atlantic Multidecadal Oscillation" (AMO) describes the large-scale fluctuations in Atlantic sea surface temperature (SST) that occur over multiple decades \cite{kerr2000north}.
+This designation is intended to underscore the distinct multidecadal nature of these oceanic variations in contrast to the inter-annual variability linked with the atmospheric North Atlantic Oscillation (NAO) \cite{enfield2001atlantic}.
 
@@
 The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}.
-The patterns emerging from the -u 10m wind component, the temperature and the directly related \qty{21}{\celcius} isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m.
+The patterns emerging from the -u 10m wind component, the temperature and the directly related \qty{21}{\celsius} isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m.
 These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes.
@@
 Upwelling related variables show distinct patterns between the two clusters.
-For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celcius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate.
+For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate.
 The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies.
@@
 \begin{center}
-\noindent\includegraphics[width=480]{fig/DepthDivTimeSeries_updated.pdf}
+\noindent\includegraphics[width=480pt]{fig/DepthDivTimeSeries_updated.pdf}
 \end{center}
@@
 u-component 10 m wind speed & 7074.0 & 0.143 & n.s.\\
-Isotherm \qty{21}{\celcius} & 5293.0 & 0.034 & **\\
+Isotherm \qty{21}{\celsius} & 5293.0 & 0.034 & **\\
 \addlinespace
@@
 This limits the size range of the observable phytoplankton and the measurable diversity through the limited sampling volume \cite{cermeno_sampling_2014}.
-We can see a marked reduction in the diversity of microphytoplankton ahead of the collapse in abundance, and similarly an increase in microphytoplankton diversity a few years ahead of the return in abundance to previous levels.
+We can see a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly an increase a few years ahead of the return in abundance to previous levels.
 This pattern results in the fact that no significant differences in diversity are observed from microscopy cell counts between two clusters, either by Shannon index or by genus richness.
@@
 Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during the low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}.
-This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and thereby reduced disturbance of the community.
+This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and the thereby reduced disturbance of the community.
 The resulting reduced compositional turnover is also observed in a decrease in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$).
@@
 % NOW let's jump to Gradient Forest and discussing CLIMATE impacts! for a final paragraph, and then we are done here!!
+To identify the variables that can best predict changes in the phytoplankton community, we used a gradient forest analysis using microscopy cell counts at the genus level as predicted variables and climate indices and in situ measurements of bottom-up drivers as predictors.
+Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series.
+Interestingly, our analysis found that the strongest predictors of changes within the community are not in situ measurements, but actually the AMO and the MEI v.2 climate indices.
+These are followed in importance by nitrate concentration and temperature in the upper \qty{100}{meters} (see Figure \ref{fig:GF}).
+
+These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, since the scaling axis that actually separates the two yearly aggregate community clusters is NMDS 1, which covaries most strongly with changes in the AMO, MEI v.2 (see Figure \ref{fig:clustering} $c$).
+Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS2.
+This indicates that the climate indices capture some variability, which is not represented in the more direct measurements of upwelling intensity.
+Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling, but also depend on precipitation and fresh water input through surface runoff, show greater variance along NMDS1.
+Both salinity and silicate concentration show some explanatory power in the gradient forest analysis, although they are weaker predictors than temperature, nitrate concentration, and the climate indices.
+
+Using microscopy cell counts as a basis for this analysis meant that the results should be interpreted in terms of changes in the microphytoplankton community, due to the size limitation of identifying smaller species via light microscopy.
+Grouping cell counts by genera does create a loss of resolution, but is based on the assumption that species within a genus generally show similar ecophysiological responses.
+The output of the gradient forest model is further grouped by broad functional types to facilitate the identification of shifts and trends.
+The phytoplankton genera that can be best predicted are mostly diatoms, but also some haptophytes, dinoflagellates, and cyanobacteria (see Figure \ref{fig:GF} $b$).
+Within these groups, we see differing threshold responses in cumulative importance along the predictor ranges.
+
+AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
+The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
+Although regional effecs may vary, a positive phase in the AMO is generally associated with periods of warmer temperatures and a shift of the ITCZ from north to south.
+This shift of the ITCZ southward increases wind-driven upwelling in the Cariaco basin \cite{taylor_ecosystem_2012}.
+The threshold responses between functional groups, which relate to a shift in the cell count abundances, show contrasting patterns.
+Diatoms and cyanobacteria show strong response thresholds for negative AMO anomalies.
+This is an interesting threshold response shown for the AMO, which, if sufficiently negative, seems to correlate to a reorganization of the physical regime in the Cariaco basin.
+The abundance of haptophytes shows a strong threshold at positive AMO anomalies, which could be related to the pattern observed from HPLC data of haptophytes mostly exhibiting blooms above the mixed layer and at a lower frequency from other functional groups \cite{pinckney_phytoplankton_2015}.
+
+The El Niño Southern Osciallation (ENSO) has a strong teleconnection to the Caribbean Sea, since during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
+\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measuerments in the Cariaco basin, and only \qtyrange{24}{36}{\%} explanantion of variance with a time lag of 12 months using linear regression.
+In contrast, we found a relatively strong predictive power for changes in the phytoplankton community with a 4-month time lag.
+We also tested a 12 month lag and found no significant improvement.
+The threshold responses show a shift in the abundances of dinoflagellates and haptophytes for a negative MEI v.2 index, which indicates La Niña conditions.
+For cyanobacteria and haptophytes, we found a pronounced threshold response for positive anomalies showing a sensitivity to ENSO events.
+Diatoms do not show a clear threshold response for the MEI v.2 index, but are affected across the entire range.
+Positive anomalies of the MEI v.2, representing strong ENSO events, are interestingly all found in Cluster 1 (see Figure \ref{fig:clustering} $e$).
+
+The in-situ measurements that best predict changes in the microphytoplankton community were nitrate concentration, temperature and salinity, in that order.
+Nitrate is a central nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco basin \cite{muller-karger_scientific_2019}.
+For dinoflagelltes and diatoms we see a strong threshold response around \qty{7}{\micro\molar}, which would indicate that strong upwelling conditions drive blooms of these functional groups.
+The "other" functional group including \textit{Mesodinium} and \textit{Eutreptria} seems to have a threshold response at even higher nitrate concentrations (around \qty{9}{\micro\molar}), indicating an effect on their abundance by very strong upwelling.
+Similarly, for temperature dinoflagellates and diatoms show a threshold around \qtyrange{21}{23}{\celsius}, which indicates higher abundances during low temperatures in the upper\qty{100}{m} driven by strong upwelling.
+For the other predictors, we do not see strong threshold responses.
+
+Gradient forest analysis clearly identified the AMO and MEI v.2 climate indices as the best predictors of microphytoplankton community changes in the Cariaco basin, in our limited set of bottom-up drivers.
+This is an interesting result, as it would be expected that more direct measures of ecosystem variability could better predict shifts in abundances.
+One possible explanation is that in situ measurements contain much stronger variability due to measurement error, local and temporal variability, and generally only provide a single snapshot for an entire month.
+Climate indices are generally detrended and integrated measurements of long-term changes, much more matching in scale to the dynamics of long-term changes within a phytoplankton community.
+Yet we do not see such strong predictive power in the ERA5 climate reanalysis-derived wind speed, which also aggregates measurements and model output over an entire month.
+Another explanation could be that the climate indices capture a dimension of variability not found in our in-situ measurements, which is also represented in our NMDS clustering (see Figure \ref{fig:clustering} $c$).
+
+Our limitiation to bottom-up drivers of community change could also limit the explanatory power of our analysis, as top-down processes such as zooplankton grazing strongly affects phytoplankton community structure.
+Due to the lack of data resolution for higher trophic levels, and the lack of zooplankton data for the start of the time series, this might require mechanistic modeling studies to investigate this aspect of the CARIACO time series further.
+Also the limitations of microscopy.
+
+Comparison to Sargasso Sea, changes in temp and upwelling and community, BUT export remains steady \cite{lomas_increased_2010, lomas_adaptive_2022}.
+Most likely mediated and strong influence of grazers, not resolved in our study.
+
+- Mention other papers correlatting AMO to ecosystems!
+
+Generally we see a strong impact of decadal-scale climatic oscilaltions on shifts in the tropical coastal ecosystem of the Cariaco basin.
+The microphytoplankton community recovered to previous levels, despite general warming trends, highlighting the resilience of ocean ecosystems under global change.
 
@@
 
-When comparing the community between the two clusters...
-- In microscopy counts we see a stark reduction in cell counts, which corresponds to a reduced concentration of chlorophyll \textit{a} biomass particularly in the top \qty{25}{meters} of the water column.
-When including the HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement (check Pinckneys discssion on this again) actually shows an increase...
-- Independent spectophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Magarita Island follow the temporal trend of the fluorometric chlorophyll reduction between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
-- The community shows a stark reduction in diatom cell counts, but also other functional groups --> shift towards smaller cell sizes, not observable in light microscopy
-
-In the Cariaco basin, wind-driven seasonal upwelling dictates nutrient availability in the upper mixed layer, which, in turn, affects phytoplankton dynamics \cite{ptacnik_diversity_2008, bopp_response_2005}.
-In particular, diatoms respond to changes in the mixing regime \cite{huisman_reduced_2006}.
-Alterations in physical mixing affected phytoplankton dynamics,
-
-- Discuss why there is a shift to smaller cell sizes at low upwelling -> increased nutrient efficiency, cite some papers here.
-- Discuss the dominant (most observed) genera per Cluster (TODO with DATA Analysis) and how that relates to the conditions
-- Discuss how similar Part 1 and 2 of Cluster 1 are.. Why are they clustered together?
-- Discuss why we do not observe a clear distinction in Diversity between the two clusters, but mention the species turnover difference and the community evenness increase.
-
 - What could have driven these shifts?
@@
 Relate this to other papers about community changes, climate effects and gradient forest applications.
-
-When comparing the environmental conditions between the two clusters...
-- When looking at distributions of measured environmental variables, there is a clear trend signifying the reduction in upwelling conditions during the period from 2004 to 2014 (cluster 2).
-- We have lower wind speed, lower Isotherm, higher Temp, lower Nuts (except Silicate -> river runoff) lower salinity -> less mixing, higher precipitation.
 

### 2025-04-01 agujournaltemplate.tex
pre-adding new figures

@@
 
-%% \begin{abstract} starts the second page
+%% \begin{abstract} starts the second page of the manuscript
 
@@
 
-Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes can not "be clearly identified as unidirectional trends driven by anthrophogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".
+Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthrophogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".
 We found that including the time series data up to its end in 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}.
@@
 We can see a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly an increase a few years ahead of the return in abundance to previous levels.
-This pattern results in the fact that no significant differences in diversity are observed from microscopy cell counts between two clusters, either by Shannon index or by genus richness.
+This pattern results in the fact that no significant differences in diversity are observed from microscopy cell counts between two clusters, either by Shannon index or by genus richness (see Figure \ref{fig:zscore} and \ref{fig:divts}).
 We do see a significant difference in the Pilou index between the two clusters.
@@
 Diatoms do not show a clear threshold response for the MEI v.2 index, but are affected across the entire range.
-Positive anomalies of the MEI v.2, representing strong ENSO events, are interestingly all found in Cluster 1 (see Figure \ref{fig:clustering} $e$).
+Positive anomalies of the MEI v.2, representing strong ENSO events, are interestingly only found in Cluster 1 (see Figure \ref{fig:clustering} $e$).
 
@@
 Nitrate is a central nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco basin \cite{muller-karger_scientific_2019}.
-For dinoflagelltes and diatoms we see a strong threshold response around \qty{7}{\micro\molar}, which would indicate that strong upwelling conditions drive blooms of these functional groups.
-The "other" functional group including \textit{Mesodinium} and \textit{Eutreptria} seems to have a threshold response at even higher nitrate concentrations (around \qty{9}{\micro\molar}), indicating an effect on their abundance by very strong upwelling.
-Similarly, for temperature dinoflagellates and diatoms show a threshold around \qtyrange{21}{23}{\celsius}, which indicates higher abundances during low temperatures in the upper\qty{100}{m} driven by strong upwelling.
+For dinoflagelltes and diatoms we see a strong threshold response around \qty{7}{\micro \mole}, which would indicate that strong upwelling conditions drive blooms of these functional groups.
+The "other" functional group including \textit{Mesodinium} and \textit{Eutreptria} seems to have a threshold response at even higher nitrate concentrations (around \qty{9}{\micro \mole}), indicating an effect on their abundance by very strong upwelling.
+Similarly, for temperature, dinoflagellates and diatoms show a threshold around \qtyrange{21}{23}{\celsius}, which indicates higher abundances during low temperatures in the upper\qty{100}{m} driven by strong upwelling.
 For the other predictors, we do not see strong threshold responses.
 
-Gradient forest analysis clearly identified the AMO and MEI v.2 climate indices as the best predictors of microphytoplankton community changes in the Cariaco basin, in our limited set of bottom-up drivers.
+Gradient forest analysis clearly identified the AMO and MEI v.2 climate indices as the best predictors of changes in the microphytoplankton community in the Cariaco basin.
 This is an interesting result, as it would be expected that more direct measures of ecosystem variability could better predict shifts in abundances.
 One possible explanation is that in situ measurements contain much stronger variability due to measurement error, local and temporal variability, and generally only provide a single snapshot for an entire month.
-Climate indices are generally detrended and integrated measurements of long-term changes, much more matching in scale to the dynamics of long-term changes within a phytoplankton community.
-Yet we do not see such strong predictive power in the ERA5 climate reanalysis-derived wind speed, which also aggregates measurements and model output over an entire month.
-Another explanation could be that the climate indices capture a dimension of variability not found in our in-situ measurements, which is also represented in our NMDS clustering (see Figure \ref{fig:clustering} $c$).
-
-Our limitiation to bottom-up drivers of community change could also limit the explanatory power of our analysis, as top-down processes such as zooplankton grazing strongly affects phytoplankton community structure.
-Due to the lack of data resolution for higher trophic levels, and the lack of zooplankton data for the start of the time series, this might require mechanistic modeling studies to investigate this aspect of the CARIACO time series further.
-Also the limitations of microscopy.
-
-Comparison to Sargasso Sea, changes in temp and upwelling and community, BUT export remains steady \cite{lomas_increased_2010, lomas_adaptive_2022}.
-Most likely mediated and strong influence of grazers, not resolved in our study.
-
-- Mention other papers correlatting AMO to ecosystems!
-
-Generally we see a strong impact of decadal-scale climatic oscilaltions on shifts in the tropical coastal ecosystem of the Cariaco basin.
-The microphytoplankton community recovered to previous levels, despite general warming trends, highlighting the resilience of ocean ecosystems under global change.
-
---------------------------------------------------------------------------
-----------
-----------
-----------
-----------
-----------
-
-- What could have driven these shifts?
-% talk about link to bottom up drivers, but also mention top down aspects...
-then go deep on Gradient Forest analysis, what that could mean.
-Relate this to other papers about community changes, climate effects and gradient forest applications.
-
-3.
-Clustering shows the return of the community to the previous cluster, confirmed by monthly NMDS, but EnvFit is interesting, as no variable clearly follows the gradient separating the two clusters (NMDS2).
-
-- Climate indices: AMO stark difference between clusters, cluster 1 is negative anomaly, cluster 2 is positive anomaly, corresponding to effects of AMO -> ITCZ etc.
-- MEIv2 less clear, but shows a preponderance of positive anomalies in cluster 1 ... why???
-% MEI index relevant (from 2019 REVIEW)
-The Caribbean experiences the most marked of the El Niño–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Niño–Southern Oscillation \cite{enfield_tropical_1997}.
-
-4. Gradient Forest clearly identifies climate indices as best predictors of community change, then nutrients etc.
-    This is really interesting, as it would be expected, that in situ-measurements are best predictor of community.
-BUT either the community + top down processes are capturing more of that variance, or we simply are missing a bottom-up variable that can explain it.
-    but that most likely requires mechanistic modeling to investigate further...
-
-    %- Discuss effect of AMO
-    % Which studies have already looked at effects of Climate on Ecosystem?
-    - AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
-The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
-    - Although regional effecs may vary, a positive phase in the AMO generally linked to periods of warmer temperatures and a shift of the ITCZ from north to south.
-
-    - Taylor 2012 looked at NAO and MEI, and mostly looked linear trends in variables (they missed the return of the community and regime).
-NAO is the north-south diplole in sea level pressure, that is primarily governed by atmospheric dynamics \cite{hurrell_north_2009}.
-NAO and AMO are l
-    - AMO is mentioned in 2019 review
-    "The observed large-scale multidecadal fluctuations in the Atlantic sea surface temperature (SST), has been referred to as the Atlantic Multidecadal Oscillation (AMO) \cite{kerr2000north} to emphasize the “multidecadal” character of this oceanic phenomenon and to distinguish it from the inter-annual variability associated with the atmospheric North Atlantic Oscillation (NAO) \cite{enfield2001atlantic}." from AMO website
-
-    %- Discuss effect of MEI
-    Since El Nino has strong teleconnection to Caribbean (see ref below from Review) we included this as variable to explain changes in community.
-From Results it is less pronounced than AMO, but still explains some variability (also ref to NMDS).
-
-    Interestingly, in Gradient Forest Model, Climate variables are the strongest predictors of changes in community (i.e. changes in genus abundances).
-    - Discuss what climate variables actually capture, and why they might be better predictors of community changes than local (directly measured) variables.
-
-    - AMO (ref to 2019 Review)
-    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al.
-2015).
-These climatic changes had a profound impact throughout the Atlantic (Alexander et al.
-2014, Nye et al.
-2014) and in the Cariaco basin \cite{taylor_ecosystem_2012, pinckney_phytoplankton_2015}.
-    - ENSO teleconnection (ref to 2019 Review)
-    The Caribbean experiences the most marked of the El Ni  ̃no–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Ni  ̃no–Southern Oscillation (Enfield & Mayer 1997).
-
-- We are not able to fully resolve....
-% "Phytoplankton communities exhibit a great deal of variation due to a combination of regulation of growth by bottom-up factors such as resource availability (Grover 1997), variation in mortality due to grazing, viral and parasitoid attack (Alpine & Cloern 1992; Mu ̈ hling et al.
-2005), and demographic drift among species (neutral variation).
-Bottom-up factors can be complex, depending on nutrient supply rates and ratios (Tilman 1982), light availability and photoinhibition (Six et al.
-2007; Alderkamp et al.
-2010), species-specific competitive ability for resources (Litchman 2007) and other factors influencing physiological responses such as temperature (Eppley 1972).
-Variation in mortality is much less frequently measured than the factors promoting growth due to the technical challenges in making measurements.
-None of our environmental variables are good proxies of grazing rates or whether the abundance of a particular population is determined by grazer control or resource supply at a particular time.
-We have ident" from Mutshinda et al.
-2013
-
-%"Taylor et al.
-2012 observed that the decline in cell abundances was disproportionate compared with the smaller decrease in total chlorophyll a, which implies that phytoplankton of smaller size were not detected in the surveys..
-This was supported by high-performance liquid chromatography analyses, which indicated a reduction in diatom concentrations and an increase in phytoplankton diversity." from 2019 Review
-
-%"A return to a phytoplankton community structure similar to that observed in the late 1990s was observed in 2014 and 2015 (Figure 5a), as upwelling intensity increased to levels similar to those seen before 2003." from 2019 Review Legacy of CARIACO
-
-- We are analyzing long term time series of data, where probably different analysts have counted the phytoplankton samples...
-%It is well known that the personal component has an influence on how well species are identified.
-Even the same person counting the samples can evolve over time, gaining more experience, which can also influence the results (Löder et al.
-2012; Nohe et al.
-2018).
-A recommendation for further improvement of the assessment is a case wise cross-validation of some samples by the monitoring agencies involved.
-Another important issue to be considered are the differences in sampling frequency between countries and also between stations (see ESM Fig. S1).
-In highly dynamic and turbid waters such as the Wadden Sea, low frequency and infrequent sampling is not sufficient to properly capture temporal signals (Fettweis et al.
-2023).
-% from Di Cavalho et al.
-2023
-- We group species and genus counts, to genus level, and use genus counts as variables the model predicts.
-This assumes that there are not very specific niches etc.
-within genus, but that they are homogenous --> caveat!
-- Discuss limitation of taxonomic diversity, i.e. not direct translation to functional diversity, but here no Data foundation to look at functional diversity.. (inspired by Borics et al. 2021)
-
-These changes are well reflected in the yearly mean anomalies shown in figure \ref{fig:zscore}.
-After 2004 we see higher SST, lower wind speed, reduced fluorometric chlorophyll \textit{a}, a collapse in functional group cell counts, all matching a shift in the AMO toward positive mean anomalies until 2014.
-However, the shift in yearly mean genus richness does not follow the abundances in the cell counts.
-Instead, we see a marked reduction following 1998 up until 2010, where richness recovers four years before the community seems to return to previous levels in 2014.
-The same trend is apparent, although to a less pronounced degree, in the other diversity metrics.
-Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance.
-%(Here check if this is due to "synechococcus" counts, that were abandoned past 2003). 
-The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy data to deduce changes in the phytoplankton community.
-This limits the size range of the observable phytoplankton and thus the measurable diversity.
-We could deduce that we see a marked reduction in the diversity of microphytoplankton ahead of the collapse in abundance, and similarly an increase in microphytoplankton diversity a few years ahead of the return in abundance to previous levels.
-
-- major thing is that previous paper only looked at microscopy to 55m (Mutshinda to 75m and Pinckney), but now we look at full 100 m!
-
-- Seasonality and Depth Distribution of biomass and diversity.
-The diversity of phytoplankton peaks in the surface ocean, similar to the patterns observed using genetic methods \cite{yeh_contrasting_2022}.
-
-2. Mismatch between Diversity variables, and cell count abundances + community clustering / ecosystem shift
-
-   The time series includes multiple measurements of the phytoplankton community, which are bulk fluorometric chlorophyll, microscopy cell counts, and high-performance liquid chromatography (HPLC) pigment analysis.
-Of these, all cover the entire time series except for the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.
-
-    The most often measured indicator of bulk phytoplankton biomass is fluorometric Chlorophyll \textit{a} concentration, which gives a general measure of biomass standing stock and distribution in the water column.
-A more advanced methodology, high-performance liquid chromatography (HPLC), allows individual pigment types to be quantified.
-These pigment distributions can be used to estimate the groups of phytoplankton that contribute to biomass.
-The most specific methodology to quantify phytoplankton communities is the counting and identification of individual phytoplankton cells, either manually under a microscope or automated in flow-through systems.
-The main drawbacks are the size limit for resolving smaller cells and, if done manually, the complexity of identifying phytoplankton species.
-
-5. Compare to what happened at other long-term time series in the ocean...
-
-% Pinckney et al. 2015:
-%The Cariaco Basin has undergone marked shifts in biogeochemistry and food web structure over the past two decades (Taylor et al., 2012; Montes et al., 2013; Bates et al., 2014; Scranton et al., 2014).
-Since the initiation of the CARIACO Ocean Time-Series, near-surface (0–55 m) chl a concentrations (measured by fluorescence of pigments extracted in methanol) and net primary production rates (NPP, measured using 14C incubations) have declined, particularly between 2004 and 2010 (Taylor et al., 2012; Scranton et al., 2014).
-This is hypothesized to be the result of a measured decrease in seasonal trade wind intensity, which has reduced coastal upwelling and led to shallower mixed layer depths (Taylor et al., 2012).
-Microscopy observations indicate that microphytoplankton assemblages (420 mm) within the euphotic zone have shifted from a community comprised mainly of diatoms, dinoflagellates, and coccolithophorids to one with increasing numbers of smaller taxa including cyanobacteria (Montes et al., 2013).
-During this period, a collapse in the sardine industry also occurred.
-The causes are not clear, but the phenomenon was accompanied by an increase in zooplankton biomass and an increase in particulate organic carbon export below the euphotic zone (Taylor et al., 2012).
-These results suggest that changes in zooplankton abundance and therefore grazing may have contributed to the decline in diatoms and to more efficient carbon export to depth.
-% Consistent with the results of Taylor et al.
-(2012), the microscopy data show large scale reductions in overall microphytoplankton (4 20 mm) cell densities and a large decrease in both diatoms and coccolithophores over the upper 55 m.
-However, the observed photopigment decline in diatom abundance is not nearly to the same extent and coccolithophores had a significant concentration increase of 60.1% over this same depth interval (Table 3).
-We argue that the reduction in cell numbers is likely only true of larger cells.
-Microscopic cell counts are inherently selective and biased toward easily identifiable species and cells that remain intact after preservation (Lund et al., 1958; Rott, 1981; Wilhelm et al., 1991; Schlüter et al., 2000).
-Furthermore, Mousing et al.
-(2014) found that small increases in temperature (as observed in the Cariaco Basin) lead to a decrease in the relative contribution of large cells in the community, regardless of ambient nutrient availability.
-Finally, depth limited microscopy data misses the downward movement of diatoms and coccolithophores deeper in the water column (i.e.
-below 56 m).
-In fact, there was also a dramatic shift in diatom speciation, with centric diatoms (Guinardia, Leptocylindrus, Skeletonema) dominating during the first half of the time-series and a mix of centric and pennate diatoms (Pseudonitzschia, Leptocylindrus, Guinardia) dominating in more recent years.
-In microscopy, smaller sample volumes may lead to underestimations of species richness or diversity (Cermeño et al., 2014).
-Nonetheless, both photopigment and microscopy measurements suggest that phytoplankton communities became more homogeneously distributed over the years, with the majority of the difference due to declining diatom populations that occur during upwelling events.
-
-% REALLY ONLY LOOK AT BOTTOM UP HERE; No Top Down effects quantified
-
-% AMO
-
-% "Shifts in global and basin-specific primary production and phytoplankton abundances are related to large-scale climatic variability in the physical environment.
-Despite the lack of a definitive period associated with the AMO, the relationship between the atmospheric variability associated with the NAO and thermohaline circulation characterized by the AMO suggests a coupling of ocean and atmospheric processes on time scales that will facilitate understanding and prediction of inter-annual weather, precipitation, fluvial output and estuarine and marine productivity.
-Several studies have documented shifts related to the AMO in oceanic (Martinez et al., 2009), coastal and continental shelf (Rodrigues et al., 2009; Schofield et al., 2008) and estuarine (Hubeny et al., 2006) productivity and responses in forage fish populations to this increase in planktonic productivity (Tourre et al., 2007).
-Correlations between the AMO indices and marine productivity have been documented at immediate (Schofield et al., 2008), historical (Martinez et al., 2009; Tourre et al., 2007), and pre-historical (Hubeny et al., 2006; Rodrigues et al., 2009) scales.
-There are several mechanisms through which this might occur.
-Stratification within the water column separates nutrients at lower depths from entering the euphotic zone, thus limiting primary productivity.
-Long-term climate oscillations may augment or suppress seasonal conditions that promote or erode stratification and regulate mixing.
-Schofield et al.
-(2008) have demonstrated links between the AMO and phytoplankton blooms in the Mid-Atlantic Bight, such that increased nutrient availability during the annual fall transition were augmented by increased winter winds associated with a positive phase in the AMO.
-Martinez et al.
-(2009) demonstrated that global phytoplankton productivity is linked to multi-decadal oscillations in both the Atlantic and Pacific, suggesting that links between chlorophyll and sea surface temperature are driven by interactions between the pycnocline and mixed surface layers.
-The influence of large-scale climate fluctuations on surface stratification has particular influence on phytoplankton productivity in the tropics and mid-latitudes, waters which are often nutrient-limited.
-Alternatively, warmer sea surface temperatures themselves may influence productivity.
-Shifts in fluvial inputs associated with altered precipitation patterns will also alter river-induced fertilization of marine productivity." from Nye et al.
-2014
-
-%"The AMO is also associated with changes in wind and current regimes (Delworth et al., 2007; Häkkinen et al., 2011).
-Thus, passive dispersal of marine organisms, many of which are planktonic or have planktonic larvae, will be affected resulting in changes in spatial distribution, survival, and population abundance." from Nye et al 2014
-
-%%% DISCUSSION DUMP:
-
-- What we want to mention, but have not included in our analysis due to lack of data at appropriate coverage or resolution, is the indicators of top-down effects on the Cariaco basin ecosystem.
-Yearly landings of the Spanish sardine (\textit{Sardinella aurita}) fisheries off the NE coast of Venezuela collapsed after 2004 and measurements of zooplankton abundances since 2001 showed a steady increase in numbers up until 2010.
-These can be viewed as consequences of the changes apparent in the phytoplankton community, but can also be important drivers of that change, for example if the collapse in fisheries was mainly driven by overfishing.
-(MENTION THAT ZOO DATA IS TOO FEW DATAPOINTS TO INCLUDE IN GRADIENT FOREST MODEL RUN)
-
-FOR THE CLUSTERING (moved from Results): This confirms previously observed patterns in the Cariaco ecosystem, where a strong shift in the community happened in 2004, but the community returned to the previous regime past 2014 \cite{taylor_ecosystem_2012, muller-karger_scientific_2019}.
+Climate indices are detrended and integrated measurements of long-term changes that are matching in scale to the dynamics of long-term changes within a phytoplankton community.
+However, we do not see such strong predictive power in the ERA5 climate reanalysis-derived mean wind speed, which also aggregates measurements and model output over an entire month.
+Another explanation could be that the climate indices capture a dimension of variability not found in our in-situ measurements, which is also represented in our NMDS clustering (see Figure \ref{fig:clustering} $c$), where both MEI v.2 and AMO scale along the second axis (NMDS 2).
+%TODO test more of my variables along these axes, then add to discussion here and maybe put plot in supplemental material.
+
+Our limitation to bottom-up drivers of community change could also limit the explanatory power of our analysis, as top-down processes such as zooplankton grazing strongly affect the structure of the phytoplankton community \cite{banas_adding_2011, acevedo-trejos_mechanisms_2015}.
+In the Cariaco basin in particular, a trophic cascade from the collapse of sardine fisheries through an increasing abundance of mesozooplaknton has been proposed to have affected the phytoplankton community \cite{muller-karger_scientific_2019}.
+Due to the lack of temporal resolution for higher trophic levels and the lack of zooplankton data for the first 6 years of the time series, we propose that mechanistic modeling studies could be well suited to investigate this aspect of the CARIACO time series further.
+
+In this study, we have found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco basin.
+Despite multiannual trends in warming, reduced productivity and a collapse in fisheries and phytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions returned.
+We see these patterns clearly in the community data retrieved from microscopy, which shows the value of the arduous work of identifying and counting phytoplankton cells.
+The strongest bottom-up variable predicting changes in the phytoplankton community is the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local ecosystems through low-frequency natural cycles.
+We hope that the work hints at the resilient nature of phytoplankton communities under global change, which should be further investigated by studying long-term ocean time series.
 
 \section{Conclusions}
+% Theoretically the final paragraph above represents the conclusion
 

### 2025-04-01 text-backup.tex
Focus Phase Day 16 - Intro Draft done, Discussion 40%
pre-adding new figures

@@
     - Other metrics, evenness, and dominance
+
+    ------
+
+- What could have driven these shifts?
+% talk about link to bottom up drivers, but also mention top down aspects...
+then go deep on Gradient Forest analysis, what that could mean.
+Relate this to other papers about community changes, climate effects and gradient forest applications.
+
+3.
+Clustering shows the return of the community to the previous cluster, confirmed by monthly NMDS, but EnvFit is interesting, as no variable clearly follows the gradient separating the two clusters (NMDS2).
+
+- Climate indices: AMO stark difference between clusters, cluster 1 is negative anomaly, cluster 2 is positive anomaly, corresponding to effects of AMO -> ITCZ etc.
+- MEIv2 less clear, but shows a preponderance of positive anomalies in cluster 1 ... why???
+% MEI index relevant (from 2019 REVIEW)
+The Caribbean experiences the most marked of the El Niño–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Niño–Southern Oscillation \cite{enfield_tropical_1997}.
+
+4. Gradient Forest clearly identifies climate indices as best predictors of community change, then nutrients etc.
+    This is really interesting, as it would be expected, that in situ-measurements are best predictor of community.
+BUT either the community + top down processes are capturing more of that variance, or we simply are missing a bottom-up variable that can explain it.
+    but that most likely requires mechanistic modeling to investigate further...
+
+    %- Discuss effect of AMO
+    % Which studies have already looked at effects of Climate on Ecosystem?
+    - AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
+The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
+    - Although regional effecs may vary, a positive phase in the AMO generally linked to periods of warmer temperatures and a shift of the ITCZ from north to south.
+
+    - Taylor 2012 looked at NAO and MEI, and mostly looked linear trends in variables (they missed the return of the community and regime).
+NAO is the north-south diplole in sea level pressure, that is primarily governed by atmospheric dynamics \cite{hurrell_north_2009}.
+NAO and AMO are l
+    - AMO is mentioned in 2019 review
+    "The observed large-scale multidecadal fluctuations in the Atlantic sea surface temperature (SST), has been referred to as the Atlantic Multidecadal Oscillation (AMO) \cite{kerr2000north} to emphasize the “multidecadal” character of this oceanic phenomenon and to distinguish it from the inter-annual variability associated with the atmospheric North Atlantic Oscillation (NAO) \cite{enfield2001atlantic}." from AMO website
+
+    %- Discuss effect of MEI
+    Since El Nino has strong teleconnection to Caribbean (see ref below from Review) we included this as variable to explain changes in community.
+From Results it is less pronounced than AMO, but still explains some variability (also ref to NMDS).
+
+    Interestingly, in Gradient Forest Model, Climate variables are the strongest predictors of changes in community (i.e. changes in genus abundances).
+    - Discuss what climate variables actually capture, and why they might be better predictors of community changes than local (directly measured) variables.
+
+    - AMO (ref to 2019 Review)
+    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al.
+2015).
+These climatic changes had a profound impact throughout the Atlantic (Alexander et al.
+2014, Nye et al.
+2014) and in the Cariaco basin \cite{taylor_ecosystem_2012, pinckney_phytoplankton_2015}.
+    - ENSO teleconnection (ref to 2019 Review)
+    The Caribbean experiences the most marked of the El Ni  ̃no–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Ni  ̃no–Southern Oscillation (Enfield & Mayer 1997).
+
+--------------------------------------------------------------------------
+----------
+----------
+----------
+----------
+----------
+
+- We are not able to fully resolve....
+% "Phytoplankton communities exhibit a great deal of variation due to a combination of regulation of growth by bottom-up factors such as resource availability (Grover 1997), variation in mortality due to grazing, viral and parasitoid attack (Alpine & Cloern 1992; Mu ̈ hling et al.
+2005), and demographic drift among species (neutral variation).
+Bottom-up factors can be complex, depending on nutrient supply rates and ratios (Tilman 1982), light availability and photoinhibition (Six et al.
+2007; Alderkamp et al.
+2010), species-specific competitive ability for resources (Litchman 2007) and other factors influencing physiological responses such as temperature (Eppley 1972).
+Variation in mortality is much less frequently measured than the factors promoting growth due to the technical challenges in making measurements.
+None of our environmental variables are good proxies of grazing rates or whether the abundance of a particular population is determined by grazer control or resource supply at a particular time.
+We have ident" from Mutshinda et al.
+2013
+
+%"Taylor et al.
+2012 observed that the decline in cell abundances was disproportionate compared with the smaller decrease in total chlorophyll a, which implies that phytoplankton of smaller size were not detected in the surveys..
+This was supported by high-performance liquid chromatography analyses, which indicated a reduction in diatom concentrations and an increase in phytoplankton diversity." from 2019 Review
+
+%"A return to a phytoplankton community structure similar to that observed in the late 1990s was observed in 2014 and 2015 (Figure 5a), as upwelling intensity increased to levels similar to those seen before 2003." from 2019 Review Legacy of CARIACO
+
+- We are analyzing long term time series of data, where probably different analysts have counted the phytoplankton samples...
+%It is well known that the personal component has an influence on how well species are identified.
+Even the same person counting the samples can evolve over time, gaining more experience, which can also influence the results (Löder et al.
+2012; Nohe et al.
+2018).
+A recommendation for further improvement of the assessment is a case wise cross-validation of some samples by the monitoring agencies involved.
+Another important issue to be considered are the differences in sampling frequency between countries and also between stations (see ESM Fig. S1).
+In highly dynamic and turbid waters such as the Wadden Sea, low frequency and infrequent sampling is not sufficient to properly capture temporal signals (Fettweis et al.
+2023).
+% from Di Cavalho et al.
+2023
+- We group species and genus counts, to genus level, and use genus counts as variables the model predicts.
+This assumes that there are not very specific niches etc.
+within genus, but that they are homogenous --> caveat!
+- Discuss limitation of taxonomic diversity, i.e. not direct translation to functional diversity, but here no Data foundation to look at functional diversity.. (inspired by Borics et al. 2021)
+
+These changes are well reflected in the yearly mean anomalies shown in figure \ref{fig:zscore}.
+After 2004 we see higher SST, lower wind speed, reduced fluorometric chlorophyll \textit{a}, a collapse in functional group cell counts, all matching a shift in the AMO toward positive mean anomalies until 2014.
+However, the shift in yearly mean genus richness does not follow the abundances in the cell counts.
+Instead, we see a marked reduction following 1998 up until 2010, where richness recovers four years before the community seems to return to previous levels in 2014.
+The same trend is apparent, although to a less pronounced degree, in the other diversity metrics.
+Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance.
+%(Here check if this is due to "synechococcus" counts, that were abandoned past 2003). 
+The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy data to deduce changes in the phytoplankton community.
+This limits the size range of the observable phytoplankton and thus the measurable diversity.
+We could deduce that we see a marked reduction in the diversity of microphytoplankton ahead of the collapse in abundance, and similarly an increase in microphytoplankton diversity a few years ahead of the return in abundance to previous levels.
+
+- major thing is that previous paper only looked at microscopy to 55m (Mutshinda to 75m and Pinckney), but now we look at full 100 m!
+
+- Seasonality and Depth Distribution of biomass and diversity.
+The diversity of phytoplankton peaks in the surface ocean, similar to the patterns observed using genetic methods \cite{yeh_contrasting_2022}.
+
+2. Mismatch between Diversity variables, and cell count abundances + community clustering / ecosystem shift
+
+   The time series includes multiple measurements of the phytoplankton community, which are bulk fluorometric chlorophyll, microscopy cell counts, and high-performance liquid chromatography (HPLC) pigment analysis.
+Of these, all cover the entire time series except for the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.
+
+    The most often measured indicator of bulk phytoplankton biomass is fluorometric Chlorophyll \textit{a} concentration, which gives a general measure of biomass standing stock and distribution in the water column.
+A more advanced methodology, high-performance liquid chromatography (HPLC), allows individual pigment types to be quantified.
+These pigment distributions can be used to estimate the groups of phytoplankton that contribute to biomass.
+The most specific methodology to quantify phytoplankton communities is the counting and identification of individual phytoplankton cells, either manually under a microscope or automated in flow-through systems.
+The main drawbacks are the size limit for resolving smaller cells and, if done manually, the complexity of identifying phytoplankton species.
+
+5. Compare to what happened at other long-term time series in the ocean...
+
+% Pinckney et al. 2015:
+%The Cariaco Basin has undergone marked shifts in biogeochemistry and food web structure over the past two decades (Taylor et al., 2012; Montes et al., 2013; Bates et al., 2014; Scranton et al., 2014).
+Since the initiation of the CARIACO Ocean Time-Series, near-surface (0–55 m) chl a concentrations (measured by fluorescence of pigments extracted in methanol) and net primary production rates (NPP, measured using 14C incubations) have declined, particularly between 2004 and 2010 (Taylor et al., 2012; Scranton et al., 2014).
+This is hypothesized to be the result of a measured decrease in seasonal trade wind intensity, which has reduced coastal upwelling and led to shallower mixed layer depths (Taylor et al., 2012).
+Microscopy observations indicate that microphytoplankton assemblages (420 mm) within the euphotic zone have shifted from a community comprised mainly of diatoms, dinoflagellates, and coccolithophorids to one with increasing numbers of smaller taxa including cyanobacteria (Montes et al., 2013).
+During this period, a collapse in the sardine industry also occurred.
+The causes are not clear, but the phenomenon was accompanied by an increase in zooplankton biomass and an increase in particulate organic carbon export below the euphotic zone (Taylor et al., 2012).
+These results suggest that changes in zooplankton abundance and therefore grazing may have contributed to the decline in diatoms and to more efficient carbon export to depth.
+% Consistent with the results of Taylor et al.
+(2012), the microscopy data show large scale reductions in overall microphytoplankton (4 20 mm) cell densities and a large decrease in both diatoms and coccolithophores over the upper 55 m.
+However, the observed photopigment decline in diatom abundance is not nearly to the same extent and coccolithophores had a significant concentration increase of 60.1% over this same depth interval (Table 3).
+We argue that the reduction in cell numbers is likely only true of larger cells.
+Microscopic cell counts are inherently selective and biased toward easily identifiable species and cells that remain intact after preservation (Lund et al., 1958; Rott, 1981; Wilhelm et al., 1991; Schlüter et al., 2000).
+Furthermore, Mousing et al.
+(2014) found that small increases in temperature (as observed in the Cariaco Basin) lead to a decrease in the relative contribution of large cells in the community, regardless of ambient nutrient availability.
+Finally, depth limited microscopy data misses the downward movement of diatoms and coccolithophores deeper in the water column (i.e.
+below 56 m).
+In fact, there was also a dramatic shift in diatom speciation, with centric diatoms (Guinardia, Leptocylindrus, Skeletonema) dominating during the first half of the time-series and a mix of centric and pennate diatoms (Pseudonitzschia, Leptocylindrus, Guinardia) dominating in more recent years.
+In microscopy, smaller sample volumes may lead to underestimations of species richness or diversity (Cermeño et al., 2014).
+Nonetheless, both photopigment and microscopy measurements suggest that phytoplankton communities became more homogeneously distributed over the years, with the majority of the difference due to declining diatom populations that occur during upwelling events.
+
+% REALLY ONLY LOOK AT BOTTOM UP HERE; No Top Down effects quantified
+
+% AMO
+
+% "Shifts in global and basin-specific primary production and phytoplankton abundances are related to large-scale climatic variability in the physical environment.
+Despite the lack of a definitive period associated with the AMO, the relationship between the atmospheric variability associated with the NAO and thermohaline circulation characterized by the AMO suggests a coupling of ocean and atmospheric processes on time scales that will facilitate understanding and prediction of inter-annual weather, precipitation, fluvial output and estuarine and marine productivity.
+Several studies have documented shifts related to the AMO in oceanic (Martinez et al., 2009), coastal and continental shelf (Rodrigues et al., 2009; Schofield et al., 2008) and estuarine (Hubeny et al., 2006) productivity and responses in forage fish populations to this increase in planktonic productivity (Tourre et al., 2007).
+Correlations between the AMO indices and marine productivity have been documented at immediate (Schofield et al., 2008), historical (Martinez et al., 2009; Tourre et al., 2007), and pre-historical (Hubeny et al., 2006; Rodrigues et al., 2009) scales.
+There are several mechanisms through which this might occur.
+Stratification within the water column separates nutrients at lower depths from entering the euphotic zone, thus limiting primary productivity.
+Long-term climate oscillations may augment or suppress seasonal conditions that promote or erode stratification and regulate mixing.
+Schofield et al.
+(2008) have demonstrated links between the AMO and phytoplankton blooms in the Mid-Atlantic Bight, such that increased nutrient availability during the annual fall transition were augmented by increased winter winds associated with a positive phase in the AMO.
+Martinez et al.
+(2009) demonstrated that global phytoplankton productivity is linked to multi-decadal oscillations in both the Atlantic and Pacific, suggesting that links between chlorophyll and sea surface temperature are driven by interactions between the pycnocline and mixed surface layers.
+The influence of large-scale climate fluctuations on surface stratification has particular influence on phytoplankton productivity in the tropics and mid-latitudes, waters which are often nutrient-limited.
+Alternatively, warmer sea surface temperatures themselves may influence productivity.
+Shifts in fluvial inputs associated with altered precipitation patterns will also alter river-induced fertilization of marine productivity." from Nye et al.
+2014
+
+%"The AMO is also associated with changes in wind and current regimes (Delworth et al., 2007; Häkkinen et al., 2011).
+Thus, passive dispersal of marine organisms, many of which are planktonic or have planktonic larvae, will be affected resulting in changes in spatial distribution, survival, and population abundance." from Nye et al 2014
+
+%%% DISCUSSION DUMP:
+
+- What we want to mention, but have not included in our analysis due to lack of data at appropriate coverage or resolution, is the indicators of top-down effects on the Cariaco basin ecosystem.
+Yearly landings of the Spanish sardine (\textit{Sardinella aurita}) fisheries off the NE coast of Venezuela collapsed after 2004 and measurements of zooplankton abundances since 2001 showed a steady increase in numbers up until 2010.
+These can be viewed as consequences of the changes apparent in the phytoplankton community, but can also be important drivers of that change, for example if the collapse in fisheries was mainly driven by overfishing.
+(MENTION THAT ZOO DATA IS TOO FEW DATAPOINTS TO INCLUDE IN GRADIENT FOREST MODEL RUN)
+
+FOR THE CLUSTERING (moved from Results): This confirms previously observed patterns in the Cariaco ecosystem, where a strong shift in the community happened in 2004, but the community returned to the previous regime past 2014 \cite{taylor_ecosystem_2012, muller-karger_scientific_2019}.

### 2025-04-02 agujournaltemplate.tex
Update Significance Levels, add comments

@@
 The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}.
+We included marks to highlight those years with missing monthly measurements, as these make the annual means less reliable, but they generally do not affect the general patterns observed, even if substituted by monthly means across the time series.
+Particularly towards the alter years of the time series, the sampling became sparse.
 The patterns emerging from the -u 10m wind component, the temperature and the directly related \qty{21}{\celsius} isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m.
@@
 % Figure 3
-To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between seasons.
+To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between the rainy and upwelling seasons.
 The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters.
@@
 For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate.
-The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies.
+The AMO index shows a separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies.
 Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences.
@@
 \begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/Artboard 1.pdf}
+\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
 \caption{Map of CARIACO Time Series location in the Cariaco basin in Venezuela in the Caribbean Ocean, located at \ang{10.50}N, \ang{64.67}W}
@@
 \end{center}
-\caption{Time series and seasonality of chlorophyll and diversity indices across four depth intervals.
+\caption{Time series and seasonality of chlorophyll \textit{a} and diversity indices across four depth intervals.
 (a,c,e,g) Annual mean across time series for years with more than 5 monthly sampling.
 Transparent dots show monthly raw values.
-(b,d,f,h) Boxplots of monthly measurements across four depth intervals grouped by season.}
+(b,d,f,h) Boxplots of monthly measurements across four depth intervals grouped by season.
+Upwelling season lasts from December to April, Rainy Season from May to November.}
 \label{fig:divts}
@@
 AMO & 3494.0 & 1.86e-14 & ***\\
-MEI v.2 & 9523.5 & 0.006 & ***\\
+MEI v.2 & 9523.5 & 0.006 & **\\
 u-component 10 m wind speed & 7074.0 & 0.143 & n.s.\\
-Isotherm \qty{21}{\celsius} & 5293.0 & 0.034 & **\\
+Isotherm \qty{21}{\celsius} & 5293.0 & 0.034 & *\\
 \addlinespace
-Temperature & 4763.0 & 0.003 & ***\\
-$NO_3$ & 6378.0 & 0.018 & **\\
-$PO_4$ & 7359.0 & 0.012 & **\\
+Temperature & 4763.0 & 0.003 & **\\
+$NO_3$ & 6378.0 & 0.018 & *\\
+$PO_4$ & 7359.0 & 0.012 & *\\
 $SiO_4$ & 5998.0 & 0.736 & n.s.\\
-Salinity & 7636.0 & 0.002 & ***\\
+Salinity & 7636.0 & 0.002 & **\\
 \addlinespace
@@
 Shannon Index & 5142.0 & 0.108 & n.s.\\
-Pielou Index & 4874.0 & 0.029 & **\\
+Pielou Index & 4874.0 & 0.029 & *\\
 \bottomrule
@@
 %TODO test more of my variables along these axes, then add to discussion here and maybe put plot in supplemental material.
+
+% FROM FRAKER ET AL. 2022
+%"Note that although this analysis identifies predictive relationships between drivers and the biological indicators, these relationships do not necessarily imply mechanistic relationships.
+As such, this analysis represents the first stages of an overall analysis (a narrowing of the potential key relationships and when they matter from a much larger dataset), and additional statistical analyses, modeling, and experimental work may be required to develop a full mechanistic understanding of driver-response relationships and ecosystem change."
 

### 2025-05-09 agujournaltemplate.tex
editing May 9th

@@
 
-\title{Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela}
+\title{Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela}
 
@@
 
-\authors{B. Post\affil{1,2}, E. Acevedo-Trejos\affil{3}, S. Chakraborty\affil{1}, A.D. Barton\affil{4}, A. Merico\affil{1}}
+\authors{Benjamin Post\affil{1,2}, Esteban Acevedo-Trejos\affil{3}, Subhendu Chakraborty\affil{1}, Andrew D. Barton\affil{4}, Agostino Merico\affil{1}}
 
@@
 
+%Comments from Subhi:
+- last 2 paragraphs, repetetive, put previously known parts into intro, just start from "the problem"
+- streamline discussion, much shorter!
+-
+
 %%% Suggested section heads:
@@
     % We need time series in order to understand effects and changes (threats) 
-    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term climate changes, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.
+    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.
     % CARIACO is one such time series
@@
 Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.
+
+    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+    %% ADD Duplicate Points here:
+    During the initial phase of the time series (1996–2004), the Cariaco basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
+In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed.
+During upwelling, the phytoplankton community was dominated by microphytoplankton, consisting primarily of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.
+
+    Following 2004, there was a transition in the physical and biogeochemical regime characterized by a significant reduction in trade winds during the upwelling period, which led to elevated sea surface temperatures and decreased mixing between the surface and deep ocean layers \cite{taylor_ecosystem_2012}.
+
+    Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis \cite{pinckney_phytoplankton_2015}.
+In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}.
+During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, which coincided with an increase in zooplankton biomass measured in the Cariaco basin \cite{taylor_ecosystem_2012}.
+
+    Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".
+    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
     %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
@@
 Data past 2012 until the end of the CARIACO time series in 2017, which was not included in previous studies of the phytoplankton community, indicates a return to stronger upwelling intensity and similar productivity levels to the late 1990s \cite{muller-karger_scientific_2019}.
+
+    %% Gaps that We Adresse:
+    % - More in-depth exploration of microscopy data & diversity time series
+    % - Unidirectional trends driven by anthropogenic climate change or low-freq natural cycles, such as AMO?
+        %Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".
+    % - Effect if ENSO on Phyotplankton community test, since previously not observed!
+        % "\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco basin, despite strong teleconnection effects."
 
@@
 Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016.
-The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$), yet there is a consistent pattern of higher Shannon diversity at lower depths for all seasons except for Fall.
+The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$).
 Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$).
@@
 This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends.
-The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index.
+The first axis separates the two clusters, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index.
 The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed.
@@
 
+% ADD HERE MORE DISCUSSION OF MODEL COMPARISON
+
 \begin{figure}
@@
 (b) Species turnover between years for annual aggregate of identified phytoplankton to a genus level.
+The three genera dominating the cell counts for the separate parts of cluster 1 and cluster 2 are given in order of abundance from top to bottom.
 (c) NMDS plot of monthly community data (cell counts) that is cube-root transformed.
@@
 Plots (d-o) below show density distributions of the environmental variables as well as diversity indices between the clustered years.
-Dashed line shows the median and significance levels of Wilcoxon sum rank tests between clusters are shown above plots.}
+Dashed line shows the median and significance levels of Wilcoxon sum rank tests between clusters are shown above plots  (***\textless0.001, **\textless0.01, *\textless0.05, n.s.=not significant).}
 \label{fig:clustering}
@@
 In the Cariaco basin, decadal-scale climate oscillations have affected the wind-driven upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
-During the initial phase of the time series (1996–2004), the Cariaco basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
-In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed.
-During upwelling, the phytoplankton community was dominated by microphytoplankton, consisting primarily of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.
-
-Following 2004, there was a transition in the physical and biogeochemical regime characterized by a significant reduction in Trade Winds during the upwelling period, which led to elevated sea surface temperatures and decreased mixing between the surface and deep ocean layers \cite{taylor_ecosystem_2012}.
+% - a few sentences on basic seasonal dynamics removed here
 These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}).
 After 2004 we see higher SST, lower wind speed, reduced fluorometric chlorophyll \textit{a}, a collapse in functional group cell counts, all matching a shift in the AMO toward positive mean anomalies until 2014.
-Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis.
-In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}.
-During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, which coincided with an increase in zooplankton biomass measured in the Cariaco basin \cite{taylor_ecosystem_2012}.
-
-Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthrophogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".
-We found that including the time series data up to its end in 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}.
-In 2014, physical conditions in the Cariaco basin returned to a regime of higher wind speeds, increased uwpelling, and higher nutrient concentrations (see Figure \ref{fig:zscore}).
+% - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries
+
+% - "We found that including the time series data until the end of 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}."
+In 2014, physical conditions in the Cariaco basin returned to a regime of higher wind speeds, increased upwelling, and higher nutrient concentrations (see Figure \ref{fig:zscore}).
 At the same time, there was an increase in chlorophyll \textit{a} biomass (see Figure \ref{fig:divts} $a$) and a return to a community more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$).
@@
 Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance.
-%(Here check if this is due to "synechococcus" counts, that were abandoned past 2003). 
+
 The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy counts to deduce changes in the phytoplankton community.
@@
 We do see a significant difference in the Pilou index between the two clusters.
-Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during the low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}.
+Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}.
 This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and the thereby reduced disturbance of the community.
 The resulting reduced compositional turnover is also observed in a decrease in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$).
-As the productivity and biomass of phytoplankton communities are mainly driven by nutrients supplied through disturbance events, this confirms the observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
+As the productivity and biomass of phytoplankton communities are mainly driven by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
 
@@
 
-These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, since the scaling axis that actually separates the two yearly aggregate community clusters is NMDS 1, which covaries most strongly with changes in the AMO, MEI v.2 (see Figure \ref{fig:clustering} $c$).
+These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, since the scaling axis that separates the two yearly aggregate community clusters is NMDS 1, which covaries most strongly with changes in the AMO, MEI v.2 (see Figure \ref{fig:clustering} $c$).
 Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS2.
@@
 
+% Move the general point on ENSO to intro! Only Discussion-relevant part here..
 The El Niño Southern Osciallation (ENSO) has a strong teleconnection to the Caribbean Sea, since during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
@@
 Nitrate is a central nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco basin \cite{muller-karger_scientific_2019}.
-For dinoflagelltes and diatoms we see a strong threshold response around \qty{7}{\micro \mole}, which would indicate that strong upwelling conditions drive blooms of these functional groups.
-The "other" functional group including \textit{Mesodinium} and \textit{Eutreptria} seems to have a threshold response at even higher nitrate concentrations (around \qty{9}{\micro \mole}), indicating an effect on their abundance by very strong upwelling.
-Similarly, for temperature, dinoflagellates and diatoms show a threshold around \qtyrange{21}{23}{\celsius}, which indicates higher abundances during low temperatures in the upper\qty{100}{m} driven by strong upwelling.
+For dinoflagellates and diatoms, we see a strong threshold response around \qty{7}{\micro \mole}, which would indicate that strong upwelling conditions drive blooms of these functional groups.
+Other functional groups including \textit{Mesodinium} and \textit{Eutreptria} also show a threshold response at even higher nitrate concentrations (around \qty{9}{\micro \mole}), indicating an effect on their abundance by very strong upwelling.
+Similarly, for temperature, dinoflagellates and diatoms show a threshold around \qtyrange{21}{23}{\celsius}, which indicates higher abundances during low temperatures in the upper \qty{100}{m} driven by strong upwelling.
 For the other predictors, we do not see strong threshold responses.
@@
 We hope that the work hints at the resilient nature of phytoplankton communities under global change, which should be further investigated by studying long-term ocean time series.
+
+% Highlight that the climate indices explain changes in the community and abundance, but not necessarily changed in diversity (as there is no sig diff beteween clusters).
+Probably need more data, other approaches to untangle effects on diversity.
 

### 2025-05-13 agujournaltemplate.tex
added new ZScore plot to manuscript
editing tables & new figures

@@
 \begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/PLOTZScores_updated.pdf}
+\noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
 \caption{Heatmap of yearly average values, normalized for comparison as z-scores.
@@
 \begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/ClusteringCompPlot_NEW 2.pdf}
+\noindent\includegraphics[width=\textwidth]{fig/Figure4_ClusteringNMDS_v2.pdf}
 \caption{(a) Clustering of microscopy data at the Genus level, using the binary Jaccard distance.
@@
 Color of dots based on which yearly cluster the monthly data is in.
-Vector overlay of ENVFIT shows underlying environmental variables that are co-varying with the community data.
-Plots (d-o) below show density distributions of the environmental variables as well as diversity indices between the clustered years.
+Vector overlay of ENVFIT shows underlying environmental variables that are co-varying with the community data.}
+\label{fig:clustering}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/Figure5_ClustCompPlot_v2.pdf}
+\caption{(Plots (a-o) below show density distributions of the environmental variables as well as diversity indices between the clustered years.
 Dashed line shows the median and significance levels of Wilcoxon sum rank tests between clusters are shown above plots  (***\textless0.001, **\textless0.01, *\textless0.05, n.s.=not significant).}
-\label{fig:clustering}
+\label{fig:clustcomp}
 \end{figure}
@@
         we tested all measurements up to a lag of 3 months. 
-        For climate variables, we tested up to a lag of 6 months. 
+        For climate variables we tested up to a lag of 6 months. 
         Time lag with maximum importance per variable are highlighted in bold and chosen for the final model run.}
-
-\ref{app:table1}
 \centering
-\begin{tabular}[t]{rrrrrrrrrr}
+\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrrrrrrrrrrr}
 \toprule
-lag & NO3 & PO4 & SiO4 & Temperature & Isotherm\_21 & Salinity\_bottles & u10 & AMO & MEIv2\\
+lag & NO_3 & PO_4 & SiO_4 & SST & \qty{21}{\celsius} isotherm & Salinity & Wind speed & AMO & MEI v.2 & Precipitation & Evaporation\\
 \midrule
-0 & \textbf{0.038} & \textbf{0.038} & \textbf{0.034} & \textbf{0.071} & \textbf{0.028} & 0.007 & 0.004 & 0.016 & 0.015\\
-1 & 0.017 & 0.011 & 0.015 & 0.002 & 0.002 & 0.004 & \textbf{0.010} & 0.014 & 0.018\\
-2 & 0.022 & 0.008 & 0.013 & 0.006 & 0.008 & \textbf{0.011} & 0.009 & \textbf{0.030} & 0.019\\
-3 & 0.016 & 0.007 & 0.004 & 0.005 & 0.007 & 0.010 & 0.007 & 0.018 & 0.018\\
-4 & - & - & - & - & - & - & 0.008 & 0.019 & \textbf{0.029}\\
-5 & - & - & - & - & - & - & 0.008 & 0.018 & 0.022\\
-6 & - & - & - & - & - & - & 0.010 & 0.018 & 0.013\\
+0 & \textbf{0.025} & \textbf{0.028} & \textbf{0.029} & \textbf{0.084} & \textbf{0.024} & 0.006 & 0.003 & 0.015 & 0.015 & 0.004 & 0.010\\
+1 & 0.013 & 0.011 & 0.014 & 0.010 & 0.003 & 0.007 & \textbf{0.007} & 0.012 & 0.015 & \textbf{0.013} & 0.011\\
+2 & 0.016 & 0.007 & 0.013 & 0.011 & 0.006 & 0.011 & 0.007 & \textbf{0.026} & 0.018 & 0.005 & 0.009\\
+3 & 0.014 & 0.007 & 0.005 & 0.007 & 0.007 & \textbf{0.016} & 0.004 & 0.017 & 0.017 & 0.008 & \textbf{0.013}\\
+4 & NA & NA & NA & NA & NA & NA & 0.006 & 0.018 & \textbf{0.025} & 0.008 & 0.005\\
+\addlinespace
+5 & NA & NA & NA & NA & NA & NA & 0.007 & 0.016 & 0.020 & 0.008 & 0.032\\
+6 & NA & NA & NA & NA & NA & NA & 0.006 & 0.017 & 0.011 & 0.006 & 0.003\\
+12 & NA & NA & NA & NA & NA & NA & NA & 0.011 & 0.012 & NA & NA\\
 \bottomrule
-\end{tabular}
+\end{tabular}}
 \end{table}
 
+\begin{table}
+
+\caption{Individual Gradient Forest model runs for each variable and the corresponding time lags. 
+        For in-situ variables, for which no coverage extends over the time series, 
+        we tested all measurements up to a lag of 3 months. 
+        For climate variables we tested up to a lag of 6 months. 
+        Time lag with maximum importance per variable are highlighted in bold and chosen for the final model run.}
+\centering
+\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrrrrrrrrrrr}
+\toprule
+lag & NO3 & PO4 & SiO4 & Temperature & Isotherm\_21 & Salinity\_bottles & u10 & AMO & MEIv2 & Precipitation & Evaporation\\
+\midrule
+0 & 0.062 & 0.11 & 0.068 & 0.092 & 0.067 & 0.175 & 0.048 & 0.013 & 0.082 & 0.048 & 0.019\\
+1 & 0.148 & NA & 0.017 & 0.019 & 0.044 & NA & 0.158 & 0.033 & 0.057 & 0.094 & 0.346\\
+2 & 0.177 & NA & NA & NA & 0.013 & NA & NA & 0.057 & 0.140 & 0.254 & 0.116\\
+3 & 0.341 & NA & NA & NA & NA & NA & NA & 0.023 & 0.143 & NA & 0.016\\
+4 & NA & NA & NA & NA & NA & NA & NA & 0.081 & 0.048 & NA & 0.209\\
+\addlinespace
+5 & NA & NA & NA & NA & NA & NA & NA & 0.260 & 0.095 & NA & NA\\
+6 & NA & NA & NA & NA & NA & NA & NA & NA & 0.046 & NA & NA\\
+12 & NA & NA & NA & NA & NA & NA & NA & NA & 0.037 & NA & NA\\
+\bottomrule
+\end{tabular}}
+\end{table}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2025-05-13 agujournaltemplate.tex

@@
 \noindent\includegraphics[width=\textwidth]{fig/Figure4_ClusteringNMDS_v2.pdf}
-\caption{(a) Clustering of microscopy data at the Genus level, using the binary Jaccard distance.
+\caption{(a) Hierarchichal lustering by complete linkage method of microscopy cell counts at the Genus level, using the binary Jaccard distance.
 The community for each year is binned together, to smooth out seasonal aspects.
@@
 
+\section{Correlation of data}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/CorrClustSupplementalPlot_v1.pdf}
+\caption{Correlation matrix with clustering of all datasets presented.
+Correlation was calcualted using a Spearman's rank correlation coefficient, the magnitude of the coefficient is indicated by the color of the circle.
+The size of the circle and significance level marking (***=p<0.001, **=p<0.01, *=p<0.05)
+for hierarchical clustering }
+\label{fig:map}
+\end{figure}
+
 \section{Gradient Forest Analysis supplement}

### 2025-05-13 agujournaltemplate.tex
stats table fixes

@@
 \caption{(Plots (a-o) below show density distributions of the environmental variables as well as diversity indices between the clustered years.
-Dashed line shows the median and significance levels of Wilcoxon sum rank tests between clusters are shown above plots  (***\textless0.001, **\textless0.01, *\textless0.05, n.s.=not significant).}
+Dashed line shows the median and significance levels of Wilcoxon sum rank tests between clusters are shown above plots  (p-value: ***\textless0.001, **\textless0.01, *\textless0.05, n.s.=not significant).}
 \label{fig:clustcomp}
@@
 \noindent\includegraphics[width=\textwidth]{fig/CorrClustSupplementalPlot_v1.pdf}
-\caption{Correlation matrix with clustering of all datasets presented.
+\caption{Correlation matrix with clustering of environmental variables, diversity metrics and cell counts.
 Correlation was calcualted using a Spearman's rank correlation coefficient, the magnitude of the coefficient is indicated by the color of the circle.
-The size of the circle and significance level marking (***=p<0.001, **=p<0.01, *=p<0.05)
-for hierarchical clustering }
+The size of the circle and significance level marking represents the calculated significance of the correlation deviating from 0 (p-value: ***\textless0.001, **\textless0.01, *\textless0.05).
+The time series were aditionally ordered by hierarchical clustering via the complete linkage method into 5 groups indicated by the black square border.}
 \label{fig:map}
@@
 
+% HERE ADD FIGURE OF GF RUN WITHOUT TIME LAGS (ALLVARS)
+
 \begin{table}
 
-\caption{Individual Gradient Forest model runs for each variable and the corresponding time lags. 
-        For in-situ variables, for which no coverage extends over the time series, 
-        we tested all measurements up to a lag of 3 months. 
-        For climate variables we tested up to a lag of 6 months. 
+\caption{Resulting $R^2$ weighted importance for running predictor variables together with time-lags to check if they provide a better fit to the data.
+Each column represents a single run, with resulting importance values for time series and monthly lags.
+The results were stable, so that repeated runs confirmed the chosen time lag.
+For in-situ variables,  we tested all measurements up to a lag of 3 months.
+For climate variables we tested up to a lag of 6 months and added a lag of 12 months for the climate indices.
 Time lag with maximum importance per variable are highlighted in bold and chosen for the final model run.}
@@
 \toprule
-lag & NO_3 & PO_4 & SiO_4 & SST & \qty{21}{\celsius} isotherm & Salinity & Wind speed & AMO & MEI v.2 & Precipitation & Evaporation\\
+lag & NO3 & PO4 & SiO4 & SST & \qty{21}{\celsius} Isotherm & Salinity & AMO & MEI v.2 & Wind speed & Precipitation & Evaporation\\
 \midrule
-0 & \textbf{0.025} & \textbf{0.028} & \textbf{0.029} & \textbf{0.084} & \textbf{0.024} & 0.006 & 0.003 & 0.015 & 0.015 & 0.004 & 0.010\\
-1 & 0.013 & 0.011 & 0.014 & 0.010 & 0.003 & 0.007 & \textbf{0.007} & 0.012 & 0.015 & \textbf{0.013} & 0.011\\
-2 & 0.016 & 0.007 & 0.013 & 0.011 & 0.006 & 0.011 & 0.007 & \textbf{0.026} & 0.018 & 0.005 & 0.009\\
-3 & 0.014 & 0.007 & 0.005 & 0.007 & 0.007 & \textbf{0.016} & 0.004 & 0.017 & 0.017 & 0.008 & \textbf{0.013}\\
-4 & NA & NA & NA & NA & NA & NA & 0.006 & 0.018 & \textbf{0.025} & 0.008 & 0.005\\
+0 & \textbf{0.027} & \textbf{0.029} & \textbf{0.026} & \textbf{0.086} & \textbf{0.024} & 0.006 & 0.015 & 0.015 & 0.003 & 0.005 & 0.011\\
+1 & 0.015 & 0.011 & 0.014 & 0.010 & 0.003 & 0.004 & 0.012 & 0.015 & 0.006 & \textbf{0.014} & 0.009\\
+2 & 0.018 & 0.007 & 0.011 & 0.013 & 0.005 & 0.010 & \textbf{0.026} & 0.018 & 0.007 & 0.005 & 0.013\\
+3 & 0.013 & 0.007 & 0.003 & 0.007 & 0.007 & \textbf{0.013} & 0.016 & 0.017 & 0.005 & 0.009 & 0.011\\
+4 & NA & NA & NA & NA & NA & NA & 0.017 & \textbf{0.026} & 0.005 & 0.008 & 0.005\\
+5 & NA & NA & NA & NA & NA & NA & 0.016 & 0.020 & \textbf{0.008} & 0.008 & \textbf{0.040}\\
+6 & NA & NA & NA & NA & NA & NA & 0.017 & 0.012 & 0.005 & 0.006 & 0.004\\
 \addlinespace
-5 & NA & NA & NA & NA & NA & NA & 0.007 & 0.016 & 0.020 & 0.008 & 0.032\\
-6 & NA & NA & NA & NA & NA & NA & 0.006 & 0.017 & 0.011 & 0.006 & 0.003\\
-12 & NA & NA & NA & NA & NA & NA & NA & 0.011 & 0.012 & NA & NA\\
+12 & NA & NA & NA & NA & NA & NA & 0.010 & 0.013 & NA & NA & NA\\
 \bottomrule
@@
 
-\begin{table}
-
-\caption{Individual Gradient Forest model runs for each variable and the corresponding time lags. 
-        For in-situ variables, for which no coverage extends over the time series, 
-        we tested all measurements up to a lag of 3 months. 
-        For climate variables we tested up to a lag of 6 months. 
-        Time lag with maximum importance per variable are highlighted in bold and chosen for the final model run.}
-\centering
-\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrrrrrrrrrrr}
-\toprule
-lag & NO3 & PO4 & SiO4 & Temperature & Isotherm\_21 & Salinity\_bottles & u10 & AMO & MEIv2 & Precipitation & Evaporation\\
-\midrule
-0 & 0.062 & 0.11 & 0.068 & 0.092 & 0.067 & 0.175 & 0.048 & 0.013 & 0.082 & 0.048 & 0.019\\
-1 & 0.148 & NA & 0.017 & 0.019 & 0.044 & NA & 0.158 & 0.033 & 0.057 & 0.094 & 0.346\\
-2 & 0.177 & NA & NA & NA & 0.013 & NA & NA & 0.057 & 0.140 & 0.254 & 0.116\\
-3 & 0.341 & NA & NA & NA & NA & NA & NA & 0.023 & 0.143 & NA & 0.016\\
-4 & NA & NA & NA & NA & NA & NA & NA & 0.081 & 0.048 & NA & 0.209\\
-\addlinespace
-5 & NA & NA & NA & NA & NA & NA & NA & 0.260 & 0.095 & NA & NA\\
-6 & NA & NA & NA & NA & NA & NA & NA & NA & 0.046 & NA & NA\\
-12 & NA & NA & NA & NA & NA & NA & NA & NA & 0.037 & NA & NA\\
-\bottomrule
-\end{tabular}}
-\end{table}
+% HERE ADD FIGURE OF GF RUN VARS FROM FINAL RUN NOT SHOWN IN FIGURE 6
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2025-05-15 agujournaltemplate.tex
Update Supplement
Start rewrite for Ago/Andrew

@@
 
-\journalname{Enter journal name here}
+\journalname{JGR: Biogeosciences}
 
@@
 Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the anomalies in the AMO index.
-% \index Previous studies proposed a shift in the community, but with the full dataset we see a shift and a return.
-The community shows a departure form cluster one around 2004, but a return to that community in 2014.
-\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, although community evenness was higher and community turnover was lower in the second cluster (low-upwelling conditions).
+\index Instead of a unidirectional trend, we see a shift around 2004 to low-upwelling conditions, and a return to the previous community and physical environment in 2014. 
+\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, but community evenness was higher and community turnover was lower in the second cluster (low-upwelling conditions).
 \item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem.
-The next strongest predictor is the MEI v2 index (lag of 4 months), with in-situ temperature and nitrate concentration following behind.
+The next strongest predictor is sea surface temperature (SST), with the MEI v2 index (lag of 4 months) and nitrate concentration following behind.
 \end{keypoints}
@@
 To consider the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all climate variables, as well as 3 months lags for in-situ measurements in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run (see Table \ref{app:table1} in the Appendix).
+
+    \begin{table}
+
+\caption{All data sources with corresponding units used as predictor variables for the gradient forest analysis.}
+\centering
+\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrrrrrrrrrrr}
+\toprule
+Variable Name & Units & Data source \\
+\midrule
+MEI v.2 & unitless & Climate Indices provided by NOAA \\
+AMO & unitless & Climate Indices provided by NSF NCAR \\
+Wind speed & m/s & ERA5 model reanalysis: 10 m u-component of wind speed \\
+Precipitation & 0.013 & ERA5 model reanalysis: total precipitation\\
+Evaporation & NA & ERA5 model reanalysis: evaporation \\
+SST & °C & CARIACO time series: CTD, Average temperature per month measured in top 10 m\\
+21° C Isotherm & m & CARIACO time series: CTD, depth at which temperature reaches 21°C\\
+Salinity & PSU & Niskin, top 100m \\
+NO3 & µM & Niskin, top 100m \\
+PO4 & µM & Niskin, top 100m \\
+SiO4 & µM & Niskin, top 100m \\
+\end{tabular}}
+\label{tab:DataSources}
+\end{table}
 
@@
 \begin{table}
+
 \caption{Wilcoxon sum rank test with continuity correction for variables between the two clusters.
 Values for in-situ data are interpolated across the top 100 meters for each individual monthly cruise sampling.}
-\label{table:clustering}
 \centering
@@
 MEI v.2 & 9523.5 & 0.006 & **\\
-u-component 10 m wind speed & 7074.0 & 0.143 & n.s.\\
-Isotherm \qty{21}{\celsius} & 5293.0 & 0.034 & *\\
+Wind speed & 7074.0 & 0.143 & n.s.\\
+Precipitation & 6251.0 & 0.004 & **\\
+Evaporation & 9979.0 & 3.68e-5 & ***\\
 \addlinespace
-Temperature & 4763.0 & 0.003 & **\\
+21 °C Isotherm & 5293.0 & 0.034 & *\\
+SST & 4712.0 & 0.001 & ***\\
+Salinity & 7636.0 & 0.002 & **\\
 $NO_3$ & 6378.0 & 0.018 & *\\
@@
 $SiO_4$ & 5998.0 & 0.736 & n.s.\\
-Salinity & 7636.0 & 0.002 & **\\
 \addlinespace
+Chlorophyll $a$ & 8229.0 & 9.63e-6 & ***\\
 Genus Richness & 5913.0 & 0.954 & n.s.\\
@@
 \begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/GF_output_plot3_new.pdf}
+\noindent\includegraphics[width=\textwidth]{fig/Figure6_GF_output_FINAL_v2.pdf}
 \caption{Gradient Forest analysis of monthly community data at a genus level.
@@
 The time series were aditionally ordered by hierarchical clustering via the complete linkage method into 5 groups indicated by the black square border.}
-\label{fig:map}
+\label{fig:sup:correlation}
 \end{figure}
@@
 
-% HERE ADD FIGURE OF GF RUN WITHOUT TIME LAGS (ALLVARS)
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/FigureA2_GF_output_Supplement_v1.pdf}
+\caption{Cumulative importance curves for all predictor variables not shown in the main figure.
+Dashed black line represents the cumulative importance across the entire community, colored lines represent functional groups.
+Gradient forest runs where done on Genus-level cell counts, integrated over the top 100 meters.}
+\label{fig:sup:GFoutput_lags_extra}
+\end{figure}
 
@@
 \end{tabular}}
+\label{tab:sup:GFlagtests}
 \end{table}
 
-% HERE ADD FIGURE OF GF RUN VARS FROM FINAL RUN NOT SHOWN IN FIGURE 6
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/FigureA3_GF_output_NOLAG_v1.pdf}
+\caption{Gradient Forest analysis of monthly community data at a genus level.
+(a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances.
+(b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance.
+(c) Cumulative importance curves for the predictor variables.
+Black line shows cumulative importance across predictor range for the entire model and all genera.
+Dashed colored lines represent the responses of individual functional groups.}
+\label{fig:sup:GFoutput_nolags}
+\end{figure}
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/FigureA4_GF_output_Supplement_NOLAG_v1.pdf}
+\caption{Cumulative importance curves for all variables not shown in the previous figure.
+Dashed black line represents the cumulative importance across the entire community, colored lines represent functional groups.
+Gradient forest runs where done on Genus-level cell counts, integrated over the top 100 meters.}
+\label{fig:sup:GFoutput_nolags_extra}
+\end{figure}
 

### 2025-05-16 agujournaltemplate.tex
Start writing abstract

@@
 \begin{keypoints}
-\item The CARIACO time series showed a highly variable phytoplankton community over the course of sampling between 1995 and 2017.
-Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the anomalies in the AMO index.
-\index Instead of a unidirectional trend, we see a shift around 2004 to low-upwelling conditions, and a return to the previous community and physical environment in 2014. 
-\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, but community evenness was higher and community turnover was lower in the second cluster (low-upwelling conditions).
-\item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem.
-The next strongest predictor is sea surface temperature (SST), with the MEI v2 index (lag of 4 months) and nitrate concentration following behind.
+\item The CARIACO time series phytoplankton community data reveals a shift and return between two clusters, defined by high- and low-upwelling conditions that correspond to anomalies in the AMO index.
+\index Between the clusters, we observe no significant changes in diversity metrics, except for lower turnover and higher community evenness under low-upwelling conditions.
+\item Gradient forest analysis identified the AMO climate index as the strongest predictor of community changes, highlighting the influence of large-scale climatic oscillations on the tropical coastal ecosystem.
 \end{keypoints}
@@
 \begin{abstract}
-The structure of the phytoplankton community was identified using microscopy in the Cariaco basin, in Venezuela, between 1995 and 2017.
-Microscopy data are limited to larger cell sizes (> 20 µm), where the community showed a large change in the abundances of functional groups and shifts in diversity patterns.
+Phytoplankton play a crucial role in Earth's biogeochemical cycles.
+Climate drives changes in the physical environment of phytoplankton communities, which can, in turn, affect the productivity and diversity of the entire ecosystem.
+This study aims to investigate the drivers and effects of changes observed in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017.
+Monthly cruises sampled microscopy cell counts, as well as biogeochemical properties, which we extend with data from AMO and MEI v.2 climate indices, as well as the ERA5 climate model reanalysis.
+
+- Focus on microscopy data
+- Look at general trends and time series of diversity
+- Find structure in community data, relate that to shifts in bottom-up drivers
+Reveals two clusters, characterized by low and high upwelling. No significant change in diversity, but changes in turnover and evenness.
+- Rank predictors of community change using gradient forest analysis
+Shows strong impact of upwelling system (i.e. temperature & nitrate), but also climate indices AMO and MEIv.2 are ranked among best predictors of changes in community.
+
+Our results highlight the effect of global climate on a tropcial coastal ecosystem, through decadal scale oscillations.
+
+The structure of the phytoplankton community was identified using microscopy . 
+Microscopy data are limited to larger cell sizes (> 20 µm), where the community exhibited significant changes in the the abundances of functional groups and shifts in diversity patterns.
+
 \end{abstract}
@@
 %1
-%\section*{Plain Language Summary}
+\section*{Plain Language Summary}
 %Enter your Plain Language Summary here or delete this section.
@@
 %https://www.agu.org/Share-and-Advocate/Share/Community/Plain-language-summary
+
+Microscopic algae drive
 
@@
     % We need time series in order to understand effects and changes (threats) 
-    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.
+    Time series of biological observations offer valuable insights into how marine ecosystems react to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.
     % CARIACO is one such time series

### 2025-05-16 text-backup.tex
Start writing abstract

@@
 FOR THE CLUSTERING (moved from Results): This confirms previously observed patterns in the Cariaco ecosystem, where a strong shift in the community happened in 2004, but the community returned to the previous regime past 2014 \cite{taylor_ecosystem_2012, muller-karger_scientific_2019}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% KEY POINTS
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%  List up to three key points (at least one is required)
+%  Key Points summarize the main points and conclusions of the article
+%  Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences
+
+% Example:
+% \begin{keypoints}
+% \item	List up to three key points (at least one is required)
+% \item	Key Points summarize the main points and conclusions of the article
+% \item	Each must be 140 characters or fewer with no special characters or punctuation and must be complete sentences
+% \end{keypoints}
+
+\begin{keypoints}
+\item The CARIACO time series showed a highly variable phytoplankton community over the course of sampling between 1995 and 2017.
+Clustering of community data from microscopy cell counts reveals two separate clusters, defined by high- and low-upwelling conditions that match the anomalies in the AMO index.
+\index Instead of a unidirectional trend, we see a shift around 2004 to low-upwelling conditions, and a return to the previous community and physical environment in 2014. 
+\item Despite the differing physical regimes and the large changes between the functional groups, there are no significant differences in genus richness and Shannon diversity between the two clusters, but community evenness was higher and community turnover was lower in the second cluster (low-upwelling conditions).
+\item A gradient forest analysis of community data reveals that the strongest predictor of shifts within the community is the AMO index with a time lag of 2 months, pointing to a strong impact of large-scale climatic oscillations in the tropical coastal ecosystem.
+The next strongest predictor is sea surface temperature (SST), with the MEI v2 index (lag of 4 months) and nitrate concentration following behind.
+\end{keypoints}

### 2025-05-19 agujournaltemplate.tex
reworking Introduction v2

@@
 Climate drives changes in the physical environment of phytoplankton communities, which can, in turn, affect the productivity and diversity of the entire ecosystem.
-This study aims to investigate the drivers and effects of changes observed in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017.
-Monthly cruises sampled microscopy cell counts, as well as biogeochemical properties, which we extend with data from AMO and MEI v.2 climate indices, as well as the ERA5 climate model reanalysis.
-
-- Focus on microscopy data
-- Look at general trends and time series of diversity
-- Find structure in community data, relate that to shifts in bottom-up drivers
-Reveals two clusters, characterized by low and high upwelling. No significant change in diversity, but changes in turnover and evenness.
-- Rank predictors of community change using gradient forest analysis
-Shows strong impact of upwelling system (i.e. temperature & nitrate), but also climate indices AMO and MEIv.2 are ranked among best predictors of changes in community.
-
-Our results highlight the effect of global climate on a tropcial coastal ecosystem, through decadal scale oscillations.
-
-The structure of the phytoplankton community was identified using microscopy . 
-Microscopy data are limited to larger cell sizes (> 20 µm), where the community exhibited significant changes in the the abundances of functional groups and shifts in diversity patterns.
+Long-term time series can capture changes in climate and the ecosystem, allowing for the untangling of effects acting on a local population.
+This study aims to describe and investigate the drivers of changes observed in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017.
+Monthly cruises sampled microscopy cell counts and biogeochemical properties, which we extend with data from AMO and MEI v.2 climate indices, as well as the ERA5 climate model reanalysis.
+We describe and relate changes in bottom-up drivers of the upwelling system with the time series of phytoplankton biomass and diversity.
+Using clustering algorithms, we investigate the structure in the community time series data, revealing two clusters that describe a shift from high to low upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014.
+Our results suggest that community diversity did not follow the same pattern but instead was diminished between 1998 and 2009.
+To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related metrics (SST and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors.
+Our study emphasizes the impact of decadal-scale climate oscillations, such as the AMO, on the phytoplankton community in a tropical coastal ecosystem through modulation of the upwelling system and highlights the resilience of ecosystems to recover from climate changes.
 
@@
 
-Microscopic algae drive
+Microscopic algae drive... (Will write this Summary last, after draft is finalized)
 
@@
 %Comments from Subhi:
-- last 2 paragraphs, repetetive, put previously known parts into intro, just start from "the problem"
-- streamline discussion, much shorter!
--
+    % - first 2 paragraphs, repetetive, put previously known parts into intro, just start from "the problem"
+    % - streamline discussion, much shorter!
+    % -
 
@@
 The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them highly suitable indicators of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
-    The structure of the community can be considered an integrator of environmental factors, as it is strongly determined by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}.
+    The structure of the community can be considered an integrator of environmental factors, as it is strongly influenced by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}.
 In addition to such bottom-up effects, there are also forces of higher trophic levels, such as zooplankton, that structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}.
@@
     % We need time series in order to understand effects and changes (threats) 
-    Time series of biological observations offer valuable insights into how marine ecosystems react to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.
+    Time series of biological observations offer valuable insights into how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.
     % CARIACO is one such time series
-    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin between 1995 and 2017 \cite{muller-karger_scientific_2019}.
+    One such project is the CARIACO time series, which has involved consistent sampling of physical and biogeochemical parameters in the Cariaco Basin between 1995 and 2017 \cite{muller-karger_scientific_2019}.
     % Describe CARIACO location
@@
 The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}.
-Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
+Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and a testbed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
     %%% mention climate impacts (AMO + ENSO) here!
@@
 A smaller wind-driven secondary upwelling often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
-During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids and nanoplankton also contributing.
-In the low-wind rainy season, the Cariaco basin exhibits oligotrophic conditions, where the portion of large phytoplankton cells is diminished, but nano- and picophytoplankton are present in considerable numbers throughout the year \cite{lorenzoni_characterization_2015}.
+During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing.
+In the low-wind rainy season, the Cariaco Basin exhibits oligotrophic conditions, where the fraction of large phytoplankton cells is diminished, but nano- and picophytoplankton are present in considerable numbers throughout the year \cite{lorenzoni_characterization_2015}.
 
@@
     %% ADD Duplicate Points here:
-    During the initial phase of the time series (1996–2004), the Cariaco basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
+    During the initial phase of the time series (1996–2004), the Cariaco Basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
 In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed.
-During upwelling, the phytoplankton community was dominated by microphytoplankton, consisting primarily of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.
+During upwelling, the phytoplankton community was dominated by microphytoplankton, primarily consisting of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.
 
@@
 In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}.
-During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, which coincided with an increase in zooplankton biomass measured in the Cariaco basin \cite{taylor_ecosystem_2012}.
-
-    Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".
+During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass measured in the Cariaco basin \cite{taylor_ecosystem_2012}.
+
+    Previous analyses of phytoplankton community data did not include time series data beyond 2011, and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO."
     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
     The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift.
-Data past 2012 until the end of the CARIACO time series in 2017, which was not included in previous studies of the phytoplankton community, indicates a return to stronger upwelling intensity and similar productivity levels to the late 1990s \cite{muller-karger_scientific_2019}.
+Data from 2012 until the end of the CARIACO time series in 2017, which were not included in previous studies of the phytoplankton community, indicate a return to stronger upwelling intensities and productivity levels similar to those of the late 1990s \cite{muller-karger_scientific_2019}.
 
@@
     % - Effect if ENSO on Phyotplankton community test, since previously not observed!
+             %This could also be part of the rational of the study. Formulated as question, e.g. "it is still unclear the effects of regional climate phenomena such as ENSO on this shift" - Esteban
         % "\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco basin, despite strong teleconnection effects."
 
+    The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
+\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance with a time lag of 12 months using linear regression.
+
     % NOw we come in!
-    In this study, we document changes in biodiversity and community structure in the Cariaco basin between 1995 and 2017 and quantify the effect of bottom-up drivers and climate indices on changes in the phytoplankton community.
-We focus on the phytoplankton cell count data by microscopy, as these data have not been fully analyzed in previous studies, and we argue that despite the shortcomings of microscopy methods, they provide a consistent measure of changes in large phytoplankton.
-Our analysis focuses on bottom-up drivers of community changes, as these data are available throughout the time series and at monthly resolution.
+    In this study, we document changes in biodiversity and community structure in the Cariaco Basin between 1995 and 2017 and quantify the effects of bottom-up drivers and climate indices on changes in the phytoplankton community.
+We focus on phytoplankton cell count data obtained through microscopy, as these data have not been fully analyzed in previous studies.
+We argue that, despite the shortcomings of microscopy methods, they provide a consistent measure of changes in large phytoplankton.
+Our analysis emphasizes bottom-up drivers of community changes, as these data are available throughout the time series and at a monthly resolution.
 This study complements the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017.
-Understanding the regional dynamics of the Cariaco basin is crucial to understanding the broader ecological changes in tropical marine ecosystems driven by changes in climate.
+Understanding the regional dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems driven by changes in climate.
 
@@
 The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}.
-We included marks to highlight those years with missing monthly measurements, as these make the annual means less reliable, but they generally do not affect the general patterns observed, even if substituted by monthly means across the time series.
-Particularly towards the alter years of the time series, the sampling became sparse.
-The patterns emerging from the -u 10m wind component, the temperature and the directly related \qty{21}{\celsius} isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m.
-These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes.
-The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing inter-annual oscillations.
-Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration.
-Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.
+We included marks to highlight those years with a significant number of missing monthly measurements (\textgreater 2 months), as these make the annual means less reliable.
+Particularly towards the later years of the time series, the sampling became sparse.
+The patterns emerging from wind speed, SST, and the directly related \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm).
+These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes (see figure \ref{fig:sup:correlation} in the appendix).
+The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing inter-annual oscillations.
+Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration.
+Silicate does not follow a clear pattern but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished diatom abundances.
 
@@
 To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between the rainy and upwelling seasons.
-The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters.
-Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$).
-The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters.
-In some years, particularly during 2000, 2005, 2007, 2008 and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$).
+The bulk of the chlorophyll \textit{a} biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters.
+Chlorophyll \textit{a} shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$).
+The yearly mean chlorophyll \textit{a} biomass shows the highest variability within the upper 25 meters.
+In some years, particularly during 2000, 2005, 2007, 2008, and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$).
 In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012.
-The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value.
+The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter months (January and February) skew the value.
 Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014.
-Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series.
-In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
-The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter.
+Chlorophyll \textit{a} biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series.
+In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters is markedly reduced, while the 50–75 meter depth interval sees a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
+The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll \textit{a} biomass in seasonality, with maximum values observed during the upwelling season in winter.
 Genus richness is consistently highest in the top \qty{25}{meters} depth interval, which also shows the most variability over the time series.
-Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011.
-The collapse in diversity following the start of the time series does not follow the signal shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}).
+Across the time series, we observe maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, which only recovers after 2011.
+The collapse in diversity following the start of the time series does not follow the pattern shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}).
 Scaling this genus diversity by the cell count abundances, i.e.
 the Shannon index, shows a similar but less pronounced dynamic.
-Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016.
+Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016.
 The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$).
@@
 Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment.
-For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements.
+To achieve this, we employed clustering analysis, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements.
 Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix.
@@
 This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years.
+
 To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$).
@@
 
-This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes.
-In the Cariaco basin, decadal-scale climate oscillations have affected the wind-driven upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
+This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes.
+In the Cariaco Basin, decadal-scale climate oscillations have influenced the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
 % - a few sentences on basic seasonal dynamics removed here
 These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}).
-After 2004 we see higher SST, lower wind speed, reduced fluorometric chlorophyll \textit{a}, a collapse in functional group cell counts, all matching a shift in the AMO toward positive mean anomalies until 2014.
+After 2004, we observe higher sea surface temperatures (SST), lower wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding with a shift in the the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014.
 % - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries
@@
 % - "We found that including the time series data until the end of 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}."
-In 2014, physical conditions in the Cariaco basin returned to a regime of higher wind speeds, increased upwelling, and higher nutrient concentrations (see Figure \ref{fig:zscore}).
-At the same time, there was an increase in chlorophyll \textit{a} biomass (see Figure \ref{fig:divts} $a$) and a return to a community more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$).
-This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine capture observed in 2016 and 2017 in a location south-east of the Isla de Magarita \cite{gomez_gaspar_variacion_2025}.
-
-When comparing the phytoplankton community between the two clusters, we see a marked reduction in cell counts, which corresponds to a reduced concentration of chlorophyll \textit{a} biomass measured by fluorometry particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}).
+In 2014, physical conditions in the Cariaco Basin returned to a regime of higher wind speeds, increased upwelling, and elevated nutrient concentrations (see Figure \ref{fig:zscore}).
+Concurrently, there was an increase in chlorophyll \textit{a} biomass (see Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$).
+This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine captures observed in 2016 and 2017 in a location southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.
+
+When comparing the phytoplankton community between the two clusters, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a} biomass measured by fluorometry, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}).
 When including HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement here actually shows an increase between the periods 1996-2000 and 2006-2010 \cite{pinckney_phytoplankton_2015}, particularly at depths below \qty{55}{meters}.
-It should be noted that the HPLC data were analyzed by multiple labs throughout the time series, with a gap in data coverage between 2000 and 2006.
-\citeA{pinckney_phytoplankton_2015} could not find a consistent method for quantifying total chlorophyll \textit{a} between the different laboratories.
+It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006.
+\citeA{pinckney_phytoplankton_2015} could not find a consistent method for quantifying total chlorophyll \textit{a} across different laboratories.
 Here, we focus on the more consistently sampled fluorometric measurements.
-Independent spectophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Magarita Island follow the temporal trend of a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
-It should be noted that changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a}, but also other related forms of chlorophyll, which can vary depending on the composition of the community \cite{welschmeyer_fluorometric_1994}.
-
-Despite some of the difficulty in reconciling HPLC pigment data and microscopy cell counts, they agree in a reduction in diatom abundance after 2004.
-Cell counts of other functional groups decreased, where HPLC data indicate an increase and redistribution of biomass to depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}.
+Independent spectrophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Margarita Island follow the temporal trend of a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
+It should be noted that changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a} but also other related forms of chlorophyll, which can vary depending on the composition of the community \cite{welschmeyer_fluorometric_1994}.
+
+Despite some difficulties in reconciling HPLC pigment data and microscopy cell counts, both methods indicate a reduction in diatom abundance after 2004.
+Cell counts of other functional groups decreased, while HPLC data indicate an increase and redistribution of biomass to depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}.
 The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community transitioned to smaller individuals that were not captured under the microscope \cite{muller-karger_scientific_2019}.
@@
 
-If we compare the two parts of the first cluster, the community did reorganize, particularly the diatoms shifted from a community dominated by centric diatoms (Guinardia, Leptocylindrus, Skeletonema) to one dominated by pennate diatoms (Pseudonitzschia) \cite{pinckney_phytoplankton_2015}.
-
-However, the shift in yearly mean genus richness does not follow the abundances in the cell counts.
-Instead, we see a marked reduction following 1998 up until 2010, where richness recovers four years before the community seems to return to previous levels in 2014.
-The same trend is apparent, although to a less pronounced degree, in the other diversity metrics.
-Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance.
-
-The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy counts to deduce changes in the phytoplankton community.
-This limits the size range of the observable phytoplankton and the measurable diversity through the limited sampling volume \cite{cermeno_sampling_2014}.
+% check sizes of cells, then work this into a paragraph with another sentence!!
+If we compare the two parts of the first cluster, the community reorganized, particularly as diatoms shifted from a community dominated by centric diatoms (Guinardia, Leptocylindrus, Skeletonema) to one dominated by pennate diatoms (Pseudonitzschia) \cite{pinckney_phytoplankton_2015}.
+
+However, the shift in yearly mean genus richness does not correspond with the abundances in the cell counts.
+Instead, we observe a marked reduction following 1998 until 2010, after which richness recovers four years prior to the community seemingly returning to previous levels in 2014.
+The same trend is evident, albeit to a lesser degree, in the other diversity metrics.
+Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and exhibit a reduced yearly mean between 2000 and 2006, during which cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance.
+%
+
+The period between 2001 and 2006 lacks coverage in HPLC data, necessitating reliance on microscopy counts to deduce changes in the phytoplankton community.
+This limitation restricts the observable phytoplankton size range and the measurable diversity through the sampling methodology \cite{cermeno_sampling_2014}.
 We can see a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly an increase a few years ahead of the return in abundance to previous levels.
@@
 % Move the general point on ENSO to intro! Only Discussion-relevant part here..
-The El Niño Southern Osciallation (ENSO) has a strong teleconnection to the Caribbean Sea, since during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
+%copied to Intro: The El Niño Southern Osciallation (ENSO) has a strong teleconnection to the Caribbean Sea, since during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
 \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measuerments in the Cariaco basin, and only \qtyrange{24}{36}{\%} explanantion of variance with a time lag of 12 months using linear regression.
@@
 
-The in-situ measurements that best predict changes in the microphytoplankton community were nitrate concentration, temperature and salinity, in that order.
-Nitrate is a central nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco basin \cite{muller-karger_scientific_2019}.
+The in-situ measurements that best predict changes in the microphytoplankton community were nitrate concentration, SST and salinity, in that order.
+Nitrate is a central nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}.
 For dinoflagellates and diatoms, we see a strong threshold response around \qty{7}{\micro \mole}, which would indicate that strong upwelling conditions drive blooms of these functional groups.
-Other functional groups including \textit{Mesodinium} and \textit{Eutreptria} also show a threshold response at even higher nitrate concentrations (around \qty{9}{\micro \mole}), indicating an effect on their abundance by very strong upwelling.
-Similarly, for temperature, dinoflagellates and diatoms show a threshold around \qtyrange{21}{23}{\celsius}, which indicates higher abundances during low temperatures in the upper \qty{100}{m} driven by strong upwelling.
-For the other predictors, we do not see strong threshold responses.
+Other functional groups, including \textit{Mesodinium} and \textit{Eutreptria} also show a threshold response at even higher nitrate concentrations (around \qty{9}{\micro \mole}), indicating an effect on their abundance from very strong upwelling.
+Similarly, for temperature, dinoflagellates and diatoms exhibit a threshold around \qtyrange{21}{23}{\celsius}, which indicates higher abundances during low temperatures in the upper \qty{100}{m} driven by strong upwelling.
+For the other predictors, we do not observe strong threshold responses.
 
 Gradient forest analysis clearly identified the AMO and MEI v.2 climate indices as the best predictors of changes in the microphytoplankton community in the Cariaco basin.
-This is an interesting result, as it would be expected that more direct measures of ecosystem variability could better predict shifts in abundances.
+This is an interesting result, as one would expect that more direct measures of ecosystem variability could better predict shifts in abundances.
 One possible explanation is that in situ measurements contain much stronger variability due to measurement error, local and temporal variability, and generally only provide a single snapshot for an entire month.
@@
 %TODO test more of my variables along these axes, then add to discussion here and maybe put plot in supplemental material.
+
+% DEFINITELY NEED TO DISCUSS PROBLEMS OF CORRELATION WITH SUCH TIME SERIES DATA
 

### 2025-05-20 agujournaltemplate.tex
rewriting Intro 10%

@@
     % Phytoplankton is important
-    Phytoplankton provides the foundation for marine food webs and plays a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
+    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
     Phytoplankton consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}.
-The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them highly suitable indicators of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
+The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them highly suitable indicators of the ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
     The structure of the community can be considered an integrator of environmental factors, as it is strongly influenced by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}.
-In addition to such bottom-up effects, there are also forces of higher trophic levels, such as zooplankton, that structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}.
-Disentangling the effects driving changes in community composition and biodiversity of phytoplankton is a difficult but crucial task to better understand our oceans.
-
-    % Phytoplankton is under threat
-    On a global scale, human activities are driving a decline in biodiversity \cite{tittensor_mid-term_2014}.
+In addition to these bottom-up effects, influences from higher trophic levels, such as zooplankton, also structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}.
+Disentangling the effects driving changes in community composition and biodiversity of phytoplankton is a challenging but crucial task for better understanding our oceans.
+
     Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}.
-The drivers and effects of changes in biodiversity need to be further investigated, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and is under greater threat of extinction, compared to temperate regions \cite{finnegan_paleontological_2015}.
+The drivers and effects of these changes require further investigation, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.
+
+    Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area, but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.
+
+    - Interestingly, the trend for coastal phytoplankton biomass has been far from diminishing.
+    Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed \cite{dai_coastal_2023}.
 
@@
     Time series of biological observations offer valuable insights into how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.
-    % CARIACO is one such time series
-    One such project is the CARIACO time series, which has involved consistent sampling of physical and biogeochemical parameters in the Cariaco Basin between 1995 and 2017 \cite{muller-karger_scientific_2019}.
+
+    A tropical coastal ecosystem that has been studied intensively is the CARIACO time series, which has involved consistent sampling of physical and biogeochemical parameters in the Cariaco Basin from 1995 to 2017 \cite{muller-karger_scientific_2019}.
+
     % Describe CARIACO location
@@
 The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}.
-Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and a testbed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
-    %%% mention climate impacts (AMO + ENSO) here!
+Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
 
@@
 In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}.
-During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass measured in the Cariaco basin \cite{taylor_ecosystem_2012}.
+During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass measured in the Cariaco Basin \cite{taylor_ecosystem_2012}.
 
@@
 \toprule
-Variable Name & Units & Data source \\
+Variable Name & Units & Mean & Range & Std. Dev. & Data source \\
 \midrule
@@
 
+- alongside a global reduction in frequency compared to the linear trend between 2006 and 2012 \cite{dai_coastal_2023}.
+% <- for this paper, check references to ENSO MEI v.2!!!
+This could be interesting!
+
 Our limitation to bottom-up drivers of community change could also limit the explanatory power of our analysis, as top-down processes such as zooplankton grazing strongly affect the structure of the phytoplankton community \cite{banas_adding_2011, acevedo-trejos_mechanisms_2015}.

### 2025-05-20 text-backup.tex
rewriting Intro 10%

@@
 \end{keypoints}
+
+    % Phytoplankton is under threat
+    On a global scale, human activities are driving a decline in biodiversity \cite{tittensor_mid-term_2014}.
+Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}.
+The drivers and effects of changes in biodiversity need to be further investigated, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and is under greater threat of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.
+
+    %%%%%%%INTRO BACKUP REWRITE MAY "="%%%%%%%%
+
+      % Phytoplankton is important
+    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
+    Phytoplankton consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}.
+The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them highly suitable indicators of the ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
+    The structure of the community can be considered an integrator of environmental factors, as it is strongly influenced by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}.
+In addition to these bottom-up effects, influences from higher trophic levels, such as zooplankton, also structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}.
+Disentangling the effects driving changes in community composition and biodiversity of phytoplankton is a challenging but crucial task for better understanding our oceans.
+
+    Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}.
+The drivers and effects of these changes require further investigation, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.
+
+    Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area, but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.
+
+    - Interestingly, the trend for coastal phytoplankton biomass has been far from diminishing.
+    Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed \cite{dai_coastal_2023}.
+
+    % We need time series in order to understand effects and changes (threats) 
+    Time series of biological observations offer valuable insights into how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.
+
+    A tropical coastal ecosystem that has been studied intensively is the CARIACO time series, which has involved consistent sampling of physical and biogeochemical parameters in the Cariaco Basin from 1995 to 2017 \cite{muller-karger_scientific_2019}.
+
+    % Describe CARIACO location
+    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}.
+The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}.
+The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}.
+Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
+
+    % Describe general seasonal cycle with functional groups etc.
+    A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward.
+A smaller wind-driven secondary upwelling often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
+During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing.
+In the low-wind rainy season, the Cariaco Basin exhibits oligotrophic conditions, where the fraction of large phytoplankton cells is diminished, but nano- and picophytoplankton are present in considerable numbers throughout the year \cite{lorenzoni_characterization_2015}.
+
+    % Describe CARIACO changes in ecosystem
+    Since the inception of the time series in 1995, the ecosystem of the Cariaco basin has undergone marked changes.
+\citeA{taylor_ecosystem_2012} analyzed time series data up to 2010 and found significant decadal trends of increase in sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors related to the impact of global warming.
+The phytoplankton community identified by microscopy showed a drastic reduction in Diatom abundance and a shift toward smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass.
+    The trends in microscopy community data were partially confirmed and expanded by \citeA{pinckney_phytoplankton_2015}.
+They observed that the drastic reduction in diatoms counted using microscopy corresponded, but to a lesser degree, to diatom abundance calculated from the pigment composition measured by high-performance liquid chromatography (HPLC).
+This is most likely explained by a shift to diatom species with smaller cell sizes, which are more difficult to detect and identify using light microscopy \cite{lorenzoni_characterization_2015}.
+In addition, the pigment data indicate a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability in the community composition.
+These changes are most likely a response to the reduced upwelling and a deepening of the euphotic zone associated with a decrease in the strength of seasonal trade winds.
+Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.
+
+    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+    %% ADD Duplicate Points here:
+    During the initial phase of the time series (1996–2004), the Cariaco Basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
+In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed.
+During upwelling, the phytoplankton community was dominated by microphytoplankton, primarily consisting of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.
+
+    Following 2004, there was a transition in the physical and biogeochemical regime characterized by a significant reduction in trade winds during the upwelling period, which led to elevated sea surface temperatures and decreased mixing between the surface and deep ocean layers \cite{taylor_ecosystem_2012}.
+
+    Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis \cite{pinckney_phytoplankton_2015}.
+In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}.
+During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass measured in the Cariaco Basin \cite{taylor_ecosystem_2012}.
+
+    Previous analyses of phytoplankton community data did not include time series data beyond 2011, and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO."
+    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+    %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
+    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift.
+Data from 2012 until the end of the CARIACO time series in 2017, which were not included in previous studies of the phytoplankton community, indicate a return to stronger upwelling intensities and productivity levels similar to those of the late 1990s \cite{muller-karger_scientific_2019}.
+
+    %% Gaps that We Adresse:
+    % - More in-depth exploration of microscopy data & diversity time series
+    % - Unidirectional trends driven by anthropogenic climate change or low-freq natural cycles, such as AMO?
+        %Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".
+    % - Effect if ENSO on Phyotplankton community test, since previously not observed!
+             %This could also be part of the rational of the study. Formulated as question, e.g. "it is still unclear the effects of regional climate phenomena such as ENSO on this shift" - Esteban
+        % "\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco basin, despite strong teleconnection effects."
+
+    The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
+\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance with a time lag of 12 months using linear regression.
+
+    % NOw we come in!
+    In this study, we document changes in biodiversity and community structure in the Cariaco Basin between 1995 and 2017 and quantify the effects of bottom-up drivers and climate indices on changes in the phytoplankton community.
+We focus on phytoplankton cell count data obtained through microscopy, as these data have not been fully analyzed in previous studies.
+We argue that, despite the shortcomings of microscopy methods, they provide a consistent measure of changes in large phytoplankton.
+Our analysis emphasizes bottom-up drivers of community changes, as these data are available throughout the time series and at a monthly resolution.
+This study complements the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017.
+Understanding the regional dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems driven by changes in climate.

### 2025-05-20 agujournaltemplate.tex
Writing Intro 80%

@@
     Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
-    Phytoplankton consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}.
-The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them highly suitable indicators of the ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
-    The structure of the community can be considered an integrator of environmental factors, as it is strongly influenced by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}.
-In addition to these bottom-up effects, influences from higher trophic levels, such as zooplankton, also structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}.
-Disentangling the effects driving changes in community composition and biodiversity of phytoplankton is a challenging but crucial task for better understanding our oceans.
-
-    Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}.
-The drivers and effects of these changes require further investigation, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.
-
+The paraphyletic group consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}.
+The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
     Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area, but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.
-
-    - Interestingly, the trend for coastal phytoplankton biomass has been far from diminishing.
-    Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed \cite{dai_coastal_2023}.
-
-    % We need time series in order to understand effects and changes (threats) 
-    Time series of biological observations offer valuable insights into how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.
+Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed in the last 20 years, although regional effects are highly variable \cite{dai_coastal_2023}.
+    Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{boyd_biological_2016, henson_future_2021}. 
+    Long-term time series of biological observations are crucial to understanding how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
+Unfortunately, such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.
 
     A tropical coastal ecosystem that has been studied intensively is the CARIACO time series, which has involved consistent sampling of physical and biogeochemical parameters in the Cariaco Basin from 1995 to 2017 \cite{muller-karger_scientific_2019}.
-
     % Describe CARIACO location
@@
 Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
-
     % Describe general seasonal cycle with functional groups etc.
@@
 
-    % Describe CARIACO changes in ecosystem
-    Since the inception of the time series in 1995, the ecosystem of the Cariaco basin has undergone marked changes.
-\citeA{taylor_ecosystem_2012} analyzed time series data up to 2010 and found significant decadal trends of increase in sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors related to the impact of global warming.
-The phytoplankton community identified by microscopy showed a drastic reduction in Diatom abundance and a shift toward smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass.
-    The trends in microscopy community data were partially confirmed and expanded by \citeA{pinckney_phytoplankton_2015}.
-They observed that the drastic reduction in diatoms counted using microscopy corresponded, but to a lesser degree, to diatom abundance calculated from the pigment composition measured by high-performance liquid chromatography (HPLC).
-This is most likely explained by a shift to diatom species with smaller cell sizes, which are more difficult to detect and identify using light microscopy \cite{lorenzoni_characterization_2015}.
-In addition, the pigment data indicate a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability in the community composition.
-These changes are most likely a response to the reduced upwelling and a deepening of the euphotic zone associated with a decrease in the strength of seasonal trade winds.
+    % PARAGRAPH THREE - CHANGES IN CARIACO
+    Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes.
+\citeA{taylor_ecosystem_2012} analyzed time series data up to 2010 and found significant decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors linked to the impact of global warming.
+During the initial phase of the time series (1996–2004), the Cariaco Basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
+Following 2004, there was a transition in the physical and biogeochemical regime characterized by a significant reduction in trade winds during the upwelling period \cite{taylor_ecosystem_2012}.
+This strongly affected the ecosystem and phytoplankton community.
+Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, as confirmed to a lesser extent by pigment data analysis \cite{pinckney_phytoplankton_2015}.
+In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{lorenzoni_characterization_2015}.
+An analysis of the pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability in community composition \cite{pinckney_phytoplankton_2015}.
+During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass measured in the Cariaco Basin \cite{taylor_ecosystem_2012}.
 Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.
 
-    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-    %% ADD Duplicate Points here:
-    During the initial phase of the time series (1996–2004), the Cariaco Basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
-In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed.
-During upwelling, the phytoplankton community was dominated by microphytoplankton, primarily consisting of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.
-
-    Following 2004, there was a transition in the physical and biogeochemical regime characterized by a significant reduction in trade winds during the upwelling period, which led to elevated sea surface temperatures and decreased mixing between the surface and deep ocean layers \cite{taylor_ecosystem_2012}.
-
-    Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis \cite{pinckney_phytoplankton_2015}.
-In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}.
-During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass measured in the Cariaco Basin \cite{taylor_ecosystem_2012}.
-
-    Previous analyses of phytoplankton community data did not include time series data beyond 2011, and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO."
-    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-    %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
+    % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
     The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift.
-Data from 2012 until the end of the CARIACO time series in 2017, which were not included in previous studies of the phytoplankton community, indicate a return to stronger upwelling intensities and productivity levels similar to those of the late 1990s \cite{muller-karger_scientific_2019}.
+Data from 2012 until the end of the CARIACO time series in 2017 indicate a return to stronger upwelling intensities and phytoplankton productivity levels similar to those of the late 1990s \cite{muller-karger_scientific_2019}.
+    % What drives the changes??? Bottom-up/ Top-down ... CLIMATE
+    Previous analyses of the phytoplankton community did not include time series data beyond 2011, and \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as
+    "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation" (AMO). 
+    From this background, we aim to 1) investigate and relate changes in the depth distribution of biomass and community diversity over the entire time series, 2) check the structure and temporal changes of the phytoplankton community and 3) check whether in-situ upwelling metrics or climatic indices are better predicotrs of community change...
+    % - microscopy data was not analyzed in detail, particularly towards time series of diversity.
+    % Question: Does phytoplankton diversity follow the trends in biomass?
+    We aim to investigate and relate changes in the depth distribution of biomass and community diversity over the entire time series.
+
+    % What drives the changes??? Bottom-up/ Top-down ... CLIMATE
+    Previous analyses of the phytoplankton community did not include time series data beyond 2011, and \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as
+    "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation" (AMO).
+
+    The impact of large-scale climatic processes on the observed shift in the community is still unclear. 
+    %ENSO MEI v.2
+    The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
+
+    \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco basin, despite strong teleconnection effects in the tropical Atlantic \cite{enfield_tropical_1997}.
+
+    % AMO??
+    AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
+The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
 
@@
 
-    % NOw we come in!
-    In this study, we document changes in biodiversity and community structure in the Cariaco Basin between 1995 and 2017 and quantify the effects of bottom-up drivers and climate indices on changes in the phytoplankton community.
-We focus on phytoplankton cell count data obtained through microscopy, as these data have not been fully analyzed in previous studies.
+    Factors act at different scales \cite{wolkovich_temporal_2014}
+    We aim to examine correlations and structures in time series data at monthly and annual time scales.
+
+    The structure of the community can be considered an integrator of environmental factors, as it is strongly influenced by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}.
+In addition to these bottom-up effects, influences from higher trophic levels, such as zooplankton, also structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}.
+
+    - What are the open questions?? Drivers of community change.
+    - Zooplankton/Sardines one important aspect, but data coverage is sparse.
+    - Climate impacts, local upwelling system, etc.
+
+    - Microscopy data in more detail (use entire depth and time)
+
+    % FINAL PARAGRAPH LEADING TO METHODS
+    In this study, we document changes in biodiversity and community structure in the Cariaco Basin between 1995 and 2017 and quantify the relationships between bottom-up drivers, climate indices and the phytoplankton community.
+We focus on cell count data obtained through microscopy, as these data have not been fully analyzed in previous studies.
 We argue that, despite the shortcomings of microscopy methods, they provide a consistent measure of changes in large phytoplankton.
@@
 This study complements the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017.
-Understanding the regional dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems driven by changes in climate.
+Understanding the regional dynamics of the Cariaco Basin is crucial for comprehending the broader ecological changes in tropical marine ecosystems driven by shifts in climate.
+
+    %%%%%%% REASSEMBLE INTRO ABOVE HERE %%%%%%%%%%%%%
+
+    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+    %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
+    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift.
+Data from 2012 until the end of the CARIACO time series in 2017, which were not included in previous studies of the phytoplankton community, indicate a return to stronger upwelling intensities and productivity levels similar to those of the late 1990s \cite{muller-karger_scientific_2019}.
 
@@
 
-AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
+%AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
 The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
+% COPIED TO INTRO
 Although regional effecs may vary, a positive phase in the AMO is generally associated with periods of warmer temperatures and a shift of the ITCZ from north to south.

### 2025-05-22 agujournaltemplate.tex
Intro rewrite done

@@
 The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
-    Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area, but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.
+    Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.
 Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed in the last 20 years, although regional effects are highly variable \cite{dai_coastal_2023}.
@@
     Long-term time series of biological observations are crucial to understanding how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
-Unfortunately, such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.
+Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.
 
@@
     The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift.
-Data from 2012 until the end of the CARIACO time series in 2017 indicate a return to stronger upwelling intensities and phytoplankton productivity levels similar to those of the late 1990s \cite{muller-karger_scientific_2019}.
+Data from 2012 until the end of the CARIACO time series in 2017 indicate a return to stronger upwelling intensities and increased numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
+This return in 2014 to conditions similar to those of the late 1990s coincided with a recovery in sardine catches observed in 2016 and 2017 in a location southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.
+    The impact of large-scale climatic processes on the observed shifts in the community is still unclear.
+The Atlantic Multidecadal Oscillation (AMO) has emerged as an important indicator of the climate system and its effects on the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
+The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
+    The El Niño Southern Oscillation (ENSO) has a teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
+However, \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin and did not test for correlations with the AMO index.
+
     % What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
-    Previous analyses of the phytoplankton community did not include time series data beyond 2011, and \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as
-    "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation" (AMO). 
-    From this background, we aim to 1) investigate and relate changes in the depth distribution of biomass and community diversity over the entire time series, 2) check the structure and temporal changes of the phytoplankton community and 3) check whether in-situ upwelling metrics or climatic indices are better predicotrs of community change...
-    % - microscopy data was not analyzed in detail, particularly towards time series of diversity.
-    % Question: Does phytoplankton diversity follow the trends in biomass?
-    We aim to investigate and relate changes in the depth distribution of biomass and community diversity over the entire time series.
-
-    % What drives the changes??? Bottom-up/ Top-down ... CLIMATE
-    Previous analyses of the phytoplankton community did not include time series data beyond 2011, and \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as
-    "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation" (AMO).
-
-    The impact of large-scale climatic processes on the observed shift in the community is still unclear. 
-    %ENSO MEI v.2
-    The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
-
-    \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco basin, despite strong teleconnection effects in the tropical Atlantic \cite{enfield_tropical_1997}.
-
-    % AMO??
-    AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
-The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
-
-%% Gaps that We Adresse:
-    % - More in-depth exploration of microscopy data & diversity time series
-    % - Unidirectional trends driven by anthropogenic climate change or low-freq natural cycles, such as AMO?
-        %Previous analysis of the phytoplankton community data did not include time series data past 2011 and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO".
-    % - Effect if ENSO on Phyotplankton community test, since previously not observed!
-             %This could also be part of the rational of the study. Formulated as question, e.g. "it is still unclear the effects of regional climate phenomena such as ENSO on this shift" - Esteban
-        % "\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco basin, despite strong teleconnection effects."
-
-    The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
-\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance with a time lag of 12 months using linear regression.
-
-    Factors act at different scales \cite{wolkovich_temporal_2014}
-    We aim to examine correlations and structures in time series data at monthly and annual time scales.
-
-    The structure of the community can be considered an integrator of environmental factors, as it is strongly influenced by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}.
-In addition to these bottom-up effects, influences from higher trophic levels, such as zooplankton, also structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}.
-
-    - What are the open questions?? Drivers of community change.
-    - Zooplankton/Sardines one important aspect, but data coverage is sparse.
-    - Climate impacts, local upwelling system, etc.
-
-    - Microscopy data in more detail (use entire depth and time)
-
-    % FINAL PARAGRAPH LEADING TO METHODS
-    In this study, we document changes in biodiversity and community structure in the Cariaco Basin between 1995 and 2017 and quantify the relationships between bottom-up drivers, climate indices and the phytoplankton community.
-We focus on cell count data obtained through microscopy, as these data have not been fully analyzed in previous studies.
-We argue that, despite the shortcomings of microscopy methods, they provide a consistent measure of changes in large phytoplankton.
-Our analysis emphasizes bottom-up drivers of community changes, as these data are available throughout the time series and at a monthly resolution.
+    \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation."
+    Previous analyses of phytoplankton community diversity did not include time series data beyond 2011, and so far there was no detailed investigation of changes in diversity based on the microscopy cell count data.
+    From this background, we aim to 1) investigate changes in the depth distribution of phytoplankton biomass and community diversity over the entire time series, 2) relate the structure of the phytoplankton community data to the shifts observed in physical and biogeochemical variables, and 3) assess how decadal-scale climate osciallations as signified by the ENSO MEI v.2 and AMO index relate to changes in the phytoplankton community.
+    Our analysis focuses exclusively on bottom-up drivers of community changes, as these data are available throughout the entire time series and at a monthly resolution.
 This study complements the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017.

### 2025-05-23 agujournaltemplate.tex
Worked over Methods Section

@@
     Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{boyd_biological_2016, henson_future_2021}. 
-    Long-term time series of biological observations are crucial to understanding how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
+    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
 Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.
@@
 This return in 2014 to conditions similar to those of the late 1990s coincided with a recovery in sardine catches observed in 2016 and 2017 in a location southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.
-    The impact of large-scale climatic processes on the observed shifts in the community is still unclear.
+    The impact of large-scale climatic processes on the observed shifts in the community remains unclear.
 The Atlantic Multidecadal Oscillation (AMO) has emerged as an important indicator of the climate system and its effects on the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
@@
 
-    %%%%%%% REASSEMBLE INTRO ABOVE HERE %%%%%%%%%%%%%
-
-    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-    %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
-    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift.
-Data from 2012 until the end of the CARIACO time series in 2017, which were not included in previous studies of the phytoplankton community, indicate a return to stronger upwelling intensities and productivity levels similar to those of the late 1990s \cite{muller-karger_scientific_2019}.
-
 \section{Materials and Methods}
@@
 \subsection{In-Situ Data}
-    Samples were collected as part of the Cariaco Ocean Time Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
+    Samples were collected as part of the Cariaco Ocean Time Series Program during monthly research vessel cruises to a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
     The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this manuscript.
@@
 
-    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at the same depths in HDPE bottles and preserved with a 5\% (final concentration) formalin solution neutralized with sodium tetraborate.
+    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at the same depths in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate.
 Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period.
@@
     %from Mutshinda et al. 2013
-    Quality control the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level, correcting spelling variations and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}).
+    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts) correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}).
 After processing the data, 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.
 
-    Data sets for Niskin bottle samples, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed February, 2024).
-    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (more than three months) in coverage.
-Therefore, we did not include Zooplankton net tow and HPLC pigment data in our analysis.
-Our study instead focused on the phytoplankton microscopy data.
+    Data sets for Niskin bottle samples, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed in February, 2024).
+    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (i.e.
+more than three months) in coverage.
+Therefore, we did not include zooplankton net tow and HPLC pigment data in our analysis.
+Our study instead focused on the phytoplankton microscopy data and bottom-up drivers of change.
 We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity.
@@
     %from Taylor et al. 2012
-    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives near-shore upwelling of nutrient-rich waters by Ekman transport \cite{taylor_ecosystem_2012}.
+    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}.
 We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling coordinates (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024).
-
-    The monthly AMO index data (10 year lowpass), which is as an area average of detrended low-pass filtered North-Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} (accessed in June, 2024).
-    %The term "Atlantic Multidecadal Oscillation" (AMO) describes the large-scale fluctuations in Atlantic sea surface temperature (SST) that occur over multiple decades \cite{kerr2000north}.
-This designation is intended to underscore the distinct multidecadal nature of these oceanic variations in contrast to the inter-annual variability linked with the atmospheric North Atlantic Oscillation (NAO) \cite{enfield2001atlantic}.
-
-    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of the surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024).
+This variable is referred to as "wind speed" throughout the manuscript for simplicity.
+
+    The monthly AMO index data (10-year low-pass), which represents an area average of detrended low-pass filtered North Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} (accessed in June, 2024).
+
+    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024).
 In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).
@@
     % Niskin, CTD and Phytoplankton data processing
-    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} using data pipelines created with tidyverse packages \cite{wickham_welcome_2019}.
-All data processing and analysis code is publicly available on GitHub \url{https://github.com/ben1post/cariaco-ms2}.
-%Note: Repo currently still set to private
+    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using tidyverse packages \cite{wickham_welcome_2019}.
+All data processing and analysis code is publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
     For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S.
 National Oceanographic Data Center \cite{johnson2006world}.
-The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis.
-For phytoplankton microscopy data, individual identified taxonomic units (Species or Genus) were interpolated, integrated, and then grouped by Genus via the sum of interpolated counts.
+The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 \meter for further analysis.
+Additionally, sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}.
+For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated, integrated, and then grouped by genus via the sum of interpolated counts.
 
 \subsection{Data analysis}    
-    To give a general overview of the data and trends in the time series, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$).
-Cruises from 1995 and 2017 were excluded, as there were too few measurements to provide a reliable yearly mean estimate.
-
-    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to make the data set more homogeneous and robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}.
-Furthermore, we calculate the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
+    To provide a general overview of the data and trends in the time series, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$).
+Cruises from 1995, 2016, and 2017 were excluded due to insufficient measurements to provide a reliable yearly mean estimate.
+
+    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to create a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}.
+Furthermore, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
 
@@
 Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}.
-We used the Bray distance matrix in a 2-dimensional ordination that reached a converging solution within 30 steps.
+We employed the Bray distance matrix in a two-dimensional ordination that reached a converging solution within 30 steps.
 To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations.
-The arrows indicate the gradient of a 2-dimensional linear trend surface.
+The arrows indicate the gradient of a two-dimensional linear trend surface.
     To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}.
-%Here I want to add "significant diferences calculation" using e.g.
-ANOVA.
+To test whether the environmental variables, chlorophyll $a$ and diversity metrics varied significantly between the two community clusters, we employed a two-sided Wilcoxon sum rank test using the "wilcox.test" function in R.
     % detailed explanation of NMDS in Winder and Hunter 2008
@@
     \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
-    To quantify the effects of bottom-up drivers in the Cariaco basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis.
+    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis.
 Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}.
@@
 Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}.
-This analysis was conducted using monthly measurements integrated to 100 m depths for observed genera counts, ensuring the inclusion of all available data.
-The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, the -u 10m wind component, the \qty{21}{\celsius} isotherm, integrated temperature and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$).
+This analysis was conducted using monthly measurements integrated to 100 m depths for observed genus counts, ensuring the inclusion of all available data.
+The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, the -u 10m wind component, the \qty{21}{\celsius} isotherm, SST, and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$).
 The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology takes into account by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}.
-To consider the temporal lag in effect, in particular of climate indices, we included time lags up to 6 months of all climate variables, as well as 3 months lags for in-situ measurements in exploratory runs, and chose time lags that scored highest in the goodness-of-fit statistic for the final model run (see Table \ref{app:table1} in the Appendix).
+To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs.
+We chose time lags that scored highest in the goodness-of-fit statistic for the final model run (see Table \ref{app:table1} in the Appendix).
+    % Note: Add some more info about Supplmentary Results and test runs here!
 
@@
 \centering
-\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrrrrrrrrrrr}
+\resizebox{\textwidth}{!}{\begin{tabular}[t]{rrl}
 \toprule
-Variable Name & Units & Mean & Range & Std. Dev. & Data source \\
+Variable Name & Units  & Data source \\
 \midrule
-MEI v.2 & unitless & Climate Indices provided by NOAA \\
-AMO & unitless & Climate Indices provided by NSF NCAR \\
-Wind speed & m/s & ERA5 model reanalysis: 10 m u-component of wind speed \\
-Precipitation & 0.013 & ERA5 model reanalysis: total precipitation\\
-Evaporation & NA & ERA5 model reanalysis: evaporation \\
-SST & °C & CARIACO time series: CTD, Average temperature per month measured in top 10 m\\
-21° C Isotherm & m & CARIACO time series: CTD, depth at which temperature reaches 21°C\\
+MEI v.2 & unitless & Climate index provided by NOAA \\
+AMO & unitless & Climate index provided by NSF NCAR \\
+Wind speed & m/s & ERA5 model reanalysis, negative 10 m u-component of wind speed \\
+Precipitation & m & ERA5 model reanalysis, total precipitation\\
+Evaporation & m & ERA5 model reanalysis, evaporation \\
+SST & °C & CTD, Average temperature measured in top 10 m\\
+21° C Isotherm & m & CTD, depth at which temperature reaches 21°C\\
 Salinity & PSU & Niskin, top 100m \\
@@
 
-This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes.
-In the Cariaco Basin, decadal-scale climate oscillations have influenced the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
+This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes.
+In the Cariaco Basin, decadal-scale climate oscillations have impacted the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
 % - a few sentences on basic seasonal dynamics removed here
 These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}).
-After 2004, we observe higher sea surface temperatures (SST), lower wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding with a shift in the the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014.
+After 2004, we observed higher sea surface temperatures (SST), lower wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014.
 % - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries
@@
 This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}.
-Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but are particularly prevalent when (as observed in the Cariaco basin) temperatures increase and the nutrient supply decreases \cite{mousing_global_2014}.
+Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but are particularly prevalent when (as observed in the Cariaco Basin) temperatures increase and the nutrient supply decreases \cite{mousing_global_2014}.
 
@@
 This limitation restricts the observable phytoplankton size range and the measurable diversity through the sampling methodology \cite{cermeno_sampling_2014}.
-We can see a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly an increase a few years ahead of the return in abundance to previous levels.
-This pattern results in the fact that no significant differences in diversity are observed from microscopy cell counts between two clusters, either by Shannon index or by genus richness (see Figure \ref{fig:zscore} and \ref{fig:divts}).
+We observe a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly an increase a few years prior to the return in abundance to previous levels.
+This pattern results in no significant differences in diversity observed from microscopy cell counts between the two clusters, either by Shannon index or by genus richness (see Figure \ref{fig:zscore} and \ref{fig:divts}).
 We do see a significant difference in the Pilou index between the two clusters.

### 2025-05-27 agujournaltemplate.tex
updated Methods and Results almost done

@@
     From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (i.e.
-more than three months) in coverage.
+no more than three months) in coverage.
 Therefore, we did not include zooplankton net tow and HPLC pigment data in our analysis.
@@
 This variable is referred to as "wind speed" throughout the manuscript for simplicity.
+The data for precipitation and evaporation are similarly extracted from the ERA5 model reanalysis dataset for the aforementioned grid cell.
+Evaporation provides an estimate of the accumulated amount of water that has evaporated from the Earth's surface, while precipitation is the estimate of total precipitation, comprising all liquid that falls to Earth's surface as rain or snow.
 
@@
 Additionally, sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}.
-For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated, integrated, and then grouped by genus via the sum of interpolated counts.
+For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.
 
@@
     To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}.
-To test whether the environmental variables, chlorophyll $a$ and diversity metrics varied significantly between the two community clusters, we employed a two-sided Wilcoxon sum rank test using the "wilcox.test" function in R.
+To test whether the environmental variables, chlorophyll $a$ and diversity metrics varied significantly between the two community clusters, we employed a two-sided Wilcoxon sum rank test using the "wilcox.test" function in R \cite{r_core_team_r_2024}.
     % detailed explanation of NMDS in Winder and Hunter 2008
@@
 This analysis was conducted using monthly measurements integrated to 100 m depths for observed genus counts, ensuring the inclusion of all available data.
-The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, the -u 10m wind component, the \qty{21}{\celsius} isotherm, SST, and salinity, as well as the concentration of nutrients ($NO_3$, $PO_4$ and $SiO_4$).
-The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology takes into account by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}.
+The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, wind speed, evaporation, precipitation, the \qty{21}{\celsius} isotherm, SST, and salinity, as well as the concentration of nutrients ($\mathrm{NO}_3$, $\mathrm{PO}_4$ and $\mathrm{SiO}_4$).
+    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}.
+    We show a model run without any time lags in the supplemental material (see Figure \ref{fig:sup:GFoutput_nolags}).
 To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs.
 We chose time lags that scored highest in the goodness-of-fit statistic for the final model run (see Table \ref{app:table1} in the Appendix).
-    % Note: Add some more info about Supplmentary Results and test runs here!
+    % Note: Perhaps add some more info about Supplmentary Results and test runs here?
 
@@
 AMO & unitless & Climate index provided by NSF NCAR \\
-Wind speed & m/s & ERA5 model reanalysis, negative 10 m u-component of wind speed \\
-Precipitation & m & ERA5 model reanalysis, total precipitation\\
-Evaporation & m & ERA5 model reanalysis, evaporation \\
-SST & °C & CTD, Average temperature measured in top 10 m\\
-21° C Isotherm & m & CTD, depth at which temperature reaches 21°C\\
-Salinity & PSU & Niskin, top 100m \\
-NO3 & µM & Niskin, top 100m \\
-PO4 & µM & Niskin, top 100m \\
-SiO4 & µM & Niskin, top 100m \\
+Wind speed & \qty{}{\meter/\second} & ERA5 model reanalysis, negative 10 m u-component of wind speed \\
+Precipitation & \qty{}{\meter} & ERA5 model reanalysis, total precipitation\\
+Evaporation & \qty{}{\meter} & ERA5 model reanalysis, evaporation \\
+SST & \qty{}{\celsius} & CTD, Average temperature measured down to \qty{10}{\meter} depth\\
+\qty{21}{\celsius} isotherm & \qty{}{\meter} & CTD, depth at which temperature reaches 21°C\\
+Salinity & PSU & Niskin, interpolated down to \qty{100}{\meter} depth \\
+$\mathrm{NO}_3$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
+$\mathrm{PO}_4$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
+$\mathrm{SiO}_4$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
 \end{tabular}}
@@
 The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}.
-We included marks to highlight those years with a significant number of missing monthly measurements (\textgreater 2 months), as these make the annual means less reliable.
-Particularly towards the later years of the time series, the sampling became sparse.
-The patterns emerging from wind speed, SST, and the directly related \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm).
-These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes (see figure \ref{fig:sup:correlation} in the appendix).
+We included marks to highlight those years with a significant number of missing monthly measurements (\textgreater 2 months and $\geq$ 4 months), as these make the annual means less reliable.
+Particularly toward the later years of the time series, the sampling became sparse.
+The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm).
+These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes (see Figure \ref{fig:sup:correlation} in the appendix).
 The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing inter-annual oscillations.
 Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration.
-Silicate does not follow a clear pattern but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished diatom abundances.
+Silicate does not follow a clear pattern but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to diminished diatom abundances.
 
@@
 The bulk of the chlorophyll \textit{a} biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters.
-Chlorophyll \textit{a} shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$).
+Chlorophyll \textit{a} exhibits marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$).
 The yearly mean chlorophyll \textit{a} biomass shows the highest variability within the upper 25 meters.
 In some years, particularly during 2000, 2005, 2007, 2008, and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$).
-In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012.
+In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period from 1996 to 2004 and only recover in 2012.
 The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter months (January and February) skew the value.
@@
 the Shannon index, shows a similar but less pronounced dynamic.
-Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016.
-The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$).
-Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$).
+Diversity remains high throughout 2000 and recovers more quickly but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016.
+The seasonal signal is lost (see Figure \ref{fig:divts} $f$).
+Community evenness, as signified by the Pielou index, shows no seasonal signal and a very large variance within seasons and across depth (see Figure \ref{fig:divts} $h$).
 The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value.
@@
 % Figure 4
-Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment.
-To achieve this, we employed clustering analysis, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements.
-Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix.
-The community is divided into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013.
-Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, the turnover is higher again.
-This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years.
-
-To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$).
+Taking the microscopy community data interpolated within the top 100 meters, we aimed to identify any apparent structure within the time series that could be related to the physical environment.
+To achieve this, we employed clustering analysis on both the annual aggregate of cell counts from microscopy and the individual monthly measurements.
+Figure \ref{fig:clustering} $a$ presents the results of clustering the annual aggregate communities using a linear complete linkage algorithm based on the binary Jaccard matrix.
+The community is divided into two groups, with the first cluster encompassing the years 1996-2004 and 2014-2016, while the second cluster includes the intermediate years from 2004-2013.
+Inter-annual species turnover is illustrated in Figure \ref{fig:clustering} $b$, revealing an increasing trend in turnover until 2004.
+Turnover subsequently decreased during the period of cluster 2, but increased again in 2014, with return to cluster 1.
+This pattern is also reflected in the hierarchical clustering results in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years.
+
+To check whether the annual aggregation of community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$).
 This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends.
-The first axis separates the two clusters, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index.
-The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed.
+The first axis separates the two clusters, with cluster 2 trending towards positive anomalies in the AMO and precipitation, while cluster 1 trends towards positive anomalies in evaporation, salinity, and the MEI v.2 index.
+The second NMDS axis appears to follow the trends in upwelling intensity, as indicated by the opposing trend arrows for nutrient concentrations, temperature, and wind speed.
 Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering.
-Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters.
-Upwelling related variables show distinct patterns between the two clusters.
-For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate.
-The AMO index shows a separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies.
-Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences.
-For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.
-
-% Figure 5
+
+% Figure 5 - NEW
+Figure \ref{fig:clustcomp} $a$ to $o$ shows density distributions for environmental variables, climate, and diversity indices between the two clusters.
+Upwelling-related variables show distinct patterns between the two clusters.
+For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} isotherm, lower temperatures, reduced evaporation, and higher salinity and nutrient concentrations, particularly for nitrate.
+The AMO index indicates a separation between the two clusters, with cluster 1 exhibiting more negative anomalies, while cluster 2 has a higher occurrence of positive anomalies.
+Although there are no significant differences in genus richness and the Shannon index between clusters, there is a slightly increased median in the Pielou index for cluster 2.
+The density distribution reveals some differences.
+For cluster 1, there is higher variance in genus richness, while cluster 2 shows greater variance and a bimodal distribution for the Shannon and Pielou indices.
+
+% Figure 6
 To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
-Preliminary model runs were used to identify if time lags of the predictor variables could better explain the data (see Appendix), and this resulted in choosing the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months.
-In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the highest predictive capacity for climate indices with time lags and upwelling-related variables such as nitrate in the upper 100 meters and wind speed scoring slightly lower.
-The \qty{21}{\celsius} Isotherm and silicate and phosphate concentrations scored lowest in predicting changes within the community.
-The way Gradient Forest predicts shifts in the community is by building a random forest model for each observed genus across the time series.
+Preliminary model runs were used to determine if time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months.
+In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity.
+The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentrations scored the lowest in predicting changes within the community.
+Gradient Forest predicts shifts in the community by building a random forest model for each observed genus across the time series.
 The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$.
-The microscopy data are biased towards larger cell sizes, and therefore most of the predicted genus cell counts are from diatoms.
-However, two Haptophytes and four Dinoflagellates, as well as the Cyanobacteria \textit{Synechococcus} do rank highly.
+The microscopy data are biased toward larger cell sizes, and therefore, most of the predicted genus cell counts are from diatoms.
+However, two haptophyte and three dinoflagellate genera rank relatively high.
 Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range.
-The black line indicates cumulative importance across the entire community and the colored dashed lines indicate importance per phytoplankton group.
-For the AMO index, we see a clear pattern in which negative anomalies highly affect the community.
-This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes.
-Nitrate concentrations between 6 and 8 \unit{\micro \mole} is where the community changed most, indicating these values as a threshold for the composition of the community.
-For wind speed, temperature, and salinity, changes occurred most at parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}) and high salinity (above \qty{36.8}{PSU}).
-An interesting aspect is that, for both climate indices, haptophytes showed the strongest cumulative importance, but for diatoms, the nitrate concentration seemed to be a stronger predictor.
-
-% ADD HERE MORE DISCUSSION OF MODEL COMPARISON
+The black line indicates cumulative importance across the entire community, and the colored dashed lines indicate importance per phytoplankton group.
+For the AMO index, we see a clear pattern in which negative anomalies affect the community.
+This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria.
+Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range where the community changed most, indicating these values as a threshold for community composition.
+See figure \ref{fig:sup:GFoutput_lags_extra} in the appendix for response curves for all variables not shown here.
+For wind speed, SST, and salinity, changes occurred most at parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}) and high salinity (above \qty{36.8}{PSU}).
 
@@
 (b,d,f,h) Boxplots of monthly measurements across four depth intervals grouped by season.
-Upwelling season lasts from December to April, Rainy Season from May to November.}
+Upwelling season lasts from December to May, Rainy Season from June to November.}
 \label{fig:divts}
@@
 
-This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes.
+This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is influenced by large-scale climatic changes.
 In the Cariaco Basin, decadal-scale climate oscillations have impacted the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
@@
 These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}).
-After 2004, we observed higher sea surface temperatures (SST), lower wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014.
+After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding to a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014.
 % - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries
@@
 This limitation restricts the observable phytoplankton size range and the measurable diversity through the sampling methodology \cite{cermeno_sampling_2014}.
-We observe a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly an increase a few years prior to the return in abundance to previous levels.
+We observe a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly, an increase a few years prior to the return in abundance to previous levels.
 This pattern results in no significant differences in diversity observed from microscopy cell counts between the two clusters, either by Shannon index or by genus richness (see Figure \ref{fig:zscore} and \ref{fig:divts}).
-We do see a significant difference in the Pilou index between the two clusters.
+We observe a significant difference in the Pilou index between the two clusters.
 Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}.
@@
 % NOW let's jump to Gradient Forest and discussing CLIMATE impacts! for a final paragraph, and then we are done here!!
-To identify the variables that can best predict changes in the phytoplankton community, we used a gradient forest analysis using microscopy cell counts at the genus level as predicted variables and climate indices and in situ measurements of bottom-up drivers as predictors.
+To identify the variables that can best predict changes in the phytoplankton community, we conducted a gradient forest analysis using microscopy cell counts at the genus level as predicted variables and climate indices, along with in situ measurements of bottom-up drivers, as predictors.
 Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series.
-Interestingly, our analysis found that the strongest predictors of changes within the community are not in situ measurements, but actually the AMO and the MEI v.2 climate indices.
-These are followed in importance by nitrate concentration and temperature in the upper \qty{100}{meters} (see Figure \ref{fig:GF}).
+Interestingly, our analysis revealed that the strongest predictors of changes within the community are not in situ measurements, but rather the AMO and the MEI v.2 climate indices.
+These are followed in importance by nitrate concentration and SST (see Figure \ref{fig:GF}).
 

### 2025-06-04 agujournaltemplate.tex
Finalizing Discussion 85%

@@
 The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm).
-These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes (see Figure \ref{fig:sup:correlation} in the appendix).
+These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in the appendix).
 The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing inter-annual oscillations.
@@
 Community evenness, as signified by the Pielou index, shows no seasonal signal and a very large variance within seasons and across depth (see Figure \ref{fig:divts} $h$).
-The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value.
+The inter-annual dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value.
 There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016.
@@
 Inter-annual species turnover is illustrated in Figure \ref{fig:clustering} $b$, revealing an increasing trend in turnover until 2004.
-Turnover subsequently decreased during the period of cluster 2, but increased again in 2014, with return to cluster 1.
+Turnover subsequently decreased during the period of cluster 2 but increased again in 2014, marking the return to cluster 1.
 This pattern is also reflected in the hierarchical clustering results in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years.
+We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster respectively.
+The diatom Thalassiosira dominated the early period and was abundant in all other periods.
+Pseudo-nitzschia dominates cluster 2, where the haptophyte Emiliana was the second most abundant genus.
+The later part of cluster 1 is dominated by Emiliana.
 
@@
 Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range where the community changed most, indicating these values as a threshold for community composition.
-See figure \ref{fig:sup:GFoutput_lags_extra} in the appendix for response curves for all variables not shown here.
+See Figure \ref{fig:sup:GFoutput_lags_extra} in the appendix for response curves for all variables not shown here.
 For wind speed, SST, and salinity, changes occurred most at parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}) and high salinity (above \qty{36.8}{PSU}).
@@
 \noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
-\caption{Map of CARIACO Time Series location in the Cariaco basin in Venezuela in the Caribbean Ocean, located at \ang{10.50}N, \ang{64.67}W}
+\caption{Map of CARIACO Ocean Time-Series location in the Cariaco basin in Venezuela in the Caribbean Ocean, located at \ang{10.50}N, \ang{64.67}W.}
 \label{fig:map}
@@
 \noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
-\caption{Heatmap of yearly average values, normalized for comparison as z-scores.
-Red indicates a larger value, blue indicates a smaller value for that particular year.
-The years 1995 (n=2) and 2017 (n=1) were removed due to lack of data, additionally we marked years with two or less and three or more missing values.
+\caption{Time series showing annual z-scores of selected indicator and driver variables.
+Red indicates negative annual anomalies and blue indicates positive annual anomalies for that year.
+Deeper shades indicate larger anomalies.
+All data was available at monthly resolution.
+The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to lack of data, additionally we marked years with more than two and four or more missing monthly values to highlight unreliable annual mean estimates.
 Diversity Indices are calculated based on microscopy counts at the Genus level.}
@@
 \caption{Time series and seasonality of chlorophyll \textit{a} and diversity indices across four depth intervals.
-(a,c,e,g) Annual mean across time series for years with more than 5 monthly sampling.
+(a,c,e,g) Large dots and lines mark annual mean across time series for years with more than 5 monthly sampling.
 Transparent dots show monthly raw values.
@@
 Color of dots based on which yearly cluster the monthly data is in.
-Vector overlay of ENVFIT shows underlying environmental variables that are co-varying with the community data.}
+Vector overlay of ENVFIT shows how environmental variables are co-varying with the community data.
+Length of vector indicates strength of relationship.}
 \label{fig:clustering}
@@
 \noindent\includegraphics[width=\textwidth]{fig/Figure5_ClustCompPlot_v2.pdf}
-\caption{(Plots (a-o) below show density distributions of the environmental variables as well as diversity indices between the clustered years.
-Dashed line shows the median and significance levels of Wilcoxon sum rank tests between clusters are shown above plots  (p-value: ***\textless0.001, **\textless0.01, *\textless0.05, n.s.=not significant).}
+\caption{(Plots (a-o) below show density distributions of the environmental variables, chlorophyll \textit{a} and diversity indices between the clustered years.
+Dashed line marks the median.
+Significance levels of Wilcoxon sum rank tests between clusters are shown above plots  (p-value: ***\textless0.001, **\textless0.01, *\textless0.05, n.s.=not significant).}
 \label{fig:clustcomp}
@@
 
-\caption{Wilcoxon sum rank test with continuity correction for variables between the two clusters.
-Values for in-situ data are interpolated across the top 100 meters for each individual monthly cruise sampling.}
+\caption{Results of two-sided Wilcoxon sum rank test with continuity correction for variables between the two clusters.
+Cluster 1 is defined as the years 1996 - 2003 and 2014 - 2016.
+Cluster 2 includes the years 2004 - 2013.}
 \centering
@@
 
-This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is influenced by large-scale climatic changes.
+% Clearly and briefly restating main findings 
+% Interpret what these findings mean in the context of the study
+% Emphasize what is novel and important (without comparison)
+% Only in following paragraphs: Compare to existing literature, highlight agreements, contradictions, and new persepctives. Discuss why and explain reasons for either.
+
+\subsection{Community reorganization and biomass shifts}
+% P1: Introductory paragraph highlighting the overall dynamics/shifts
+This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is in turn influenced by large-scale climatic changes.
 In the Cariaco Basin, decadal-scale climate oscillations have impacted the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
-% - a few sentences on basic seasonal dynamics removed here
 These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}).
 After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding to a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014.
-% - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries
-
-% - "We found that including the time series data until the end of 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}."
+We found that including the time series data until the end of 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}.
 In 2014, physical conditions in the Cariaco Basin returned to a regime of higher wind speeds, increased upwelling, and elevated nutrient concentrations (see Figure \ref{fig:zscore}).
@@
 
+% P2: Discuss community reorganisation, biomass shifts vs HPLC data
+We found that the annual aggregates of community data showed two clusters, one including the time series up until 2004 and the end of the time series past 2013, and the other consisting of the intermediate time period between 2004 and 2013 (see Figure \ref{fig:clustering}).
 When comparing the phytoplankton community between the two clusters, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a} biomass measured by fluorometry, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}).
-When including HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement here actually shows an increase between the periods 1996-2000 and 2006-2010 \cite{pinckney_phytoplankton_2015}, particularly at depths below \qty{55}{meters}.
+The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community transitioned to smaller individuals that were not captured under the microscope \cite{muller-karger_scientific_2019}.
+This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}.
+Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but are particularly prevalent when temperatures increase and the nutrient supply decreases \cite{mousing_global_2014}, as was observed in the Cariaco Basin.
+
+% Disucssion of HPLC vs Microscopy!
+When comparing microscopy counts to previous analyses of HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement here actually shows an increase between the periods 1996-2000 and 2006-2010 \cite{pinckney_phytoplankton_2015}, particularly at depths below \qty{55}{meters}.
+Cell counts of all functional groups decreased, while HPLC data indicate an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}.
 It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006.
@@
 Independent spectrophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Margarita Island follow the temporal trend of a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
-It should be noted that changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a} but also other related forms of chlorophyll, which can vary depending on the composition of the community \cite{welschmeyer_fluorometric_1994}.
-
-Despite some difficulties in reconciling HPLC pigment data and microscopy cell counts, both methods indicate a reduction in diatom abundance after 2004.
-Cell counts of other functional groups decreased, while HPLC data indicate an increase and redistribution of biomass to depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}.
-The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community transitioned to smaller individuals that were not captured under the microscope \cite{muller-karger_scientific_2019}.
-This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}.
-Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but are particularly prevalent when (as observed in the Cariaco Basin) temperatures increase and the nutrient supply decreases \cite{mousing_global_2014}.
-
-% check sizes of cells, then work this into a paragraph with another sentence!!
-If we compare the two parts of the first cluster, the community reorganized, particularly as diatoms shifted from a community dominated by centric diatoms (Guinardia, Leptocylindrus, Skeletonema) to one dominated by pennate diatoms (Pseudonitzschia) \cite{pinckney_phytoplankton_2015}.
-
+Changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a} but also other related forms of chlorophyll, which can vary depending on the composition of the community \cite{welschmeyer_fluorometric_1994}.
+
+% P3: Shift in Diversity does not match shift in biomass/cell counts, why?
 However, the shift in yearly mean genus richness does not correspond with the abundances in the cell counts.
-Instead, we observe a marked reduction following 1998 until 2010, after which richness recovers four years prior to the community seemingly returning to previous levels in 2014.
+Instead, we observe a marked reduction from 1998 until 2010, after which richness recovers four years prior to the community seemingly returning to previous levels in 2014.
 The same trend is evident, albeit to a lesser degree, in the other diversity metrics.
-Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and exhibit a reduced yearly mean between 2000 and 2006, during which cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance.
-%
-
+Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and exhibit a reduced yearly mean between 2000 and 2006, during which cell counts recorded an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance.
 The period between 2001 and 2006 lacks coverage in HPLC data, necessitating reliance on microscopy counts to deduce changes in the phytoplankton community.
@@
 This pattern results in no significant differences in diversity observed from microscopy cell counts between the two clusters, either by Shannon index or by genus richness (see Figure \ref{fig:zscore} and \ref{fig:divts}).
-We observe a significant difference in the Pilou index between the two clusters.
+We observe a significant difference in the Pielou's index between the two clusters.
 Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}.
-This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and the thereby reduced disturbance of the community.
-The resulting reduced compositional turnover is also observed in a decrease in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$).
-As the productivity and biomass of phytoplankton communities are mainly driven by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
-
-% NOW let's jump to Gradient Forest and discussing CLIMATE impacts! for a final paragraph, and then we are done here!!
+This increase in evenness, as quantified by Pielou's index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community.
+The resulting decrease in compositional turnover is also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$).
+As the productivity and biomass of phytoplankton communities are primarily driven by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
+
+\subsection{Drivers of change: Gradient forest analysis}
+% P4: Introduce GF Analysis and discuss results
 To identify the variables that can best predict changes in the phytoplankton community, we conducted a gradient forest analysis using microscopy cell counts at the genus level as predicted variables and climate indices, along with in situ measurements of bottom-up drivers, as predictors.
+Gradient forest is an extension of the random forest algorithm, designed to identify thresholds and patterns in species abundance or composition along environmental gradients \cite{ellis_gradient_2012}.
 Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series.
-Interestingly, our analysis revealed that the strongest predictors of changes within the community are not in situ measurements, but rather the AMO and the MEI v.2 climate indices.
-These are followed in importance by nitrate concentration and SST (see Figure \ref{fig:GF}).
-
-These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, since the scaling axis that separates the two yearly aggregate community clusters is NMDS 1, which covaries most strongly with changes in the AMO, MEI v.2 (see Figure \ref{fig:clustering} $c$).
-Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS2.
-This indicates that the climate indices capture some variability, which is not represented in the more direct measurements of upwelling intensity.
-Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling, but also depend on precipitation and fresh water input through surface runoff, show greater variance along NMDS1.
-Both salinity and silicate concentration show some explanatory power in the gradient forest analysis, although they are weaker predictors than temperature, nitrate concentration, and the climate indices.
-
-Using microscopy cell counts as a basis for this analysis meant that the results should be interpreted in terms of changes in the microphytoplankton community, due to the size limitation of identifying smaller species via light microscopy.
-Grouping cell counts by genera does create a loss of resolution, but is based on the assumption that species within a genus generally show similar ecophysiological responses.
+Interestingly, our analysis revealed that the strongest predictor of changes within the community was not an in situ measurement but rather the AMO climate index.
+The AMO is followed in importance by SST, the MEI v.2 index, and nitrate concentration (see Figure \ref{fig:GF}).
+
+These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separates the two yearly aggregate community clusters is NMDS 1, which co-varies most strongly with changes in the AMO, MEI v.2, salinity, evaporation, and precipitation (see Figure \ref{fig:clustering} $c$).
+Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS 2.
+This indicates that the climate indices capture some variability that is not represented in the more direct measurements of upwelling intensity.
+Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, show greater variance along NMDS 1.
+Salinity, evaporation, and silicate concentration exhibit some explanatory power in the gradient forest analysis, although they are weaker predictors than SST, nitrate concentration, and the climate indices.
+
+Using microscopy cell counts as a basis for this analysis means that the results should be interpreted in terms of changes in the microphytoplankton community, due to the size limitations of identifying species via light microscopy.
+Grouping cell counts by genera does create a loss of resolution but is based on the assumption that species within a genus generally exhibit similar eco-physiological responses.
 The output of the gradient forest model is further grouped by broad functional types to facilitate the identification of shifts and trends.
-The phytoplankton genera that can be best predicted are mostly diatoms, but also some haptophytes, dinoflagellates, and cyanobacteria (see Figure \ref{fig:GF} $b$).
-Within these groups, we see differing threshold responses in cumulative importance along the predictor ranges.
-
-%AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
-The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
-% COPIED TO INTRO
-Although regional effecs may vary, a positive phase in the AMO is generally associated with periods of warmer temperatures and a shift of the ITCZ from north to south.
-This shift of the ITCZ southward increases wind-driven upwelling in the Cariaco basin \cite{taylor_ecosystem_2012}.
-The threshold responses between functional groups, which relate to a shift in the cell count abundances, show contrasting patterns.
-Diatoms and cyanobacteria show strong response thresholds for negative AMO anomalies.
-This is an interesting threshold response shown for the AMO, which, if sufficiently negative, seems to correlate to a reorganization of the physical regime in the Cariaco basin.
-The abundance of haptophytes shows a strong threshold at positive AMO anomalies, which could be related to the pattern observed from HPLC data of haptophytes mostly exhibiting blooms above the mixed layer and at a lower frequency from other functional groups \cite{pinckney_phytoplankton_2015}.
+The phytoplankton genera that can be best predicted are mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (see Figure \ref{fig:GF} $b$).
+Within these groups, we observe differing threshold responses in cumulative importance along the predictor ranges.
+
+% P5: Climate impacts
+Although regional effects may vary, a positive phase in the AMO is generally associated with periods of warmer temperatures and a shift of the ITCZ from north to south.
+This shift of the ITCZ southward increases wind-driven upwelling in the Cariaco Basin \cite{taylor_ecosystem_2012}.
+The threshold responses between functional groups, which relate to shifts in cell count abundances, show contrasting patterns.
+Diatoms and cyanobacteria exhibit strong response thresholds for negative AMO anomalies.
+This interesting threshold response observed for the AMO, which, if sufficiently negative, seems to correlate with a reorganization of the physical regime in the Cariaco Basin.
+The abundance of haptophytes shows a strong threshold at positive AMO anomalies, which could be related to the pattern observed from HPLC data that haptophytes predominantly exhibit blooms above the mixed layer and at a lower frequency than other functional groups \cite{pinckney_phytoplankton_2015}.
 
 % Move the general point on ENSO to intro! Only Discussion-relevant part here..
-%copied to Intro: The El Niño Southern Osciallation (ENSO) has a strong teleconnection to the Caribbean Sea, since during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
-\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measuerments in the Cariaco basin, and only \qtyrange{24}{36}{\%} explanantion of variance with a time lag of 12 months using linear regression.
-In contrast, we found a relatively strong predictive power for changes in the phytoplankton community with a 4-month time lag.
-We also tested a 12 month lag and found no significant improvement.
-The threshold responses show a shift in the abundances of dinoflagellates and haptophytes for a negative MEI v.2 index, which indicates La Niña conditions.
-For cyanobacteria and haptophytes, we found a pronounced threshold response for positive anomalies showing a sensitivity to ENSO events.
-Diatoms do not show a clear threshold response for the MEI v.2 index, but are affected across the entire range.
-Positive anomalies of the MEI v.2, representing strong ENSO events, are interestingly only found in Cluster 1 (see Figure \ref{fig:clustering} $e$).
-
-The in-situ measurements that best predict changes in the microphytoplankton community were nitrate concentration, SST and salinity, in that order.
-Nitrate is a central nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}.
-For dinoflagellates and diatoms, we see a strong threshold response around \qty{7}{\micro \mole}, which would indicate that strong upwelling conditions drive blooms of these functional groups.
+The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea.
+During ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
+\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance with a time lag of 12 months using linear regression.
+In contrast, we found relatively strong predictive power for changes in the phytoplankton community with a 4-month time lag.
+We also tested a 12-month lag and found no significant improvement.
+The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for a negative MEI v.2 index, which suggests La Niña conditions.
+For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events.
+Diatoms do not exhibit a clear threshold response for the MEI v.2 index but are affected across the entire range.
+Interestingly, positive anomalies of the MEI v.2, representing strong ENSO events, are found mostly in Cluster 1 (see Figure \ref{fig:clustering} $e$).
+
+The in-situ measurements that best predict changes in the microphytoplankton community were SST, nitrate concentration, and salinity, in that order.
+Nitrate is a crucial nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}.
+For dinoflagellates and diatoms, we observe a strong threshold response around \qty{7}{\micro \mole}, which indicates that strong upwelling conditions drive blooms of these functional groups.
 Other functional groups, including \textit{Mesodinium} and \textit{Eutreptria} also show a threshold response at even higher nitrate concentrations (around \qty{9}{\micro \mole}), indicating an effect on their abundance from very strong upwelling.
@@
 
-Gradient forest analysis clearly identified the AMO and MEI v.2 climate indices as the best predictors of changes in the microphytoplankton community in the Cariaco basin.
+Gradient forest analysis clearly identified the AMO and MEI v.2 climate indices as the best predictors of changes in the microphytoplankton community in the Cariaco Basin.
 This is an interesting result, as one would expect that more direct measures of ecosystem variability could better predict shifts in abundances.
-One possible explanation is that in situ measurements contain much stronger variability due to measurement error, local and temporal variability, and generally only provide a single snapshot for an entire month.
+One possible explanation is that in situ measurements contain significantly stronger variability due to measurement error, as well as local and temporal variability, and generally provide only a single snapshot for an entire month.
 Climate indices are detrended and integrated measurements of long-term changes that are matching in scale to the dynamics of long-term changes within a phytoplankton community.
@@
 
+\subsection{Limitations of microscopy cell counts and gradient forest}
+% P6: DISCUSS LIMITATIONS OF STUDY!!
 % DEFINITELY NEED TO DISCUSS PROBLEMS OF CORRELATION WITH SUCH TIME SERIES DATA
@@
 
+\section{Conclusions}
+- Our analysis shows that...
+
+- We also show that...
+
+- But we cannot confirm or test. so we would recommend X for the future.
+
+% Highlight that the climate indices explain changes in the community and abundance, but not necessarily changed in diversity (as there is no sig diff beteween clusters).
+Probably need more data, other approaches to untangle effects on diversity.
 In this study, we have found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco basin.
@@
 We hope that the work hints at the resilient nature of phytoplankton communities under global change, which should be further investigated by studying long-term ocean time series.
-
-% Highlight that the climate indices explain changes in the community and abundance, but not necessarily changed in diversity (as there is no sig diff beteween clusters).
-Probably need more data, other approaches to untangle effects on diversity.
-
-\section{Conclusions}
-% Theoretically the final paragraph above represents the conclusion
-
-- Our analysis shows that...
-
-- We also show that...
-
-- But we cannot confirm or test. so we would recommend X for the future.
 

### 2025-06-04 text-backup.tex
Finalizing Discussion 85%

@@
 Understanding the regional dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems driven by changes in climate.
+
+    %%%%%%%DISCUSSION BACKUP REWRITE MAY "="%%%%%%%%
+
+\section{Discussion}
+
+This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is in turn influenced by large-scale climatic changes.
+In the Cariaco Basin, decadal-scale climate oscillations have impacted the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
+% - a few sentences on basic seasonal dynamics removed here
+These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}).
+After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding to a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014.
+% - a few sentences about shift to smaller taxa and collapse in Sardine Fisheries
+
+% - "We found that including the time series data until the end of 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}."
+In 2014, physical conditions in the Cariaco Basin returned to a regime of higher wind speeds, increased upwelling, and elevated nutrient concentrations (see Figure \ref{fig:zscore}).
+Concurrently, there was an increase in chlorophyll \textit{a} biomass (see Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$).
+This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine captures observed in 2016 and 2017 in a location southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.
+
+When comparing the phytoplankton community between the two clusters, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a} biomass measured by fluorometry, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}).
+When including HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement here actually shows an increase between the periods 1996-2000 and 2006-2010 \cite{pinckney_phytoplankton_2015}, particularly at depths below \qty{55}{meters}.
+It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006.
+\citeA{pinckney_phytoplankton_2015} could not find a consistent method for quantifying total chlorophyll \textit{a} across different laboratories.
+Here, we focus on the more consistently sampled fluorometric measurements.
+Independent spectrophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Margarita Island follow the temporal trend of a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
+It should be noted that changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a} but also other related forms of chlorophyll, which can vary depending on the composition of the community \cite{welschmeyer_fluorometric_1994}.
+
+Despite some difficulties in reconciling HPLC pigment data and microscopy cell counts, both methods indicate a reduction in diatom abundance after 2004.
+Cell counts of other functional groups decreased, while HPLC data indicate an increase and redistribution of biomass to depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}.
+The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community transitioned to smaller individuals that were not captured under the microscope \cite{muller-karger_scientific_2019}.
+This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}.
+Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but are particularly prevalent when (as observed in the Cariaco Basin) temperatures increase and the nutrient supply decreases \cite{mousing_global_2014}.
+
+% check sizes of cells, then work this into a paragraph with another sentence!!
+If we compare the two parts of the first cluster, the community reorganized, particularly as diatoms shifted from a community dominated by centric diatoms (Guinardia, Leptocylindrus, Skeletonema) to one dominated by pennate diatoms (Pseudonitzschia) \cite{pinckney_phytoplankton_2015}.
+
+However, the shift in yearly mean genus richness does not correspond with the abundances in the cell counts.
+Instead, we observe a marked reduction following 1998 until 2010, after which richness recovers four years prior to the community seemingly returning to previous levels in 2014.
+The same trend is evident, albeit to a lesser degree, in the other diversity metrics.
+Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and exhibit a reduced yearly mean between 2000 and 2006, during which cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance.
+%
+
+The period between 2001 and 2006 lacks coverage in HPLC data, necessitating reliance on microscopy counts to deduce changes in the phytoplankton community.
+This limitation restricts the observable phytoplankton size range and the measurable diversity through the sampling methodology \cite{cermeno_sampling_2014}.
+We observe a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly, an increase a few years prior to the return in abundance to previous levels.
+This pattern results in no significant differences in diversity observed from microscopy cell counts between the two clusters, either by Shannon index or by genus richness (see Figure \ref{fig:zscore} and \ref{fig:divts}).
+We observe a significant difference in the Pilou index between the two clusters.
+Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}.
+This increase in evenness, as quantified by Pilou's index, is most likely explained by the lack of strong upwelling events and the thereby reduced disturbance of the community.
+The resulting reduced compositional turnover is also observed in a decrease in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$).
+As the productivity and biomass of phytoplankton communities are mainly driven by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
+
+% NOW let's jump to Gradient Forest and discussing CLIMATE impacts! for a final paragraph, and then we are done here!!
+To identify the variables that can best predict changes in the phytoplankton community, we conducted a gradient forest analysis using microscopy cell counts at the genus level as predicted variables and climate indices, along with in situ measurements of bottom-up drivers, as predictors.
+Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series.
+Interestingly, our analysis revealed that the strongest predictors of changes within the community are not in situ measurements, but rather the AMO and the MEI v.2 climate indices.
+These are followed in importance by nitrate concentration and SST (see Figure \ref{fig:GF}).
+
+These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, since the scaling axis that separates the two yearly aggregate community clusters is NMDS 1, which covaries most strongly with changes in the AMO, MEI v.2 (see Figure \ref{fig:clustering} $c$).
+Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS2.
+This indicates that the climate indices capture some variability, which is not represented in the more direct measurements of upwelling intensity.
+Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling, but also depend on precipitation and fresh water input through surface runoff, show greater variance along NMDS1.
+Both salinity and silicate concentration show some explanatory power in the gradient forest analysis, although they are weaker predictors than temperature, nitrate concentration, and the climate indices.
+
+Using microscopy cell counts as a basis for this analysis meant that the results should be interpreted in terms of changes in the microphytoplankton community, due to the size limitation of identifying smaller species via light microscopy.
+Grouping cell counts by genera does create a loss of resolution, but is based on the assumption that species within a genus generally show similar ecophysiological responses.
+The output of the gradient forest model is further grouped by broad functional types to facilitate the identification of shifts and trends.
+The phytoplankton genera that can be best predicted are mostly diatoms, but also some haptophytes, dinoflagellates, and cyanobacteria (see Figure \ref{fig:GF} $b$).
+Within these groups, we see differing threshold responses in cumulative importance along the predictor ranges.
+
+%AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
+The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
+% COPIED TO INTRO
+Although regional effecs may vary, a positive phase in the AMO is generally associated with periods of warmer temperatures and a shift of the ITCZ from north to south.
+This shift of the ITCZ southward increases wind-driven upwelling in the Cariaco basin \cite{taylor_ecosystem_2012}.
+The threshold responses between functional groups, which relate to a shift in the cell count abundances, show contrasting patterns.
+Diatoms and cyanobacteria show strong response thresholds for negative AMO anomalies.
+This is an interesting threshold response shown for the AMO, which, if sufficiently negative, seems to correlate to a reorganization of the physical regime in the Cariaco basin.
+The abundance of haptophytes shows a strong threshold at positive AMO anomalies, which could be related to the pattern observed from HPLC data of haptophytes mostly exhibiting blooms above the mixed layer and at a lower frequency from other functional groups \cite{pinckney_phytoplankton_2015}.
+
+% Move the general point on ENSO to intro! Only Discussion-relevant part here..
+%copied to Intro: The El Niño Southern Osciallation (ENSO) has a strong teleconnection to the Caribbean Sea, since during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
+\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measuerments in the Cariaco basin, and only \qtyrange{24}{36}{\%} explanantion of variance with a time lag of 12 months using linear regression.
+In contrast, we found a relatively strong predictive power for changes in the phytoplankton community with a 4-month time lag.
+We also tested a 12 month lag and found no significant improvement.
+The threshold responses show a shift in the abundances of dinoflagellates and haptophytes for a negative MEI v.2 index, which indicates La Niña conditions.
+For cyanobacteria and haptophytes, we found a pronounced threshold response for positive anomalies showing a sensitivity to ENSO events.
+Diatoms do not show a clear threshold response for the MEI v.2 index, but are affected across the entire range.
+Positive anomalies of the MEI v.2, representing strong ENSO events, are interestingly only found in Cluster 1 (see Figure \ref{fig:clustering} $e$).
+
+The in-situ measurements that best predict changes in the microphytoplankton community were nitrate concentration, SST and salinity, in that order.
+Nitrate is a central nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}.
+For dinoflagellates and diatoms, we see a strong threshold response around \qty{7}{\micro \mole}, which would indicate that strong upwelling conditions drive blooms of these functional groups.
+Other functional groups, including \textit{Mesodinium} and \textit{Eutreptria} also show a threshold response at even higher nitrate concentrations (around \qty{9}{\micro \mole}), indicating an effect on their abundance from very strong upwelling.
+Similarly, for temperature, dinoflagellates and diatoms exhibit a threshold around \qtyrange{21}{23}{\celsius}, which indicates higher abundances during low temperatures in the upper \qty{100}{m} driven by strong upwelling.
+For the other predictors, we do not observe strong threshold responses.
+
+Gradient forest analysis clearly identified the AMO and MEI v.2 climate indices as the best predictors of changes in the microphytoplankton community in the Cariaco basin.
+This is an interesting result, as one would expect that more direct measures of ecosystem variability could better predict shifts in abundances.
+One possible explanation is that in situ measurements contain much stronger variability due to measurement error, local and temporal variability, and generally only provide a single snapshot for an entire month.
+Climate indices are detrended and integrated measurements of long-term changes that are matching in scale to the dynamics of long-term changes within a phytoplankton community.
+However, we do not see such strong predictive power in the ERA5 climate reanalysis-derived mean wind speed, which also aggregates measurements and model output over an entire month.
+Another explanation could be that the climate indices capture a dimension of variability not found in our in-situ measurements, which is also represented in our NMDS clustering (see Figure \ref{fig:clustering} $c$), where both MEI v.2 and AMO scale along the second axis (NMDS 2).
+%TODO test more of my variables along these axes, then add to discussion here and maybe put plot in supplemental material.
+
+% DEFINITELY NEED TO DISCUSS PROBLEMS OF CORRELATION WITH SUCH TIME SERIES DATA
+
+% FROM FRAKER ET AL. 2022
+%"Note that although this analysis identifies predictive relationships between drivers and the biological indicators, these relationships do not necessarily imply mechanistic relationships.
+As such, this analysis represents the first stages of an overall analysis (a narrowing of the potential key relationships and when they matter from a much larger dataset), and additional statistical analyses, modeling, and experimental work may be required to develop a full mechanistic understanding of driver-response relationships and ecosystem change."
+
+- alongside a global reduction in frequency compared to the linear trend between 2006 and 2012 \cite{dai_coastal_2023}.
+% <- for this paper, check references to ENSO MEI v.2!!!
+This could be interesting!
+
+Our limitation to bottom-up drivers of community change could also limit the explanatory power of our analysis, as top-down processes such as zooplankton grazing strongly affect the structure of the phytoplankton community \cite{banas_adding_2011, acevedo-trejos_mechanisms_2015}.
+In the Cariaco basin in particular, a trophic cascade from the collapse of sardine fisheries through an increasing abundance of mesozooplaknton has been proposed to have affected the phytoplankton community \cite{muller-karger_scientific_2019}.
+Due to the lack of temporal resolution for higher trophic levels and the lack of zooplankton data for the first 6 years of the time series, we propose that mechanistic modeling studies could be well suited to investigate this aspect of the CARIACO time series further.
+
+In this study, we have found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco basin.
+Despite multiannual trends in warming, reduced productivity and a collapse in fisheries and phytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions returned.
+We see these patterns clearly in the community data retrieved from microscopy, which shows the value of the arduous work of identifying and counting phytoplankton cells.
+The strongest bottom-up variable predicting changes in the phytoplankton community is the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local ecosystems through low-frequency natural cycles.
+We hope that the work hints at the resilient nature of phytoplankton communities under global change, which should be further investigated by studying long-term ocean time series.
+
+% Highlight that the climate indices explain changes in the community and abundance, but not necessarily changed in diversity (as there is no sig diff beteween clusters).
+Probably need more data, other approaches to untangle effects on diversity.

### 2025-06-04 agujournaltemplate.tex
Discussion 95 % done

@@
 We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster respectively.
-The diatom Thalassiosira dominated the early period and was abundant in all other periods.
-Pseudo-nitzschia dominates cluster 2, where the haptophyte Emiliana was the second most abundant genus.
-The later part of cluster 1 is dominated by Emiliana.
+The diatom \textit{Thalassiosira} dominated the early period and was abundant in all other periods.
+\textit{Pseudo-nitzschia} dominates cluster 2, where the haptophyte \textit{Emiliana} was the second most abundant genus.
+The later part of cluster 1 is dominated by \textit{Emiliana}.
 
@@
 
+% Subhendu's Comment:
 % Clearly and briefly restating main findings 
@@
 % P4: Introduce GF Analysis and discuss results
-To identify the variables that can best predict changes in the phytoplankton community, we conducted a gradient forest analysis using microscopy cell counts at the genus level as predicted variables and climate indices, along with in situ measurements of bottom-up drivers, as predictors.
-Gradient forest is an extension of the random forest algorithm, designed to identify thresholds and patterns in species abundance or composition along environmental gradients \cite{ellis_gradient_2012}.
+To identify the variables that best predict changes in the phytoplankton community, we conducted a gradient forest analysis using time series of climate indices and in situ measurements to predict genus-level abundances from microscopy cell counts.
+Gradient forest is an extension of the random forest algorithm, designed to identify thresholds and patterns in community composition along environmental gradients \cite{ellis_gradient_2012}.
 Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series.
@@
 
-These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separates the two yearly aggregate community clusters is NMDS 1, which co-varies most strongly with changes in the AMO, MEI v.2, salinity, evaporation, and precipitation (see Figure \ref{fig:clustering} $c$).
-Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS 2.
-This indicates that the climate indices capture some variability that is not represented in the more direct measurements of upwelling intensity.
-Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, show greater variance along NMDS 1.
-Salinity, evaporation, and silicate concentration exhibit some explanatory power in the gradient forest analysis, although they are weaker predictors than SST, nitrate concentration, and the climate indices.
-
-Using microscopy cell counts as a basis for this analysis means that the results should be interpreted in terms of changes in the microphytoplankton community, due to the size limitations of identifying species via light microscopy.
-Grouping cell counts by genera does create a loss of resolution but is based on the assumption that species within a genus generally exhibit similar eco-physiological responses.
-The output of the gradient forest model is further grouped by broad functional types to facilitate the identification of shifts and trends.
-The phytoplankton genera that can be best predicted are mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (see Figure \ref{fig:GF} $b$).
-Within these groups, we observe differing threshold responses in cumulative importance along the predictor ranges.
-
 % P5: Climate impacts
-Although regional effects may vary, a positive phase in the AMO is generally associated with periods of warmer temperatures and a shift of the ITCZ from north to south.
-This shift of the ITCZ southward increases wind-driven upwelling in the Cariaco Basin \cite{taylor_ecosystem_2012}.
+Although regional effects may vary, a negative phase in the AMO is generally associated with periods of colder temperatures and a southward shift of the ITCZ \cite{knight_climate_2006, colna2017latitudinal}.
+In the Cariaco Basin, a southward shift of the ITCZ correlates with increased wind-driven upwelling \cite{taylor_ecosystem_2012}.
 The threshold responses between functional groups, which relate to shifts in cell count abundances, show contrasting patterns.
-Diatoms and cyanobacteria exhibit strong response thresholds for negative AMO anomalies.
-This interesting threshold response observed for the AMO, which, if sufficiently negative, seems to correlate with a reorganization of the physical regime in the Cariaco Basin.
-The abundance of haptophytes shows a strong threshold at positive AMO anomalies, which could be related to the pattern observed from HPLC data that haptophytes predominantly exhibit blooms above the mixed layer and at a lower frequency than other functional groups \cite{pinckney_phytoplankton_2015}.
-
-% Move the general point on ENSO to intro! Only Discussion-relevant part here..
+Diatoms, dinoflagellates, and haptophytes exhibit strong response thresholds for negative AMO anomalies, which could point to the effects of the resulting stronger upwelling conditions.
+The abundance of haptophytes additionally shows a threshold at positive AMO anomalies, which may be related to the pattern observed from HPLC data that haptophytes predominantly exhibit blooms above the mixed layer and at a lower frequency than other functional groups \cite{pinckney_phytoplankton_2015}.
+
 The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea.
 During ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
-\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance with a time lag of 12 months using linear regression.
-In contrast, we found relatively strong predictive power for changes in the phytoplankton community with a 4-month time lag.
-We also tested a 12-month lag and found no significant improvement.
-The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for a negative MEI v.2 index, which suggests La Niña conditions.
+\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months.
+In contrast, we found relatively strong predictive power for changes in the phytoplankton community for the MEI v.2 index with a 4-month time lag.
+We also tested a 12-month lag and found no significant improvement (see Table \ref{tab:sup:GFlagtests} in the appendix).
+The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for negative MEI v.2 index anomalies, which represent La Niña conditions.
 For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events.
 Diatoms do not exhibit a clear threshold response for the MEI v.2 index but are affected across the entire range.
-Interestingly, positive anomalies of the MEI v.2, representing strong ENSO events, are found mostly in Cluster 1 (see Figure \ref{fig:clustering} $e$).
+Interestingly, positive anomalies of the MEI v.2, representing strong ENSO events, are mostly found in Cluster 1 (see Figure \ref{fig:clustering} $e$).
 
@@
 Nitrate is a crucial nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}.
-For dinoflagellates and diatoms, we observe a strong threshold response around \qty{7}{\micro \mole}, which indicates that strong upwelling conditions drive blooms of these functional groups.
-Other functional groups, including \textit{Mesodinium} and \textit{Eutreptria} also show a threshold response at even higher nitrate concentrations (around \qty{9}{\micro \mole}), indicating an effect on their abundance from very strong upwelling.
-Similarly, for temperature, dinoflagellates and diatoms exhibit a threshold around \qtyrange{21}{23}{\celsius}, which indicates higher abundances during low temperatures in the upper \qty{100}{m} driven by strong upwelling.
-For the other predictors, we do not observe strong threshold responses.
-
-Gradient forest analysis clearly identified the AMO and MEI v.2 climate indices as the best predictors of changes in the microphytoplankton community in the Cariaco Basin.
-This is an interesting result, as one would expect that more direct measures of ecosystem variability could better predict shifts in abundances.
-One possible explanation is that in situ measurements contain significantly stronger variability due to measurement error, as well as local and temporal variability, and generally provide only a single snapshot for an entire month.
-Climate indices are detrended and integrated measurements of long-term changes that are matching in scale to the dynamics of long-term changes within a phytoplankton community.
-However, we do not see such strong predictive power in the ERA5 climate reanalysis-derived mean wind speed, which also aggregates measurements and model output over an entire month.
-Another explanation could be that the climate indices capture a dimension of variability not found in our in-situ measurements, which is also represented in our NMDS clustering (see Figure \ref{fig:clustering} $c$), where both MEI v.2 and AMO scale along the second axis (NMDS 2).
-%TODO test more of my variables along these axes, then add to discussion here and maybe put plot in supplemental material.
+For diatoms, we observe a strong threshold response above \qty{7}{\micro \molar}, indicating that intense upwelling conditions drive blooms, resulting in predictable changes in cell counts.
+Similarly, for SST, haptophytes and diatoms exhibit a threshold around \qtyrange{22}{24}{\celsius}, indicating thresholds at low surface temperatures driven by stronger upwelling.
+
+These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separates the two yearly aggregate community clusters is NMDS 1, which covaries most strongly with changes in the AMO, MEI v.2, salinity, evaporation, and precipitation (see Figure \ref{fig:clustering} $c$).
+Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS 2.
+Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, show greater variance along NMDS 1.
+Salinity, evaporation, and silicate concentration exhibit some explanatory power in the gradient forest analysis, although they are weaker predictors than SST, nitrate concentration, and the climate indices (see Figure \ref{fig:GF}).
+This indicates that the climate indices capture variability not represented in the more direct measurements of upwelling intensity.
 
@@
 
+% 1.  Discuss Limitations of Light microscopy cell counts in CARIACO Time Series
+Using microscopy cell counts as a basis for this analysis means that the results should be interpreted in terms of changes in the microphytoplankton community, due to the size limitations of identifying species via light microscopy.
+Grouping cell counts by genus does create a loss of resolution but is based on the assumption that species within a genus generally exhibit similar eco-physiological responses.
+The output of the gradient forest model is further grouped by broad functional types to facilitate the identification of shifts and trends.
+The phytoplankton genera that can be best predicted are mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (see Figure \ref{fig:GF} $b$).
+This is very likely not a complete picture of the phytoplankton community -> compare to HPLC data.
+
+% 2. Discuss limitations of GF analysis
 % FROM FRAKER ET AL. 2022
@@
 As such, this analysis represents the first stages of an overall analysis (a narrowing of the potential key relationships and when they matter from a much larger dataset), and additional statistical analyses, modeling, and experimental work may be required to develop a full mechanistic understanding of driver-response relationships and ecosystem change."
-
-- alongside a global reduction in frequency compared to the linear trend between 2006 and 2012 \cite{dai_coastal_2023}.
-% <- for this paper, check references to ENSO MEI v.2!!!
-This could be interesting!
-
+Without adjusting for time lags, SST and nitrate concentration are the strongest predictors of community change (see Figure \ref{fig:sup:GFoutput_nolags} in the appendix).
+Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and the AMO index as the strongest predictors yield a relatively low value around \qty{0.03}{}.
+This is to be expected with the data we are working with, and in line with values reported in other studies utilizing the gradient forest algorithm \cite{}.
+
+- There is some correction for correlation between the predictors, and we check the time lags to test for impeding effects of autocorrelation, but still the model does not take the temporal aspect into account in building the random forest models underlying the gradient forest analysis.
+It can however detect non-linear responses between variables.
+
+% 3. Limitation to bottom-up drivers and climate indices
 Our limitation to bottom-up drivers of community change could also limit the explanatory power of our analysis, as top-down processes such as zooplankton grazing strongly affect the structure of the phytoplankton community \cite{banas_adding_2011, acevedo-trejos_mechanisms_2015}.
-In the Cariaco basin in particular, a trophic cascade from the collapse of sardine fisheries through an increasing abundance of mesozooplaknton has been proposed to have affected the phytoplankton community \cite{muller-karger_scientific_2019}.
-Due to the lack of temporal resolution for higher trophic levels and the lack of zooplankton data for the first 6 years of the time series, we propose that mechanistic modeling studies could be well suited to investigate this aspect of the CARIACO time series further.
+In the Cariaco Basin in particular, a trophic cascade from the collapse of sardine fisheries through an increasing abundance of mesozooplankton has been proposed to have affected the phytoplankton community \cite{muller-karger_scientific_2019}.
+Due to the lack of temporal resolution for higher trophic levels and the absence of zooplankton data for the first six years of the time series, we propose that mechanistic modeling studies could be well-suited to investigate this aspect of the CARIACO time series further.
 
@@
 
-% Highlight that the climate indices explain changes in the community and abundance, but not necessarily changed in diversity (as there is no sig diff beteween clusters).
-Probably need more data, other approaches to untangle effects on diversity.
-In this study, we have found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco basin.
-Despite multiannual trends in warming, reduced productivity and a collapse in fisheries and phytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions returned.
-We see these patterns clearly in the community data retrieved from microscopy, which shows the value of the arduous work of identifying and counting phytoplankton cells.
+In this study, we found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco Basin.
+
+Despite multiannual trends in warming, reduced productivity, and a collapse in fisheries and phytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions were restored.
+
+We observe these patterns clearly in the community data obtained from microscopy, which demonstrates the value of the rigorous work involved in identifying and counting phytoplankton cells.
+
 The strongest bottom-up variable predicting changes in the phytoplankton community is the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local ecosystems through low-frequency natural cycles.
+We can concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state, particularly for the phytoplankton community of the Cariaco Basin.
+
+The analysis could be extended by including top-down effects on the phytoplankton community, particularly zooplankton and fisheries data. 
 We hope that the work hints at the resilient nature of phytoplankton communities under global change, which should be further investigated by studying long-term ocean time series.

### 2025-06-05 agujournaltemplate.tex
First Full Draft done

@@
 
-% Subhendu's Comment:
-% Clearly and briefly restating main findings 
-% Interpret what these findings mean in the context of the study
-% Emphasize what is novel and important (without comparison)
-% Only in following paragraphs: Compare to existing literature, highlight agreements, contradictions, and new persepctives. Discuss why and explain reasons for either.
-
 \subsection{Community reorganization and biomass shifts}
@@
 The in-situ measurements that best predict changes in the microphytoplankton community were SST, nitrate concentration, and salinity, in that order.
-Nitrate is a crucial nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}.
+Nitrate is a crucial nutrient for phytoplankton growth and has been identified as the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}.
 For diatoms, we observe a strong threshold response above \qty{7}{\micro \molar}, indicating that intense upwelling conditions drive blooms, resulting in predictable changes in cell counts.
@@
 \subsection{Limitations of microscopy cell counts and gradient forest}
-% P6: DISCUSS LIMITATIONS OF STUDY!!
-% DEFINITELY NEED TO DISCUSS PROBLEMS OF CORRELATION WITH SUCH TIME SERIES DATA
+% P6: Discuss limitations of study and methodology
 
@@
 The phytoplankton genera that can be best predicted are mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (see Figure \ref{fig:GF} $b$).
-This is very likely not a complete picture of the phytoplankton community -> compare to HPLC data.
-
-% 2. Discuss limitations of GF analysis
-% FROM FRAKER ET AL. 2022
-%"Note that although this analysis identifies predictive relationships between drivers and the biological indicators, these relationships do not necessarily imply mechanistic relationships.
-As such, this analysis represents the first stages of an overall analysis (a narrowing of the potential key relationships and when they matter from a much larger dataset), and additional statistical analyses, modeling, and experimental work may be required to develop a full mechanistic understanding of driver-response relationships and ecosystem change."
+This is certainly not a complete picture of the phytoplankton community at the CARIACO time series station.
+The HPLC pigment data captures all size classes of phytoplankton, but as previously discussed, does not cover the entire time series and was analyzed by four different laboratories \cite{pinckney_phytoplankton_2015}.
+With the caveat that the microscopy counts capture microphytoplankton exclusively, the data can be brought into agreement.
+Where the cell counts collapsed, a shift to smaller species occurred, which also agrees with size spectra calculated from the HPLC pigment data \cite{lorenzoni_characterization_2015}.
+
+% 2. Discuss limitations of GF analysis & Limitation to bottom-up drivers and climate indices
+Given the limited scope of the microscopy genus counts, it is necessary to identify the limitations of the gradient forest algorithm employed to quantify predictors of community change.
+Many of the predictors in ecological time series and in this study are highly correlated (see Figure \ref{fig:sup:correlation} in the appendix).
+This can lead to the identification of a correlated predictor that has no direct effect on community changes.
+The gradient forest methodology addresses this by conditionally permuting subsets of the correlated predictors, which can mitigate some of this effect \cite{ellis_gradient_2012}.
+Additionally, random forest models do not consider temporal relationships between data points, which is why we included tests for the predictive relationships with time lags of our predictors.
 Without adjusting for time lags, SST and nitrate concentration are the strongest predictors of community change (see Figure \ref{fig:sup:GFoutput_nolags} in the appendix).
 Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and the AMO index as the strongest predictors yield a relatively low value around \qty{0.03}{}.
-This is to be expected with the data we are working with, and in line with values reported in other studies utilizing the gradient forest algorithm \cite{}.
-
-- There is some correction for correlation between the predictors, and we check the time lags to test for impeding effects of autocorrelation, but still the model does not take the temporal aspect into account in building the random forest models underlying the gradient forest analysis.
-It can however detect non-linear responses between variables.
-
-% 3. Limitation to bottom-up drivers and climate indices
-Our limitation to bottom-up drivers of community change could also limit the explanatory power of our analysis, as top-down processes such as zooplankton grazing strongly affect the structure of the phytoplankton community \cite{banas_adding_2011, acevedo-trejos_mechanisms_2015}.
-In the Cariaco Basin in particular, a trophic cascade from the collapse of sardine fisheries through an increasing abundance of mesozooplankton has been proposed to have affected the phytoplankton community \cite{muller-karger_scientific_2019}.
-Due to the lack of temporal resolution for higher trophic levels and the absence of zooplankton data for the first six years of the time series, we propose that mechanistic modeling studies could be well-suited to investigate this aspect of the CARIACO time series further.
+This is expected given the relatively noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the gradient forest algorithm \cite{pitcher_example_2012, roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}.
+It is important to consider that although predictive relationships between drivers and ecological predictors are identified, this does not necessarily imply a mechanistic relationship.
+The scope of this analysis is exploratory data analysis aimed at narrowing down potential key relationships.
+Additional statistical analyses, modeling, and experimental work are required to develop a clearer mechanistic understanding of the drivers of ecosystem change in the Cariaco Basin.
+
+Our limitation to bottom-up drivers of community change may also restrict the explanatory power of our analysis, as top-down processes such as zooplankton grazing significantly impact the structure of the phytoplankton community \cite{banas_adding_2011, acevedo-trejos_mechanisms_2015}.
+In the Cariaco Basin specifically, a trophic cascade stemming from the collapse of sardine fisheries coupled with an increasing abundance of mesozooplankton has been proposed to have influenced the phytoplankton community \cite{muller-karger_scientific_2019}.
+Due to the lack of temporal resolution for higher trophic levels and the absence of zooplankton data for the first six years of the time series, we suggest that mechanistic modeling studies could be well-suited to further investigate this aspect of the CARIACO time series.
 
 \section{Conclusions}
-- Our analysis shows that...
-
-- We also show that...
-
-- But we cannot confirm or test. so we would recommend X for the future.
-
 In this study, we found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco Basin.
-
-Despite multiannual trends in warming, reduced productivity, and a collapse in fisheries and phytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions were restored.
-
-We observe these patterns clearly in the community data obtained from microscopy, which demonstrates the value of the rigorous work involved in identifying and counting phytoplankton cells.
-
+Specifically, microscopy cell counts reveal a shift and return between two clusters, defined by high- and low-upwelling conditions that correspond to anomalies in the AMO index.
+Between the clusters, we observe no significant changes in diversity metrics, except for lower turnover and higher community evenness under low-upwelling conditions.
+Despite multiannual trends in warming, reduced productivity, and a collapse in fisheries and microphytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions were restored.
 The strongest bottom-up variable predicting changes in the phytoplankton community is the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local ecosystems through low-frequency natural cycles.
 We can concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state, particularly for the phytoplankton community of the Cariaco Basin.
-
 The analysis could be extended by including top-down effects on the phytoplankton community, particularly zooplankton and fisheries data. 

### 2025-06-25 agujournaltemplate.tex
post-Ago Review working at Methods

@@
 \begin{keypoints}
-\item The CARIACO time series phytoplankton community data reveals a shift and return between two clusters, defined by high- and low-upwelling conditions that correspond to anomalies in the AMO index.
-\index Between the clusters, we observe no significant changes in diversity metrics, except for lower turnover and higher community evenness under low-upwelling conditions.
-\item Gradient forest analysis identified the AMO climate index as the strongest predictor of community changes, highlighting the influence of large-scale climatic oscillations on the tropical coastal ecosystem.
+\item The CARIACO phytoplankton community time series reveals two clusters associated to high- and low-upwelling conditions and corresponding to anomalies in the AMO index. 
+\index No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low-upwelling conditions.
+\item Gradient forest analysis identifies the AMO climate index as the strongest predictor of community change, highlighting the influence of large-scale climatic oscillations on this tropical coastal ecosystem.
 \end{keypoints}
@@
 \begin{abstract}
-Phytoplankton play a crucial role in Earth's biogeochemical cycles.
-Climate drives changes in the physical environment of phytoplankton communities, which can, in turn, affect the productivity and diversity of the entire ecosystem.
-Long-term time series can capture changes in climate and the ecosystem, allowing for the untangling of effects acting on a local population.
-This study aims to describe and investigate the drivers of changes observed in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017.
-Monthly cruises sampled microscopy cell counts and biogeochemical properties, which we extend with data from AMO and MEI v.2 climate indices, as well as the ERA5 climate model reanalysis.
-We describe and relate changes in bottom-up drivers of the upwelling system with the time series of phytoplankton biomass and diversity.
-Using clustering algorithms, we investigate the structure in the community time series data, revealing two clusters that describe a shift from high to low upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014.
-Our results suggest that community diversity did not follow the same pattern but instead was diminished between 1998 and 2009.
-To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related metrics (SST and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors.
-Our study emphasizes the impact of decadal-scale climate oscillations, such as the AMO, on the phytoplankton community in a tropical coastal ecosystem through modulation of the upwelling system and highlights the resilience of ecosystems to recover from climate changes.
+Phytoplankton communities play a crucial role in oceans' biogeochemical cycles.
+Climate drives changes in the physical environment that phytoplankton communities experience and can therefore impact the productivity and diversity of marine ecosystems.
+Long-term time series can capture changes in climate and ecosystems, allowing us to unravel the effects on local communities.
+This study aims to describe and investigate the drivers of changes in  phytoplankton community data (biomass and diversity) collected in the Cariaco Basin, Venezuela, between 1995 and 2017.
+The monthly cruises sampled water for microscopy cell counts and determination of biogeochemical properties, which we integrated with data from AMO and MEI v.2 climate indices and ERA5 climate model reanalysis.
+Using clustering algorithms, we investigate the structure of the community data, revealing two temporal clusters associated with high and low upwelling conditions (from 1995 to 2004 and from 2005 to 2017, respectively) and showing a return to community biomass and physical conditions similar to those characterizing the period 1995-2004.
+However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009.
+To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ, short-term, upwelling-related variables (SST and nitrate concentration) and long-term, climate indices (AMO and MEI v.2) were the strongest predictors.
+Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of the ecosystem to recover from climate change.
 
@@
     Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
-The paraphyletic group consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}.
-The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
-    Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.
-Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed in the last 20 years, although regional effects are highly variable \cite{dai_coastal_2023}.
-    Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{boyd_biological_2016, henson_future_2021}. 
-    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
+This paraphyletic community consists of a diverse group of organisms  exhibiting highly variable eco-physiological properties \cite{appeltans_magnitude_2012}.
+The short generation time and high diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
+    Coastal margins, where phytoplankton typically thrive, constitute only \qty{\sim 10}{\%} of the global ocean surface  but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.
+
+    Data from remote sensing, suggest a global increase in the frequency and spatial extent of coastal phytoplankton blooms over the last 20 years, but regional effects are highly variable \cite{dai_coastal_2023} and difficult to capture.
+    Climate change is predicted to have a lasting impact on phytoplankton communities, including long-term shifts in community composition and alterations of diversity \cite{acevedo-trejos_glimpse_2014, boyd_biological_2016, henson_future_2021}.
+    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to short-term cyclic processes and to long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
 Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.
 
-    A tropical coastal ecosystem that has been studied intensively is the CARIACO time series, which has involved consistent sampling of physical and biogeochemical parameters in the Cariaco Basin from 1995 to 2017 \cite{muller-karger_scientific_2019}.
+    A tropical coastal location that has been studied intensively in the last decades is the CARIACO basin, where a wide range of physical and biogeochemical parameters have been consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}.
     % Describe CARIACO location
-    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}.
-The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}.
-The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}.
-Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
+    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world oceans \cite{edgcomb_accessing_2011}.
+The basin is a continental shelf depression comprising two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}.
+The water below the sill is poorly ventilated.
+This underwater topography in combination with a high surface productivity, associated to aerobic degradation of fast sinking organic matter, creates permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}.
+Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for understanding the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
+
     % Describe general seasonal cycle with functional groups etc.
-    A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward.
-A smaller wind-driven secondary upwelling often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
+    The Cariaco Basin is characterised by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward.
+A smaller wind-driven secondary upwelling event occurs often in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
 During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing.
-In the low-wind rainy season, the Cariaco Basin exhibits oligotrophic conditions, where the fraction of large phytoplankton cells is diminished, but nano- and picophytoplankton are present in considerable numbers throughout the year \cite{lorenzoni_characterization_2015}.
+In the low-wind, rainy season, the Cariaco Basin exhibits oligotrophic conditions and the fraction of large phytoplankton cells is reduced, albeit nano- and picophytoplankton cells are present in relatively high numbers throughout the year \cite{lorenzoni_characterization_2015}.
 
@@
     Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes.
-\citeA{taylor_ecosystem_2012} analyzed time series data up to 2010 and found significant decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors linked to the impact of global warming.
-During the initial phase of the time series (1996–2004), the Cariaco Basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
-Following 2004, there was a transition in the physical and biogeochemical regime characterized by a significant reduction in trade winds during the upwelling period \cite{taylor_ecosystem_2012}.
-This strongly affected the ecosystem and phytoplankton community.
-Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, as confirmed to a lesser extent by pigment data analysis \cite{pinckney_phytoplankton_2015}.
-In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{lorenzoni_characterization_2015}.
-An analysis of the pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability in community composition \cite{pinckney_phytoplankton_2015}.
-During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass measured in the Cariaco Basin \cite{taylor_ecosystem_2012}.
+\citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found significant decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to global warming.
+During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
+After 2004, a significant reduction in the strength of trade winds during the upwelling period \cite{taylor_ecosystem_2012} caused a regime change in the physical and biogeochemical system.
+This strongly affected the phytoplankton community and the whole ecosystem.
+Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms, as confirmed, albeit not with the same high significance, by analysis on pigment data  \cite{pinckney_phytoplankton_2015}.
+In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly occurring in stratified water columns \cite{lorenzoni_characterization_2015}.
+An analysis of pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.
+During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass \cite{taylor_ecosystem_2012}.
 Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.
@@
     % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
-    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift.
-Data from 2012 until the end of the CARIACO time series in 2017 indicate a return to stronger upwelling intensities and increased numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
-This return in 2014 to conditions similar to those of the late 1990s coincided with a recovery in sardine catches observed in 2016 and 2017 in a location southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.
-    The impact of large-scale climatic processes on the observed shifts in the community remains unclear.
-The Atlantic Multidecadal Oscillation (AMO) has emerged as an important indicator of the climate system and its effects on the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
+    The shift in phytoplankton community composition associated to increased temperature and to reduced upwelling intensities, does not appear to be a permanent change in regime.
+Data from 2012 to 2017, the end phase of the CARIACO time series, indicates a return to strong upwelling intensities and to high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
+The return to conditions similar to those of the late 1990s coincides with a recovery in sardine catches observed in 2016 and 2017 in an area southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.
+    The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear.
+The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
 The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
-    The El Niño Southern Oscillation (ENSO) has a teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
-However, \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin and did not test for correlations with the AMO index.
+    The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
+However, using the time-series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in-situ observations.
+In addition, the AMO index was included in the analyses of the Cariaco Basin.
 
     % What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
-    \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation."
-    Previous analyses of phytoplankton community diversity did not include time series data beyond 2011, and so far there was no detailed investigation of changes in diversity based on the microscopy cell count data.
-    From this background, we aim to 1) investigate changes in the depth distribution of phytoplankton biomass and community diversity over the entire time series, 2) relate the structure of the phytoplankton community data to the shifts observed in physical and biogeochemical variables, and 3) assess how decadal-scale climate osciallations as signified by the ENSO MEI v.2 and AMO index relate to changes in the phytoplankton community.
-    Our analysis focuses exclusively on bottom-up drivers of community changes, as these data are available throughout the entire time series and at a monthly resolution.
-This study complements the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017.
-Understanding the regional dynamics of the Cariaco Basin is crucial for comprehending the broader ecological changes in tropical marine ecosystems driven by shifts in climate.
+    \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change", and indicated that they may "simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation".
+    Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have been undertaken yet on possible changes in diversity based on microscopy cell counts.
+    Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period 1995\,--\,2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO)  and changes in the phytoplankton community.
+    Our analysis focuses exclusively on bottom-up drivers of community changes, because these data are available throughout the entire time series and at a monthly resolution.
+Our study builds on and expands substantially the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017.
+Understanding the regional dynamics of the Cariaco Basin is crucial for unveiling theecological changes affecting this tropical marine ecosystem and the variables driving these changes.
 
@@
     The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon.
-The Cariaco Basin is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north at a mean depth of around \qty{100}{m}.
-This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the accumulation of anoxic bottom waters below \qty{\sim 200}{m} depth.
+The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north, running almost parallel to the coastline, at a mean depth of around \qty{100}{m}.
+This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 200}{m} depth.
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
+\caption{Bathimetric map of the CARIACO Basin in the Caribbean Ocean, off the coast of Venezuela, indicating the CARIACO Mooring, the station of the Ocean Time-Series (located at \ang{10.50}N, \ang{64.67}W).
+A sill, running almost parallel to the coast at an average depth of 100\,m, from the Isla de Magarita and the Isla de la Tortuga and beyond, encloses the Cariaco basin from the North.}
+\label{fig:map}
+\end{figure}
 
 \subsection{In-Situ Data}
-    Samples were collected as part of the Cariaco Ocean Time Series Program during monthly research vessel cruises to a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
-    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this manuscript.
+    Samples were collected as part of the Cariaco Ocean Time Series Program by means of monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
+    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this work.
 Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela.
-Details of the history of sampling procedures in CARIACO, as well as sampling schemes, analytical methods, and data quality control and inter-calibration procedures, have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.
+Details about the sampling procedures adopted in CARIACO, sampling schemes, analytical methods, data quality control, and inter-calibration procedures, have been already described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.
 
@@
 
-    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to the present by Dr.
-K.
-Fanning (University of South Florida, St.
-Petersburg, FL).
+    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to present by Dr.
+Kent Fanning (University of South Florida, St.
+Petersburg, USA).
 Samples from eight cruises were analyzed by both laboratories.
-Data for $PO_4$ and $SiO_4$ were in strong agreement (within ±5\%), whereas the $NO_x$ and $NH_4$ measurements showed discrepancies.
+Data for PO$_4$ and SiO$_4$ were in strong agreement (within ±5\%), whereas  NO$_\text{x}$ and NH$_4$ measurements showed discrepancies.
 \cite{taylor_ecosystem_2012}.
-For our analysis, we used a merged dataset for all nutrients, where coincidental monthly measurements were averaged.
+We used a merged dataset for all nutrients, in which coincidental monthly measurements are averaged.
 
@@
 
-    Data sets for Niskin bottle samples, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047} (accessed in February, 2024).
+    Data collected with Niskin bottles, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}.
     From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (i.e.
 no more than three months) in coverage.
-Therefore, we did not include zooplankton net tow and HPLC pigment data in our analysis.
-Our study instead focused on the phytoplankton microscopy data and bottom-up drivers of change.
-We compared these data with Niskin bottle nutrient, temperature, and salinity data, as well as climate data and indices, focusing on bottom-up drivers of community variability and diversity.
-In total, we used a complete set of 254 monthly data points, on which all variables were available.
+In total, we used a complete set of 254 monthly data points, for which all variables were available.
 
@@
     Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}.
-We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling coordinates (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024).
-This variable is referred to as "wind speed" throughout the manuscript for simplicity.
-The data for precipitation and evaporation are similarly extracted from the ERA5 model reanalysis dataset for the aforementioned grid cell.
-Evaporation provides an estimate of the accumulated amount of water that has evaporated from the Earth's surface, while precipitation is the estimate of total precipitation, comprising all liquid that falls to Earth's surface as rain or snow.
-
-    The monthly AMO index data (10-year low-pass), which represents an area average of detrended low-pass filtered North Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} (accessed in June, 2024).
-
-    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024).
+We extracted the -u wind component at \qty{10}{\meter} from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
+For simplicity, we refer to this variable as "wind speed" throughout the article.
+The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell.
+Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate for total precipitation, comprising all water that falls on Earth's surface as rain or snow.
+
+    The monthly AMO index data (10-year low-pass), which represents an area-averaged, detrended low-pass filtered North Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.
+
+    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of surface wind, and outgoing long-wave radiation (OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/}.
 In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).
@@
     Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using tidyverse packages \cite{wickham_welcome_2019}.
-All data processing and analysis code is publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
+All the codes we produced for data processing and data analysis are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
     For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S.
@@
 \subsection{Data analysis}    
-    To provide a general overview of the data and trends in the time series, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$).
-Cruises from 1995, 2016, and 2017 were excluded due to insufficient measurements to provide a reliable yearly mean estimate.
+    To provide an overview of the data available  and the general trends shown by the data, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$).
+Cruises from 1995, 2016, and 2017 were excluded because insufficient coverage in these years prevented us from producing reliable yearly mean estimates.
 
@@
 
-    \subsubsection{Phytoplankton Community Structure Analysis}
-    To investigate the structure of the phytoplankton time series data, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth.
+    \subsubsection{Analysis of Phytoplankton Community Structure}
+    To investigate the structure of the phytoplankton community over time, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth.
 Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}.
@@
 To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations.
-The arrows indicate the gradient of a two-dimensional linear trend surface.
     To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}.
@@
 Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}.
-Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor.
+Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then it averages these to determine the overall importance of each predictor.
 The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover.
 Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}.
-This analysis was conducted using monthly measurements integrated to 100 m depths for observed genus counts, ensuring the inclusion of all available data.
-The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, wind speed, evaporation, precipitation, the \qty{21}{\celsius} isotherm, SST, and salinity, as well as the concentration of nutrients ($\mathrm{NO}_3$, $\mathrm{PO}_4$ and $\mathrm{SiO}_4$).
+This analysis was conducted using monthly measurements integrated over a depth of 100 m for observed genus counts, ensuring the inclusion of all available data.
+The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO]}_3$, $\mathrm{[PO]}_4$ and $\mathrm{[SiO]}_4$).
     The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}.
-    We show a model run without any time lags in the supplemental material (see Figure \ref{fig:sup:GFoutput_nolags}).
+    We show a model run without any time lags in  supplementary material (Figure \ref{fig:sup:GFoutput_nolags}).
 To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs.
-We chose time lags that scored highest in the goodness-of-fit statistic for the final model run (see Table \ref{app:table1} in the Appendix).
+We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{app:table1} in Appendix).
     % Note: Perhaps add some more info about Supplmentary Results and test runs here?
@@
 
-\caption{All data sources with corresponding units used as predictor variables for the gradient forest analysis.}
+\caption{Data and corresponding units used as predictor variables for the gradient forest analysis.}
 \centering
@@
 
-\section{Draft Results}
+\section{Results}
 
 % Figure 2
-The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}.
-We included marks to highlight those years with a significant number of missing monthly measurements (\textgreater 2 months and $\geq$ 4 months), as these make the annual means less reliable.
+Figure \ref{fig:zscore} shows the normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group.
+We included marks to highlight years with a significant number of missing monthly measurements ($=$ 2, 3 months and $\geq$ 4 months), since these make the annual means less reliable.
 Particularly toward the later years of the time series, the sampling became sparse.
 The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm).
-These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in the appendix).
+These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in Appendix).
 The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing inter-annual oscillations.
-Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration.
-Silicate does not follow a clear pattern but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to diminished diatom abundances.
+Annual mean anomalies of nitrate and phosphate appear to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration.
+Silicate does not follow a clear pattern but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to diminished diatom abundances.
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
+\caption{Time series showing annual z-scores of selected indicator and driver variables.
+Red indicates negative annual anomalies and blue indicates positive annual anomalies.
+Deeper shades indicate larger anomalies.
+All data are available at monthly resolution.
+The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to lack of data, additionally we marked years with two or three and four or more missing values to highlight unreliable annual mean estimates.
+Diversity Indices are calculated based on microscopy counts at the Genus level.}
+\label{fig:zscore}
+\end{figure}
 
 % Figure 3
-To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between the rainy and upwelling seasons.
-The bulk of the chlorophyll \textit{a} biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters.
-Chlorophyll \textit{a} exhibits marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$).
+Figure \ref{fig:divts} shows the depth distribution and dynamics of chlorophyll \textit{a} and community diversity indices, also in relation to rainy and upwelling seasons.
+The bulk of the chlorophyll \textit{a} biomass ($\sim$ 80\,\%) was measured in the upper 50 meters of the water column, with approximately 15\,\% measured between 50 and 75 meters.
+Chlorophyll \textit{a} exhibits marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, Figure \ref{fig:divts}$b$).
 The yearly mean chlorophyll \textit{a} biomass shows the highest variability within the upper 25 meters.
-In some years, particularly during 2000, 2005, 2007, 2008, and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$).
-In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period from 1996 to 2004 and only recover in 2012.
+In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range (Figure \ref{fig:divts}$a$).
+In general, yearly mean anomalies within the upper 75 meters appear to be high during the period from 1996 to 2004 and only reduces in 2012.
 The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter months (January and February) skew the value.
-Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014.
+Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\,\% of the mean over the entire time series for these months), which recovers in 2014.
 Chlorophyll \textit{a} biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series.
-In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters is markedly reduced, while the 50–75 meter depth interval sees a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
-The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll \textit{a} biomass in seasonality, with maximum values observed during the upwelling season in winter.
-Genus richness is consistently highest in the top \qty{25}{meters} depth interval, which also shows the most variability over the time series.
-Across the time series, we observe maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, which only recovers after 2011.
-The collapse in diversity following the start of the time series does not follow the pattern shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}).
+In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters is markedly reduced, while the 50–75 meter depth interval shows a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
+The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons in winter.
+Genus richness is consistently the highest diversity index in the top \qty{25}{meters} depth interval, and shows also the highest variability over the time series.
+Across the time series, we observe maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, with a recovery after 2011.
+The  reduction in diversity following the start of the time series does not follow the pattern shown by chlorophyll \textit{a}, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}).
 Scaling this genus diversity by the cell count abundances, i.e.
@@
 The seasonal signal is lost (see Figure \ref{fig:divts} $f$).
-Community evenness, as signified by the Pielou index, shows no seasonal signal and a very large variance within seasons and across depth (see Figure \ref{fig:divts} $h$).
-The inter-annual dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value.
+Community evenness, expressed by the Pielou index, shows no seasonal signal and a very large variance within seasons and through depth (see Figure \ref{fig:divts} $h$).
+The inter-annual dynamics of evenness are remarkably stable across all depth intervals, clustered around an intermediate 0.5 value.
 There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016.
 Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.
-
-% Figure 4
-Taking the microscopy community data interpolated within the top 100 meters, we aimed to identify any apparent structure within the time series that could be related to the physical environment.
-To achieve this, we employed clustering analysis on both the annual aggregate of cell counts from microscopy and the individual monthly measurements.
-Figure \ref{fig:clustering} $a$ presents the results of clustering the annual aggregate communities using a linear complete linkage algorithm based on the binary Jaccard matrix.
-The community is divided into two groups, with the first cluster encompassing the years 1996-2004 and 2014-2016, while the second cluster includes the intermediate years from 2004-2013.
-Inter-annual species turnover is illustrated in Figure \ref{fig:clustering} $b$, revealing an increasing trend in turnover until 2004.
-Turnover subsequently decreased during the period of cluster 2 but increased again in 2014, marking the return to cluster 1.
-This pattern is also reflected in the hierarchical clustering results in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years.
-We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster respectively.
-The diatom \textit{Thalassiosira} dominated the early period and was abundant in all other periods.
-\textit{Pseudo-nitzschia} dominates cluster 2, where the haptophyte \textit{Emiliana} was the second most abundant genus.
-The later part of cluster 1 is dominated by \textit{Emiliana}.
-
-To check whether the annual aggregation of community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$).
-This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends.
-The first axis separates the two clusters, with cluster 2 trending towards positive anomalies in the AMO and precipitation, while cluster 1 trends towards positive anomalies in evaporation, salinity, and the MEI v.2 index.
-The second NMDS axis appears to follow the trends in upwelling intensity, as indicated by the opposing trend arrows for nutrient concentrations, temperature, and wind speed.
-Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering.
-
-% Figure 5 - NEW
-Figure \ref{fig:clustcomp} $a$ to $o$ shows density distributions for environmental variables, climate, and diversity indices between the two clusters.
-Upwelling-related variables show distinct patterns between the two clusters.
-For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} isotherm, lower temperatures, reduced evaporation, and higher salinity and nutrient concentrations, particularly for nitrate.
-The AMO index indicates a separation between the two clusters, with cluster 1 exhibiting more negative anomalies, while cluster 2 has a higher occurrence of positive anomalies.
-Although there are no significant differences in genus richness and the Shannon index between clusters, there is a slightly increased median in the Pielou index for cluster 2.
-The density distribution reveals some differences.
-For cluster 1, there is higher variance in genus richness, while cluster 2 shows greater variance and a bimodal distribution for the Shannon and Pielou indices.
-
-% Figure 6
-To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
-Preliminary model runs were used to determine if time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months.
-In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity.
-The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentrations scored the lowest in predicting changes within the community.
-Gradient Forest predicts shifts in the community by building a random forest model for each observed genus across the time series.
-The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$.
-The microscopy data are biased toward larger cell sizes, and therefore, most of the predicted genus cell counts are from diatoms.
-However, two haptophyte and three dinoflagellate genera rank relatively high.
-Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range.
-The black line indicates cumulative importance across the entire community, and the colored dashed lines indicate importance per phytoplankton group.
-For the AMO index, we see a clear pattern in which negative anomalies affect the community.
-This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria.
-Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range where the community changed most, indicating these values as a threshold for community composition.
-See Figure \ref{fig:sup:GFoutput_lags_extra} in the appendix for response curves for all variables not shown here.
-For wind speed, SST, and salinity, changes occurred most at parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}) and high salinity (above \qty{36.8}{PSU}).
-
-\begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
-\caption{Map of CARIACO Ocean Time-Series location in the Cariaco basin in Venezuela in the Caribbean Ocean, located at \ang{10.50}N, \ang{64.67}W.}
-\label{fig:map}
-\end{figure}
-
-\begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
-\caption{Time series showing annual z-scores of selected indicator and driver variables.
-Red indicates negative annual anomalies and blue indicates positive annual anomalies for that year.
-Deeper shades indicate larger anomalies.
-All data was available at monthly resolution.
-The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to lack of data, additionally we marked years with more than two and four or more missing monthly values to highlight unreliable annual mean estimates.
-Diversity Indices are calculated based on microscopy counts at the Genus level.}
-\label{fig:zscore}
-\end{figure}
 
@@
 \end{center}
-\caption{Time series and seasonality of chlorophyll \textit{a} and diversity indices across four depth intervals.
+\caption{Time series and seasonality of chlorophyll \textit{a} and diversity indices at four depth intervals.
 (a,c,e,g) Large dots and lines mark annual mean across time series for years with more than 5 monthly sampling.
-Transparent dots show monthly raw values.
-(b,d,f,h) Boxplots of monthly measurements across four depth intervals grouped by season.
-Upwelling season lasts from December to May, Rainy Season from June to November.}
+Small, transparent dots show monthly raw values.
+(b,d,f,h) Boxplots of monthly measurements at four depth intervals grouped by season.
+Upwelling season goes from December to May, Rainy Season from June to November.}
 \label{fig:divts}
@@
 
+% Figure 4
+We undertook clustering analysis on annually aggregated cell counts from microscopy and on individual monthly measurements.
+Figure \ref{fig:clustering}$a$ shows the results of clustering the annually aggregated communities using a linear complete linkage algorithm based on the binary Jaccard matrix.
+The distinct clusters clearly emerge from these analysis, , the first cluster encompassing the years 1996-2004 and 2014-2016 and the second cluster including the intermediate years from 2004 to 2013.
+Inter-annual species turnover is shown in Figure \ref{fig:clustering}$b$, it reveals an increasing trend in turnover from 1996 to 2004.
+Turnover subsequently decreases during the period underpinning cluster 2 and increases again in 2014, marking a return to cluster 1.
+This pattern is also reflected in the hierarchical clustering results, Figure \ref{fig:clustering}$a$, where cluster 2 shows a lower distance in the clustering tree between the years.
+We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster, respectively.
+The diatom \textit{Thalassiosira} dominated the early period and was abundant in all other periods.
+\textit{Pseudo-nitzschia} dominates cluster 2, with the haptophyte \textit{Emiliana} being the second most abundant genus.
+The later part of cluster 1 is dominated by \textit{Emiliana}.
+
+To check whether the annual aggregation of community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (Figure \ref{fig:clustering}$c$).
+The NMDS clustering is overlaid with environmental data and climate indices in the attempt of unveiling some of the underlying trends.
+The horizontal axis separates the two clusters, with cluster 2 located over positive anomalies in AMO and precipitation and cluster 1 located over positive anomalies in evaporation, salinity, and MEI v.2.
+The second NMDS axis appears to follow the trends in upwelling intensity, as indicated by the arrows poinintg in opposite directions for nutrient concentrations, temperature, and wind speed.
+Overall, the NMDS clustering confirms the pattern shown by clustering on annually aggregated data.
+
 \begin{figure}
 \noindent\includegraphics[width=\textwidth]{fig/Figure4_ClusteringNMDS_v2.pdf}
-\caption{(a) Hierarchichal lustering by complete linkage method of microscopy cell counts at the Genus level, using the binary Jaccard distance.
-The community for each year is binned together, to smooth out seasonal aspects.
+\caption{(a) Hierarchical clustering by complete linkage method of microscopy cell counts at the Genus level, using the binary Jaccard distance.
+The community for each year is binned together, to smooth out seasonal features.
 (b) Species turnover between years for annual aggregate of identified phytoplankton to a genus level.
@@
 Stress: 0.261, 2 dimensions, converging solution.
-Color of dots based on which yearly cluster the monthly data is in.
-Vector overlay of ENVFIT shows how environmental variables are co-varying with the community data.
+The dots are color-coded based on the cluster they belong to.
+Vector overlay of ENVFIT shows how environmental variables are co-varying with community data.
 Length of vector indicates strength of relationship.}
@@
 
+% Figure 5 - NEW
+Figure \ref{fig:clustcomp} shows density distributions of environmental variables, chlorophyll \textit{a}, and diversity indices for the two clusters.
+Upwelling-related variables show distinct patterns between the two clusters.
+For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} isotherm, lower temperatures, reduced evaporation, and higher salinity and nutrient concentrations, particularly for nitrate.
+The AMO index shows a pronounced separation between the two clusters, with cluster 1 exhibiting more negative anomalies, while cluster 2 showing higher occurrence of positive anomalies.
+Although there are no significant differences in genus richness and the Shannon index between clusters, there is a slightly increased median in the Pielou index for cluster 2.
+The density distribution reveals some differences.
+Cluster 1 shows higher variance for genus richness, while cluster 2  shows greater variance and a bimodal distribution for Shannon and Pielou indices.
+
 \begin{figure}
 \noindent\includegraphics[width=\textwidth]{fig/Figure5_ClustCompPlot_v2.pdf}
-\caption{(Plots (a-o) below show density distributions of the environmental variables, chlorophyll \textit{a} and diversity indices between the clustered years.
-Dashed line marks the median.
+\caption{Density distributions of environmental variables, chlorophyll \textit{a}, and diversity indices for the two clusters.
+The dashed lines mark the median of each distribution.
 Significance levels of Wilcoxon sum rank tests between clusters are shown above plots  (p-value: ***\textless0.001, **\textless0.01, *\textless0.05, n.s.=not significant).}
@@
 
+% Figure 6
+To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
+Preliminary model runs were used to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months.
+Figure \ref{fig:GF}$a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity.
+The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentrations scored the lowest in predicting changes within the community.
+Gradient Forest predicts shifts in the community by building a random forest model for each observed genus across the time series.
+The predictive power of the models for each genus is shown in Figure \ref{fig:GF}$b$.
+The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are from diatoms.
+However, two haptophyte and three dinoflagellate genera rank relatively high.
+Figure \ref{fig:GF}$c$ shows the cumulative importance of the predictor variables across the parameter range.
+The black, dashed lines indicate cumulative importance across the entire community, and the solid, dashed lines indicate importance per phytoplankton group.
+For the AMO index, we see a clear pattern in which negative anomalies affect the community.
+This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria.
+Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range within which the community changed most, indicating these values as a threshold for community composition.
+Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows esponse curves for all variables not shown here.
+For wind speed, SST, and salinity, changes occurred mostly within parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}), and high salinity (above \qty{36.8}{PSU}).
+
+\begin{figure}
+\noindent\includegraphics[width=\textwidth]{fig/Figure6_GF_output_FINAL_v2.pdf}
+\caption{Gradient Forest analysis of monthly community data at a genus level.
+(a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances.
+(b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance.
+(c) Cumulative importance curves for the predictor variables.
+Black line shows cumulative importance across predictor range for the entire model and all genera.
+Dashed colored lines represent the responses of individual functional groups.}
+\label{fig:GF}
+\end{figure}
+
 \begin{table}
-
 \caption{Results of two-sided Wilcoxon sum rank test with continuity correction for variables between the two clusters.
-Cluster 1 is defined as the years 1996 - 2003 and 2014 - 2016.
-Cluster 2 includes the years 2004 - 2013.}
+Cluster 1 covers the periods 1996-2003 and 2014-2016 and cluster 2 covers the period 2004-2013.}
 \centering
@@
 
-\begin{figure}
-\noindent\includegraphics[width=\textwidth]{fig/Figure6_GF_output_FINAL_v2.pdf}
-\caption{Gradient Forest analysis of monthly community data at a genus level.
-(a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances.
-(b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance.
-(c) Cumulative importance curves for the predictor variables.
-Black line shows cumulative importance across predictor range for the entire model and all genera.
-Dashed colored lines represent the responses of individual functional groups.}
-\label{fig:GF}
-\end{figure}
-
 \section{Discussion}
@@
 % P1: Introductory paragraph highlighting the overall dynamics/shifts
-This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is in turn influenced by large-scale climatic changes.
-In the Cariaco Basin, decadal-scale climate oscillations have impacted the wind-driven upwelling regime, resulting in confounding effects on the composition and biodiversity of the phytoplankton community.
-These changes are well reflected in the yearly mean anomalies (see Figure \ref{fig:zscore}) and the density distribution of monthly measurements (see Figure \ref{fig:clustering}).
-After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and a collapse in functional group cell counts, all corresponding to a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014.
-We found that including the time series data until the end of 2017 in our analysis clearly shows that the observed shifts were reversible, confirming the observations by \citeA{muller-karger_scientific_2019}.
-In 2014, physical conditions in the Cariaco Basin returned to a regime of higher wind speeds, increased upwelling, and elevated nutrient concentrations (see Figure \ref{fig:zscore}).
-Concurrently, there was an increase in chlorophyll \textit{a} biomass (see Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$).
-This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine captures observed in 2016 and 2017 in a location southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.
+This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes.
+In the Cariaco Basin, decadal-scale climate oscillations impacted the wind-driven upwelling regime, producing confounding effects on the composition and biodiversity of the phytoplankton community.
+These changes are well visible in the yearly mean anomalies (Figure \ref{fig:zscore}) and the density distribution of monthly measurements (Figure \ref{fig:clustering}).
+After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident to a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies up to 2014.
+By considering the time series data until the end of 2017, the temporally complete dataset, we find that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}.
+In 2014, physical conditions in the Cariaco Basin returned to a regime of increased wind speed, increased upwelling, and elevated nutrient concentrations (Figure \ref{fig:zscore}).
+Concurrently, we observe an increase in chlorophyll \textit{a} (Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (Figure \ref{fig:clustering} $a$ and $c$).
 
 % P2: Discuss community reorganisation, biomass shifts vs HPLC data
-We found that the annual aggregates of community data showed two clusters, one including the time series up until 2004 and the end of the time series past 2013, and the other consisting of the intermediate time period between 2004 and 2013 (see Figure \ref{fig:clustering}).
-When comparing the phytoplankton community between the two clusters, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a} biomass measured by fluorometry, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}).
-The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community transitioned to smaller individuals that were not captured under the microscope \cite{muller-karger_scientific_2019}.
+The annual aggregates of community data showed two regimes, one covering the periods 1996 to 2004 and 2013-2017 and the other covering the period between 2004 and 2013 (see Figure \ref{fig:clustering}).
+When comparing the two regimes with respect to phytoplankton community data, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a}, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}).
+The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community might have transitioned to smaller cells, which sizes were probably below the detection limit of optical microscopes \cite{muller-karger_scientific_2019}.
 This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}.
-Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but are particularly prevalent when temperatures increase and the nutrient supply decreases \cite{mousing_global_2014}, as was observed in the Cariaco Basin.
+Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but shifts towards smaller cells are particularly prevalent when temperatures increase and nutrient supply decreases \cite{mousing_global_2014}, as we observed in the Cariaco Basin.
 
 % Disucssion of HPLC vs Microscopy!
-When comparing microscopy counts to previous analyses of HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement here actually shows an increase between the periods 1996-2000 and 2006-2010 \cite{pinckney_phytoplankton_2015}, particularly at depths below \qty{55}{meters}.
-Cell counts of all functional groups decreased, while HPLC data indicate an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}.
+In addition to our focus on the most temporally complete dataset, another important difference between our work and previous analyses of the Cariaco time series like, for example, the study of \citeauthor{pinckney_phytoplankton_2015}, is that, for unveiling changes in the community composition of phytoplankton, we concentrated on  microscopy counts, whereas  \citeauthor{pinckney_phytoplankton_2015} considered HPLC pigment data.
+The patterns we found in the microscopy data are less clear in the HPLC pigment data, which instead show an increase in chlorophyll \textit{a} between the periods 1996-2000 and 2006-2010, particularly at depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}.
+Cell counts of all functional groups decrease, while HPLC data in  \citeauthor{pinckney_phytoplankton_2015} indicate an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{meters} .
 It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006.
-\citeA{pinckney_phytoplankton_2015} could not find a consistent method for quantifying total chlorophyll \textit{a} across different laboratories.
-Here, we focus on the more consistently sampled fluorometric measurements.
-Independent spectrophotometric measurements of chlorophyll \textit{a} concentration at \qty{20}{meters} depth off the coast of Margarita Island follow the temporal trend of a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
-Changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a} but also other related forms of chlorophyll, which can vary depending on the composition of the community \cite{welschmeyer_fluorometric_1994}.
+Additionally, \citeauthor{pinckney_phytoplankton_2015} could not use a unique, shared method for quantifying chlorophyll \textit{a} across different laboratories.
+Here, we focus on consistent fluorometric measurements.
+Consistently with our results, independent spectrophotometric measurements of chlorophyll \textit{a} concentrations at \qty{20}{meters} depth, off the coast of Margarita Island, show a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
+Changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a} but also other related forms of chlorophyll, which can vary depending on the composition of the phytoplankton community \cite{welschmeyer_fluorometric_1994}.
 
 % P3: Shift in Diversity does not match shift in biomass/cell counts, why?
-However, the shift in yearly mean genus richness does not correspond with the abundances in the cell counts.
-Instead, we observe a marked reduction from 1998 until 2010, after which richness recovers four years prior to the community seemingly returning to previous levels in 2014.
+The shift we find in yearly mean genus richness, however, does not correspond with abundances in cell counts.
+Instead, we observe a marked reduction from 1998 to 2010, after which richness recovers four years prior to 2014, when the community returns, seemingly, to previous levels, those characteristics of regime 1.
 The same trend is evident, albeit to a lesser degree, in the other diversity metrics.
-Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and exhibit a reduced yearly mean between 2000 and 2006, during which cell counts recorded an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance.
-The period between 2001 and 2006 lacks coverage in HPLC data, necessitating reliance on microscopy counts to deduce changes in the phytoplankton community.
-This limitation restricts the observable phytoplankton size range and the measurable diversity through the sampling methodology \cite{cermeno_sampling_2014}.
-We observe a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly, an increase a few years prior to the return in abundance to previous levels.
-This pattern results in no significant differences in diversity observed from microscopy cell counts between the two clusters, either by Shannon index or by genus richness (see Figure \ref{fig:zscore} and \ref{fig:divts}).
-We observe a significant difference in the Pielou's index between the two clusters.
-Both HPLC pigment and microscopy measurements agree that the phytoplankton community became more evenly distributed during low-upwelling conditions of cluster 2 \cite{pinckney_phytoplankton_2015}.
-This increase in evenness, as quantified by Pielou's index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community.
-The resulting decrease in compositional turnover is also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$).
-As the productivity and biomass of phytoplankton communities are primarily driven by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
-
-\subsection{Drivers of change: Gradient forest analysis}
+Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and exhibit a reduced yearly mean between 2000 and 2006, during which cell counts recorded an increase in nanoflagellates and a pronounced peak in cyanobacteria.
+ We observe a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly an increase a few years prior to the return in abundance to previous levels.
+Microscopy cell counts do not show any statistical difference in diversity between the two  regimes, either by Shannon index or by genus richness (Figures \ref{fig:zscore} and \ref{fig:divts}).
+We observe a significant difference in the Pielou's index between the two regime.
+The HPLC pigment trend unveiled by and Our analysis on microscopy measurements suggesting that the phytoplankton community becomes more evenly distributed during low-upwelling conditions of regime 2 are consistent with the analysis of \citeauthor{pinckney_phytoplankton_2015} based on HPLC pigment data.
+This increase in evenness, quantified by the Pielou index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community.
+The resulting decrease in compositional turnover is also observed in a decline in interannual species turnover during regime 2 (see Figure \ref{fig:clustering}$b$).
+As the productivity and biomass of phytoplankton communities are driven primarily by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
+
+\subsection{Drivers of change}
 % P4: Introduce GF Analysis and discuss results
@@
 Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series.
-Interestingly, our analysis revealed that the strongest predictor of changes within the community was not an in situ measurement but rather the AMO climate index.
-The AMO is followed in importance by SST, the MEI v.2 index, and nitrate concentration (see Figure \ref{fig:GF}).
+Interestingly, our analysis revealed that the strongest predictor of change within the community was not an in situ measurement but rather the AMO climate index.
+The AMO is followed in importance by SST, MEI v.2 index, and nitrate concentration (see Figure \ref{fig:GF}).
 
@@
 The threshold responses between functional groups, which relate to shifts in cell count abundances, show contrasting patterns.
-Diatoms, dinoflagellates, and haptophytes exhibit strong response thresholds for negative AMO anomalies, which could point to the effects of the resulting stronger upwelling conditions.
-The abundance of haptophytes additionally shows a threshold at positive AMO anomalies, which may be related to the pattern observed from HPLC data that haptophytes predominantly exhibit blooms above the mixed layer and at a lower frequency than other functional groups \cite{pinckney_phytoplankton_2015}.
-
-The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea.
+Diatoms, dinoflagellates, and haptophytes exhibit strong response thresholds for negative AMO anomalies, which could point to influences of stronger upwelling conditions.
+The abundance of haptophytes shows, additionally, a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} using HPLC data of haptophytes predominantly exhibiting blooms above the mixed layer and at a lower frequency than other functional groups.
+
+The El Niño Southern Oscillation (ENSO) exhibits a strong influence on the Caribbean Sea.
 During ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
 \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months.
-In contrast, we found relatively strong predictive power for changes in the phytoplankton community for the MEI v.2 index with a 4-month time lag.
-We also tested a 12-month lag and found no significant improvement (see Table \ref{tab:sup:GFlagtests} in the appendix).
+In contrast, we found a relatively strong predictive power for changes in the phytoplankton community for the MEI v.2 index with a 4-month time lag.
+We also tested a 12-month lag and found no significant improvement (Table \ref{tab:sup:GFlagtests} in Appendix).
+
 The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for negative MEI v.2 index anomalies, which represent La Niña conditions.
@@
 Diatoms do not exhibit a clear threshold response for the MEI v.2 index but are affected across the entire range.
-Interestingly, positive anomalies of the MEI v.2, representing strong ENSO events, are mostly found in Cluster 1 (see Figure \ref{fig:clustering} $e$).
+Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in regime 1 (see Figure \ref{fig:clustering} $e$).
 
@@
 For diatoms, we observe a strong threshold response above \qty{7}{\micro \molar}, indicating that intense upwelling conditions drive blooms, resulting in predictable changes in cell counts.
-Similarly, for SST, haptophytes and diatoms exhibit a threshold around \qtyrange{22}{24}{\celsius}, indicating thresholds at low surface temperatures driven by stronger upwelling.
-
-These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separates the two yearly aggregate community clusters is NMDS 1, which covaries most strongly with changes in the AMO, MEI v.2, salinity, evaporation, and precipitation (see Figure \ref{fig:clustering} $c$).
+Similarly, haptophytes and diatoms exhibit a threshold around \qtyrange{22}{24}{\celsius}, indicating thresholds at low surface temperatures driven by stronger upwelling.
+
+These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separates the two regimes is NMDS 1, which covaries most strongly with changes in AMO, MEI v.2, salinity, evaporation, and precipitation (Figure \ref{fig:clustering} $c$).
 Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS 2.
 Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, show greater variance along NMDS 1.
-Salinity, evaporation, and silicate concentration exhibit some explanatory power in the gradient forest analysis, although they are weaker predictors than SST, nitrate concentration, and the climate indices (see Figure \ref{fig:GF}).
-This indicates that the climate indices capture variability not represented in the more direct measurements of upwelling intensity.
-
-\subsection{Limitations of microscopy cell counts and gradient forest}
+Salinity, evaporation, and silicate concentration exhibit some explanatory power in the gradient forest analysis, although they are weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}).
+This implies that the climate indices explain a variability that more direct measurements of upwelling intensity do not capture.
+
+\subsection{Limitations}
 % P6: Discuss limitations of study and methodology
@@
 % 1.  Discuss Limitations of Light microscopy cell counts in CARIACO Time Series
-Using microscopy cell counts as a basis for this analysis means that the results should be interpreted in terms of changes in the microphytoplankton community, due to the size limitations of identifying species via light microscopy.
-Grouping cell counts by genus does create a loss of resolution but is based on the assumption that species within a genus generally exhibit similar eco-physiological responses.
+Using microscopy cell counts as a basis for our analyses means that the results should be interpreted in terms of changes in the microphytoplankton community.
+Our choice of grouping cell counts by genus is based on the assumption that species within a genus generally exhibit similar eco-physiological responses.
 The output of the gradient forest model is further grouped by broad functional types to facilitate the identification of shifts and trends.
-The phytoplankton genera that can be best predicted are mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (see Figure \ref{fig:GF} $b$).
-This is certainly not a complete picture of the phytoplankton community at the CARIACO time series station.
-The HPLC pigment data captures all size classes of phytoplankton, but as previously discussed, does not cover the entire time series and was analyzed by four different laboratories \cite{pinckney_phytoplankton_2015}.
+The phytoplankton genera that can be best predicted are mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (Figure \ref{fig:GF} $b$).
+This does not represent a complete picture of the phytoplankton community of the CARIACO station.
+The HPLC pigment data used by \citeauthor{pinckney_phytoplankton_2015} capture all size classes of phytoplankton, but as previously discussed, does not cover the entire time series and the dataset was assembled by four different laboratories.
 With the caveat that the microscopy counts capture microphytoplankton exclusively, the data can be brought into agreement.
-Where the cell counts collapsed, a shift to smaller species occurred, which also agrees with size spectra calculated from the HPLC pigment data \cite{lorenzoni_characterization_2015}.
+Drastic declines in cell counts reflect shifts to smaller species, an aspect consistent with size spectra calculated by previous studies using HPLC pigment data \cite{lorenzoni_characterization_2015}.
 
 % 2. Discuss limitations of GF analysis & Limitation to bottom-up drivers and climate indices
-Given the limited scope of the microscopy genus counts, it is necessary to identify the limitations of the gradient forest algorithm employed to quantify predictors of community change.
-Many of the predictors in ecological time series and in this study are highly correlated (see Figure \ref{fig:sup:correlation} in the appendix).
+The restricted scope of the microscopy genus counts poses limitations to the gradient forest algorithm we employed to quantify predictors of community change.
+Many of the predictors in ecological time series and in this study are highly correlated (Figure \ref{fig:sup:correlation} in Appendix).
 This can lead to the identification of a correlated predictor that has no direct effect on community changes.
-The gradient forest methodology addresses this by conditionally permuting subsets of the correlated predictors, which can mitigate some of this effect \cite{ellis_gradient_2012}.
+However, the gradient forest methodology addresses this by conditionally permuting subsets of the correlated predictors, which can partly mitigate this effect \cite{ellis_gradient_2012}.
 Additionally, random forest models do not consider temporal relationships between data points, which is why we included tests for the predictive relationships with time lags of our predictors.
-Without adjusting for time lags, SST and nitrate concentration are the strongest predictors of community change (see Figure \ref{fig:sup:GFoutput_nolags} in the appendix).
-Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and the AMO index as the strongest predictors yield a relatively low value around \qty{0.03}{}.
+Without adjusting for time lags, SST and nitrate concentration are the strongest predictors of community change (Figure \ref{fig:sup:GFoutput_nolags} in Appendix).
+Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and AMO index as the strongest predictors yield a relatively low value around \qty{0.03}{}.
 This is expected given the relatively noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the gradient forest algorithm \cite{pitcher_example_2012, roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}.
 It is important to consider that although predictive relationships between drivers and ecological predictors are identified, this does not necessarily imply a mechanistic relationship.
-The scope of this analysis is exploratory data analysis aimed at narrowing down potential key relationships.
-Additional statistical analyses, modeling, and experimental work are required to develop a clearer mechanistic understanding of the drivers of ecosystem change in the Cariaco Basin.
-
-Our limitation to bottom-up drivers of community change may also restrict the explanatory power of our analysis, as top-down processes such as zooplankton grazing significantly impact the structure of the phytoplankton community \cite{banas_adding_2011, acevedo-trejos_mechanisms_2015}.
-In the Cariaco Basin specifically, a trophic cascade stemming from the collapse of sardine fisheries coupled with an increasing abundance of mesozooplankton has been proposed to have influenced the phytoplankton community \cite{muller-karger_scientific_2019}.
-Due to the lack of temporal resolution for higher trophic levels and the absence of zooplankton data for the first six years of the time series, we suggest that mechanistic modeling studies could be well-suited to further investigate this aspect of the CARIACO time series.
+The scope of our analysis, like any statistical analysis, is exploratory and aims at narrowing down potential key relationships.
+Mathematical modeling and experimental work are required to develop a clearer mechanistic understanding of the drivers of ecosystem change in the Cariaco Basin.
+
+Our focus on bottom-up drivers of community change may also restrict the explanatory power of our analysis, as top-down processes such as zooplankton grazing may impact on the structure of the phytoplankton community \cite{banas_adding_2011, acevedo-trejos_mechanisms_2015}.
+Due to the scarcity of data covering higher trophic levels and the absence of zooplankton data for the first six years of the time series, we suggest that mechanistic modeling is the ideal approach for investigating the relative impact of top-down and bottom-up drivers.
 
 \section{Conclusions}
-In this study, we found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco Basin.
-Specifically, microscopy cell counts reveal a shift and return between two clusters, defined by high- and low-upwelling conditions that correspond to anomalies in the AMO index.
-Between the clusters, we observe no significant changes in diversity metrics, except for lower turnover and higher community evenness under low-upwelling conditions.
-Despite multiannual trends in warming, reduced productivity, and a collapse in fisheries and microphytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions were restored.
-The strongest bottom-up variable predicting changes in the phytoplankton community is the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local ecosystems through low-frequency natural cycles.
-We can concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state, particularly for the phytoplankton community of the Cariaco Basin.
-The analysis could be extended by including top-down effects on the phytoplankton community, particularly zooplankton and fisheries data. 
-We hope that the work hints at the resilient nature of phytoplankton communities under global change, which should be further investigated by studying long-term ocean time series.
+We found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco Basin.
+Specifically, microscopy cell counts reveal the presence of two distinct regimes (and a return, during the final years, to regime 1) defined by high- and low-upwelling conditions that correspond to anomalies in the AMO index.
+We observe no significant changes in diversity metrics between the two regimes, except for lower turnover and higher community evenness under low-upwelling conditions.
+Despite multiannual trends in warming, reduced productivity, and drastic declines in microphytoplankton cell counts, the phytoplankton community returned to regime 1 levels once the climatic and local physical conditions were restored.
+The strongest bottom-up variable predicting changes in the phytoplankton community resulted the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local conditions through low-frequency natural cycles.
+We concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state for the phytoplankton community of the Cariaco Basin.
+Notwithstanding a number of limitations, the incipient return of the Cariaco basin to a pre-2004 regime hints at the resilient nature of phytoplankton communities under global change, an aspect that points to the relevance of long-term time series and deserves further attention.
+Future studies should focus on the relative impact of top-down and bottom-up drivers.
 

### 2025-06-26 agujournaltemplate.tex
Restored Ago's comments to go through them more carefully this time

@@
 Phytoplankton communities play a crucial role in oceans' biogeochemical cycles.
-Climate drives changes in the physical environment that phytoplankton communities experience and can therefore impact the productivity and diversity of marine ecosystems.
-Long-term time series can capture changes in climate and ecosystems, allowing us to unravel the effects on local communities.
+Climate drives changes in the physical environment that phytoplankton communities experience and can impact, thus, on the productivity and diversity of marine ecosystems.
+Long-term time series can capture changes in climate and ecosystems, allowing us to untangling of effects on local communities.
 This study aims to describe and investigate the drivers of changes in  phytoplankton community data (biomass and diversity) collected in the Cariaco Basin, Venezuela, between 1995 and 2017.
-The monthly cruises sampled water for microscopy cell counts and determination of biogeochemical properties, which we integrated with data from AMO and MEI v.2 climate indices and ERA5 climate model reanalysis.
-Using clustering algorithms, we investigate the structure of the community data, revealing two temporal clusters associated with high and low upwelling conditions (from 1995 to 2004 and from 2005 to 2017, respectively) and showing a return to community biomass and physical conditions similar to those characterizing the period 1995-2004.
+The monthly cruises sampled water for microscopy cell counts and determination of biogeochemical properties, which we integrated with data from AMO and MEI v.2 climate indices and ERA5 climate model reanalysis data.
+Using clustering algorithms, we investigate the structure of the community data, revealing two temporal clusters associated to high and low upwelling conditions (from 1995 to 2004 and from 2005 to 2017, respectively) and showing a return to community biomass and physical conditions similar to those characterizing the period 1995-2004.
 However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009.
@@
     Climate change is predicted to have a lasting impact on phytoplankton communities, including long-term shifts in community composition and alterations of diversity \cite{acevedo-trejos_glimpse_2014, boyd_biological_2016, henson_future_2021}.
-    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to short-term cyclic processes and to long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
+    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to short-termcyclic processes and to long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
 Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.
@@
 Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela.
-Details about the sampling procedures adopted in CARIACO, sampling schemes, analytical methods, data quality control, and inter-calibration procedures, have been already described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.
+Details about the sampling procedures adopted in CARIACO, sampling schemes, analytical methods,  data quality control and inter-calibration procedures, have been already described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.
 
@@
 
-    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to present by Dr.
-Kent Fanning (University of South Florida, St.
+    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to  present by Kent Fanning (University of South Florida, St.
 Petersburg, USA).
@@
     Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}.
-We extracted the -u wind component at \qty{10}{\meter} from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
+We extracted the -u wind component at 10\,m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
 For simplicity, we refer to this variable as "wind speed" throughout the article.
@@
 To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs.
-We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{app:table1} in Appendix).
+We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix).
     % Note: Perhaps add some more info about Supplmentary Results and test runs here?
@@
 
+The analysis could be extended by including top-down effects on the phytoplankton community, particularly zooplankton and fisheries data. 
+We hope that the work hints at the resilient nature of phytoplankton communities under global change, which should be further investigated by studying long-term ocean time series.
+
 %%

### 2025-06-27 agujournaltemplate.tex
Revised Abstract - post Ago 1st Review

@@
 \begin{keypoints}
-\item The CARIACO phytoplankton community time series reveals two clusters associated to high- and low-upwelling conditions and corresponding to anomalies in the AMO index. 
+\item The CARIACO phytoplankton community time series reveals two clusters, separated by a shift from high- to low-upwelling conditions after 2004 and a return to high-upwelling conditions following 2013.
 \index No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low-upwelling conditions.
@@
 Phytoplankton communities play a crucial role in oceans' biogeochemical cycles.
-Climate drives changes in the physical environment that phytoplankton communities experience and can impact, thus, on the productivity and diversity of marine ecosystems.
-Long-term time series can capture changes in climate and ecosystems, allowing us to untangling of effects on local communities.
-This study aims to describe and investigate the drivers of changes in  phytoplankton community data (biomass and diversity) collected in the Cariaco Basin, Venezuela, between 1995 and 2017.
-The monthly cruises sampled water for microscopy cell counts and determination of biogeochemical properties, which we integrated with data from AMO and MEI v.2 climate indices and ERA5 climate model reanalysis data.
-Using clustering algorithms, we investigate the structure of the community data, revealing two temporal clusters associated to high and low upwelling conditions (from 1995 to 2004 and from 2005 to 2017, respectively) and showing a return to community biomass and physical conditions similar to those characterizing the period 1995-2004.
+Climate drives changes in the physical environment that phytoplankton communities experience, thus impacting the productivity and diversity of marine ecosystems.
+Long-term time series can capture changes in climate and ecosystems, allowing us to untangle the effects on local communities.
+This study aims to describe and investigate the changes in phytoplankton community in the Cariaco Basin, Venezuela, between 1995 and 2017.
+Water samples were taken during monthly cruises to determine microscopy cell counts and biogeochemical properties, which we integrated with data from the AMO and MEI v.2 climate indices and ERA5 climate model reanalysis data.
+Using clustering algorithms, we investigate the structure of the community data, revealing two clusters that describe a shift from high- to low-upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014.
 However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009.
-To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ, short-term, upwelling-related variables (SST and nitrate concentration) and long-term, climate indices (AMO and MEI v.2) were the strongest predictors.
-Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of the ecosystem to recover from climate change.
+To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related variables (SST and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors.
+Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of ecosystems to recover from the effects of climate change.
 

### 2025-06-29 agujournaltemplate.tex
Ago Review Intro revised

@@
     Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
-This paraphyletic community consists of a diverse group of organisms  exhibiting highly variable eco-physiological properties \cite{appeltans_magnitude_2012}.
+Phytoplankton are a paraphyletic group of photo-autotrophic organisms exhibiting highly variable eco-physiological properties \cite{appeltans_magnitude_2012}.
 The short generation time and high diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
-    Coastal margins, where phytoplankton typically thrive, constitute only \qty{\sim 10}{\%} of the global ocean surface  but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.
-
-    Data from remote sensing, suggest a global increase in the frequency and spatial extent of coastal phytoplankton blooms over the last 20 years, but regional effects are highly variable \cite{dai_coastal_2023} and difficult to capture.
-    Climate change is predicted to have a lasting impact on phytoplankton communities, including long-term shifts in community composition and alterations of diversity \cite{acevedo-trejos_glimpse_2014, boyd_biological_2016, henson_future_2021}.
-    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to short-termcyclic processes and to long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
+    Coastal margins, where phytoplankton typically thrive, constitute only \qty{\sim 10}{\%} of the global ocean's surface but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.
+
+    Data from remote sensing suggest a global increase in the frequency and spatial extent of coastal phytoplankton blooms over the last 20 years, but regional effects are highly variable \cite{dai_coastal_2023}.
+    Climate change is predicted to have a lasting impact on phytoplankton communities, including shifts in community composition and diversity \cite{acevedo-trejos_glimpse_2014, boyd_biological_2016, henson_future_2021}.
+    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to short-term processes and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
 Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.
 
-    A tropical coastal location that has been studied intensively in the last decades is the CARIACO basin, where a wide range of physical and biogeochemical parameters have been consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}.
+    A tropical coastal location that has been studied intensively in the last decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters have been consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}.
     % Describe CARIACO location
@@
 The water below the sill is poorly ventilated.
-This underwater topography in combination with a high surface productivity, associated to aerobic degradation of fast sinking organic matter, creates permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}.
+This underwater topography in combination with a high surface productivity and aerobic degradation of sinking organic matter, creates permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}.
 Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for understanding the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
@@
     % Describe general seasonal cycle with functional groups etc.
-    The Cariaco Basin is characterised by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward.
+    The Cariaco Basin is characterized by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward.
 A smaller wind-driven secondary upwelling event occurs often in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
 During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing.
-In the low-wind, rainy season, the Cariaco Basin exhibits oligotrophic conditions and the fraction of large phytoplankton cells is reduced, albeit nano- and picophytoplankton cells are present in relatively high numbers throughout the year \cite{lorenzoni_characterization_2015}.
+In the low-wind, rainy season, the Cariaco Basin exhibits oligotrophic conditions, and the fraction of large phytoplankton cells is reduced.
+However, nano- and picophytoplankton cells are present in relatively high numbers throughout the year \cite{lorenzoni_characterization_2015}.
 
@@
 After 2004, a significant reduction in the strength of trade winds during the upwelling period \cite{taylor_ecosystem_2012} caused a regime change in the physical and biogeochemical system.
-This strongly affected the phytoplankton community and the whole ecosystem.
-Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms, as confirmed, albeit not with the same high significance, by analysis on pigment data  \cite{pinckney_phytoplankton_2015}.
-In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly occurring in stratified water columns \cite{lorenzoni_characterization_2015}.
-An analysis of pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.
-During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass \cite{taylor_ecosystem_2012}.
-Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.
+This strongly affected the phytoplankton community and the entire ecosystem.
+Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms, as confirmed, albeit not with the same high significance, by analysis of pigment data \cite{pinckney_phytoplankton_2015}.
+In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly found in stratified water columns \cite{lorenzoni_characterization_2015}.
+An analysis of pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in the seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.
 
     % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
-    The shift in phytoplankton community composition associated to increased temperature and to reduced upwelling intensities, does not appear to be a permanent change in regime.
+    The shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities does not appear to be a permanent change in regime.
 Data from 2012 to 2017, the end phase of the CARIACO time series, indicates a return to strong upwelling intensities and to high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
-The return to conditions similar to those of the late 1990s coincides with a recovery in sardine catches observed in 2016 and 2017 in an area southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.
     The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear.
@@
 The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
-    The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
-However, using the time-series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in-situ observations.
-In addition, the AMO index was included in the analyses of the Cariaco Basin.
+    The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea, given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
+However, using the time series available only until 2010, \citeA{taylor_ecosystem_2012} found no significant correlation between MEI v.2 and in situ observations, and did not test for correlations with the AMO index.
 
     % What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
-    \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change", and indicated that they may "simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation".
+    \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation".
     Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have been undertaken yet on possible changes in diversity based on microscopy cell counts.
-    Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period 1995\,--\,2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO)  and changes in the phytoplankton community.
+    Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 - 2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community.
     Our analysis focuses exclusively on bottom-up drivers of community changes, because these data are available throughout the entire time series and at a monthly resolution.
 Our study builds on and expands substantially the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017.
-Understanding the regional dynamics of the Cariaco Basin is crucial for unveiling theecological changes affecting this tropical marine ecosystem and the variables driving these changes.
+Understanding the regional dynamics of the Cariaco Basin is crucial for unveiling the ecological changes affecting this tropical marine ecosystem and the variables driving these changes.
 
@@
 The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north, running almost parallel to the coastline, at a mean depth of around \qty{100}{m}.
-This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 200}{m} depth.
+This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 275}{m} depth.
 

### 2025-06-29 text-backup.tex
Ago Review Intro revised

@@
+% TOP DOWN ASPECTS FROM INTRO:
+
+During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass \cite{taylor_ecosystem_2012}.
+Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.
+
+The return to conditions similar to those of the late 1990s coincides with a recovery in sardine catches observed in 2016 and 2017 in an area southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% BELOW HERE ARE INTRO DRAFT TEXTS! %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2025-06-29 agujournaltemplate.tex
pre-figure edit

@@
     The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon.
-The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north, running almost parallel to the coastline, at a mean depth of around \qty{100}{m}.
+The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north, running almost parallel to the coastline at a mean depth of around \qty{100}{m}.
 This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 275}{m} depth.

### 2025-06-29 agujournaltemplate.tex
Reviewed Method and Results

@@
 \caption{Bathimetric map of the CARIACO Basin in the Caribbean Ocean, off the coast of Venezuela, indicating the CARIACO Mooring, the station of the Ocean Time-Series (located at \ang{10.50}N, \ang{64.67}W).
-A sill, running almost parallel to the coast at an average depth of 100\,m, from the Isla de Magarita and the Isla de la Tortuga and beyond, encloses the Cariaco basin from the North.}
+A sill, running almost parallel to the coast at an average depth of \qty{100}{m}, from the Isla de Magarita and the Isla la Tortuga and beyond, encloses the Cariaco basin from the North.}
 \label{fig:map}
@@
     Samples were collected as part of the Cariaco Ocean Time Series Program by means of monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
-    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, in addition to other deployments and samples not used in this work.
-Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela.
-Details about the sampling procedures adopted in CARIACO, sampling schemes, analytical methods,  data quality control and inter-calibration procedures, have been already described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.
+    The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work.
+Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Isla de Magarita, Venezuela.
+Details about the sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.
 
@@
 
-    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to  present by Kent Fanning (University of South Florida, St.
+    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to  present by Dr.
+K.
+Fanning (University of South Florida, St.
 Petersburg, USA).
@@
     Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}.
-We extracted the -u wind component at 10\,m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
+We extracted the -u wind component at \qty{10}{\meter} from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
 For simplicity, we refer to this variable as "wind speed" throughout the article.
@@
     As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to create a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}.
-Furthermore, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
+Furthermore, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
 
@@
 We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix).
-    % Note: Perhaps add some more info about Supplmentary Results and test runs here?
 
@@
 Figure \ref{fig:clustering}$a$ shows the results of clustering the annually aggregated communities using a linear complete linkage algorithm based on the binary Jaccard matrix.
-The distinct clusters clearly emerge from these analysis, , the first cluster encompassing the years 1996-2004 and 2014-2016 and the second cluster including the intermediate years from 2004 to 2013.
+Two distinct clusters emerge from this analysis, with the first cluster encompassing the years 1996-2004 and 2014-2016, and the second cluster including the intermediate years from 2005 to 2013.
 Inter-annual species turnover is shown in Figure \ref{fig:clustering}$b$, it reveals an increasing trend in turnover from 1996 to 2004.
@@
 The horizontal axis separates the two clusters, with cluster 2 located over positive anomalies in AMO and precipitation and cluster 1 located over positive anomalies in evaporation, salinity, and MEI v.2.
-The second NMDS axis appears to follow the trends in upwelling intensity, as indicated by the arrows poinintg in opposite directions for nutrient concentrations, temperature, and wind speed.
+The second NMDS axis appears to follow the trends in upwelling intensity, as indicated by the vectors pointing in opposite directions for nutrient concentrations, temperature, and wind speed.
 Overall, the NMDS clustering confirms the pattern shown by clustering on annually aggregated data.
@@
 Figure \ref{fig:clustcomp} shows density distributions of environmental variables, chlorophyll \textit{a}, and diversity indices for the two clusters.
-Upwelling-related variables show distinct patterns between the two clusters.
+Upwelling-related variables exhibit distinct patterns between the two clusters.
 For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} isotherm, lower temperatures, reduced evaporation, and higher salinity and nutrient concentrations, particularly for nitrate.
-The AMO index shows a pronounced separation between the two clusters, with cluster 1 exhibiting more negative anomalies, while cluster 2 showing higher occurrence of positive anomalies.
-Although there are no significant differences in genus richness and the Shannon index between clusters, there is a slightly increased median in the Pielou index for cluster 2.
+The AMO index shows a pronounced separation between the two clusters, with cluster 1 exhibiting more negative anomalies, while cluster 2 shows a higher occurrence of positive anomalies.
+Although there are no significant differences in genus richness and the Shannon index between clusters, there is a slightly increased median in the Pielou index for cluster 2 (Table \ref{tab:ClustCompWilcox}).
 The density distribution reveals some differences.
-Cluster 1 shows higher variance for genus richness, while cluster 2  shows greater variance and a bimodal distribution for Shannon and Pielou indices.
+Cluster 1 shows higher variance for genus richness, while cluster 2 shows greater variance and a bimodal distribution for the Shannon and Pielou indices.
 
@@
 Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range within which the community changed most, indicating these values as a threshold for community composition.
-Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows esponse curves for all variables not shown here.
+Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows response curves for all variables not shown here.
 For wind speed, SST, and salinity, changes occurred mostly within parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}), and high salinity (above \qty{36.8}{PSU}).
@@
 \caption{Results of two-sided Wilcoxon sum rank test with continuity correction for variables between the two clusters.
-Cluster 1 covers the periods 1996-2003 and 2014-2016 and cluster 2 covers the period 2004-2013.}
+Cluster 1 covers the periods 1996-2004 and 2014-2016 and cluster 2 covers the period 2005-2013.}
 \centering
@@
 \end{tabular}
+
+\label{tab:ClustCompWilcox}
 \end{table}

### 2025-06-30 agujournaltemplate.tex
Ago revision completed all comments/corrections

@@
 \item The CARIACO phytoplankton community time series reveals two clusters, separated by a shift from high- to low-upwelling conditions after 2004 and a return to high-upwelling conditions following 2013.
-\index No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low-upwelling conditions.
+\item No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low-upwelling conditions.
 \item Gradient forest analysis identifies the AMO climate index as the strongest predictor of community change, highlighting the influence of large-scale climatic oscillations on this tropical coastal ecosystem.
@@
 Long-term time series can capture changes in climate and ecosystems, allowing us to untangle the effects on local communities.
-This study aims to describe and investigate the changes in phytoplankton community in the Cariaco Basin, Venezuela, between 1995 and 2017.
+This study aims to describe and investigate the changes in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017.
 Water samples were taken during monthly cruises to determine microscopy cell counts and biogeochemical properties, which we integrated with data from the AMO and MEI v.2 climate indices and ERA5 climate model reanalysis data.
@@
 
-    A tropical coastal location that has been studied intensively in the last decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters have been consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}.
+    A tropical coastal location that has been studied intensively in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters have been consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}.
     % Describe CARIACO location
@@
 The water below the sill is poorly ventilated.
-This underwater topography in combination with a high surface productivity and aerobic degradation of sinking organic matter, creates permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}.
+This underwater topography, in combination with high surface productivity and aerobic degradation of sinking organic matter, creates permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}.
 Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for understanding the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
@@
 
-    The monthly AMO index data (10-year low-pass), which represents an area-averaged, detrended low-pass filtered North Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.
+    The monthly AMO index data (10-year low-pass), which represents area-averaged, detrended low-pass filtered North Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.
 
@@
 \subsection{Data analysis}    
-    To provide an overview of the data available and the general trends shown by the data, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$).
+    To provide an overview of the available data and the general trends, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$).
 Cruises from 1995, 2016, and 2017 were excluded because insufficient coverage in these years prevented us from producing reliable yearly mean estimates.
@@
 To check whether the annual aggregation of community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (Figure \ref{fig:clustering}$c$).
-The NMDS clustering is overlaid with environmental data and climate indices in the attempt of unveiling some of the underlying trends.
-The horizontal axis separates the two clusters, with cluster 2 located over positive anomalies in AMO and precipitation and cluster 1 located over positive anomalies in evaporation, salinity, and MEI v.2.
+The NMDS clustering is overlaid with environmental data and climate indices in an attempt to unveil some of the underlying trends.
+The horizontal axis separates the two clusters, with cluster 2 located over positive anomalies in AMO and precipitation, and cluster 1 located over positive anomalies in evaporation, salinity, and MEI v.2.
 The second NMDS axis appears to follow the trends in upwelling intensity, as indicated by the vectors pointing in opposite directions for nutrient concentrations, temperature, and wind speed.
@@
 These changes are well visible in the yearly mean anomalies (Figure \ref{fig:zscore}) and the density distribution of monthly measurements (Figure \ref{fig:clustering}).
-After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident to a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies up to 2014.
-By considering the time series data until the end of 2017, the temporally complete dataset, we find that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}.
+After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies up to 2014.
+By considering the phytoplankton community time series until the end of 2017, the temporally complete dataset, we find that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}.
 In 2014, physical conditions in the Cariaco Basin returned to a regime of increased wind speed, increased upwelling, and elevated nutrient concentrations (Figure \ref{fig:zscore}).
@@
 % P2: Discuss community reorganisation, biomass shifts vs HPLC data 
-The annual aggregates of community data showed two regimes, one covering the periods 1996 to 2004 and 2013-2017 and the other covering the period between 2004 and 2013 (see Figure \ref{fig:clustering}).
-When comparing the two regimes with respect to phytoplankton community data, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a}, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}).
-The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community might have transitioned to smaller cells, which sizes were probably below the detection limit of optical microscopes \cite{muller-karger_scientific_2019}.
+When comparing the two regimes, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a}, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}).
+The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community might have transitioned to smaller cells, which were probably below the detection size limit of optical microscopes \cite{muller-karger_scientific_2019}.
 This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}.
@@
 % Disucssion of HPLC vs Microscopy!
-In addition to our focus on the most temporally complete dataset, another important difference between our work and previous analyses of the Cariaco time series like, for example, the study of \citeauthor{pinckney_phytoplankton_2015}, is that, for unveiling changes in the community composition of phytoplankton, we concentrated on  microscopy counts, whereas  \citeauthor{pinckney_phytoplankton_2015} considered HPLC pigment data.
-The patterns we found in the microscopy data are less clear in the HPLC pigment data, which instead show an increase in chlorophyll \textit{a} between the periods 1996-2000 and 2006-2010, particularly at depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}.
-Cell counts of all functional groups decrease, while HPLC data in  \citeauthor{pinckney_phytoplankton_2015} indicate an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{meters} .
+A similar pattern emerges when comparing our observations with High-Performance Liquid Chromatography (HPLC) data from the Cariaco Basin, which were analyzed and presented by \citeA{pinckney_phytoplankton_2015}.
+The HPLC data suggest an increase in chlorophyll \textit{a} between the periods 1996-2000 and 2006-2010, particularly at depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}.
+Despite cell counts of all functional groups decreasing, the HPLC data indicate an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{\meter}.
+This is a further strong indication that the phytoplankton community shifted to smaller sizes, undetected using light microscopy.
 It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006.
-Additionally, \citeauthor{pinckney_phytoplankton_2015} could not use a unique, shared method for quantifying chlorophyll \textit{a} across different laboratories.
-Here, we focus on consistent fluorometric measurements.
-Consistently with our results, independent spectrophotometric measurements of chlorophyll \textit{a} concentrations at \qty{20}{meters} depth, off the coast of Margarita Island, show a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
-Changes in the composition of the phytoplankton community can affect chlorophyll \textit{a} concentrations, as most measurement methods do not capture only chlorophyll \textit{a} but also other related forms of chlorophyll, which can vary depending on the composition of the phytoplankton community \cite{welschmeyer_fluorometric_1994}.
+\citeA{pinckney_phytoplankton_2015} could not find a consistent method for quantifying total chlorophyll \textit{a} across different laboratories.
+This is why we focused our analysis on the more consistent and temporally complete microscopy cell count and fluorometric chlorophyll \textit{a} data.
+Consistent with our results, independent spectrophotometric measurements of chlorophyll \textit{a} concentrations at \qty{20}{meters} depth, off the coast of Margarita Island, show a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
 
 % P3: Shift in Diversity does not match shift in biomass/cell counts, why?
-The shift we find in yearly mean genus richness, however, does not correspond with abundances in cell counts.
-Instead, we observe a marked reduction from 1998 to 2010, after which richness recovers four years prior to 2014, when the community returns, seemingly, to previous levels, those characteristics of regime 1.
+The temporal dynamics we find in yearly mean genus richness, however, do not correspond directly to the observed regimes.
+Instead, we see a marked reduction from 1998 to 2010, after which richness recovers four years prior to 2014, when the community returns to levels characteristic of regime 1.
 The same trend is evident, albeit to a lesser degree, in the other diversity metrics.
 Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and exhibit a reduced yearly mean between 2000 and 2006, during which cell counts recorded an increase in nanoflagellates and a pronounced peak in cyanobacteria.
- We observe a marked reduction in the genus richness of microphytoplankton ahead of the collapse in abundance, and similarly an increase a few years prior to the return in abundance to previous levels.
-Microscopy cell counts do not show any statistical difference in diversity between the two  regimes, either by Shannon index or by genus richness (Figures \ref{fig:zscore} and \ref{fig:divts}).
-We observe a significant difference in the Pielou's index between the two regime.
-The HPLC pigment trend unveiled by and Our analysis on microscopy measurements suggesting that the phytoplankton community becomes more evenly distributed during low-upwelling conditions of regime 2 are consistent with the analysis of \citeauthor{pinckney_phytoplankton_2015} based on HPLC pigment data.
+We did not find a significant difference in diversity between the two regimes, either by Shannon index or by genus richness (Figures \ref{fig:zscore} and \ref{fig:divts}).
+However, we did observe a significant difference in Pielou's index between the two regimes.
+The results suggest that the phytoplankton community becomes more evenly distributed during low-upwelling conditions of regime 2, which is consistent with the analysis of \citeauthor{pinckney_phytoplankton_2015} based on HPLC pigment data.
 This increase in evenness, quantified by the Pielou index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community.
@@
 The threshold responses between functional groups, which relate to shifts in cell count abundances, show contrasting patterns.
-Diatoms, dinoflagellates, and haptophytes exhibit strong response thresholds for negative AMO anomalies, which could point to influences of stronger upwelling conditions.
-The abundance of haptophytes shows, additionally, a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} using HPLC data of haptophytes predominantly exhibiting blooms above the mixed layer and at a lower frequency than other functional groups.
-
-The El Niño Southern Oscillation (ENSO) exhibits a strong influence on the Caribbean Sea.
+Diatoms, dinoflagellates, and haptophytes exhibit strong response thresholds for negative AMO anomalies, which could point to the influence of stronger upwelling conditions.
+The abundance of haptophytes, additionally, shows a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} based on HPLC data that haptophytes predominantly exhibit blooms above the mixed layer and at a lower frequency than other functional groups.
+
+The El Niño Southern Oscillation (ENSO) exhibits a strong teleconnection to the Caribbean Sea.
 During ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
-\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months.
+\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months.
 In contrast, we found a relatively strong predictive power for changes in the phytoplankton community for the MEI v.2 index with a 4-month time lag.
 We also tested a 12-month lag and found no significant improvement (Table \ref{tab:sup:GFlagtests} in Appendix).
-
 The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for negative MEI v.2 index anomalies, which represent La Niña conditions.
@@
 Salinity, evaporation, and silicate concentration exhibit some explanatory power in the gradient forest analysis, although they are weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}).
-This implies that the climate indices explain a variability that more direct measurements of upwelling intensity do not capture.
+This implies that the climate indices explain variability that more direct measurements of upwelling intensity do not capture.
 
@@
 This does not represent a complete picture of the phytoplankton community of the CARIACO station.
-The HPLC pigment data used by \citeauthor{pinckney_phytoplankton_2015} capture all size classes of phytoplankton, but as previously discussed, does not cover the entire time series and the dataset was assembled by four different laboratories.
-With the caveat that the microscopy counts capture microphytoplankton exclusively, the data can be brought into agreement.
+The HPLC pigment data used by \citeauthor{pinckney_phytoplankton_2015} capture all size classes of phytoplankton, but, as previously discussed, it does not cover the entire time series and the dataset was assembled by four different laboratories.
+With the caveat that the microscopy counts capture microphytoplankton almost exclusively, the discrepancies to the fluorometric chlorophyll \textit{a} measurements can be explained.
 Drastic declines in cell counts reflect shifts to smaller species, an aspect consistent with size spectra calculated by previous studies using HPLC pigment data \cite{lorenzoni_characterization_2015}.
@@
 % 2. Discuss limitations of GF analysis & Limitation to bottom-up drivers and climate indices
-The restricted scope of the microscopy genus counts poses limitations to the gradient forest algorithm we employed to quantify predictors of community change.
-Many of the predictors in ecological time series and in this study are highly correlated (Figure \ref{fig:sup:correlation} in Appendix).
+The limited scope of the microscopy cell counts also affects the conclusions that can be drawn from the Gradient Forest analysis, as these results can only be interpreted as concerning microphytoplankton.
+Another potential caveat of our methodology is that many of the predictors in ecological time series and in this study are highly correlated (Figure \ref{fig:sup:correlation} in Appendix).
 This can lead to the identification of a correlated predictor that has no direct effect on community changes.
@@
 
-Our focus on bottom-up drivers of community change may also restrict the explanatory power of our analysis, as top-down processes such as zooplankton grazing may impact on the structure of the phytoplankton community \cite{banas_adding_2011, acevedo-trejos_mechanisms_2015}.
-Due to the scarcity of data covering higher trophic levels and the absence of zooplankton data for the first six years of the time series, we suggest that mechanistic modeling is the ideal approach for investigating the relative impact of top-down and bottom-up drivers.
+Our focus on bottom-up drivers of community change may also restrict the explanatory power of our analysis, as top-down processes such as zooplankton grazing may impact the structure of the phytoplankton community \cite{frank_ups_2007, banas_adding_2011, acevedo-trejos_mechanisms_2015}.
+Due to the scarcity of data covering higher trophic levels and the absence of zooplankton data for the first six years of the time series, we suggest that mechanistic modeling is the ideal approach for investigating the relative impacts of both top-down and bottom-up drivers.
 
@@
 We found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco Basin.
-Specifically, microscopy cell counts reveal the presence of two distinct regimes (and a return, during the final years, to regime 1) defined by high- and low-upwelling conditions that correspond to anomalies in the AMO index.
+Specifically, microscopy cell counts reveal the presence of two distinct regimes defined by high- and low-upwelling conditions that correspond to anomalies in the AMO index.
+The community changes do not indicate a permanent regime shift, as a return to regime 1 is observed towards the end of the time series.
 We observe no significant changes in diversity metrics between the two regimes, except for lower turnover and higher community evenness under low-upwelling conditions.
-Despite multiannual trends in warming, reduced productivity, and drastic declines in microphytoplankton cell counts, the phytoplankton community returned to regime 1 levels once the climatic and local physical conditions were restored.
-The strongest bottom-up variable predicting changes in the phytoplankton community resulted the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local conditions through low-frequency natural cycles.
+Despite multiannual trends in warming, reduced productivity, and drastic declines in microphytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions were restored.
+The strongest bottom-up variable predicting changes in the phytoplankton community was the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local conditions through low-frequency natural cycles.
 We concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state for the phytoplankton community of the Cariaco Basin.
-Notwithstanding a number of limitations, the incipient return of the Cariaco basin to a pre-2004 regime hints at the resilient nature of phytoplankton communities under global change, an aspect that points to the relevance of long-term time series and deserves further attention.
-Future studies should focus on the relative impact of top-down and bottom-up drivers.
-
-The analysis could be extended by including top-down effects on the phytoplankton community, particularly zooplankton and fisheries data. 
-We hope that the work hints at the resilient nature of phytoplankton communities under global change, which should be further investigated by studying long-term ocean time series.
+Notwithstanding a number of limitations, our analysis points to the incipient return of the Cariaco Basin to a pre-2004 regime, hinting at the resilient nature of phytoplankton communities under global change, an aspect that emphasizes the relevance of long-term time series and deserves further attention.
+To extend our approach, future studies should include the impact of top-down drivers and investigate their explanatory power regarding the phytoplankton community changes observed in the Cariaco Basin.
 

### 2025-07-15 agujournaltemplate.tex
Andrew's Feedback

@@
 \begin{abstract}
-Phytoplankton communities play a crucial role in oceans' biogeochemical cycles.
+Phytoplankton communities play a crucial role in ocean biogeochemical cycles.
 Climate drives changes in the physical environment that phytoplankton communities experience, thus impacting the productivity and diversity of marine ecosystems.
-Long-term time series can capture changes in climate and ecosystems, allowing us to untangle the effects on local communities.
+Long-term time series of XXX can capture changes in climate and ecosystems, allowing us to understand the effects of environmental change on local communities.
 This study aims to describe and investigate the changes in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017.
-Water samples were taken during monthly cruises to determine microscopy cell counts and biogeochemical properties, which we integrated with data from the AMO and MEI v.2 climate indices and ERA5 climate model reanalysis data.
-Using clustering algorithms, we investigate the structure of the community data, revealing two clusters that describe a shift from high- to low-upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014.
+Water samples were taken during monthly cruises to determine plankton community composition and local water column conditions, which we integrated climate indice and meteorological data.
+Using clustering algorithms, we investigate the structure of the plankton community data, revealing two clusters that described a shift from high- to low-upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014.
 However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009.
-To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related variables (SST and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors.
+To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related variables (sea surface temperature and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors.
 Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of ecosystems to recover from the effects of climate change.
@@
     % Phytoplankton is important
-    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
-Phytoplankton are a paraphyletic group of photo-autotrophic organisms exhibiting highly variable eco-physiological properties \cite{appeltans_magnitude_2012}.
-The short generation time and high diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
-    Coastal margins, where phytoplankton typically thrive, constitute only \qty{\sim 10}{\%} of the global ocean's surface but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.
-
-    Data from remote sensing suggest a global increase in the frequency and spatial extent of coastal phytoplankton blooms over the last 20 years, but regional effects are highly variable \cite{dai_coastal_2023}.
-    Climate change is predicted to have a lasting impact on phytoplankton communities, including shifts in community composition and diversity \cite{acevedo-trejos_glimpse_2014, boyd_biological_2016, henson_future_2021}.
-    Long-term time series of biological observations are crucial for understanding how marine ecosystems respond to short-term processes and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
-Such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.
-
-    A tropical coastal location that has been studied intensively in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters have been consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}.
-    % Describe CARIACO location
+    Phytoplankton provide the foundation for marine food webs and play crucial roles in biogeochemical processes, \cite{falkowski_biogeochemical_1998}.
+Their short generation times and high assemblage diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state and dynamics \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
+Long-term time series of biological observations are crucial for understanding how phytoplankton communities respond to short-term processes and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
+Such time series are rare, particularly in the tropics \cite{clarke_does_2017}.
+
+A tropical coastal location that has been studied intensively in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters were consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}.
 The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world oceans \cite{edgcomb_accessing_2011}.
-The basin is a continental shelf depression comprising two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}.
-The water below the sill is poorly ventilated.
-This underwater topography, in combination with high surface productivity and aerobic degradation of sinking organic matter, creates permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}.
-Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for understanding the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
-
-    % Describe general seasonal cycle with functional groups etc.
 The Cariaco Basin is characterized by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward.
@@
     Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes.
-\citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found significant decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to global warming.
+\citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found  decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to global warming.
 During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
@@
 In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly found in stratified water columns \cite{lorenzoni_characterization_2015}.
-An analysis of pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in the seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.
+An analysis of pigment data indicated a deepening of the chlorophyll maximum, an increase in species diversity, and a reduction in the seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.
 
     % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
-    The shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities does not appear to be a permanent change in regime.
-Data from 2012 to 2017, the end phase of the CARIACO time series, indicates a return to strong upwelling intensities and to high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
+    However, the shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities in year-start-year-end did not appear to be a permanent change in regime.
+Data from 2012 to 2017, the end phase of the CARIACO time series, indicated a return to strong upwelling intensities and to high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
 The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear.
@@
 The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north, running almost parallel to the coastline at a mean depth of around \qty{100}{m}.
-This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 275}{m} depth.
+This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 275}{m} depth.
 
@@
 \noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
-\caption{Bathimetric map of the CARIACO Basin in the Caribbean Ocean, off the coast of Venezuela, indicating the CARIACO Mooring, the station of the Ocean Time-Series (located at \ang{10.50}N, \ang{64.67}W).
+\caption{Bathymetric map of the CARIACO Basin in the Caribbean Ocean.
+The station is located at \ang{10.50}N, \ang{64.67}W).
 A sill, running almost parallel to the coast at an average depth of \qty{100}{m}, from the Isla de Magarita and the Isla la Tortuga and beyond, encloses the Cariaco basin from the North.}
@@
 
-\subsection{In-Situ Data}
-    Samples were collected as part of the Cariaco Ocean Time Series Program by means of monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
+\subsection{In-situ biological and environmental data}
+    <Add a topic sentence stating what in situ data are use for.> Samples were collected as part of the Cariaco Ocean Time Series Program on monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
     The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work.
-Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Isla de Magarita, Venezuela.
 Details about the sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.
@@
     All samples were collected using a 12-bottle rosette system equipped with a CTD (SeaBird, model SBE 25) \cite{astor_yrene_m_handbook_2013}.
-Water samples for chlorophyll a and nutrient analysis were taken monthly during the first cast of each cruise.
+Water samples for chlorophyll-a and nutrient analysis were taken monthly during the first cast of each cruise.
 These samples were collected at eight standard depths within the upper 100 meters (1, 7, 15, 25, 35, 55, 75, and 100 m).
 
-    Between late 1995 and early 1998, nutrient data was provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to  present by Dr.
+    Between late 1995 and early 1998, nutrient data, including X, Y, and Z, were provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to  present by Dr.
 K.
@@
     For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at the same depths in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate.
-Quantitative analysis was performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period.
+Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period.
 Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
     %from Mutshinda et al. 2013
-    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts) correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}).
-After processing the data, 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.
+    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}).
+482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.
 
     Data collected with Niskin bottles, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}.
-    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (i.e.
-no more than three months) in coverage.
-In total, we used a complete set of 254 monthly data points, for which all variables were available.
-
-\subsection{Climate data}
+    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than three months) in coverage.
+In total, we used a complete set of 254 monthly data points from YearX-YearY, for which all variables were available.
+
+\subsection{Meteorological and climate data}
     %from Taylor et al. 2012
-    Zonal wind stress (-u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters via Ekman transport \cite{rueda-roa_southern_2013}.
-We extracted the -u wind component at \qty{10}{\meter} from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
+    In addition to oceanographic environmental data, we also examine meteorological and climatic data that are known to influence ocean conditions and marine ecosystems.
+Zonal wind stress (u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters  \cite{rueda-roa_southern_2013}.
+We extracted the u wind component at \qty{10}{\meter} from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
 For simplicity, we refer to this variable as "wind speed" throughout the article.
@@
 
-    The monthly AMO index data (10-year low-pass), which represents area-averaged, detrended low-pass filtered North Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.
-
-    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of surface wind, and outgoing long-wave radiation (OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/}.
-In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).
+    In addition the wind data, we examined two climatic indices, each of which provide information on the basin-scale climatic conditions and consequently the broad-scale forcing of marine ecosystems.
+The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies over what area, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.
+
+    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP); sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/}.
+In our analysis, bi-monthly values were aligned with the tail month (December-January is treated as the value for January).
 
@@
 
-    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to create a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}.
-Furthermore, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
+    We calculated three metrics of phytoplankton community diversity.
+First, we calculated genus richness, which reduces taxonomic resolution cut creates a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}.
+Second and third, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, respectively, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
 
@@
 This analysis was conducted using monthly measurements integrated over a depth of 100 m for observed genus counts, ensuring the inclusion of all available data.
-The drivers/predictor variables considered in the analysis were the AMO index, the MEI v.2 index, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO]}_3$, $\mathrm{[PO]}_4$ and $\mathrm{[SiO]}_4$).
+The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO]}_3$, $\mathrm{[PO]}_4$ and $\mathrm{[SiO]}_4$).
     The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}.
@@
 
+Create subsections in results. This one could be "Interannual changes in the environment and ecosystem"
 % Figure 2
-Figure \ref{fig:zscore} shows the normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group.
-We included marks to highlight years with a significant number of missing monthly measurements ($=$ 2, 3 months and $\geq$ 4 months), since these make the annual means less reliable.
-Particularly toward the later years of the time series, the sampling became sparse.
+Coordinated, interannual changes occur in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}).
 The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm).
@@
 Annual mean anomalies of nitrate and phosphate appear to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration.
-Silicate does not follow a clear pattern but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to diminished diatom abundances.
+Silicate does not follow a clear pattern but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to reduced abundances of diatoms.
+<I think the most obvious pattern in Fig.
+2 has not yet been discussed: Chl and phyto are high in 1996-2004, then low in 2004-2013, followed by a switch in 2014-2015.
+Diversity is high when chl/phyto are low.
+I think you'll want to describe these patterns as they are interesting.>
 
@@
 \noindent\includegraphics[width=\textwidth]{fig/Figure2_ZScores_v2.pdf}
-\caption{Time series showing annual z-scores of selected indicator and driver variables.
+\caption{Time series showing annual z-scores of selected environmental, climatological, and ecological variables.
 Red indicates negative annual anomalies and blue indicates positive annual anomalies.
 Deeper shades indicate larger anomalies.
-All data are available at monthly resolution.
-The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to lack of data, additionally we marked years with two or three and four or more missing values to highlight unreliable annual mean estimates.
-Diversity Indices are calculated based on microscopy counts at the Genus level.}
+The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to insufficient monthly data to generate robust annual anomalies.
+Additionally we marked years with two or three and four or more missing monthly values.
+Diversity indices were calculated based on microscopy counts at the genus level.}
 \label{fig:zscore}
@@
 Figure \ref{fig:divts} shows the depth distribution and dynamics of chlorophyll \textit{a} and community diversity indices, also in relation to rainy and upwelling seasons.
-The bulk of the chlorophyll \textit{a} biomass ($\sim$ 80\,\%) was measured in the upper 50 meters of the water column, with approximately 15\,\% measured between 50 and 75 meters.
-Chlorophyll \textit{a} exhibits marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, Figure \ref{fig:divts}$b$).
-The yearly mean chlorophyll \textit{a} biomass shows the highest variability within the upper 25 meters.
-In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range (Figure \ref{fig:divts}$a$).
-In general, yearly mean anomalies within the upper 75 meters appear to be high during the period from 1996 to 2004 and only reduces in 2012.
-The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter months (January and February) skew the value.
-Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\,\% of the mean over the entire time series for these months), which recovers in 2014.
-Chlorophyll \textit{a} biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series.
-In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters is markedly reduced, while the 50–75 meter depth interval shows a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
+The bulk of the chlorophyll \textit{a}  ($\sim$ 80\,\%) was measured in the upper 50 meters of the water column, with approximately 15\,\% measured between 50 and 75 meters.
+Chlorophyll \textit{a} exhibited marked seasonality, with the highest concentration in the upper water column during the winter months (December–February, Figure \ref{fig:divts}$b$).
+The yearly mean chlorophyll \textit{a} concentration showed the highest variability within the upper 25 meters.
+In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deep chlorophyll maximum in those years (Figure \ref{fig:divts}$a$).
+In general, yearly mean anomalies within the upper 75 meters appear to be high during the period from 1996 to 2004 and only reduced in 2012.
+The year 2013 was an outlier, particularly for the upper 25 meters, as the missing monthly sampling from the winter months (January and February) skewed the value downwards.
+Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower cholorophyll (65\,\% of the mean over the entire time series for these months), which recovers in 2014.
+Chlorophyll \textit{a} concentration below 50 meters depth showed a relatively stable distribution throughout the year and across the time series.
+In fall, during the rainy season, mean chlorophyll \textit{a} concentration in the upper 25 meters is markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
+
 The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons in winter.
@@
 The seasonal signal is lost (see Figure \ref{fig:divts} $f$).
+
 Community evenness, expressed by the Pielou index, shows no seasonal signal and a very large variance within seasons and through depth (see Figure \ref{fig:divts} $h$).
@@
 
+Create subsection like "Clustering analysis yields contrasting ecological regimes"
 % Figure 4

### 2025-07-21 agujournaltemplate.tex
In process- editing Andrews Comments

@@
 Climate drives changes in the physical environment that phytoplankton communities experience, thus impacting the productivity and diversity of marine ecosystems.
-Long-term time series of XXX can capture changes in climate and ecosystems, allowing us to understand the effects of environmental change on local communities.
+Long-term time series of phytoplankton can capture changes in climate and ecosystems, allowing us to understand the effects of environmental change on local communities.
 This study aims to describe and investigate the changes in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017.
-Water samples were taken during monthly cruises to determine plankton community composition and local water column conditions, which we integrated climate indice and meteorological data.
-Using clustering algorithms, we investigate the structure of the plankton community data, revealing two clusters that described a shift from high- to low-upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014.
+Water samples were taken during monthly cruises to determine plankton community composition and local water column conditions, which we integrated with climate indices and meteorological data.
+Using clustering algorithms, we investigate the structure of the phytoplankton community data, revealing two clusters that described a shift from high- to low-upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014.
 However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009.
@@
 
-Microscopic algae drive... (Will write this Summary last, after draft is finalized)
+% NO MORE THAN 200 Words
+% Topic Overview
+Phytoplankton are microscopic algae that form the foundation of marine food webs and ecosystems.
+Their growth and community composition are impacted by the local physical and chemical environment, which is determined by currents and climate, for example, through processes such as wind-driven upwelling of nutrients from the deeper ocean.
+Time series data of phytoplankton biomass and community composition can be very helpful in untangling the link between global and local climate and its effects on the phytoplankton community.
+% Paper Overview
+In this study, we set out to analyze changes in the phytoplankton community between 1995 and 2017 at a time series location in a large coastal basin in Venezuela, the Cariaco Basin.
+Previous analysis of the data found strong shifts within the phytoplankton community that could be related to impacts of global warming.
+Our main question in approaching the data was whether we can confirm this shift in community composition or whether other climate processes can explain the observed patterns.
+% Findings Summary
+We found that the phytoplankton community could be separated into two groups.
+The first group includes an early and late period (1995-2003 and 2014-2017) that was characterized by robust phytoplankton growth driven by relatively strong upwelling of nutrients.
+The second group is the intermediate period, where a reduction in upwelling and phytoplankton biomass was observed.
+Additionally, we related physical and climatic variables, such as sea surface temperature and climate indices that relate to long-term oscillations in the North Atlantic, to the changes in community composition.
+By allowing for a time lag, we found that the Atlantic Meridional Oscillation (AMO) index was the strongest predictor of community changes, with temperature and
+% Key Takeaways
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%Comments from Subhi:
-    % - first 2 paragraphs, repetetive, put previously known parts into intro, just start from "the problem"
-    % - streamline discussion, much shorter!
-    % -
 
@@
 A tropical coastal location that has been studied intensively in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters were consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}.
-The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world oceans \cite{edgcomb_accessing_2011}.
+The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world's oceans \cite{edgcomb_accessing_2011}.
 The Cariaco Basin is characterized by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward.
@@
 In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly found in stratified water columns \cite{lorenzoni_characterization_2015}.
-An analysis of pigment data indicated a deepening of the chlorophyll maximum, an increase in species diversity, and a reduction in the seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.
+An analysis of pigment data indicated a deepening of the chlorophyll maximum, an increase in pigment diversity, and a reduction in the seasonal variability of community composition \cite{pinckney_phytoplankton_2015}.
 
 % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
-    However, the shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities in year-start-year-end did not appear to be a permanent change in regime.
-Data from 2012 to 2017, the end phase of the CARIACO time series, indicated a return to strong upwelling intensities and to high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
+However, the shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities from 2004 to 2013 did not appear to be a permanent change in regime.
+Data from 2013 to 2017, the end phase of the CARIACO time series, indicated a return to strong upwelling intensities and high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
 The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear.
@@
 \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation".
-Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have been undertaken yet on possible changes in diversity based on microscopy cell counts.
-Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 - 2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community.
-Our analysis focuses exclusively on bottom-up drivers of community changes, because these data are available throughout the entire time series and at a monthly resolution.
-Our study builds on and expands substantially the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017.
+Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have yet been undertaken on possible changes in diversity based on microscopy cell counts.
+Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community.
+Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution.
+Our study builds on and substantially expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017.
 Understanding the regional dynamics of the Cariaco Basin is crucial for unveiling the ecological changes affecting this tropical marine ecosystem and the variables driving these changes.
@@
     The Cariaco Ocean Time Series Program was established to explore the relationship between ocean surface processes and the sinking flux of particulate carbon.
-The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin, situated on the northeastern Venezuelan continental shelf, with a sill to the north, running almost parallel to the coastline at a mean depth of around \qty{100}{m}.
+The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin situated on the northeastern Venezuelan continental shelf, with a sill to the north running almost parallel to the coastline at a mean depth of around \qty{100}{m}.
 This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 275}{m} depth.
@@
 \noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
-\caption{Bathymetric map of the CARIACO Basin in the Caribbean Ocean.
-The station is located at \ang{10.50}N, \ang{64.67}W).
-A sill, running almost parallel to the coast at an average depth of \qty{100}{m}, from the Isla de Magarita and the Isla la Tortuga and beyond, encloses the Cariaco basin from the North.}
+\caption{Bathymetric map of the CARIACO Basin in the Caribbean Ocean. The station is located at \ang{10.50}N, \ang{64.67}W).}
 \label{fig:map}
@@
 \subsection{In-situ biological and environmental data}
-    <Add a topic sentence stating what in situ data are use for.> Samples were collected as part of the Cariaco Ocean Time Series Program on monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.66}W) \cite{muller-karger_scientific_2019}.
+    The CARIACO Ocean Time Series Program collected biological and environmental data, which we utilized here to examine connections between environmental and ecological change in the region.
+Samples were collected on monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.67}W) \cite{muller-karger_scientific_2019}.
     The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work.
@@
     All samples were collected using a 12-bottle rosette system equipped with a CTD (SeaBird, model SBE 25) \cite{astor_yrene_m_handbook_2013}.
-Water samples for chlorophyll-a and nutrient analysis were taken monthly during the first cast of each cruise.
+Water samples for chlorophyll \textit{a} and nutrient analysis were taken monthly during the first cast of each cruise.
 These samples were collected at eight standard depths within the upper 100 meters (1, 7, 15, 25, 35, 55, 75, and 100 m).
@@
 
-    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at the same depths in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate.
+    For the identification of phytoplankton species by microscopy, \qty{500}{\milli\liter} seawater samples were collected at depths of 1, 7, 15, 25, 35, 55, 75, and 100 m in HDPE bottles and preserved with a \qty{5}{\%} (final concentration) formalin solution neutralized with sodium tetraborate.
 Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period.
@@
     Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}).
-482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.
-
-    Data collected with Niskin bottles, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}.
+A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.
+
+    All in-situ data, inlcuding Niskin bottle, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}.
     From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than three months) in coverage.
-In total, we used a complete set of 254 monthly data points from YearX-YearY, for which all variables were available.
+In total, we used a complete set of 254 monthly data points from 1995 to 2017, for which all variables were available.
 
@@
     In addition to oceanographic environmental data, we also examine meteorological and climatic data that are known to influence ocean conditions and marine ecosystems.
-Zonal wind stress (u, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters  \cite{rueda-roa_southern_2013}.
-We extracted the u wind component at \qty{10}{\meter} from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
+Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}.
+We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
 For simplicity, we refer to this variable as "wind speed" throughout the article.
 The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell.
-Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate for total precipitation, comprising all water that falls on Earth's surface as rain or snow.
-
-    In addition the wind data, we examined two climatic indices, each of which provide information on the basin-scale climatic conditions and consequently the broad-scale forcing of marine ecosystems.
-The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies over what area, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.
-
-    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP); sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR)) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/}.
+Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.
+
+    In addition to the wind data, we examined two climatic indices, each of which provides information on the basin-scale climatic conditions and, consequently, the broad-scale forcing of marine ecosystems.
+The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies (\ang{0}N-\ang{60}N), were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.
+
+    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/}.
 In our analysis, bi-monthly values were aligned with the tail month (December-January is treated as the value for January).
@@
 \subsection{Data analysis}    
-    To provide an overview of the available data and the general trends, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$).
-Cruises from 1995, 2016, and 2017 were excluded because insufficient coverage in these years prevented us from producing reliable yearly mean estimates.
+    To provide an overview of the available data and highlight the interannual trends and changes in the time series data, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$).
+Cruises from 1995, 2016, and 2017 were excluded because insufficient coverage (i.e.
+more than six missing monthly samples) in these years prevented us from producing reliable yearly mean estimates.
 
     We calculated three metrics of phytoplankton community diversity.
-First, we calculated genus richness, which reduces taxonomic resolution cut creates a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}.
+First, we calculated genus richness, which reduces taxonomic resolution but creates a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}.
 Second and third, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, respectively, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
@@
 Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}.
-Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then it averages these to determine the overall importance of each predictor.
+Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor.
 The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover.
@@
 These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in Appendix).
-The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing inter-annual oscillations.
+The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing interannual oscillations.
 Annual mean anomalies of nitrate and phosphate appear to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration.
@@
 The bulk of the chlorophyll \textit{a} ($\sim$ 80\,\%) was measured in the upper 50 meters of the water column, with approximately 15\,\% measured between 50 and 75 meters.
-Chlorophyll \textit{a} exhibited marked seasonality, with the highest concentration in the upper water column during the winter months (December–February, Figure \ref{fig:divts}$b$).
-The yearly mean chlorophyll \textit{a} concentration showed the highest variability within the upper 25 meters.
+Chlorophyll \textit{a} exhibited marked seasonality, with the highest biomass in the upper water column during the early upwelling months (December–February, Figure \ref{fig:divts}$b$).
+The yearly mean chlorophyll \textit{a} biomass showed the highest variability within the upper 25 meters.
 In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deep chlorophyll maximum in those years (Figure \ref{fig:divts}$a$).
@@
 Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower cholorophyll (65\,\% of the mean over the entire time series for these months), which recovers in 2014.
-Chlorophyll \textit{a} concentration below 50 meters depth showed a relatively stable distribution throughout the year and across the time series.
-In fall, during the rainy season, mean chlorophyll \textit{a} concentration in the upper 25 meters is markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
+Chlorophyll \textit{a} biomass below 50 meters depth showed a relatively stable distribution throughout the year and across the time series.
+In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters is markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
 

### 2025-07-21 text-backup.tex
In process- editing Andrews Comments

@@
 The return to conditions similar to those of the late 1990s coincides with a recovery in sardine catches observed in 2016 and 2017 in an area southeast of Isla de Margarita \cite{gomez_gaspar_variacion_2025}.
+
+%% COASTAL MARGIN FROM INTRO:
+
+Coastal margins, where phytoplankton typically thrive, constitute only \qty{\sim 10}{\%} of the global ocean's surface but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.
+Data from remote sensing suggest a global increase in the frequency and spatial extent of coastal phytoplankton blooms over the last 20 years, but regional effects are highly variable \cite{dai_coastal_2023}.
+Climate change is predicted to have a lasting impact on phytoplankton communities, including shifts in community composition and diversity \cite{acevedo-trejos_glimpse_2014, boyd_biological_2016, henson_future_2021}.
+
+, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}
+
+The basin is a continental shelf depression comprising two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}.
+The water below the sill is poorly ventilated.
+This underwater topography, in combination with high surface productivity and aerobic degradation of sinking organic matter, creates permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}.
+Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for understanding the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
+% Describe general seasonal cycle with functional groups etc.
 

### 2025-07-23 agujournaltemplate.tex
Reworked Abstract + Plain language summary

@@
 \begin{abstract}
-Phytoplankton communities play a crucial role in ocean biogeochemical cycles.
-Climate drives changes in the physical environment that phytoplankton communities experience, thus impacting the productivity and diversity of marine ecosystems.
+
+Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
 Long-term time series of phytoplankton can capture changes in climate and ecosystems, allowing us to understand the effects of environmental change on local communities.
-This study aims to describe and investigate the changes in the phytoplankton community of the Cariaco Basin, Venezuela, between 1995 and 2017.
-Water samples were taken during monthly cruises to determine plankton community composition and local water column conditions, which we integrated with climate indices and meteorological data.
-Using clustering algorithms, we investigate the structure of the phytoplankton community data, revealing two clusters that described a shift from high- to low-upwelling conditions in 2004, but also show a return to a community and conditions similar to those before the shift in 2014.
-However,  community diversity did not follow the same pattern, showing a decrease between 1998 and 2009.
-To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that a combination of in situ upwelling-related variables (sea surface temperature and nitrate concentration) and climate indices (AMO and MEI v.2) were the strongest predictors.
-Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the capability of ecosystems to recover from the effects of climate change.
+This study investigates changes in the phytoplankton community of the Cariaco Basin, Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data.
+We considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in wind that drive upwelling in the Cariaco Basin.
+Cluster analysis identified two distinct community states associated with a shift from high to low upwelling conditions in 2004, followed by a return to a community and conditions similar to those before the shift in 2014.
+In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
+To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2 index.
+Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the resilient nature of the tropical marine ecosystem under global change.
 
@@
 
-% NO MORE THAN 200 Words
 % Topic Overview
-Phytoplankton are microscopic algae that form the foundation of marine food webs and ecosystems.
-Their growth and community composition are impacted by the local physical and chemical environment, which is determined by currents and climate, for example, through processes such as wind-driven upwelling of nutrients from the deeper ocean.
-Time series data of phytoplankton biomass and community composition can be very helpful in untangling the link between global and local climate and its effects on the phytoplankton community.
+Phytoplankton are microscopic algae that form the foundation of the ocean food web.
+Because they respond quickly to environmental changes, studying them can show how marine ecosystems are affected by climate.
 % Paper Overview
-In this study, we set out to analyze changes in the phytoplankton community between 1995 and 2017 at a time series location in a large coastal basin in Venezuela, the Cariaco Basin.
-Previous analysis of the data found strong shifts within the phytoplankton community that could be related to impacts of global warming.
-Our main question in approaching the data was whether we can confirm this shift in community composition or whether other climate processes can explain the observed patterns.
+This study looked at how phytoplankton communities in the Cariaco Basin, off the coast of Venezuela, changed from 1995 to 2017.
+We compared microscopy cell counts and monthly water samples with weather data and two major climate indicators: the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to local wind and currents.
 % Findings Summary
-We found that the phytoplankton community could be separated into two groups.
-The first group includes an early and late period (1995-2003 and 2014-2017) that was characterized by robust phytoplankton growth driven by relatively strong upwelling of nutrients.
-The second group is the intermediate period, where a reduction in upwelling and phytoplankton biomass was observed.
-Additionally, we related physical and climatic variables, such as sea surface temperature and climate indices that relate to long-term oscillations in the North Atlantic, to the changes in community composition.
-By allowing for a time lag, we found that the Atlantic Meridional Oscillation (AMO) index was the strongest predictor of community changes, with temperature and
+The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning to the previous state by 2014.
+Interestingly, species diversity declined between 1998 and 2009, although upwelling had not yet changed.
+Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate levels in the water column, were the strongest predictors of changes in the phytoplankton community.
 % Key Takeaways
+These findings contribute to a broader understanding of how climate variability shapes marine phytoplankton communities and underscore the importance of sustained long-term observations in detecting and interpreting ecological responses to climate change.
 
@@
 In the Cariaco Basin, decadal-scale climate oscillations impacted the wind-driven upwelling regime, producing confounding effects on the composition and biodiversity of the phytoplankton community.
-These changes are well visible in the yearly mean anomalies (Figure \ref{fig:zscore}) and the density distribution of monthly measurements (Figure \ref{fig:clustering}).
+These changes are clearly visible in the yearly mean anomalies (Figure \ref{fig:zscore}) and the density distribution of monthly measurements (Figure \ref{fig:clustcomp}).
 After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies up to 2014.
@@
 Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in regime 1 (see Figure \ref{fig:clustering} $e$).
+- Strong El Niño events increase the strength of currents in the Caribbean sea through modulating the easterly wind anomalies \cite{huang_enso_2023}.
 

### 2025-07-25 agujournaltemplate.tex
added genus level table for gradient forest input data

@@
 
-A tropical coastal location that has been studied intensively in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters were consistently collected from 1995 to 2017 \cite{muller-karger_scientific_2019}.
+A tropical coastal location that has been intensively studied in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters were collected consistently between 1995 and 2017 \cite{muller-karger_scientific_2019}.
 The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world's oceans \cite{edgcomb_accessing_2011}.
 The Cariaco Basin is characterized by a marked seasonal cycle of upwelling, which is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward.
-A smaller wind-driven secondary upwelling event occurs often in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
+A smaller wind-driven secondary upwelling event often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
 During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing.
@@
 \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation".
-Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have yet been undertaken on possible changes in diversity based on microscopy cell counts.
+Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts.
 Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community.
@@
 \subsection{In-situ biological and environmental data}
-    The CARIACO Ocean Time Series Program collected biological and environmental data, which we utilized here to examine connections between environmental and ecological change in the region.
+    The CARIACO Ocean Time Series Program collected biological and environmental data, which we used here to examine the connections between environmental and ecological change in the region.
 Samples were collected on monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.67}W) \cite{muller-karger_scientific_2019}.
@@
 
-    Between late 1995 and early 1998, nutrient data, including X, Y, and Z, were provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to  present by Dr.
+    Between late 1995 and early 1998, nutrient data, including nitrate, phosphate, and silicate, were provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and from mid-1997 to present by Dr.
 K.
@@
 Petersburg, USA).
-Samples from eight cruises were analyzed by both laboratories.
-Data for PO$_4$ and SiO$_4$ were in strong agreement (within ±5\%), whereas  NO$_\text{x}$ and NH$_4$ measurements showed discrepancies.
-\cite{taylor_ecosystem_2012}.
+Samples from eight cruises were analyzed by both laboratories \cite{taylor_ecosystem_2012}.
 We used a merged dataset for all nutrients, in which coincidental monthly measurements are averaged.
@@
     %from Mutshinda et al. 2013
-    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (\url{marinespecies.org}).
+    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (WoRMS, \url{marinespecies.org}).
 A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.
-
-    All in-situ data, inlcuding Niskin bottle, CTD, and phytoplankton taxonomy were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}.
+The identified genera were further grouped into functional groups based on their taxonomic classification in WoRMS.
+
+    All in situ data, including the Niskin bottle, CTD, and phytoplankton taxonomy, were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}.
     From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than three months) in coverage.
@@
 \subsection{Data analysis}    
-    To provide an overview of the available data and highlight the interannual trends and changes in the time series data, we calculated annual means for relevant variables and normalized the data range as z-scores ($z = \frac{(x – \mu)}{\sigma}$).
-Cruises from 1995, 2016, and 2017 were excluded because insufficient coverage (i.e.
-more than six missing monthly samples) in these years prevented us from producing reliable yearly mean estimates.
-
-    We calculated three metrics of phytoplankton community diversity.
-First, we calculated genus richness, which reduces taxonomic resolution but creates a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}.
-Second and third, we calculated the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, respectively, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
+    In the analysis, we set out to find if there is a structure in the distribution and temporal evolution of diversity metrics and the community composition and to identify potential bottom-up drivers of this change.
+    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where \mu is the mean across all available data points).
+Z scores from 1995, 2016, and 2017 were not shown because insufficient coverage (i.e.
+more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.
+
+    Cell counts of phytoplankton per functional group were assembled per monthly sampling, summed over all individual interpolated counts at the genus or species level, and finally averaged over a year.
+Nanoflagellate counts were reported as such in the data and not further identified.
+To characterize phytoplankton community structure, we calculated genus richness, Shannon diversity (H'), and Pielou’s evenness (J') to assess taxonomic variety, community complexity, and the distribution of abundances among genera, respectively.
+We calculated genus richness as the sum of all identified genera present in a monthly sample.
+Genus richness provides a lower taxonomic resolution than species richness, but creates a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}.
+The Shannon index was calculated as $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness as $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
+Diversity metrics were calculated for each monthly sample and averaged throughout the year.
 
@@
     To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis.
-Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables (here: monthly observed genus counts) to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}.
+Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}.
 Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor.
@@
 Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}.
-This analysis was conducted using monthly measurements integrated over a depth of 100 m for observed genus counts, ensuring the inclusion of all available data.
-The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO]}_3$, $\mathrm{[PO]}_4$ and $\mathrm{[SiO]}_4$).
+We used monthly genus level abundances integrated over the top \qty{100}{\meter} as indicator variables and limited the data to only those genera that were counted more than 10 times over the entire time series.
+This produced a data set of 193 monthly samples and 89 observed genera abundances, where all predictor variables were also available.
+The genus abundances were transformed by the natural logarithm plus one prior to model runs.
+The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO_3]}$, $\mathrm{[PO_4]}$ and $\mathrm{[SiO_4]}$).
+All predictor variables are in some way related to the local climate and upwelling system and could be potential bottom-up drivers of community change, some more directly (e.g.
+the isotherm and nutrient concentrations) and some more indirectly, such as the climate indices and meteorological data.
     The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}.
-    We show a model run without any time lags in  supplementary material (Figure \ref{fig:sup:GFoutput_nolags}).
-To consider the temporal lag in effect, particularly of climate indices, we included time lags of up to 6 months for all climate variables, as well as 3-month lags for in-situ measurements in exploratory runs.
+Gradient forest runs were performed with 1500 bootstrapped tress and a correlation threshold of 0.5.
+    We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}).
+To consider the temporal lag in effect, particularly of climate indices, we tested for an improved predictive power of time lags of up to 6 months for all climate variables, as well as 3-month lags for in situ measurements in exploratory runs.
 We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix).
+All the code and data used in this analysis and to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
 
@@
 
-Create subsections in results. This one could be "Interannual changes in the environment and ecosystem"
+\subsection{Interannual changes in the environment and ecosystem}
 % Figure 2
@@
 % Figure 6
-To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
+To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the monthly microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
 Preliminary model runs were used to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months.
@@
 The predictive power of the models for each genus is shown in Figure \ref{fig:GF}$b$.
-The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are from diatoms.
+The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are diatoms.
 However, two haptophyte and three dinoflagellate genera rank relatively high.
+For the presented run a total of 89 genera were used as indicator variables, of which 46 genera could be predicted using the given predictor variables, which corresponds to more than \qty{95}{\%} of counted cells.
 Figure \ref{fig:GF}$c$ shows the cumulative importance of the predictor variables across the parameter range.
@@
 
+%
+\begin{table}
+\caption{Table of all identified genus that were counted more than 10 times across the time series (n=89). The monthly cell counts per genus were used as input data to the gradient forest analysis.}
+% Please add the following required packages to your document preamble:
+% \usepackage{booktabs}
+
+\begin{tabular}{@{}ll@{}}
+\textbf{Diatoms} &                                                            \\
+\textit{Amphora}           & C.G. Ehrenberg ex F.T. Kützing, 1844                       \\
+\textit{Asterionella}      & A.H. Hassall, 1850                                         \\
+\textit{Asterionellopsis}  & F.E. Round in F.E. Round, R.M. Crawford \& D.G. Mann, 1990 \\
+\textit{Bacteriastrum}     & G. Shadbolt, 1854                                          \\
+\textit{Bellerochea}       & H. Van Heurck, 1885                                        \\
+\textit{Cerataulina}       & H. Peragallo ex F. Schütt in Engler \& Prantl, 1896        \\
+\textit{Chaetoceros}       & C.G. Ehrenberg, 1844                                       \\
+\textit{Climacodium}       & A. Grunow, 1868                                            \\
+\textit{Corethron}         & A.F. Castracane, 1886                                      \\
+\textit{Coscinodiscus}     & C.G. Ehrenberg, 1839                                       \\
+\textit{Cylindrotheca}     & L. Rabenhorst, 1859                                        \\
+\textit{Dactyliosolen}     & A.F. Castracane, 1886                                      \\
+\textit{Detonula}          & F. Schütt ex G.B. De Toni, 1894                            \\
+\textit{Diploneis}         & (C. G. Ehrenberg) P.T. Cleve, 1894                         \\
+\textit{Eucampia}          & C.G. Ehrenberg, 1839                                       \\
+\textit{Grammatophora}     & C.G. Ehrenberg, 1840                                       \\
+\textit{Guinardia}         & H. Peragallo, 1892                                         \\
+\textit{Haslea}            & R. Simonsen, 1974                                          \\
+\textit{Helicotheca}       & M. Ricard, 1987                                            \\
+\textit{Hemiaulus}         & P.A.C. Heiberg, 1863                                       \\
+\textit{Lauderia}          & P.T. Cleve, 1873                                           \\
+\textit{Leptocylindrus}    & P.T. Cleve in C.G.J. Petersen, 1889                        \\
+\textit{Lioloma}           & Hasle, 1997                                                \\
+\textit{Navicula}          & J.B.M. Bory de Saint-Vincent, 1822                         \\
+\textit{Nitzschia}         & A.H. Hassall, 1845                                         \\
+\textit{Odontella}         & C.A. Agardh, 1832                                          \\
+\textit{Pinnularia}        & C.G. Ehrenberg, 1843                                       \\
+\textit{Pinnunavis}        & H. Okuno, 1975                                             \\
+\textit{Planktoniella}     & (C.G.Wallich) Schütt, 1892                                 \\
+\textit{Pleurosigma}       & W. Smith, 1852                                             \\
+\textit{Proboscia}         & G. Sundström, 1986                                         \\
+\textit{Pseudo-nitzschia}  & H. Peragallo in H. Peragallo \& M. Peragallo, 1900         \\
+\textit{Rhizosolenia}      & T. Brightwell, 1858                                        \\
+\textit{Skeletonema}       & R.K. Greville, 1865                                        \\
+\textit{Sundstroemia}      & Medlin, Boonprakob, Lundholm \& Moestrup, 2021             \\
+\textit{Tabularia}         & (F.T. Kutzing) D.M. Williams \& F.E. Round, 1986           \\
+\textit{Thalassionema}     & A. Grunow ex C. Mereschkowsky, 1902                        \\
+\textit{Thalassiosira}     & P.T. Cleve, 1873 emend. Hasle, 1973                        \\
+\textbf{Cyanobacteria}     &                                                            \\
+\textit{Anabaena}          & Bory de Saint-Vincent ex Bornet \& Flahault, 1886          \\
+\textit{Johannesbaptistia} & G.De Toni, 1934                                            \\
+\textit{Lyngbya}           & C.Agardh ex Gomont, 1892                                   \\
+\textit{Synechococcus}     & Nägeli, 1849                                               \\
+\textit{Trichodesmium}     & Ehrenberg ex Gomont, 1892                                  \\
+\textbf{Dinoflagellates}    &                                                            \\
+\textit{Akashiwo}          & Gert Hansen \& Moestrup, 2000                              \\
+\textit{Alexandrium}       & Halim, 1960                                                \\
+\textit{Amphidinium}       & Claperède \& Lachmann, 1859                                \\
+\textit{Blepharocysta}     & Ehrenberg, 1873                                            \\
+\textit{Cochlodinium}      & Schütt, 1896                                               \\
+\textit{Cystodinium}       & Klebs, 1912                                                \\
+
+\end{tabular}
+
+\label{tab:sup:GenusFuncGroups}
+\end{table}
+
+\begin{table}
+
+\caption{Table \ref{tab:sup:GenusFuncGroups} cont.:}
+\begin{tabular}{@{}ll@{}}
+\textbf{Dinoflagellates} cont.    &                                                            \\
+\textit{Dinophysis}        & Ehrenberg, 1839                                            \\
+\textit{Diplopsalis}       & R.S.Bergh, 1881                                            \\
+\textit{Glenodinium}       & Ehrenberg, 1836     
+\\
+\textit{Gonyaulax}         & Diesing, 1866                                              \\
+\textit{Gymnodinium}       & F.Stein, 1878                                              \\
+\textit{Gyrodinium}        & Kofoid \& Swezy, 1921                                      \\
+\textit{Heterocapsa}       & F.Stein, 1883                                              \\
+\textit{Lingulodinium}     & D.Wall, 1967                                               \\
+\textit{Mesoporos}         & Lillick, 1937                                              \\
+\textit{Ornithocercus}     & Stein, 1883                                                \\
+\textit{Oxytoxum}          & Stein, 1883                                                \\
+\textit{Phalacroma}        & Stein, 1883                                                \\
+\textit{Oxyphysis}         & Kofoid, 1926                                               \\
+\textit{Podolampas}        & F.Stein, 1883                                              \\
+\textit{Pronoctiluca}      & Fabre-Domergue, 1889                                       \\
+\textit{Prorocentrum}      & Ehrenberg, 1834                                            \\
+\textit{Exuviella}         & Cienkowski, 1881                                           \\
+\textit{Protoperidinium}   & Bergh, 1881                                                \\
+\textit{Pyrocystis}        & J.Murray ex Haeckel, 1890                                  \\
+\textit{Pyrophacus}        & F.Stein, 1883                                              \\
+\textit{Scrippsiella}      & Balech ex A.R.Loeblich III, 1965                           \\
+\textit{Sphaerodinium}     & Woloszynska, 1916                                          \\
+\textit{Triadinium}        & Dodge, 1981                                                \\
+\textit{Tripos}            & Bory de Saint-Vincent, 1823                                \\
+\textbf{Haptophytes}        &                                                            \\
+\textit{Calcidiscus}       & Kamptner, 1950                                             \\
+\textit{Calciopappus}      & Gaarder \& Ramsfjell, 1954                                 \\
+\textit{Calciosolenia}     & Gran, 1912                                                 \\
+\textit{Calyptrosphaera}   & Lohmann, 1902                                              \\
+\textit{Coccolithus}       & E.H.L.Schwarz, 1894                                        \\
+\textit{Cyclolithella}     & Loeblich \& Tappan, 1963                                   \\
+\textit{Emiliania}         & W.W.Hay \& H.P.Mohler, 1967                                \\
+\textit{Michaelsarsia}     & Gran, 1912                                                 \\
+\textit{Ophiaster}         & Gran, 1912                                                 \\
+\textit{Syracosphaera}     & Lohmann, 1902                                              \\
+\textit{Umbellosphaera}    & Paasche, 1955                                              \\
+\textit{Umbilicosphaera}   & Lohmann, 1902                                              \\
+\textbf{Others}            &                                                            \\
+\textit{Dictyocha}         & Ehrenberg, 1837                                            \\
+\textit{Eutreptia}         & Perty, 1852                                                \\
+\textit{Eutreptiella}      & A.M.Cunha, 1913                                            \\
+\textit{Mesodinium}        & von Stein                                                 
+\end{tabular}
+\end{table}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2025-07-27 agujournaltemplate.tex
Results checked and Open Research Statement added

@@
 The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning to the previous state by 2014.
-Interestingly, species diversity declined between 1998 and 2009, although upwelling had not yet changed.
+Interestingly, community diversity declined between 1998 and 2009, although upwelling had not yet changed.
 Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate levels in the water column, were the strongest predictors of changes in the phytoplankton community.
@@
     The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work.
-Details about the sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.
+Details about sampling procedures, analytical methods, data quality control, and inter-calibration procedures have already been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.
 
@@
 
-    All in situ data, including the Niskin bottle, CTD, and phytoplankton taxonomy, were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}.
-    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than three months) in coverage.
+    All in situ data, including the Niskin bottle \cite{mullerkarger2019niskin}, CTD \cite{mullerkarger2019ctd}, and phytoplankton taxonomy \cite{troccoli2019phytoplankton}, were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}.
+    From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than six months) in coverage.
 In total, we used a complete set of 254 monthly data points from 1995 to 2017, for which all variables were available.
@@
 Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}.
-We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
+We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
 For simplicity, we refer to this variable as "wind speed" throughout the article.
@@
     In addition to the wind data, we examined two climatic indices, each of which provides information on the basin-scale climatic conditions and, consequently, the broad-scale forcing of marine ecosystems.
-The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies (\ang{0}N-\ang{60}N), were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.
-
-    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W), were obtained from \url{https://psl.noaa.gov/enso/mei/}.
+The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies (\ang{0}N-\ang{60}N) \cite{trenberth2023amo}, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.
+
+    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}, were obtained from \url{https://psl.noaa.gov/enso/mei/}.
 In our analysis, bi-monthly values were aligned with the tail month (December-January is treated as the value for January).
@@
 The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 \meter for further analysis.
-Additionally, sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}.
+Furthermore, the sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}.
 For phytoplankton microscopy data, individually identified taxonomic units (species or genus) were interpolated and integrated across depth, and finally grouped by genus via the sum of interpolated counts.
@@
     In the analysis, we set out to find if there is a structure in the distribution and temporal evolution of diversity metrics and the community composition and to identify potential bottom-up drivers of this change.
-    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where \mu is the mean across all available data points).
+    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where $\mu$ is the mean across all available data points).
 Z scores from 1995, 2016, and 2017 were not shown because insufficient coverage (i.e.
@@
     The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}.
-Gradient forest runs were performed with 1500 bootstrapped tress and a correlation threshold of 0.5.
+Gradient forest runs were performed with 1500 bootstrapped trees and a correlation threshold of 0.5.
     We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}).
@@
     \begin{table}
-
     \caption{Data and corresponding units used as predictor variables for the gradient forest analysis.}
@@
     % Figure 2
-Coordinated, interannual changes occur in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}).
+    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}).
 The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm).
 These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in Appendix).
-The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series but show differing interannual oscillations.
-Annual mean anomalies of nitrate and phosphate appear to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration.
-Silicate does not follow a clear pattern but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to reduced abundances of diatoms.
-<I think the most obvious pattern in Fig.
-2 has not yet been discussed: Chl and phyto are high in 1996-2004, then low in 2004-2013, followed by a switch in 2014-2015.
-Diversity is high when chl/phyto are low.
-I think you'll want to describe these patterns as they are interesting.>
+The yearly mean nutrient concentrations have a maximum positive anomaly at the beginning of the time series and are highly correlated.
+Annual mean anomalies of nitrate and phosphate are generally high during stronger upwelling, but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration.
+Silicate does not follow the pattern of nitrate or phosphate, but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to reduced abundances of diatoms.
+
+    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}).
+Distinct patterns emerge in wind speed, evaporation, precipitation, SST, and the \qty{21}{\celsius} isotherm depth.
+Periods of positive wind anomalies (blue) correspond to negative anomalies in SST and upwelling (red, indicated by a shallower isotherm depth), reflecting stronger upwelling conditions.
+These patterns are consistent with the Atlantic Multidecadal Oscillation (AMO), where a positive AMO phase is associated with weaker upwelling regimes (see also Figure~\ref{fig:sup:correlation} in the Appendix).
+Yearly mean nutrient concentrations show strong positive anomalies at the beginning of the time series and are generally well correlated.
+Nitrate and phosphate anomalies tend to be higher during periods of stronger upwelling but remain relatively low between 2002 and 2014, with nitrate showing more interannual variability.
+Silicate anomalies, in contrast, do not follow the same pattern as nitrate and phosphate.
+Instead, elevated concentrations are observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
+    Chlorophyll \textit{a} and functional group cell counts display positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift.
+This shift occurs during the period between 2003 and 2013, when the AMO index presents mostly positive anomalies, while upwelling-related variables indicate a weak upwelling regime (low wind, high SST, deep isotherm).
+Although data are sparse, nutrients and biological variables show a strong increase following 2013, coinciding with a return to negative anomalies in the AMO index and more upwelling-favorable conditions.
+    The annual means of diversity metrics do not follow this trend, but are reduced between 1998 and 2008.
+Shannon diversity specifically appears elevated past 2006 and the general pattern that emerges is that mean diversity is lower in years, where mean chlorophyll \textit{a} is high.
+However, in 1996 and 1997 positive anomalies in cell counts, chlorophyll \textit{a} and diversity metrics coincide.
 
@@
 Deeper shades indicate larger anomalies.
+Chlorophyll and functional group counts were log-transformed prior to normalisation.
 The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to insufficient monthly data to generate robust annual anomalies.
-Additionally we marked years with two or three and four or more missing monthly values.
-Diversity indices were calculated based on microscopy counts at the genus level.}
+Additionally we marked years with two or three and four or more missing monthly values.}
     \label{fig:zscore}
@@
 The bulk of the chlorophyll \textit{a} ($\sim$ 80\,\%) was measured in the upper 50 meters of the water column, with approximately 15\,\% measured between 50 and 75 meters.
-Chlorophyll \textit{a} exhibited marked seasonality, with the highest biomass in the upper water column during the early upwelling months (December–February, Figure \ref{fig:divts}$b$).
+Chlorophyll \textit{a} exhibited marked seasonality, with the highest biomass in the upper water column during the early upwelling season (December–February, Figure \ref{fig:divts}$b$).
 The yearly mean chlorophyll \textit{a} biomass showed the highest variability within the upper 25 meters.
 In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deep chlorophyll maximum in those years (Figure \ref{fig:divts}$a$).
-In general, yearly mean anomalies within the upper 75 meters appear to be high during the period from 1996 to 2004 and only reduced in 2012.
-The year 2013 was an outlier, particularly for the upper 25 meters, as the missing monthly sampling from the winter months (January and February) skewed the value downwards.
-Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower cholorophyll (65\,\% of the mean over the entire time series for these months), which recovers in 2014.
+The year 2013 was an outlier, particularly for the upper 25 meters, as missing monthly sampling from the upwelling season (January and February) skewed the value downwards.
+Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower cholorophyll (65\,\% of the mean over the entire time series for these months), which recovered in 2014.
 Chlorophyll \textit{a} biomass below 50 meters depth showed a relatively stable distribution throughout the year and across the time series.
-In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters is markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
+In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters was markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
 
     The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons in winter.
-Genus richness is consistently the highest diversity index in the top \qty{25}{meters} depth interval, and shows also the highest variability over the time series.
+Genus richness is consistently the highest diversity index in the top \qty{25}{meters} depth interval, also showing the highest variability over the time series.
 Across the time series, we observe maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, with a recovery after 2011.
 The reduction in diversity following the start of the time series does not follow the pattern shown by chlorophyll \textit{a}, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}).
-Scaling this genus diversity by the cell count abundances, i.e.
+Scaling genus richness by the cell count abundances, i.e.
 the Shannon index, shows a similar but less pronounced dynamic.
@@
 
-Create subsection like "Clustering analysis yields contrasting ecological regimes"
+\subsection{Clustering analysis yields contrasting ecological regimes}
     % Figure 4
@@
 
+\subsection{AMO and upwelling-related variables as drivers of community changes}
     % Figure 6
@@
 However, two haptophyte and three dinoflagellate genera rank relatively high.
-For the presented run a total of 89 genera were used as indicator variables, of which 46 genera could be predicted using the given predictor variables, which corresponds to more than \qty{95}{\%} of counted cells.
+For the presented run a total of 89 genera were used as indicator variables, of which 46 genera could be predicted using the given predictor variables, which corresponds to more than \qty{95}{\%} of counted cells (see Table \ref{tab:sup:GenusFuncGroups} for a complete list of identified genera).
     Figure \ref{fig:GF}$c$ shows the cumulative importance of the predictor variables across the parameter range.
@@
     % P2: Discuss community reorganisation, biomass shifts vs HPLC data 
-When comparing the two regimes, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a}, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}).
+    When comparing the two clusters, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a}, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}).
 The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community might have transitioned to smaller cells, which were probably below the detection size limit of optical microscopes \cite{muller-karger_scientific_2019}.
@@
     % P3: Shift in Diversity does not match shift in biomass/cell counts, why?
-The temporal dynamics we find in yearly mean genus richness, however, do not correspond directly to the observed regimes.
-Instead, we see a marked reduction from 1998 to 2010, after which richness recovers four years prior to 2014, when the community returns to levels characteristic of regime 1.
+    The temporal dynamics we find in yearly mean genus richness, however, do not correspond directly to the observed clusters.
+Instead, we see a marked reduction from 1998 to 2010, after which richness recovers four years prior to 2014, when the community returns to levels characteristic of cluster 1.
 The same trend is evident, albeit to a lesser degree, in the other diversity metrics.
 Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and exhibit a reduced yearly mean between 2000 and 2006, during which cell counts recorded an increase in nanoflagellates and a pronounced peak in cyanobacteria.
-We did not find a significant difference in diversity between the two regimes, either by Shannon index or by genus richness (Figures \ref{fig:zscore} and \ref{fig:divts}).
-However, we did observe a significant difference in Pielou's index between the two regimes.
-The results suggest that the phytoplankton community becomes more evenly distributed during low-upwelling conditions of regime 2, which is consistent with the analysis of \citeauthor{pinckney_phytoplankton_2015} based on HPLC pigment data.
+    We did not find a significant difference in diversity between the two clusters, either by Shannon index or by genus richness (Figures \ref{fig:zscore} and \ref{fig:divts}).
+However, we did observe a significant difference in Pielou's index between the two clusters.
+The results suggest that the phytoplankton community becomes more evenly distributed during low-upwelling conditions of cluster 2, which is consistent with the analysis of \citeauthor{pinckney_phytoplankton_2015} based on HPLC pigment data.
 This increase in evenness, quantified by the Pielou index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community.
-The resulting decrease in compositional turnover is also observed in a decline in interannual species turnover during regime 2 (see Figure \ref{fig:clustering}$b$).
+The resulting decrease in compositional turnover is also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering}$b$).
 As the productivity and biomass of phytoplankton communities are driven primarily by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
@@
 Diatoms do not exhibit a clear threshold response for the MEI v.2 index but are affected across the entire range.
-Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in regime 1 (see Figure \ref{fig:clustering} $e$).
-- Strong El Niño events increase the strength of currents in the Caribbean sea through modulating the easterly wind anomalies \cite{huang_enso_2023}.
+Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in cluster 1 (see Figure \ref{fig:clustering} $e$).
+This is consistent with previous findings, that strong El Niño events (as indicated by a positive MEI) increase the strength of currents in the Caribbean sea through modulating the easterly wind anomalies, thereby increasing upwelling in the Cariaco Basin \cite{huang_enso_2023}.
 
@@
 
-These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separates the two regimes is NMDS 1, which covaries most strongly with changes in AMO, MEI v.2, salinity, evaporation, and precipitation (Figure \ref{fig:clustering} $c$).
+    These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separates the two clusters is NMDS 1, which covaries most strongly with changes in AMO, MEI v.2, salinity, evaporation, and precipitation (Figure \ref{fig:clustering} $c$).
 Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS 2.
@@
     We found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco Basin.
-Specifically, microscopy cell counts reveal the presence of two distinct regimes defined by high- and low-upwelling conditions that correspond to anomalies in the AMO index.
-The community changes do not indicate a permanent regime shift, as a return to regime 1 is observed towards the end of the time series.
-We observe no significant changes in diversity metrics between the two regimes, except for lower turnover and higher community evenness under low-upwelling conditions.
+Specifically, microscopy cell counts reveal the presence of two distinct clusters defined by high- and low-upwelling conditions that correspond to anomalies in the AMO index.
+The community changes do not indicate a permanent regime shift, as a return to the early cluster is observed towards the end of the time series.
+We observe no significant changes in diversity metrics between the two clusters, except for lower turnover and higher community evenness under low-upwelling conditions.
     Despite multiannual trends in warming, reduced productivity, and drastic declines in microphytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions were restored.
-The strongest bottom-up variable predicting changes in the phytoplankton community was the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local conditions through low-frequency natural cycles.
+    The strongest variable predicting changes in the phytoplankton community was the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local conditions through low-frequency natural cycles.
 We concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state for the phytoplankton community of the Cariaco Basin.
@@
 \section*{Open Research Section}
-This section MUST contain a statement that describes where the data supporting the conclusions can be obtained.
-Data cannot be listed as ''Available from authors'' or stored solely in supporting information.
-Citations to archived data should be included in your reference list.
-Wiley will publish it as a separate section on the paper’s page.
-Examples and complete information are here:
-https://www.agu.org/Publish with AGU/Publish/Author Resources/Data for Authors
-
-\section*{As Applicable – Inclusion in Global Research Statement}
-The Authorship: Inclusion in Global Research policy aims to promote greater equity and transparency in research collaborations.
+All data used in this manuscript is publicly available with open access.
+The CARIACO time series data we used was accessed from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) under the doi \url{10.1575/1912/bco-dmo.3095.1}.
+This includes the Niskin bottle dataset for monthly in-situ measurements of chlorophyll \textit{a} and nutrient data \url{https://www.bco-dmo.org/dataset/3093}, the CTD dataset \url{https://www.bco-dmo.org/dataset/3092}, and the phytoplankton taxonomy dataset \url{https://www.bco-dmo.org/dataset/3095}.
+The climate variables, such as wind speed, precipitation and evaporation, were extracted from the "ERA5 monthly averaged data on single levels from 1940 to present" dataset \url{https://cds.climate.copernicus.eu/cdsapp#!/dataset/10.24381/cds.f17050d7}.
+The AMO index was retrieved from the Climate Data Guide provided by the NSF National Center for Atmospheric Research at \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} and the MEI v.2 was accessed from NOAA Physical Sciences Laboratory (PSL) at \url{https://psl.noaa.gov/enso/mei/}.
+All the code and data used in this analysis and to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
+%TODO add Zenodo link before submission
+
+%\section*{As Applicable – Inclusion in Global Research Statement}
+%The Authorship: Inclusion in Global Research policy aims to promote greater equity and transparency in research collaborations.
 AGU Publications encourage research collaborations between regions, countries, and communities and expect authors to include their local collaborators as co-authors when they meet the AGU Publications authorship criteria (described here: https://www.agu.org/publications/authors/policies\#authorship).

### 2025-07-27 agujournaltemplate.tex
Double checked and reread - off to Ago again

@@
 
-\title{Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela}
+\title{Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela}
 
@@
 Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
-Long-term time series of phytoplankton can capture changes in climate and ecosystems, allowing us to understand the effects of environmental change on local communities.
+Long-term time series can capture this variability, allowing us to untangle the effects of environmental change on local communities.
 This study investigates changes in the phytoplankton community of the Cariaco Basin, Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data.
-We considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in wind that drive upwelling in the Cariaco Basin.
+To investigate impacts of larger scale climate, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in wind that drive upwelling in the Cariaco Basin.
 Cluster analysis identified two distinct community states associated with a shift from high to low upwelling conditions in 2004, followed by a return to a community and conditions similar to those before the shift in 2014.
 In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
-To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2 index.
-Our study emphasizes the impact of long-term (decadal) climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the resilient nature of the tropical marine ecosystem under global change.
+To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2.
+Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the resilient nature of the tropical marine ecosystem under global change.
 
@@
 This study looked at how phytoplankton communities in the Cariaco Basin, off the coast of Venezuela, changed from 1995 to 2017.
-We compared microscopy cell counts and monthly water samples with weather data and two major climate indicators: the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to local wind and currents.
+We compared microscopy cell counts and monthly water samples with weather data and two major climate indicators: the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to local weather and currents.
 % Findings Summary
@@
 % Phytoplankton is important
-Phytoplankton provide the foundation for marine food webs and play crucial roles in biogeochemical processes, \cite{falkowski_biogeochemical_1998}.
+Phytoplankton provide the foundation for marine food webs and play crucial roles in biogeochemical processes \cite{falkowski_biogeochemical_1998}.
 Their short generation times and high assemblage diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state and dynamics \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
-Long-term time series of biological observations are crucial for understanding how phytoplankton communities respond to short-term processes and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
+Long-term time series of biological observations are crucial for understanding how phytoplankton communities respond to both short- and long-term changes in the physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
 Such time series are rare, particularly in the tropics \cite{clarke_does_2017}.
@@
 Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes.
-\citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to global warming.
+\citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to effects of global warming.
 During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
@@
 This strongly affected the phytoplankton community and the entire ecosystem.
-Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms, as confirmed, albeit not with the same high significance, by analysis of pigment data \cite{pinckney_phytoplankton_2015}.
+Microscopy cell counts indicated a drastic reduction of large phytoplankton, particularly diatoms \cite{pinckney_phytoplankton_2015}.
 In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, groups commonly found in stratified water columns \cite{lorenzoni_characterization_2015}.
@@
 However, the shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities from 2004 to 2013 did not appear to be a permanent change in regime.
-Data from 2013 to 2017, the end phase of the CARIACO time series, indicated a return to strong upwelling intensities and high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
+Data from 2013 to 2017, the final years of the CARIACO (CArbon Retention In A Colored Ocean) time series, indicated a return to strong upwelling intensities and high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
 The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear.
 The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
-The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.
+The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean Sea \cite{stephenson_changes_2014}.
 The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea, given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
@@
 Our study builds on and substantially expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017.
-Understanding the regional dynamics of the Cariaco Basin is crucial for unveiling the ecological changes affecting this tropical marine ecosystem and the variables driving these changes.
+In doing so, we hope to provide a more comprehensive assessment of long-term phytoplankton community dynamics in the Cariaco Basin, offering new insights into the ecological responses to environmental variability across decadal scales.
 
@@
     %from Mutshinda et al. 2013
-    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (WoRMS, \url{marinespecies.org}).
+    Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts that are shown in Figure \ref{fig:zscore}), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (WoRMS, \url{marinespecies.org}).
 A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.
@@
 \subsection{Data analysis}    
-    In the analysis, we set out to find if there is a structure in the distribution and temporal evolution of diversity metrics and the community composition and to identify potential bottom-up drivers of this change.
+    In the analysis, we set out to find if there is a structure in the distribution and temporal evolution of diversity metrics and the community composition, as well as to identify potential bottom-up drivers of this change.
     To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where $\mu$ is the mean across all available data points).
@@
     % Figure 2
-    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}).
-The patterns emerging from wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth clearly show the years of positive anomalies of wind (blue) and negative anomalies for SST (red) and upwelling (red, i.e., shallower depth of the isotherm).
-These patterns are reflected in the mode of the AMO, where a positive anomaly correlates with weak upwelling regimes (see also Figure \ref{fig:sup:correlation} in Appendix).
-The yearly mean nutrient concentrations have a maximum positive anomaly at the beginning of the time series and are highly correlated.
-Annual mean anomalies of nitrate and phosphate are generally high during stronger upwelling, but remain relatively low between 2002 and 2014, with some variability in mean phosphate concentration.
-Silicate does not follow the pattern of nitrate or phosphate, but shows an elevated mean concentration between 2008 and 2013, which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion due to reduced abundances of diatoms.
-
-    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables considered, the logarithmic transformation of chlorophyll \textit{a}, and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}).
+    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables, as well as log-transformed chlorophyll \textit{a} and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}).
 Distinct patterns emerge in wind speed, evaporation, precipitation, SST, and the \qty{21}{\celsius} isotherm depth.
 Periods of positive wind anomalies (blue) correspond to negative anomalies in SST and upwelling (red, indicated by a shallower isotherm depth), reflecting stronger upwelling conditions.
-These patterns are consistent with the Atlantic Multidecadal Oscillation (AMO), where a positive AMO phase is associated with weaker upwelling regimes (see also Figure~\ref{fig:sup:correlation} in the Appendix).
+These patterns are consistent with the Atlantic Multidecadal Oscillation (AMO), where a positive AMO phase is correlated with weaker upwelling regimes (see also Figure~\ref{fig:sup:correlation} in the Appendix).
 Yearly mean nutrient concentrations show strong positive anomalies at the beginning of the time series and are generally well correlated.
-Nitrate and phosphate anomalies tend to be higher during periods of stronger upwelling but remain relatively low between 2002 and 2014, with nitrate showing more interannual variability.
-Silicate anomalies, in contrast, do not follow the same pattern as nitrate and phosphate.
-Instead, elevated concentrations are observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
-    Chlorophyll \textit{a} and functional group cell counts display positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift.
+Nitrate and phosphate anomalies tended to be higher during periods of stronger upwelling but remained relatively low between 2002 and 2014, with nitrate showing more interannual variability.
+Silicate anomalies, in contrast, did not follow the same pattern as nitrate and phosphate.
+Instead, elevated concentrations were observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
+    Chlorophyll \textit{a} and functional group cell counts show positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift.
 This shift occurs during the period between 2003 and 2013, when the AMO index presents mostly positive anomalies, while upwelling-related variables indicate a weak upwelling regime (low wind, high SST, deep isotherm).
@@
 Deeper shades indicate larger anomalies.
-Chlorophyll and functional group counts were log-transformed prior to normalisation.
-The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to insufficient monthly data to generate robust annual anomalies.
+Chlorophyll and functional group counts were log-transformed prior to normalization.
+The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to insufficient monthly data to generate robust annual means.
 Additionally we marked years with two or three and four or more missing monthly values.}
@@
 The yearly mean chlorophyll \textit{a} biomass showed the highest variability within the upper 25 meters.
-In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deep chlorophyll maximum in those years (Figure \ref{fig:divts}$a$).
+In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deeper chlorophyll maximum in those years (Figure \ref{fig:divts}$a$).
 The year 2013 was an outlier, particularly for the upper 25 meters, as missing monthly sampling from the upwelling season (January and February) skewed the value downwards.
-Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower cholorophyll (65\,\% of the mean over the entire time series for these months), which recovered in 2014.
+Nevertheless, the monthly chlorophyll \textit{a} measurements for the remainder of the year 2013 were markedly lower (65\,\% of the mean over the entire time series for these months), which recovered in 2014.
 Chlorophyll \textit{a} biomass below 50 meters depth showed a relatively stable distribution throughout the year and across the time series.
@@
 
-    The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons in winter.
-Genus richness is consistently the highest diversity index in the top \qty{25}{meters} depth interval, also showing the highest variability over the time series.
-Across the time series, we observe maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, with a recovery after 2011.
-The reduction in diversity following the start of the time series does not follow the pattern shown by chlorophyll \textit{a}, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}).
+    The calculated genus diversity from the microscopy cell counts generally followed the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons in winter.
+Genus richness was consistently highest in the top \qty{25}{meters} depth interval, also showing the highest variability over the time series.
+Across the time series, we observed maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, followed by a recovery after 2011.
+The reduction in diversity following the start of the time series did not follow the pattern shown by chlorophyll \textit{a}, which remained relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}).
 Scaling genus richness by the cell count abundances, i.e.
-the Shannon index, shows a similar but less pronounced dynamic.
-Diversity remains high throughout 2000 and recovers more quickly but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016.
-The seasonal signal is lost (see Figure \ref{fig:divts} $f$).
+the Shannon index, showed a similar but less pronounced dynamic.
+Diversity remained high throughout 2000 and recovered more quickly but showed marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016.
+The seasonal signal was lost (see Figure \ref{fig:divts} $f$).
 
@@
 The inter-annual dynamics of evenness are remarkably stable across all depth intervals, clustered around an intermediate 0.5 value.
-There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016.
-Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.
+There were marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009, and 2016.
+Throughout 2003 to 2013 evenness was almost consistently highest in the deepest interval below 75 meters, where the community appeared to have been more evenly distributed.
 
@@
     Inter-annual species turnover is shown in Figure \ref{fig:clustering}$b$, it reveals an increasing trend in turnover from 1996 to 2004.
-Turnover subsequently decreases during the period underpinning cluster 2 and increases again in 2014, marking a return to cluster 1.
+Turnover subsequently decreased during the period underpinning cluster 2 and increased again in 2014, marking a return to cluster 1.
 This pattern is also reflected in the hierarchical clustering results, Figure \ref{fig:clustering}$a$, where cluster 2 shows a lower distance in the clustering tree between the years.
@@
 The diatom \textit{Thalassiosira} dominated the early period and was abundant in all other periods.
-\textit{Pseudo-nitzschia} dominates cluster 2, with the haptophyte \textit{Emiliana} being the second most abundant genus.
+\textit{Pseudo-nitzschia} dominated cluster 2, with the haptophyte \textit{Emiliana} being the second most abundant genus.
 The later part of cluster 1 is dominated by \textit{Emiliana}.
@@
 Upwelling-related variables exhibit distinct patterns between the two clusters.
-For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celsius} isotherm, lower temperatures, reduced evaporation, and higher salinity and nutrient concentrations, particularly for nitrate.
+For example, cluster 1 experienced more frequent high wind events, a shallower \qty{21}{\celsius} isotherm, lower temperatures, reduced evaporation, and higher salinity and nutrient concentrations, particularly for nitrate.
 The AMO index shows a pronounced separation between the two clusters, with cluster 1 exhibiting more negative anomalies, while cluster 2 shows a higher occurrence of positive anomalies.
@@
     To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the monthly microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
-Preliminary model runs were used to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months.
+Preliminary model runs were made to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months.
 Figure \ref{fig:GF}$a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity.
@@
 However, two haptophyte and three dinoflagellate genera rank relatively high.
-For the presented run a total of 89 genera were used as indicator variables, of which 46 genera could be predicted using the given predictor variables, which corresponds to more than \qty{95}{\%} of counted cells (see Table \ref{tab:sup:GenusFuncGroups} for a complete list of identified genera).
+For the presented run a total of 89 genera (each observed more than 10 times over the time series) were used as indicator variables, of which 46 genera could be predicted using the given predictor variables, which corresponds to more than \qty{95}{\%} of counted cells (see Table \ref{tab:sup:GenusFuncGroups} for a complete list of identified genera).
     Figure \ref{fig:GF}$c$ shows the cumulative importance of the predictor variables across the parameter range.
@@
 After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies up to 2014.
-    By considering the phytoplankton community time series until the end of 2017, the temporally complete dataset, we find that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}.
+    By considering the phytoplankton community time series until the end of 2017, the temporally complete dataset, we found that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}.
 In 2014, physical conditions in the Cariaco Basin returned to a regime of increased wind speed, increased upwelling, and elevated nutrient concentrations (Figure \ref{fig:zscore}).
-Concurrently, we observe an increase in chlorophyll \textit{a} (Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (Figure \ref{fig:clustering} $a$ and $c$).
+Concurrently, we observed an increase in chlorophyll \textit{a} (Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (Figure \ref{fig:clustering} $a$ and $c$).
 
     % P2: Discuss community reorganisation, biomass shifts vs HPLC data 
-    When comparing the two clusters, we observe a marked reduction in cell counts, which corresponds to a decreased concentration of chlorophyll \textit{a}, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}).
+    When comparing the two clusters, we observed a marked reduction in cell counts, which corresponded to a decreased concentration of chlorophyll \textit{a}, particularly in the top \qty{25}{meters} of the water column (see Figure \ref{fig:divts}).
 The decline in cell counts was not proportional to the more modest decrease in total chlorophyll \textit{a}, suggesting that the phytoplankton community might have transitioned to smaller cells, which were probably below the detection size limit of optical microscopes \cite{muller-karger_scientific_2019}.
@@
     % Disucssion of HPLC vs Microscopy!
-    A similar pattern emerges when comparing our observations with High-Performance Liquid Chromatography (HPLC) data from the Cariaco Basin, which were analyzed and presented by \citeA{pinckney_phytoplankton_2015}.
-The HPLC data suggest an increase in chlorophyll \textit{a} between the periods 1996-2000 and 2006-2010, particularly at depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}.
-Despite cell counts of all functional groups decreasing, the HPLC data indicate an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{\meter}.
-This is a further strong indication that the phytoplankton community shifted to smaller sizes, undetected using light microscopy.
+    A similar pattern emerged when comparing our observations with those based on High-Performance Liquid Chromatography (HPLC) data from the Cariaco Basin, which were analyzed and presented by \citeA{pinckney_phytoplankton_2015}.
+However, there were some contrasting observations.
+The HPLC data suggested an increase in chlorophyll \textit{a} between the periods 1996-2000 and 2006-2010, particularly at depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}.
 It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006.
@@
 This is why we focused our analysis on the more consistent and temporally complete microscopy cell count and fluorometric chlorophyll \textit{a} data.
-    Consistent with our results, independent spectrophotometric measurements of chlorophyll \textit{a} concentrations at \qty{20}{meters} depth, off the coast of Margarita Island, show a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
+    Consistent with our results, independent spectrophotometric measurements of chlorophyll \textit{a} concentrations at \qty{20}{meters} depth, off the coast of Margarita Island, showed a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
+    Additionally, despite cell counts of all functional groups decreasing, the HPLC data indicated an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{\meter}.
+This provides a further strong indication that the phytoplankton community shifted to smaller sizes, undetected using light microscopy.
 
     % P3: Shift in Diversity does not match shift in biomass/cell counts, why?
-    The temporal dynamics we find in yearly mean genus richness, however, do not correspond directly to the observed clusters.
-Instead, we see a marked reduction from 1998 to 2010, after which richness recovers four years prior to 2014, when the community returns to levels characteristic of cluster 1.
-The same trend is evident, albeit to a lesser degree, in the other diversity metrics.
-Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and exhibit a reduced yearly mean between 2000 and 2006, during which cell counts recorded an increase in nanoflagellates and a pronounced peak in cyanobacteria.
+    The temporal dynamics found in yearly mean genus richness, however, did not correspond directly to the observed clusters.
+Instead, there was a marked reduction from 1998 to 2010, after which richness recovered four years prior to 2014, when the community returned to levels characteristic of cluster 1.
+The same trend was evident, albeit to a lesser degree, in the other diversity metrics.
+Generally, mean annual anomalies in Shannon diversity and Pielou's evenness correlated with genus richness and exhibited a reduced yearly mean between 2000 and 2006, during which cell counts recorded an increase in nanoflagellates and a pronounced peak in cyanobacteria.
     We did not find a significant difference in diversity between the two clusters, either by Shannon index or by genus richness (Figures \ref{fig:zscore} and \ref{fig:divts}).
 However, we did observe a significant difference in Pielou's index between the two clusters.
-The results suggest that the phytoplankton community becomes more evenly distributed during low-upwelling conditions of cluster 2, which is consistent with the analysis of \citeauthor{pinckney_phytoplankton_2015} based on HPLC pigment data.
+The results suggest that the phytoplankton community became more evenly distributed during low-upwelling conditions of cluster 2, which is consistent with the analysis of \citeauthor{pinckney_phytoplankton_2015} based on HPLC pigment data.
 This increase in evenness, quantified by the Pielou index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community.
-The resulting decrease in compositional turnover is also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering}$b$).
-As the productivity and biomass of phytoplankton communities are driven primarily by nutrients supplied through disturbance events, this confirms the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
+The resulting decrease in compositional turnover was also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering}$b$).
+As the productivity and biomass of phytoplankton communities are driven primarily by nutrients supplied through disturbance events, this is in line with the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
 
@@
 Interestingly, our analysis revealed that the strongest predictor of change within the community was not an in situ measurement but rather the AMO climate index.
-The AMO is followed in importance by SST, MEI v.2 index, and nitrate concentration (see Figure \ref{fig:GF}).
+The AMO is followed in importance by SST, MEI v.2 index, and nitrate concentration as predictive variables (see Figure \ref{fig:GF}).
 
@@
 In the Cariaco Basin, a southward shift of the ITCZ correlates with increased wind-driven upwelling \cite{taylor_ecosystem_2012}.
-The threshold responses between functional groups, which relate to shifts in cell count abundances, show contrasting patterns.
-Diatoms, dinoflagellates, and haptophytes exhibit strong response thresholds for negative AMO anomalies, which could point to the influence of stronger upwelling conditions.
-The abundance of haptophytes, additionally, shows a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} based on HPLC data that haptophytes predominantly exhibit blooms above the mixed layer and at a lower frequency than other functional groups.
+The threshold responses between functional groups, which relate to shifts in cell count abundances, showed contrasting patterns.
+Diatoms, dinoflagellates, and haptophytes exhibited strong response thresholds for negative AMO anomalies, which pointed to the influence of stronger upwelling conditions.
+The abundance of haptophytes, additionally, showed a threshold effect at positive AMO anomalies, which may have been related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} based on HPLC data that haptophytes predominantly exhibited blooms above the mixed layer and at a lower frequency than other functional groups.
 
@@
 
-    The in-situ measurements that best predict changes in the microphytoplankton community were SST, nitrate concentration, and salinity, in that order.
+    The in-situ measurements that best predicted changes in the microphytoplankton community were SST, nitrate concentration, and salinity, in that order.
 Nitrate is a crucial nutrient for phytoplankton growth and has been identified as the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}.
-For diatoms, we observe a strong threshold response above \qty{7}{\micro \molar}, indicating that intense upwelling conditions drive blooms, resulting in predictable changes in cell counts.
-Similarly, haptophytes and diatoms exhibit a threshold around \qtyrange{22}{24}{\celsius}, indicating thresholds at low surface temperatures driven by stronger upwelling.
-
-    These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separates the two clusters is NMDS 1, which covaries most strongly with changes in AMO, MEI v.2, salinity, evaporation, and precipitation (Figure \ref{fig:clustering} $c$).
-Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS 2.
-Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, show greater variance along NMDS 1.
-Salinity, evaporation, and silicate concentration exhibit some explanatory power in the gradient forest analysis, although they are weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}).
+For diatoms, a strong threshold response was observed above \qty{7}{\micro \molar}, indicating that intense upwelling conditions drove blooms, resulting in predictable changes in cell counts.
+Similarly, haptophytes and diatoms exhibited a threshold around \qtyrange{22}{24}{\celsius}, indicating thresholds at low surface temperatures driven by stronger upwelling.
+
+    These results matched the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separated the two clusters was NMDS 1, which covaried most strongly with changes in AMO, MEI v.2, salinity, evaporation, and precipitation (Figure \ref{fig:clustering} $c$).
+Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations varied more strongly along NMDS 2.
+Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, showed greater variance along NMDS 1.
+Salinity, evaporation, and silicate concentration exhibited some explanatory power in the gradient forest analysis, although they were weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}).
 This implies that the climate indices explain variability that more direct measurements of upwelling intensity do not capture.
@@
     % 1.  Discuss Limitations of Light microscopy cell counts in CARIACO Time Series
-    Using microscopy cell counts as a basis for our analyses means that the results should be interpreted in terms of changes in the microphytoplankton community.
-Our choice of grouping cell counts by genus is based on the assumption that species within a genus generally exhibit similar eco-physiological responses.
-The output of the gradient forest model is further grouped by broad functional types to facilitate the identification of shifts and trends.
-The phytoplankton genera that can be best predicted are mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (Figure \ref{fig:GF} $b$).
+    Using microscopy cell counts as a basis for our analyses meant that the results should be interpreted in terms of changes in the microphytoplankton community.
+Our choice of grouping cell counts by genus was based on the assumption that species within a genus generally exhibit similar eco-physiological responses.
+The output of the gradient forest model was further grouped by broad functional types to facilitate the identification of shifts and trends.
+The phytoplankton genera that could be best predicted were mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (Figure \ref{fig:GF} $b$).
 This does not represent a complete picture of the phytoplankton community of the CARIACO station.
-The HPLC pigment data used by \citeauthor{pinckney_phytoplankton_2015} capture all size classes of phytoplankton, but, as previously discussed, it does not cover the entire time series and the dataset was assembled by four different laboratories.
-With the caveat that the microscopy counts capture microphytoplankton almost exclusively, the discrepancies to the fluorometric chlorophyll \textit{a} measurements can be explained.
-Drastic declines in cell counts reflect shifts to smaller species, an aspect consistent with size spectra calculated by previous studies using HPLC pigment data \cite{lorenzoni_characterization_2015}.
+The HPLC pigment data used by \citeauthor{pinckney_phytoplankton_2015} captured all size classes of phytoplankton, but, as previously discussed, it did not cover the entire time series and the dataset was assembled by four different laboratories.
+With the caveat that the microscopy counts captured microphytoplankton almost exclusively, the discrepancies with the HPLC pigment analysis can be explained.
+Drastic declines in cell counts reflected shifts to smaller species, an aspect consistent with size spectra calculated by previous studies using HPLC pigment data \cite{lorenzoni_characterization_2015}.
 
     % 2. Discuss limitations of GF analysis & Limitation to bottom-up drivers and climate indices
-    The limited scope of the microscopy cell counts also affects the conclusions that can be drawn from the Gradient Forest analysis, as these results can only be interpreted as concerning microphytoplankton.
+    The limited scope of the microscopy cell counts also affected the conclusions that could be drawn from the Gradient Forest analysis, as these results could only be interpreted as concerning microphytoplankton.
     Another potential caveat of our methodology is that many of the predictors in ecological time series and in this study are highly correlated (Figure \ref{fig:sup:correlation} in Appendix).
@@
 Without adjusting for time lags, SST and nitrate concentration are the strongest predictors of community change (Figure \ref{fig:sup:GFoutput_nolags} in Appendix).
-Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and AMO index as the strongest predictors yield a relatively low value around \qty{0.03}{}.
-This is expected given the relatively noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the gradient forest algorithm \cite{pitcher_example_2012, roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}.
+Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and the AMO index yielded a relatively low value around \qty{0.03}{}.
+This is expected given the noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the gradient forest algorithm \cite{pitcher_example_2012, roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}.
 It is important to consider that although predictive relationships between drivers and ecological predictors are identified, this does not necessarily imply a mechanistic relationship.
@@
 
-    Our focus on bottom-up drivers of community change may also restrict the explanatory power of our analysis, as top-down processes such as zooplankton grazing may impact the structure of the phytoplankton community \cite{frank_ups_2007, banas_adding_2011, acevedo-trejos_mechanisms_2015}.
+    Our focus on bottom-up drivers of community change may have also restricted the explanatory power of our analysis, as top-down processes such as zooplankton grazing may have impacted the structure of the phytoplankton community \cite{frank_ups_2007, banas_adding_2011, acevedo-trejos_mechanisms_2015}.
 Due to the scarcity of data covering higher trophic levels and the absence of zooplankton data for the first six years of the time series, we suggest that mechanistic modeling is the ideal approach for investigating the relative impacts of both top-down and bottom-up drivers.
@@
     We found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco Basin.
-Specifically, microscopy cell counts reveal the presence of two distinct clusters defined by high- and low-upwelling conditions that correspond to anomalies in the AMO index.
-The community changes do not indicate a permanent regime shift, as a return to the early cluster is observed towards the end of the time series.
-We observe no significant changes in diversity metrics between the two clusters, except for lower turnover and higher community evenness under low-upwelling conditions.
+Specifically, microscopy cell counts revealed the presence of two distinct clusters defined by high- and low-upwelling conditions that corresponded to anomalies in the AMO index.
+The community changes did not indicate a permanent regime shift, as a return to the early cluster was observed towards the end of the time series.
+We observed no significant changes in diversity metrics between the two clusters, except for lower turnover and higher community evenness under low-upwelling conditions.
     Despite multiannual trends in warming, reduced productivity, and drastic declines in microphytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions were restored.

### 2025-07-28 agujournaltemplate.tex
Final edits, acknowledgements etc.

@@
     %from Taylor et al. 2012
-    In addition to oceanographic environmental data, we also examine meteorological and climatic data that are known to influence ocean conditions and marine ecosystems.
+    In addition to the in situ data, we also examine meteorological and climatic data that are known to influence ocean conditions and marine ecosystems.
 Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}.
-We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
+We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
 For simplicity, we refer to this variable as "wind speed" throughout the article.
@@
 National Oceanographic Data Center \cite{johnson2006world}.
-The interpolated values were depth integrated to 100 meters and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 \meter for further analysis.
+The interpolated values were depth integrated to \qty{100}{\meter} and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis.
 Furthermore, the sea surface temperature (SST) was calculated from CTD temperature measurements averaged over the top \qty{10}{\meter}.
@@
     In the analysis, we set out to find if there is a structure in the distribution and temporal evolution of diversity metrics and the community composition, as well as to identify potential bottom-up drivers of this change.
-    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where $\mu$ is the mean across all available data points).
-Z scores from 1995, 2016, and 2017 were not shown because insufficient coverage (i.e.
+    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where $x$ is the yearly mean, $\mu$ is the mean and $\sigma$ is the standard deviation of all yearly means).
+Data from 1995, 2016, and 2017 were not included because insufficient coverage (i.e.
 more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.
@@
 Genus richness provides a lower taxonomic resolution than species richness, but creates a more homogeneous metric that is robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}.
-The Shannon index was calculated as $H' = -\sum_i p_i \log_{e} p_i$ and Pielou's evenness as $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.
+The Shannon index was calculated as $H' = -\sum_i [p_i \log_{e}(p_i)]$ and Pielou's evenness as $J' = -\sum_i[ p_i \log_{e}( p_i )/\log_{e}(H')]$, where $p_i$ represents the number of genera observed for each interpolated and integrated monthly sampling.
 Diversity metrics were calculated for each monthly sample and averaged throughout the year.
@@
 \acknowledgments
-Thank you to all who have helped, communicated, provided data. Claudia Benitez-Nelson, Digna Rueda-Roa, James Pinckney.
+We thank Digna Rueda-Roa, Claudia Benitez-Nelson, and James Pinckney for their valuable input during the early stages of this research.
 

### 2025-08-04 agujournaltemplate.tex
Ago final comments

@@
 
-\authors{Benjamin Post\affil{1,2}, Esteban Acevedo-Trejos\affil{3}, Subhendu Chakraborty\affil{1}, Andrew D. Barton\affil{4}, Agostino Merico\affil{1}}
+\authors{Benjamin Post\affil{1,2}, Esteban Acevedo-Trejos\affil{3}, Subhendu Chakraborty\affil{1}, Andrew D. Barton\affil{4}, Agostino Merico\affil{1,2,5}}
 
@@
 \affiliation{4}{Scripps Institution of Oceanography and Department of Ecology, Behavior and Evolution, University of California San Diego, La Jolla, CA, United States}
+\affiliation{5}{Faculty of Biology and Chemistry, University of Bremen, Germany}
 %(repeat as many times as is necessary)
@@
 \begin{keypoints}
-\item The CARIACO phytoplankton community time series reveals two clusters, separated by a shift from high- to low-upwelling conditions after 2004 and a return to high-upwelling conditions following 2013.
-\item No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low-upwelling conditions.
+\item The CARIACO phytoplankton community time series is characterized by two distinct temporal clusters separated by a shift in 2004 from high to low upwelling and by a return to high upwelling conditions in 2014.
+\item No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low upwelling conditions.
 \item Gradient forest analysis identifies the AMO climate index as the strongest predictor of community change, highlighting the influence of large-scale climatic oscillations on this tropical coastal ecosystem.
@@
 Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
-Long-term time series can capture this variability, allowing us to untangle the effects of environmental change on local communities.
-This study investigates changes in the phytoplankton community of the Cariaco Basin, Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data.
-To investigate impacts of larger scale climate, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in wind that drive upwelling in the Cariaco Basin.
-Cluster analysis identified two distinct community states associated with a shift from high to low upwelling conditions in 2004, followed by a return to a community and conditions similar to those before the shift in 2014.
+Long-term time series can capture this variability, allowing us to unravel the effects of environmental change on local communities.
+This study investigates changes in the phytoplankton community of the Cariaco Basin, off the coast of Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data.
+To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in wind that drive upwelling in the Cariaco Basin.
+Cluster analysis identified two distinct community states separated by a shift from high to low upwelling conditions in 2004.
+In 2014, community and environmental conditions returned to a state similar to that before the shift in 2004.
 In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
-To quantify the effect of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2.
-Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and highlights the resilient nature of the tropical marine ecosystem under global change.
+To quantify the effects of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2.
+Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.
 
@@
 % Topic Overview
-Phytoplankton are microscopic algae that form the foundation of the ocean food web.
-Because they respond quickly to environmental changes, studying them can show how marine ecosystems are affected by climate.
+Phytoplankton are photosynthesizing unicellular microorganisms that form the foundation of the ocean food web.
+Since they respond quickly to environmental changes, studying them can show how marine ecosystems are affected by climate.
 % Paper Overview
-This study looked at how phytoplankton communities in the Cariaco Basin, off the coast of Venezuela, changed from 1995 to 2017.
+This study investigated how phytoplankton communities in the Cariaco Basin, off the coast of Venezuela, changed from 1995 to 2017.
 We compared microscopy cell counts and monthly water samples with weather data and two major climate indicators: the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to local weather and currents.
 % Findings Summary
-The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning to the previous state by 2014.
-Interestingly, community diversity declined between 1998 and 2009, although upwelling had not yet changed.
-Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate levels in the water column, were the strongest predictors of changes in the phytoplankton community.
+The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning in 2014 to the state antecedent 2004.
+Interestingly, community diversity declined between 1998 and 2009 independently from changes in the upwelling regime.
+Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate concentrations in the water column, were the strongest predictors of changes in phytoplankton community.
 % Key Takeaways
@@
 Phytoplankton provide the foundation for marine food webs and play crucial roles in biogeochemical processes \cite{falkowski_biogeochemical_1998}.
-Their short generation times and high assemblage diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state and dynamics \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
+Their short generation times and high taxonomic and functional group diversity allow phytoplankton communities to respond relatively quickly to environmental changes, making them suitable indicators of ecosystem state and dynamics \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
 Long-term time series of biological observations are crucial for understanding how phytoplankton communities respond to both short- and long-term changes in the physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}.
@@
 % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
-However, the shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities from 2004 to 2013 did not appear to be a permanent change in regime.
-Data from 2013 to 2017, the final years of the CARIACO (CArbon Retention In A Colored Ocean) time series, indicated a return to strong upwelling intensities and high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
+The shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities from 2004 to 2013 did not appear to be a permanent change in regime.
+Data from 2014 to 2017, the final years of the CARIACO (CArbon Retention In A Colored Ocean) time series, indicated a return to strong upwelling intensities and high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
 The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear.
@@
 % What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
-\citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation".
-Previous analyses of phytoplankton community diversity in the Cariaco Basin did not include time series data beyond 2011, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts.
-Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community data and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community.
+\citeA{taylor_ecosystem_2012} noted that the observed changes could not be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation".
+But, to date, all analyses of phytoplankton community diversity in the Cariaco Basin do not include time series data beyond 2011, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts.
+Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community.
 Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution.
 Our study builds on and substantially expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017.
-In doing so, we hope to provide a more comprehensive assessment of long-term phytoplankton community dynamics in the Cariaco Basin, offering new insights into the ecological responses to environmental variability across decadal scales.
+In doing so, we aim to provide a more comprehensive assessment of long-term phytoplankton community dynamics in the Cariaco Basin, offering new insights into the ecological responses to environmental variability across decadal scales.
 
@@
 The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin situated on the northeastern Venezuelan continental shelf, with a sill to the north running almost parallel to the coastline at a mean depth of around \qty{100}{m}.
-This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the BATS and HOT sites, resulting in the formation of anoxic conditions below \qty{\sim 275}{m} depth.
+This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the BATS and HOT sites, resulting in the formation of anoxic conditions below a depth of  \qty{\sim 275}{m}.
 
@@
 \noindent\includegraphics[width=\textwidth]{fig/Map_CAR.pdf}
-\caption{Bathymetric map of the CARIACO Basin in the Caribbean Ocean. The station is located at \ang{10.50}N, \ang{64.67}W).}
+\caption{Bathymetric map of the CARIACO Basin in the Caribbean Ocean. The CARIACO mooring station is located at \ang{10.50}N, \ang{64.67}W).}
 \label{fig:map}
@@
     The CARIACO Ocean Time Series Program collected biological and environmental data, which we used here to examine the connections between environmental and ecological change in the region.
-Samples were collected on monthly research cruises at a station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.67}W) \cite{muller-karger_scientific_2019}.
+Samples were collected on monthly research cruises at a mooring station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.67}W) \cite{muller-karger_scientific_2019}.
     The CARIACO field program consisted of a monthly core oceanographic cruise with Niskin bottle and CTD deployments, along with other deployments and samples not utilized in this work.
@@
 
-    All in situ data, including the Niskin bottle \cite{mullerkarger2019niskin}, CTD \cite{mullerkarger2019ctd}, and phytoplankton taxonomy \cite{troccoli2019phytoplankton}, were retrieved from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}.
+    All in situ data, including the Niskin bottle \cite{mullerkarger2019niskin}, CTD \cite{mullerkarger2019ctd}, and phytoplankton taxonomy \cite{troccoli2019phytoplankton}, were obtained from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}.
     From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than six months) in coverage.
@@
     %from Taylor et al. 2012
-    In addition to the in situ data, we also examine meteorological and climatic data that are known to influence ocean conditions and marine ecosystems.
+    In addition to the in situ data, we also examined meteorological and climatic data that are known to influence ocean conditions and marine ecosystems.
 Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}.
@@
 \subsection{Data analysis}    
-    In the analysis, we set out to find if there is a structure in the distribution and temporal evolution of diversity metrics and the community composition, as well as to identify potential bottom-up drivers of this change.
+    In the analysis, we set out to find if there was a structure in the distribution and temporal evolution of diversity metrics and community composition and, additionally, to identify potential bottom-up drivers of this change.
     To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where $x$ is the yearly mean, $\mu$ is the mean and $\sigma$ is the standard deviation of all yearly means).
@@
     \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
-    To quantify the effects of bottom-up drivers in the Cariaco Basin on shifts in the phytoplankton community, we employed a Gradient Forest analysis.
+    To quantify the effects of bottom-up drivers on shifts in the phytoplankton community of the Cariaco Basin, we employed a Gradient Forest analysis.
 Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}.
@@
     The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}.
-Gradient forest runs were performed with 1500 bootstrapped trees and a correlation threshold of 0.5.
+Gradient Forest runs were performed with 1500 bootstrapped trees and a correlation threshold of 0.5.
     We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}).
@@
 We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix).
-All the code and data used in this analysis and to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
+All the codes and data used in this analysis and used to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
 
     \begin{table}
-    \caption{Data and corresponding units used as predictor variables for the gradient forest analysis.}
+    \caption{Data and corresponding units used as predictor variables for the Gradient Forest analysis.}
     \centering
@@
     Salinity & PSU & Niskin, interpolated down to \qty{100}{\meter} depth \\
-    $\mathrm{NO}_3$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
-    $\mathrm{PO}_4$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
-    $\mathrm{SiO}_4$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
+    $\mathrm{[NO_3]}$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
+    $\mathrm{[PO_4]}$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
+    $\mathrm{[SiO_4]}$  & µM & Niskin, interpolated down to \qty{100}{\meter} depth \\
     \end{tabular}}
@@
     % Figure 2
-    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables, as well as log-transformed chlorophyll \textit{a} and phytoplankton cell counts per functional group (Figure \ref{fig:zscore}).
-Distinct patterns emerge in wind speed, evaporation, precipitation, SST, and the \qty{21}{\celsius} isotherm depth.
+    Coordinated, interannual changes are observable in normalized yearly means of all environmental variables, in log-transformed chlorophyll \textit{a}, and in phytoplankton cell counts per functional group (Figure \ref{fig:zscore}).
+Distinct patterns emerge in wind speed, evaporation, precipitation, SST, and \qty{21}{\celsius} isotherm depth.
 Periods of positive wind anomalies (blue) correspond to negative anomalies in SST and upwelling (red, indicated by a shallower isotherm depth), reflecting stronger upwelling conditions.
@@
 This shift occurs during the period between 2003 and 2013, when the AMO index presents mostly positive anomalies, while upwelling-related variables indicate a weak upwelling regime (low wind, high SST, deep isotherm).
-Although data are sparse, nutrients and biological variables show a strong increase following 2013, coinciding with a return to negative anomalies in the AMO index and more upwelling-favorable conditions.
+Although data are sparse, nutrients and biological variables show a strong increase from 2014, coinciding with a return to negative anomalies in the AMO index and to more upwelling-favorable conditions.
     The annual means of diversity metrics do not follow this trend, but are reduced between 1998 and 2008.
-Shannon diversity specifically appears elevated past 2006 and the general pattern that emerges is that mean diversity is lower in years, where mean chlorophyll \textit{a} is high.
+Shannon diversity specifically appears elevated past 2006 and the general pattern that emerges is that mean diversity is lower in years when mean chlorophyll \textit{a} is high.
 However, in 1996 and 1997 positive anomalies in cell counts, chlorophyll \textit{a} and diversity metrics coincide.
@@
 The years 1995 (n=2) and 2016 (n=5), 2017 (n=1) were removed due to insufficient monthly data to generate robust annual means.
-Additionally we marked years with two or three and four or more missing monthly values.}
+Additionally, we marked years with two or three and four or more missing monthly values.}
     \label{fig:zscore}
@@
 (b,d,f,h) Boxplots of monthly measurements at four depth intervals grouped by season.
-Upwelling season goes from December to May, Rainy Season from June to November.}
+Upwelling season goes from December to May, rainy season from June to November.}
     \label{fig:divts}
@@
     \noindent\includegraphics[width=\textwidth]{fig/Figure4_ClusteringNMDS_v2.pdf}
-    \caption{(a) Hierarchical clustering by complete linkage method of microscopy cell counts at the Genus level, using the binary Jaccard distance.
+    \caption{Clustering analysis.
+(a) Hierarchical clustering by complete linkage method of microscopy cell counts at the Genus level, using the binary Jaccard distance.
 The community for each year is binned together, to smooth out seasonal features.
@@
 The dots are color-coded based on the cluster they belong to.
-Vector overlay of ENVFIT shows how environmental variables are co-varying with community data.
+Vector overlay of ENVFIT shows how environmental variables co-vary with community data.
 Length of vector indicates strength of relationship.}
@@
 The density distribution reveals some differences.
-Cluster 1 shows higher variance for genus richness, while cluster 2 shows greater variance and a bimodal distribution for the Shannon and Pielou indices.
+Cluster 1 shows higher variance for genus richness, whereas cluster 2 shows greater variance and a bimodal distribution for the Shannon and Pielou indices.
 
@@
 
-\subsection{AMO and upwelling-related variables as drivers of community changes}
+\subsection{AMO and upwelling-related variables drive community changes}
     % Figure 6
@@
 Figure \ref{fig:GF}$a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity.
-The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentrations scored the lowest in predicting changes within the community.
+The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentration scored the lowest in predicting changes within the community.
 Gradient Forest predicts shifts in the community by building a random forest model for each observed genus across the time series.
@@
     % P1: Introductory paragraph highlighting the overall dynamics/shifts
-    This study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes.
+    This study shows  a multi-annual oscillatory nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes.
 In the Cariaco Basin, decadal-scale climate oscillations impacted the wind-driven upwelling regime, producing confounding effects on the composition and biodiversity of the phytoplankton community.
     These changes are clearly visible in the yearly mean anomalies (Figure \ref{fig:zscore}) and the density distribution of monthly measurements (Figure \ref{fig:clustcomp}).
-After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies up to 2014.
-    By considering the phytoplankton community time series until the end of 2017, the temporally complete dataset, we found that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}.
+After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all in coincidence with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014.
+    By considering the phytoplankton community time series up to the end of 2017, the temporally complete dataset, we found that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}.
 In 2014, physical conditions in the Cariaco Basin returned to a regime of increased wind speed, increased upwelling, and elevated nutrient concentrations (Figure \ref{fig:zscore}).
@@
 This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}.
-Increases in temperature can lead to a decrease in the relative contribution of large cells in the community, irrespective of nutrient availability, but shifts towards smaller cells are particularly prevalent when temperatures increase and nutrient supply decreases \cite{mousing_global_2014}, as we observed in the Cariaco Basin.
+Increases in temperature can lead to a decrease in the relative contribution of larger cells in the community, irrespective of nutrient availability, but shifts towards smaller cells are particularly prevalent when temperatures increase and nutrient supply decreases \cite{mousing_global_2014}, as we observed in the Cariaco Basin.
 
     % Disucssion of HPLC vs Microscopy!
-    A similar pattern emerged when comparing our observations with those based on High-Performance Liquid Chromatography (HPLC) data from the Cariaco Basin, which were analyzed and presented by \citeA{pinckney_phytoplankton_2015}.
-However, there were some contrasting observations.
-The HPLC data suggested an increase in chlorophyll \textit{a} between the periods 1996-2000 and 2006-2010, particularly at depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}.
+    A similar general pattern emerged when comparing our results with those based on High-Performance Liquid Chromatography (HPLC) data from the Cariaco Basin, which were analyzed and presented by \citeA{pinckney_phytoplankton_2015}.
+However, some differences emerged between our results and those obtained with the HPLC data.
+The HPLC data showed an increase in chlorophyll \textit{a} between the periods 1996-2000 and 2006-2010, particularly at depths below \qty{55}{meters} \cite{pinckney_phytoplankton_2015}.
 It should be noted that the HPLC data were analyzed by multiple laboratories throughout the time series, with a gap in data coverage between 2000 and 2006.
-\citeA{pinckney_phytoplankton_2015} could not find a consistent method for quantifying total chlorophyll \textit{a} across different laboratories.
-This is why we focused our analysis on the more consistent and temporally complete microscopy cell count and fluorometric chlorophyll \textit{a} data.
+Total chlorophyll \textit{a} could not be quantified with a consistent method across the different laboratories \cite{pinckney_phytoplankton_2015}.
+This is the reason why we focused our analysis on the more consistent and temporally complete microscopy cell count and fluorometric chlorophyll \textit{a} data.
     Consistent with our results, independent spectrophotometric measurements of chlorophyll \textit{a} concentrations at \qty{20}{meters} depth, off the coast of Margarita Island, showed a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
-    Additionally, despite cell counts of all functional groups decreasing, the HPLC data indicated an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{\meter}.
-This provides a further strong indication that the phytoplankton community shifted to smaller sizes, undetected using light microscopy.
+    Additionally, despite cell counts of all functional groups decreased, the HPLC data indicated an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{\meter}.
+This provides a further strong indication that the phytoplankton community shifted to smaller sizes, an aspect that was not detected by previous studies using light microscopy data.
 
@@
 However, we did observe a significant difference in Pielou's index between the two clusters.
-The results suggest that the phytoplankton community became more evenly distributed during low-upwelling conditions of cluster 2, which is consistent with the analysis of \citeauthor{pinckney_phytoplankton_2015} based on HPLC pigment data.
+The results suggest that the phytoplankton community became more evenly distributed during the low upwelling conditions of cluster 2, which is consistent with the analysis of \citeauthor{pinckney_phytoplankton_2015} based on HPLC pigment data.
 This increase in evenness, quantified by the Pielou index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community.
 The resulting decrease in compositional turnover was also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering}$b$).
-As the productivity and biomass of phytoplankton communities are driven primarily by nutrients supplied through disturbance events, this is in line with the previously observed inverse relationship between evenness and community performance \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
+As the productivity and biomass of phytoplankton communities are driven primarily by nutrients supplied through disturbance events, this is in line with the inverse relationship between evenness and community performance observed in previous studies \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
 
@@
     % P4: Introduce GF Analysis and discuss results
-    To identify the variables that best predict changes in the phytoplankton community, we conducted a gradient forest analysis using time series of climate indices and in situ measurements to predict genus-level abundances from microscopy cell counts.
-Gradient forest is an extension of the random forest algorithm, designed to identify thresholds and patterns in community composition along environmental gradients \cite{ellis_gradient_2012}.
+    To identify the variables that best predict changes in the phytoplankton community, we conducted a Gradient Forest analysis using time series of climate indices and in situ measurements to predict genus-level abundances from microscopy cell counts.
+Gradient Forest is an extension of the random forest algorithm, designed to identify thresholds and patterns in community composition along environmental gradients \cite{ellis_gradient_2012}.
 Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series.
@@
 Diatoms, dinoflagellates, and haptophytes exhibited strong response thresholds for negative AMO anomalies, which pointed to the influence of stronger upwelling conditions.
-The abundance of haptophytes, additionally, showed a threshold effect at positive AMO anomalies, which may have been related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} based on HPLC data that haptophytes predominantly exhibited blooms above the mixed layer and at a lower frequency than other functional groups.
+The abundance of haptophytes, additionally, showed a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} based on HPLC data showing haptophytes blooming predominantly above the mixed layer and at a lower frequency than other functional groups.
 
@@
 During ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
-\citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months.
+\citeA{taylor_ecosystem_2012} found no strong correlation between MEI v.2 and other variables in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months.
 In contrast, we found a relatively strong predictive power for changes in the phytoplankton community for the MEI v.2 index with a 4-month time lag.
-We also tested a 12-month lag and found no significant improvement (Table \ref{tab:sup:GFlagtests} in Appendix).
-    The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for negative MEI v.2 index anomalies, which represent La Niña conditions.
+We also tested a 12-month time lag and found no significant improvement (Table \ref{tab:sup:GFlagtests} in Appendix).
+    The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for negative MEI v.2 anomalies, which represent La Niña conditions.
 For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events.
-Diatoms do not exhibit a clear threshold response for the MEI v.2 index but are affected across the entire range.
+Diatoms do not exhibit a clear threshold response to the MEI v.2 index but are affected across the entire range.
 Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in cluster 1 (see Figure \ref{fig:clustering} $e$).
@@
 
-    These results matched the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separated the two clusters was NMDS 1, which covaried most strongly with changes in AMO, MEI v.2, salinity, evaporation, and precipitation (Figure \ref{fig:clustering} $c$).
+    These results matched the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separated the two clusters was NMDS 1, which co-varied most strongly with changes in AMO, MEI v.2, salinity, evaporation, and precipitation (Figure \ref{fig:clustering} $c$).
 Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations varied more strongly along NMDS 2.
 Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, showed greater variance along NMDS 1.
-Salinity, evaporation, and silicate concentration exhibited some explanatory power in the gradient forest analysis, although they were weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}).
+Salinity, evaporation, and silicate concentration exhibited some explanatory power in the Gradient Forest analysis, although they were weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}).
 This implies that the climate indices explain variability that more direct measurements of upwelling intensity do not capture.
@@
     % 1.  Discuss Limitations of Light microscopy cell counts in CARIACO Time Series
-    Using microscopy cell counts as a basis for our analyses meant that the results should be interpreted in terms of changes in the microphytoplankton community.
+    Using microscopy cell counts as a basis for our analyses allowed us to focus on changes in the microphytoplankton community.
 Our choice of grouping cell counts by genus was based on the assumption that species within a genus generally exhibit similar eco-physiological responses.
-The output of the gradient forest model was further grouped by broad functional types to facilitate the identification of shifts and trends.
+The output of the Gradient Forest model was further grouped by broad functional types to facilitate the identification of shifts and trends.
 The phytoplankton genera that could be best predicted were mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (Figure \ref{fig:GF} $b$).
-This does not represent a complete picture of the phytoplankton community of the CARIACO station.
-The HPLC pigment data used by \citeauthor{pinckney_phytoplankton_2015} captured all size classes of phytoplankton, but, as previously discussed, it did not cover the entire time series and the dataset was assembled by four different laboratories.
+This does not represent a complete picture of the phytoplankton community of the CARIACO basin.
+The HPLC pigment data used by \citeauthor{pinckney_phytoplankton_2015} captured all size classes of phytoplankton but, as we mentioned above, it had a limited temporal coverage and the dataset was assembled by four different laboratories.
 With the caveat that the microscopy counts captured microphytoplankton almost exclusively, the discrepancies with the HPLC pigment analysis can be explained.
@@
     % 2. Discuss limitations of GF analysis & Limitation to bottom-up drivers and climate indices
-    The limited scope of the microscopy cell counts also affected the conclusions that could be drawn from the Gradient Forest analysis, as these results could only be interpreted as concerning microphytoplankton.
-    Another potential caveat of our methodology is that many of the predictors in ecological time series and in this study are highly correlated (Figure \ref{fig:sup:correlation} in Appendix).
+    The limited scope of the microscopy cell counts also affected the conclusions that could be drawn from the Gradient Forest analysis, as these results are limited to microphytoplankton. 
+    Another potential caveat of our methodology is that many of the predictors in ecological time series in general and, specifically, in this study, are correlated (Figure \ref{fig:sup:correlation} in Appendix).
 This can lead to the identification of a correlated predictor that has no direct effect on community changes.
-However, the gradient forest methodology addresses this by conditionally permuting subsets of the correlated predictors, which can partly mitigate this effect \cite{ellis_gradient_2012}.
+However, the Gradient Forest methodology addresses this by conditionally permuting subsets of the correlated predictors, which can partly mitigate the problem \cite{ellis_gradient_2012}.
 Additionally, random forest models do not consider temporal relationships between data points, which is why we included tests for the predictive relationships with time lags of our predictors.
 Without adjusting for time lags, SST and nitrate concentration are the strongest predictors of community change (Figure \ref{fig:sup:GFoutput_nolags} in Appendix).
-Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and the AMO index yielded a relatively low value around \qty{0.03}{}.
-This is expected given the noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the gradient forest algorithm \cite{pitcher_example_2012, roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}.
-It is important to consider that although predictive relationships between drivers and ecological predictors are identified, this does not necessarily imply a mechanistic relationship.
+Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both the SST and the AMO index yielded a relatively low value around \qty{0.03}{}.
+This is expected given the noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the Gradient Forest algorithm \cite{pitcher_example_2012, roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}.
+It is important to consider that although predictive relationships between drivers and ecological variables are identified, this does not necessarily imply a mechanistic causal relationship .
 The scope of our analysis, like any statistical analysis, is exploratory and aims at narrowing down potential key relationships.
@@
     Our focus on bottom-up drivers of community change may have also restricted the explanatory power of our analysis, as top-down processes such as zooplankton grazing may have impacted the structure of the phytoplankton community \cite{frank_ups_2007, banas_adding_2011, acevedo-trejos_mechanisms_2015}.
-Due to the scarcity of data covering higher trophic levels and the absence of zooplankton data for the first six years of the time series, we suggest that mechanistic modeling is the ideal approach for investigating the relative impacts of both top-down and bottom-up drivers.
+Due to the scarcity of data covering higher trophic levels and the absence of zooplankton data for the first six years of the time series, we suggest that mechanistic modeling is the ideal approach for investigating the relative impacts of top-down and bottom-up drivers.
 
@@
 Specifically, microscopy cell counts revealed the presence of two distinct clusters defined by high- and low-upwelling conditions that corresponded to anomalies in the AMO index.
-The community changes did not indicate a permanent regime shift, as a return to the early cluster was observed towards the end of the time series.
+The community changes did not indicate a permanent regime shift, a return to the early cluster was observed towards the end of the time series.
 We observed no significant changes in diversity metrics between the two clusters, except for lower turnover and higher community evenness under low-upwelling conditions.
-    Despite multiannual trends in warming, reduced productivity, and drastic declines in microphytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions were restored.
+    Despite multi-annual trends in warming, reduced productivity, and drastic declines in micro-phytoplankton cell counts, the phytoplankton community returned to previous levels once the long-term climatic and local physical conditions were restored.
     The strongest variable predicting changes in the phytoplankton community was the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local conditions through low-frequency natural cycles.
 We concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state for the phytoplankton community of the Cariaco Basin.
-    Notwithstanding a number of limitations, our analysis points to the incipient return of the Cariaco Basin to a pre-2004 regime, hinting at the resilient nature of phytoplankton communities under global change, an aspect that emphasizes the relevance of long-term time series and deserves further attention.
-    To extend our approach, future studies should include the impact of top-down drivers and investigate their explanatory power regarding the phytoplankton community changes observed in the Cariaco Basin.
+    Notwithstanding a number of limitations, our analysis points to the incipient return of the Cariaco Basin to a pre-2004 regime, hinting at an oscillatory nature of phytoplankton communities under global change, an aspect that emphasizes the relevance of long-term time series and deserves further attention.
+    Future studies, ideally based on mathematical modeling, could focus on the relative impact of top-down vs. bottom-up drivers,  aspects remaining largely unexplored in the Cariaco Basin.
 
@@
 
-\section{Gradient Forest Analysis supplement}
+\section{Gradient Forest analysis supplement}
 
@@
 \section*{Open Research Section}
-All data used in this manuscript is publicly available with open access.
+All data used in this manuscript are publicly available through open access.
 The CARIACO time series data we used was accessed from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) under the doi \url{10.1575/1912/bco-dmo.3095.1}.

### 2025-08-05 agujournaltemplate.tex
Post-Ago comments, just a few small things remain before final checks

@@
 Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period.
-Phytoplankton identification and enumeration by microscopy were restricted to microphytoplankton (cells $>$ \qty{20}{\micro \meter}) \cite{mutshinda_environmental_2013}.
     %from Mutshinda et al. 2013
@@
     % P1: Introductory paragraph highlighting the overall dynamics/shifts
-    This study shows  a multi-annual oscillatory nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes.
+    This study shows the multi-annual oscillatory nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes.
 In the Cariaco Basin, decadal-scale climate oscillations impacted the wind-driven upwelling regime, producing confounding effects on the composition and biodiversity of the phytoplankton community.
     These changes are clearly visible in the yearly mean anomalies (Figure \ref{fig:zscore}) and the density distribution of monthly measurements (Figure \ref{fig:clustcomp}).
-After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all in coincidence with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014.
+After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014.
     By considering the phytoplankton community time series up to the end of 2017, the temporally complete dataset, we found that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}.
@@
 This shift to smaller cell sizes and a reduction in diatoms is consistent with predicted effects of warming and reduced mixing on phytoplankton communities \cite{bopp_response_2005}.
-Increases in temperature can lead to a decrease in the relative contribution of larger cells in the community, irrespective of nutrient availability, but shifts towards smaller cells are particularly prevalent when temperatures increase and nutrient supply decreases \cite{mousing_global_2014}, as we observed in the Cariaco Basin.
+Increases in temperature can lead to a decrease in the relative contribution of larger cells in the community, irrespective of nutrient availability, but shifts towards smaller cells are particularly prevalent when temperatures increase and nutrient supply decreases \cite{acevedo-trejos_glimpse_2014, mousing_global_2014}, as we observed in the Cariaco Basin.
 
@@
     Consistent with our results, independent spectrophotometric measurements of chlorophyll \textit{a} concentrations at \qty{20}{meters} depth, off the coast of Margarita Island, showed a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
-    Additionally, despite cell counts of all functional groups decreased, the HPLC data indicated an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{\meter}.
-This provides a further strong indication that the phytoplankton community shifted to smaller sizes, an aspect that was not detected by previous studies using light microscopy data.
+    Additionally, despite decreased cell counts of all functional groupsing, the HPLC data indicated an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{\meter}.
+This provides a further strong indication that the phytoplankton community shifted to smaller cell sizes, which were not detected using light microscopy.
 
@@
 Diatoms, dinoflagellates, and haptophytes exhibited strong response thresholds for negative AMO anomalies, which pointed to the influence of stronger upwelling conditions.
-The abundance of haptophytes, additionally, showed a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} based on HPLC data showing haptophytes blooming predominantly above the mixed layer and at a lower frequency than other functional groups.
+The abundance of haptophytes, additionally, showed a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} based on HPLC data showing haptophytes blooming more frequently during the time period of 2006 to 2010, which was characterized by mostly positive AMO anomalies.
 
@@
 The phytoplankton genera that could be best predicted were mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (Figure \ref{fig:GF} $b$).
-This does not represent a complete picture of the phytoplankton community of the CARIACO basin.
+This does not represent a complete picture of the phytoplankton community of the Cariaco basin.
 The HPLC pigment data used by \citeauthor{pinckney_phytoplankton_2015} captured all size classes of phytoplankton but, as we mentioned above, it had a limited temporal coverage and the dataset was assembled by four different laboratories.
@@
 This can lead to the identification of a correlated predictor that has no direct effect on community changes.
-However, the Gradient Forest methodology addresses this by conditionally permuting subsets of the correlated predictors, which can partly mitigate the problem \cite{ellis_gradient_2012}.
+However, the Gradient Forest methodology addresses this by conditionally permuting subsets of the correlated predictors, which can partly mitigate this problem \cite{ellis_gradient_2012}.
 Additionally, random forest models do not consider temporal relationships between data points, which is why we included tests for the predictive relationships with time lags of our predictors.
 Without adjusting for time lags, SST and nitrate concentration are the strongest predictors of community change (Figure \ref{fig:sup:GFoutput_nolags} in Appendix).
-Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both the SST and the AMO index yielded a relatively low value around \qty{0.03}{}.
+Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and the AMO index yielded a relatively low value around \qty{0.03}{}.
 This is expected given the noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the Gradient Forest algorithm \cite{pitcher_example_2012, roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}.
-It is important to consider that although predictive relationships between drivers and ecological variables are identified, this does not necessarily imply a mechanistic causal relationship .
+It is important to consider that although predictive relationships between drivers and ecological variables are identified, this does not necessarily imply a mechanistic causal relationship.
 The scope of our analysis, like any statistical analysis, is exploratory and aims at narrowing down potential key relationships.
@@
 Specifically, microscopy cell counts revealed the presence of two distinct clusters defined by high- and low-upwelling conditions that corresponded to anomalies in the AMO index.
-The community changes did not indicate a permanent regime shift, a return to the early cluster was observed towards the end of the time series.
+The community changes did not indicate a permanent regime shift, since a return to the early cluster was observed towards the end of the time series.
 We observed no significant changes in diversity metrics between the two clusters, except for lower turnover and higher community evenness under low-upwelling conditions.
@@
 \section*{Open Research Section}
-All data used in this manuscript are publicly available through open access.
+All data used in this manuscript are publicly available.
 The CARIACO time series data we used was accessed from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) under the doi \url{10.1575/1912/bco-dmo.3095.1}.

### 2025-08-06 agujournaltemplate.tex
FINAL VERSION for checks by coauthors

@@
 \usepackage{siunitx}
+\usepackage[section]{placeins}
 
@@
 Species counts were performed at the Universidad de Oriente, Venezuela, following the Utermöhl method \cite{hasle1978inverted}, using \qty{100}{\milli\liter} sedimentation chambers with a 48-hour settling period.
-    %from Mutshinda et al. 2013
+
     Quality control of the microscopy data was performed by assembling the taxonomic identifications made across all cruises, removing observations that were not identified at the species or genus level (with the exception of nanoflagellate cell counts that are shown in Figure \ref{fig:zscore}), correcting spelling variations, and grouping synonyms using taxonomic information from the World Register of Marine Species (WoRMS, \url{marinespecies.org}).
@@
     The El Niño Southern Oscillation (ENSO) exhibits a strong teleconnection to the Caribbean Sea.
-During ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
+During El Niño (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
+This would imply a general reduction in upwelling strength during El Niño events in the Cariaco Basin, which has indeed been observed during the period from 1997 to 1998 \cite{romero_seasonal_2009}.
+The MEI v.2 is significantly positively correlated to local evaporation and salinity and negatively correlated to the AMO index, however it is not significantly correlated to local upwelling-related variables (see Figure \ref{fig:sup:correlation} in Appendix).
+The link here is less clear; while generally trade wind strength and SST are anti-correlated in the Caribbean Sea, positive trade wind anomalies have been observed during during El Niño conditions and the reverse during La Niña \cite{chang_coupled_2013}.
 \citeA{taylor_ecosystem_2012} found no strong correlation between MEI v.2 and other variables in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months.
@@
 We also tested a 12-month time lag and found no significant improvement (Table \ref{tab:sup:GFlagtests} in Appendix).
-    The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for negative MEI v.2 anomalies, which represent La Niña conditions.
+    The threshold responses indicate a shift in the abundances of dinoflagellates for negative MEI v.2 anomalies, which represent La Niña conditions.
+Findings from Bayesian modeling showed that dinoflagellates responding positively to warmer, stratified conditions in the Cariaco Basin \cite{mutshinda_environmental_2013}.
 For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events.
-Diatoms do not exhibit a clear threshold response to the MEI v.2 index but are affected across the entire range.
-Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in cluster 1 (see Figure \ref{fig:clustering} $e$).
-This is consistent with previous findings, that strong El Niño events (as indicated by a positive MEI) increase the strength of currents in the Caribbean sea through modulating the easterly wind anomalies, thereby increasing upwelling in the Cariaco Basin \cite{huang_enso_2023}.
+Diatoms do not exhibit a clear threshold response, but show a slightly stronger response in the positive range of the MEI v.2.
+Interestingly, positive anomalies in the MEI v.2, representing strong ENSO events, are mostly found in the high upwelling regime of cluster 1 (see Figure \ref{fig:clustering} $e$).
+This could be explained by previous findings, that strong El Niño events (as indicated by a positive MEI) increase the strength of currents in the Caribbean sea through modulating the easterly wind anomalies, thereby potentially increasing upwelling strength in the Cariaco Basin \cite{huang_enso_2023}.
 
     The in-situ measurements that best predicted changes in the microphytoplankton community were SST, nitrate concentration, and salinity, in that order.
+Haptophytes and diatoms exhibited a steep threshold around \qtyrange{22}{24}{\celsius}, indicating the strong effect on cell counts at low surface temperatures that are indicative of strong upwelling.
 Nitrate is a crucial nutrient for phytoplankton growth and has been identified as the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}.
-For diatoms, a strong threshold response was observed above \qty{7}{\micro \molar}, indicating that intense upwelling conditions drove blooms, resulting in predictable changes in cell counts.
-Similarly, haptophytes and diatoms exhibited a threshold around \qtyrange{22}{24}{\celsius}, indicating thresholds at low surface temperatures driven by stronger upwelling.
+For diatoms, a strong threshold response was observed above \qty{7}{\micro M}, indicating that intense upwelling conditions drove blooms, resulting in predictable changes in cell counts.
 
@@
 \appendix
-
+\newpage
 \section{Correlation of data}
 
-\begin{figure}
+\begin{figure}[h!]
 \noindent\includegraphics[width=\textwidth]{fig/CorrClustSupplementalPlot_v1.pdf}
@@
 
+\newpage
 \section{Gradient Forest analysis supplement}
 
-\begin{figure}
+\begin{figure}[h!]
 \noindent\includegraphics[width=\textwidth]{fig/FigureA2_GF_output_Supplement_v1.pdf}
@@
 
+%\end{tabular}}
+
+%\label{tab:sup:GenusFuncGroups}
+%\end{table}
+
+%\begin{table}
+
+%\caption{Table \ref{tab:sup:GenusFuncGroups} cont.:}
+%\begin{tabular}{@{}ll@{}}
+
 %
@@
 
+%\resizebox*{\textheight}{!}{
+\resizebox*{!}{\dimexpr\textheight-\lineskip\relax}{%
 \begin{tabular}{@{}ll@{}}
@@
 
-\end{tabular}
-
-\label{tab:sup:GenusFuncGroups}
-\end{table}
-
-\begin{table}
-
-\caption{Table \ref{tab:sup:GenusFuncGroups} cont.:}
-\begin{tabular}{@{}ll@{}}
-\textbf{Dinoflagellates} cont.    &                                                            \\
 \textit{Dinophysis}        & Ehrenberg, 1839                                            \\
@@
 \textit{Mesodinium}        & von Stein                                                 
-\end{tabular}
+\end{tabular}}
+
+\label{tab:sup:GenusFuncGroups}
 \end{table}
@@
 \acknowledgments
-We thank Digna Rueda-Roa, Claudia Benitez-Nelson, and James Pinckney for their valuable input during the early stages of this research.
+We thank Digna Rueda-Roa, Claudia Benitez-Nelson, and James Pinckney for their valuable input during the very early stages of this research.
+Additionally we want to thank all researchers and staff involved in the CARIACO Time Series for the considerable effort of collecting the time series data and providing it openly to the scientific community.
+
+AI tools (specifically Open AI's ChatGPT) were used to aid in spelling and grammar corrections in writing this study.
 

### 2025-08-13 agujournaltemplate.tex
backup before final edits before submission

@@
 Across the time series, we observed maximum richness at the start of the time series (1996-1997), with a subsequent reduction in diversity across all depths, followed by a recovery after 2011.
-The reduction in diversity following the start of the time series did not follow the pattern shown by chlorophyll \textit{a}, which remained relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}).
+The reduction in diversity following the start of the time series did not follow the pattern shown by chlorophyll \textit{a}, which remained relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts} $a$).
 Scaling genus richness by the cell count abundances, i.e.
@@
     We undertook clustering analysis on annually aggregated cell counts from microscopy and on individual monthly measurements.
-Figure \ref{fig:clustering}$a$ shows the results of clustering the annually aggregated communities using a linear complete linkage algorithm based on the binary Jaccard matrix.
+Figure \ref{fig:clustering}$a$ shows the results of clustering annually aggregated communities using a linear complete linkage algorithm based on the binary Jaccard matrix.
 Two distinct clusters emerge from this analysis, with the first cluster encompassing the years 1996-2004 and 2014-2016, and the second cluster including the intermediate years from 2005 to 2013.
-    Inter-annual species turnover is shown in Figure \ref{fig:clustering}$b$, it reveals an increasing trend in turnover from 1996 to 2004.
+    Inter-annual species turnover is shown in Figure \ref{fig:clustering}$b$, revealing an increasing trend in turnover from 1996 to 2004.
 Turnover subsequently decreased during the period underpinning cluster 2 and increased again in 2014, marking a return to cluster 1.
@@
 Although there are no significant differences in genus richness and the Shannon index between clusters, there is a slightly increased median in the Pielou index for cluster 2 (Table \ref{tab:ClustCompWilcox}).
-The density distribution reveals some differences.
-Cluster 1 shows higher variance for genus richness, whereas cluster 2 shows greater variance and a bimodal distribution for the Shannon and Pielou indices.
+The density distributions highlight additional contrasts; cluster 1 shows higher variance for genus richness, whereas cluster 2 shows greater variance and a bimodal distribution for the Shannon and Pielou indices.
 

### 2025-08-13 agujournaltemplate.tex
Implemented final comments

@@
 To quantify the effects of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2.
-Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.
+Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a possible cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.
 
@@
 \citeA{taylor_ecosystem_2012} noted that the observed changes could not be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation".
-But, to date, all analyses of phytoplankton community diversity in the Cariaco Basin do not include time series data beyond 2011, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts.
+But, to our knowledge, all analyses of phytoplankton community diversity in the Cariaco Basin do not include time series data beyond 2012, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts.
 Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community.
@@
     In the analysis, we set out to find if there was a structure in the distribution and temporal evolution of diversity metrics and community composition and, additionally, to identify potential bottom-up drivers of this change.
-    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores ($z = \frac{(x – \mu)}{\sigma}$, where $x$ is the yearly mean, $\mu$ is the mean and $\sigma$ is the standard deviation of all yearly means).
+    To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores, $z = \frac{(x – \mu)}{\sigma}$, where $x$ is the yearly mean, $\mu$ is the mean and $\sigma$ is the standard deviation of all yearly means.
 Data from 1995, 2016, and 2017 were not included because insufficient coverage (i.e.
@@
     \subsubsection{Assessing Environmental Drivers of Phytoplankton Dynamics}
-    To quantify the effects of bottom-up drivers on shifts in the phytoplankton community of the Cariaco Basin, we employed a Gradient Forest analysis.
-Gradient Forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}.
-Gradient Forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor.
+    To quantify the effects of bottom-up drivers on shifts in the phytoplankton community of the Cariaco Basin, we employed a gradient forest analysis.
+Gradient forest is a community-level extension of random forest analysis that combines regression trees calculated for indicator variables to assess turnover in indicator variables across predictor variables \cite{pitcher_example_2012, large_critical_2015, tam_comparing_2017}.
+Gradient forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor.
 The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover.
@@
 the isotherm and nutrient concentrations) and some more indirectly, such as the climate indices and meteorological data.
-    The environmental variables from ocean time series tend to be strongly correlated, which the Gradient Forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}.
-Gradient Forest runs were performed with 1500 bootstrapped trees and a correlation threshold of 0.5.
+    The environmental variables from ocean time series tend to be strongly correlated, which the gradient forest methodology accounts for by implementing a conditional permutation of the correlated predictors \cite{ellis_gradient_2012}.
+Gradient forest runs were performed with 1500 bootstrapped trees and a correlation threshold of 0.5.
     We show a model run without any time lags in the supplementary material (Figure \ref{fig:sup:GFoutput_nolags}).
@@
     \begin{table}
-    \caption{Data and corresponding units used as predictor variables for the Gradient Forest analysis.}
+    \caption{Data and corresponding units used as predictor variables for the gradient forest analysis.}
     \centering
@@
 The bulk of the chlorophyll \textit{a} ($\sim$ 80\,\%) was measured in the upper 50 meters of the water column, with approximately 15\,\% measured between 50 and 75 meters.
-Chlorophyll \textit{a} exhibited marked seasonality, with the highest biomass in the upper water column during the early upwelling season (December–February, Figure \ref{fig:divts}$b$).
+Chlorophyll \textit{a} exhibited marked seasonality, with the highest biomass in the upper water column during the early upwelling season (December–February, Figure \ref{fig:divts} $b$).
 The yearly mean chlorophyll \textit{a} biomass showed the highest variability within the upper 25 meters.
-In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deeper chlorophyll maximum in those years (Figure \ref{fig:divts}$a$).
+In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deeper chlorophyll maximum in those years (Figure \ref{fig:divts} $a$).
 The year 2013 was an outlier, particularly for the upper 25 meters, as missing monthly sampling from the upwelling season (January and February) skewed the value downwards.
@@
 The dots are color-coded based on the cluster they belong to.
-Vector overlay of ENVFIT shows how environmental variables co-vary with community data.
+Vectors show how environmental variables co-vary with community data.
 Length of vector indicates strength of relationship.}
@@
     % Figure 6
-    To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the monthly microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
+    To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a gradient forest analysis on the monthly microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
 Preliminary model runs were made to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months.
@@
 The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentration scored the lowest in predicting changes within the community.
-Gradient Forest predicts shifts in the community by building a random forest model for each observed genus across the time series.
+Gradient forest predicts shifts in the community by building a random forest model for each observed genus across the time series.
 The predictive power of the models for each genus is shown in Figure \ref{fig:GF}$b$.
@@
     \noindent\includegraphics[width=\textwidth]{fig/Figure6_GF_output_FINAL_v2.pdf}
-    \caption{Gradient Forest analysis of monthly community data at a genus level.
+    \caption{Gradient forest analysis of monthly community data at a genus level.
 (a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances.
@@
 However, we did observe a significant difference in Pielou's index between the two clusters.
-The results suggest that the phytoplankton community became more evenly distributed during the low upwelling conditions of cluster 2, which is consistent with the analysis of \citeauthor{pinckney_phytoplankton_2015} based on HPLC pigment data.
+The results suggest that the phytoplankton community became more evenly distributed during the low upwelling conditions of cluster 2, which is consistent with the analysis of \citeA{pinckney_phytoplankton_2015} based on HPLC pigment data.
 This increase in evenness, quantified by the Pielou index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community.
@@
     % P4: Introduce GF Analysis and discuss results
-    To identify the variables that best predict changes in the phytoplankton community, we conducted a Gradient Forest analysis using time series of climate indices and in situ measurements to predict genus-level abundances from microscopy cell counts.
-Gradient Forest is an extension of the random forest algorithm, designed to identify thresholds and patterns in community composition along environmental gradients \cite{ellis_gradient_2012}.
+    To identify the variables that best predict changes in the phytoplankton community, we conducted a gradient forest analysis using time series of climate indices and in situ measurements to predict genus-level abundances from microscopy cell counts.
+Gradient forest is an extension of the random forest algorithm, designed to identify thresholds and patterns in community composition along environmental gradients \cite{ellis_gradient_2012}.
 Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series.
@@
 Diatoms, dinoflagellates, and haptophytes exhibited strong response thresholds for negative AMO anomalies, which pointed to the influence of stronger upwelling conditions.
-The abundance of haptophytes, additionally, showed a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeauthor{pinckney_phytoplankton_2015} based on HPLC data showing haptophytes blooming more frequently during the time period of 2006 to 2010, which was characterized by mostly positive AMO anomalies.
+The abundance of haptophytes, additionally, showed a threshold effect at positive AMO anomalies, which may be related to the pattern observed by \citeA{pinckney_phytoplankton_2015} based on HPLC data showing haptophytes blooming more frequently during the time period of 2006 to 2010, which was characterized by mostly positive AMO anomalies.
 
@@
 Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, showed greater variance along NMDS 1.
-Salinity, evaporation, and silicate concentration exhibited some explanatory power in the Gradient Forest analysis, although they were weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}).
+Salinity, evaporation, and silicate concentration exhibited some explanatory power in the gradient forest analysis, although they were weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}).
 This implies that the climate indices explain variability that more direct measurements of upwelling intensity do not capture.
@@
 Our choice of grouping cell counts by genus was based on the assumption that species within a genus generally exhibit similar eco-physiological responses.
-The output of the Gradient Forest model was further grouped by broad functional types to facilitate the identification of shifts and trends.
+The output of the gradient forest model was further grouped by broad functional types to facilitate the identification of shifts and trends.
 The phytoplankton genera that could be best predicted were mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (Figure \ref{fig:GF} $b$).
 This does not represent a complete picture of the phytoplankton community of the Cariaco basin.
-The HPLC pigment data used by \citeauthor{pinckney_phytoplankton_2015} captured all size classes of phytoplankton but, as we mentioned above, it had a limited temporal coverage and the dataset was assembled by four different laboratories.
+The HPLC pigment data used by \citeA{pinckney_phytoplankton_2015} captured all size classes of phytoplankton but, as we mentioned above, it had a limited temporal coverage and the dataset was assembled by four different laboratories.
 With the caveat that the microscopy counts captured microphytoplankton almost exclusively, the discrepancies with the HPLC pigment analysis can be explained.
@@
     % 2. Discuss limitations of GF analysis & Limitation to bottom-up drivers and climate indices
-    The limited scope of the microscopy cell counts also affected the conclusions that could be drawn from the Gradient Forest analysis, as these results are limited to microphytoplankton. 
+    The limited scope of the microscopy cell counts also affected the conclusions that could be drawn from the gradient forest analysis, as these results are limited to microphytoplankton. 
     Another potential caveat of our methodology is that many of the predictors in ecological time series in general and, specifically, in this study, are correlated (Figure \ref{fig:sup:correlation} in Appendix).
 This can lead to the identification of a correlated predictor that has no direct effect on community changes.
-However, the Gradient Forest methodology addresses this by conditionally permuting subsets of the correlated predictors, which can partly mitigate this problem \cite{ellis_gradient_2012}.
+However, the gradient forest methodology addresses this by conditionally permuting subsets of the correlated predictors, which can partly mitigate this problem \cite{ellis_gradient_2012}.
 Additionally, random forest models do not consider temporal relationships between data points, which is why we included tests for the predictive relationships with time lags of our predictors.
@@
 Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and the AMO index yielded a relatively low value around \qty{0.03}{}.
-This is expected given the noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the Gradient Forest algorithm \cite{pitcher_example_2012, roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}.
+This is expected given the noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the gradient forest algorithm \cite{pitcher_example_2012, roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}.
 It is important to consider that although predictive relationships between drivers and ecological variables are identified, this does not necessarily imply a mechanistic causal relationship.
@@
 \newpage
-\section{Gradient Forest analysis supplement}
+\section{Gradient forest analysis supplement}
 
@@
 \noindent\includegraphics[width=\textwidth]{fig/FigureA3_GF_output_NOLAG_v1.pdf}
-\caption{Gradient Forest analysis of monthly community data at a genus level.
+\caption{Gradient forest analysis of monthly community data at a genus level.
 (a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances.

### 2025-08-14 agujournaltemplate.tex
Final Text Revision

@@
 This study investigates changes in the phytoplankton community of the Cariaco Basin, off the coast of Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data.
-To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in wind that drive upwelling in the Cariaco Basin.
+To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in local weather and upwelling in the Cariaco Basin.
 Cluster analysis identified two distinct community states separated by a shift from high to low upwelling conditions in 2004.
@@
 % Findings Summary
-The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning in 2014 to the state antecedent 2004.
+The results showed a major shift in the phytoplankton community in 2004, linked to a decrease in upwelling, with conditions partially returning in 2014 to the state preceding 2004.
 Interestingly, community diversity declined between 1998 and 2009 independently from changes in the upwelling regime.
-Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate concentrations in the water column, were the strongest predictors of changes in phytoplankton community.
+Further analysis showed that the AMO index, along with sea surface temperature, MEI v.2 and nitrate concentrations in the water column, were the strongest predictors of changes in the phytoplankton community.
 % Key Takeaways
@@
 
-A tropical coastal location that has been intensively studied in the last few decades is the Cariaco Basin, where a wide range of physical and biogeochemical parameters were collected consistently between 1995 and 2017 \cite{muller-karger_scientific_2019}.
+A tropical coastal location that has been intensively studied in the last few decades is the Cariaco Basin, where a wide range of physical and biological parameters were collected consistently between 1995 and 2017 \cite{muller-karger_scientific_2019}.
 The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within one of the largest anoxic basins in the world's oceans \cite{edgcomb_accessing_2011}.
@@
 Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes.
-\citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to effects of global warming.
+\citeA{taylor_ecosystem_2012} analyzed the time series data up to 2010 and found decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors attributed to effects of global climate change.
 During the initial phase of the time series (1996–2004), the Cariaco Basin underwent consistent and pronounced seasonal upwelling cycles, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}.
-After 2004, a significant reduction in the strength of trade winds during the upwelling period \cite{taylor_ecosystem_2012} caused a regime change in the physical and biogeochemical system.
+After 2004, a significant reduction in the strength of trade winds during the upwelling period caused a regime change in the physical and biogeochemical system \cite{taylor_ecosystem_2012}.
 This strongly affected the phytoplankton community and the entire ecosystem.
@@
 % What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
-\citeA{taylor_ecosystem_2012} noted that the observed changes could not be clearly identified as "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation".
-But, to our knowledge, all analyses of phytoplankton community diversity in the Cariaco Basin do not include time series data beyond 2012, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts.
+\citeA{taylor_ecosystem_2012} noted that the observed changes could not be clearly identified as ``unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation''.
+To our knowledge, all analyses of phytoplankton community diversity in the Cariaco Basin do not include time series data beyond 2012, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts.
 Given these knowledge gaps, we aim to investigate 1) changes in the depth distribution of phytoplankton biomass and community diversity over the period from 1995 to 2017, 2) changes in the structure of the phytoplankton community and potential connections with shifts observed in physical and biogeochemical variables, and 3) potential connections between long-term climatic oscillations (ENSO MEI v.2 and AMO) and changes in the phytoplankton community.
 Our analysis focuses exclusively on bottom-up drivers of community changes because these data are available throughout the entire time series and at a monthly resolution.
-Our study builds on and substantially expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017.
-In doing so, we aim to provide a more comprehensive assessment of long-term phytoplankton community dynamics in the Cariaco Basin, offering new insights into the ecological responses to environmental variability across decadal scales.
+Our study builds on and expands the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017.
+In doing so, we aim to provide a more comprehensive assessment of long-term phytoplankton community dynamics in the Cariaco Basin, offering new insights into the local ecological responses to environmental variability across decadal scales.
 
@@
 The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin situated on the northeastern Venezuelan continental shelf, with a sill to the north running almost parallel to the coastline at a mean depth of around \qty{100}{m}.
-This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the BATS and HOT sites, resulting in the formation of anoxic conditions below a depth of  \qty{\sim 275}{m}.
+This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the Bermuda Atlantic Time-series Study (BATS) and Hawaii Ocean Time-series (HOT) sites, resulting in the formation of anoxic conditions below a depth of  \qty{\sim 275}{m}.
 
@@
 \subsection{In-situ biological and environmental data}
-    The CARIACO Ocean Time Series Program collected biological and environmental data, which we used here to examine the connections between environmental and ecological change in the region.
+    The CARIACO Ocean Time Series Program collected biological, chemical and physical data, which we used here to examine the connections between environmental and ecological change in the region.
 Samples were collected on monthly research cruises at a mooring station located in the eastern Cariaco Basin (\ang{10.50}N, \ang{64.67}W) \cite{muller-karger_scientific_2019}.
@@
 A total of 482 phytoplankton species were observed over the course of the time series, which were grouped into 223 phytoplankton genera.
-The identified genera were further grouped into functional groups based on their taxonomic classification in WoRMS.
+The identified genera were further grouped into functional groups based on their taxonomic classification in WoRMS (see Table \ref{tab:sup:GenusFuncGroups} in Appendix).
 
@@
 We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
-For simplicity, we refer to this variable as "wind speed" throughout the article.
+For simplicity, we refer to this variable as ``wind speed'' throughout the article.
 The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell.
@@
     The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}, were obtained from \url{https://psl.noaa.gov/enso/mei/}.
-In our analysis, bi-monthly values were aligned with the tail month (December-January is treated as the value for January).
+In our analysis, the bi-monthly values were aligned with the tail month (December-January is treated as the value for January).
 
@@
     % Niskin, CTD and Phytoplankton data processing
-    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using tidyverse packages \cite{wickham_welcome_2019}.
+    Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using ``tidyverse'' packages \cite{wickham_welcome_2019}.
 All the codes we produced for data processing and data analysis are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
-    For all discrete depth measurements, the data were interpolated using the "oce" package \cite{kelley_oce_2023} with the adaptive "unesco" algorithm defined by the U.S.
+    For all discrete depth measurements, the data were interpolated using the ``oce'' package \cite{kelley_oce_2023} with the adaptive ``unesco'' algorithm defined by the U.S.
 National Oceanographic Data Center \cite{johnson2006world}.
@@
     \subsubsection{Analysis of Phytoplankton Community Structure}
-    To investigate the structure of the phytoplankton community over time, we used a complete linkage clustering method via the "hclust" function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix of annual aggregates of observed genera integrated over 100 m depth.
-    Additionally, we performed non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering on the monthly integrated counts that were square-root transformed using "metaMDS" from the "vegan" package \cite{oksanen_vegan_2024}.
-We employed the Bray distance matrix in a two-dimensional ordination that reached a converging solution within 30 steps.
-To relate the multidimensional structuring of the community data to variability in environmental data, we used the "envfit" function from the "vegan" package to fit the environmental vectors onto the ordination with 999 permutations.
-    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the "ggpubr" package \cite{kassambara_ggpubr_2023}.
-To test whether the environmental variables, chlorophyll $a$ and diversity metrics varied significantly between the two community clusters, we employed a two-sided Wilcoxon sum rank test using the "wilcox.test" function in R \cite{r_core_team_r_2024}.
+    To examine temporal changes in the phytoplankton community structure in the Cariaco Basin, we applied complete linkage hierarchical clustering using the ``hclust'' function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix This analysis was based on presence-absence data of observed genera, over the upper 100 meters of the water column and aggregated annually.
+By using binary data and temporal aggregation, we aimed to reduce the influence of extreme values in the cell counts and focus on broad compositional shifts in the community.
+    Additionally, we performed non-metric multidimensional scaling (NMDS) ordination on the monthly counts, depth-integrated to \qty{100}{\meter}, that were square-root transformed using ``metaMDS'' from the ``vegan'' package \cite{oksanen_vegan_2024}.
+The NMDS analysis is based on the Bray distance matrix with a two-dimensional ordination that reached a converging solution within 30 steps.
+To relate the multidimensional structuring of the community data to variability in environmental data, we used the ``envfit'' function from the ``vegan'' package to fit the environmental vectors onto the ordination with 999 permutations.
+In contrast to the annual clustering, the NMDS ordination uses the individual monthly cell counts and takes abundance of genera into account.
+    To contrast the environmental parameters and diversity data between the community clusters resulting from the yearly aggregated clustering, we created density distribution plots contrasting the monthly measurements between clusters using the ``ggpubr'' package \cite{kassambara_ggpubr_2023}.
+To test whether the environmental variables, chlorophyll $a$ and diversity metrics varied significantly between the two community clusters, we employed a two-sided Wilcoxon sum rank test using the ``wilcox.test'' function in R \cite{r_core_team_r_2024}.
     % detailed explanation of NMDS in Winder and Hunter 2008
@@
 Gradient forest calculates a goodness-of-fit statistic for each predictor variable and indicator variable, then averages these to determine the overall importance of each predictor.
-The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community turnover.
+The regression tree split values also identify the ranges of predictors in which significant changes occur in the indicator variables, signaling a threshold for phytoplankton community changes.
 Cumulative plots of the predictive importance of each predictor variable can reveal thresholds for ecosystem drivers, as steep increases in cumulative importance signal a threshold for community turnover in the phytoplankton community \cite{tam_comparing_2017}.
@@
 The genus abundances were transformed by the natural logarithm plus one prior to model runs.
-The drivers/predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO_3]}$, $\mathrm{[PO_4]}$ and $\mathrm{[SiO_4]}$).
+The predictor variables considered in the analysis were the AMO and MEI v.2 indices, wind speed, evaporation, precipitation, \qty{21}{\celsius} isotherm, SST, salinity, and nutrient concentrations ($\mathrm{[NO_3]}$, $\mathrm{[PO_4]}$ and $\mathrm{[SiO_4]}$; see Table \ref{tab:DataSources} for a full overview).
 All predictor variables are in some way related to the local climate and upwelling system and could be potential bottom-up drivers of community change, some more directly (e.g.
@@
 Instead, elevated concentrations were observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
-    Chlorophyll \textit{a} and functional group cell counts show positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift.
-This shift occurs during the period between 2003 and 2013, when the AMO index presents mostly positive anomalies, while upwelling-related variables indicate a weak upwelling regime (low wind, high SST, deep isotherm).
+    Chlorophyll \textit{a} and functional group cell counts show positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift in the community.
+This shift persists during the period between 2003 and 2013, when the AMO index presents mostly positive anomalies, while upwelling-related variables indicate a weak upwelling regime (low wind, high SST, deep isotherm).
 Although data are sparse, nutrients and biological variables show a strong increase from 2014, coinciding with a return to negative anomalies in the AMO index and to more upwelling-favorable conditions.
@@
     \caption{Time series and seasonality of chlorophyll \textit{a} and diversity indices at four depth intervals.
-(a,c,e,g) Large dots and lines mark annual mean across time series for years with more than 5 monthly sampling.
-Small, transparent dots show monthly raw values.
+(a,c,e,g) Large dots and lines mark annual mean across time series for years with more than 5 monthly samples.
+Small, transparent dots show monthly values.
 (b,d,f,h) Boxplots of monthly measurements at four depth intervals grouped by season.
-Upwelling season goes from December to May, rainy season from June to November.}
+Upwelling season includes the months from December to May, rainy season from June to November.}
     \label{fig:divts}
@@
     We undertook clustering analysis on annually aggregated cell counts from microscopy and on individual monthly measurements.
-Figure \ref{fig:clustering}$a$ shows the results of clustering annually aggregated communities using a linear complete linkage algorithm based on the binary Jaccard matrix.
+Figure \ref{fig:clustering} $a$ shows the results of clustering annually aggregated communities using a linear complete linkage algorithm based on the binary Jaccard matrix.
 Two distinct clusters emerge from this analysis, with the first cluster encompassing the years 1996-2004 and 2014-2016, and the second cluster including the intermediate years from 2005 to 2013.
-    Inter-annual species turnover is shown in Figure \ref{fig:clustering}$b$, revealing an increasing trend in turnover from 1996 to 2004.
+    Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, revealing an increasing trend in turnover from 1996 to 2004.
 Turnover subsequently decreased during the period underpinning cluster 2 and increased again in 2014, marking a return to cluster 1.
-This pattern is also reflected in the hierarchical clustering results, Figure \ref{fig:clustering}$a$, where cluster 2 shows a lower distance in the clustering tree between the years.
+This pattern is also reflected in the hierarchical clustering results, Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years.
 We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster, respectively.
@@
 
-    To check whether the annual aggregation of community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (Figure \ref{fig:clustering}$c$).
+    To check whether the annual aggregation of community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (Figure \ref{fig:clustering} $c$).
 The NMDS clustering is overlaid with environmental data and climate indices in an attempt to unveil some of the underlying trends.
@@
 The second NMDS axis appears to follow the trends in upwelling intensity, as indicated by the vectors pointing in opposite directions for nutrient concentrations, temperature, and wind speed.
-Overall, the NMDS clustering confirms the pattern shown by clustering on annually aggregated data.
+Overall, the NMDS clustering confirms the pattern shown by clustering the annually aggregated data.
 
@@
 The community for each year is binned together, to smooth out seasonal features.
-(b) Species turnover between years for annual aggregate of identified phytoplankton to a genus level.
+(b) Species turnover between years for annual aggregate of identified phytoplankton at the genus level.
 The three genera dominating the cell counts for the separate parts of cluster 1 and cluster 2 are given in order of abundance from top to bottom.
@@
 Stress: 0.261, 2 dimensions, converging solution.
-The dots are color-coded based on the cluster they belong to.
-Vectors show how environmental variables co-vary with community data.
-Length of vector indicates strength of relationship.}
+The fill color of dots indicates which cluster the monthly values belong, cluster 1 in blue and cluster 2 in red.
+Vectors show how environmental variables co-vary with the community data.
+The length of the vector indicates the strength of the relationship.}
     \label{fig:clustering}
@@
 
-\subsection{AMO and upwelling-related variables drive community changes}
-    % Figure 6
-    To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a gradient forest analysis on the monthly microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
-Preliminary model runs were made to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months.
-Figure \ref{fig:GF}$a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity.
-The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentration scored the lowest in predicting changes within the community.
-Gradient forest predicts shifts in the community by building a random forest model for each observed genus across the time series.
-The predictive power of the models for each genus is shown in Figure \ref{fig:GF}$b$.
-The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are diatoms.
-However, two haptophyte and three dinoflagellate genera rank relatively high.
-For the presented run a total of 89 genera (each observed more than 10 times over the time series) were used as indicator variables, of which 46 genera could be predicted using the given predictor variables, which corresponds to more than \qty{95}{\%} of counted cells (see Table \ref{tab:sup:GenusFuncGroups} for a complete list of identified genera).
-    Figure \ref{fig:GF}$c$ shows the cumulative importance of the predictor variables across the parameter range.
-The black, dashed lines indicate cumulative importance across the entire community, and the solid, dashed lines indicate importance per phytoplankton group.
-For the AMO index, we see a clear pattern in which negative anomalies affect the community.
-This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria.
-Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range within which the community changed most, indicating these values as a threshold for community composition.
-Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows response curves for all variables not shown here.
-For wind speed, SST, and salinity, changes occurred mostly within parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}), and high salinity (above \qty{36.8}{PSU}).
-
-    \begin{figure}
-    \noindent\includegraphics[width=\textwidth]{fig/Figure6_GF_output_FINAL_v2.pdf}
-    \caption{Gradient forest analysis of monthly community data at a genus level.
-(a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances.
-(b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance.
-(c) Cumulative importance curves for the predictor variables.
-Black line shows cumulative importance across predictor range for the entire model and all genera.
-Dashed colored lines represent the responses of individual functional groups.}
-    \label{fig:GF}
-    \end{figure}
-
     \begin{table}
-    \caption{Results of two-sided Wilcoxon sum rank test with continuity correction for variables between the two clusters.
+    \caption{Results of two-sided Wilcoxon sum rank test with continuity correction for monthly values of the variables between the two clusters.
 Cluster 1 covers the periods 1996-2004 and 2014-2016 and cluster 2 covers the period 2005-2013.}
@@
 
+\subsection{AMO and upwelling-related variables drive community changes}
+    % Figure 6
+    To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a gradient forest analysis on the monthly microscopy cell counts interpolated over the top 100 meters and grouped by identified genus.
+Preliminary model runs were made to determine whether time lags of the predictor variables could better explain the data (see Appendix), resulting in the selection of the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months.
+Figure \ref{fig:GF} $a$ shows the predictor variables ranked by their weighted importance, with the AMO and MEI v.2 indices with time lags, as well as upwelling-related variables such as SST and nitrate in the upper 100 meters, demonstrating the highest predictive capacity.
+The \qty{21}{\celsius} isotherm, precipitation, and phosphate concentration scored the lowest in predicting changes within the community.
+Gradient forest predicts shifts in the community by building a random forest model for each observed genus across the time series.
+The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$.
+The microscopy data are biased toward larger cell sizes and, therefore, most of the predicted genus cell counts are diatoms.
+However, two haptophyte and three dinoflagellate genera rank relatively high.
+For the presented run a total of 89 genera (each observed more than 10 times over the time series) were used as indicator variables, of which 46 genera could be predicted using the given predictor variables, which corresponds to more than \qty{95}{\%} of counted cells (see Table \ref{tab:sup:GenusFuncGroups} for a complete list of identified genera).
+    Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range.
+The black, dashed lines indicate cumulative importance across the entire community, and the solid, dashed lines indicate importance per phytoplankton group.
+For the AMO index, we see a clear pattern in which negative anomalies affect the community.
+This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria.
+Nitrate concentrations between 6 and 8 \unit{\micro \mole} represent the range within which the community changed most, indicating these values as a threshold for community composition.
+Figure \ref{fig:sup:GFoutput_lags_extra} in Appendix shows response curves for all variables not shown here.
+Variables that show a pronounced threshold response are evaporation and salinity.
+For evaporation the threshold is at larger values of evaporation, which are most likely linked to periods of strong stratification, and salinity at both extreme ends of the range, signifying effects on cell counts of diatoms, dinoflagellates and cyanobacteria at strong upwelling and stratified conditions.
+
+    \begin{figure}
+    \noindent\includegraphics[width=\textwidth]{fig/Figure6_GF_output_FINAL_v2.pdf}
+    \caption{Gradient forest analysis of monthly community data at a genus level.
+(a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances.
+(b) Phytoplankton genera that could be predicted ranked according to the predictive power via the weighted importance.
+(c) Cumulative importance curves for the predictor variables.
+Black line shows cumulative importance across predictor range for the entire model and all genera.
+Dashed colored lines represent the responses of individual functional groups.}
+    \label{fig:GF}
+    \end{figure}
+
 \section{Discussion}
@@
     % P1: Introductory paragraph highlighting the overall dynamics/shifts
-    This study shows the multi-annual oscillatory nature of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes.
+    This study shows the long-term dynamics of a tropical coastal marine ecosystem driven by local environmental variability, which is, in turn, influenced by large-scale climatic changes.
 In the Cariaco Basin, decadal-scale climate oscillations impacted the wind-driven upwelling regime, producing confounding effects on the composition and biodiversity of the phytoplankton community.
-    These changes are clearly visible in the yearly mean anomalies (Figure \ref{fig:zscore}) and the density distribution of monthly measurements (Figure \ref{fig:clustcomp}).
+    These changes are clearly visible in the yearly mean anomalies and the density distribution of monthly measurements.
 After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014.
     By considering the phytoplankton community time series up to the end of 2017, the temporally complete dataset, we found that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}.
-In 2014, physical conditions in the Cariaco Basin returned to a regime of increased wind speed, increased upwelling, and elevated nutrient concentrations (Figure \ref{fig:zscore}).
-Concurrently, we observed an increase in chlorophyll \textit{a} (Figure \ref{fig:divts} $a$) and a return to a community composition more similar to the period between 1996 and 2004 (Figure \ref{fig:clustering} $a$ and $c$).
+In 2014, physical conditions in the Cariaco Basin returned to a regime of increased wind speed, increased upwelling, and elevated nutrient concentrations.
+Concurrently, we observed an increase in chlorophyll \textit{a} and a return to a community composition more similar to the period between 1996 and 2004.
 
@@
 This increase in evenness, quantified by the Pielou index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community.
-The resulting decrease in compositional turnover was also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering}$b$).
+The resulting decrease in compositional turnover was also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$).
 As the productivity and biomass of phytoplankton communities are driven primarily by nutrients supplied through disturbance events, this is in line with the inverse relationship between evenness and community performance observed in previous studies \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
@@
 Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series.
-Interestingly, our analysis revealed that the strongest predictor of change within the community was not an in situ measurement but rather the AMO climate index.
-The AMO is followed in importance by SST, MEI v.2 index, and nitrate concentration as predictive variables (see Figure \ref{fig:GF}).
+Interestingly, our analysis revealed that the strongest predictor of change within the community was not an in situ measurement but rather the AMO index.
+The AMO index is followed in importance by SST, the MEI v.2, and nitrate concentration as predictive variables (see Figure \ref{fig:GF}).
 
     % P5: Climate impacts
-    Although regional effects may vary, a negative phase in the AMO is generally associated with periods of colder temperatures and a southward shift of the ITCZ \cite{knight_climate_2006, colna2017latitudinal}.
+    Although regional effects may vary, a negative phase in the AMO is generally associated with periods of colder surface temperatures and a southward shift of the ITCZ \cite{knight_climate_2006, colna2017latitudinal}.
 In the Cariaco Basin, a southward shift of the ITCZ correlates with increased wind-driven upwelling \cite{taylor_ecosystem_2012}.
@@
 Our choice of grouping cell counts by genus was based on the assumption that species within a genus generally exhibit similar eco-physiological responses.
-The output of the gradient forest model was further grouped by broad functional types to facilitate the identification of shifts and trends.
+The output of the gradient forest model was further grouped by broad functional types to facilitate the identification of broader ecological shifts and trends.
 The phytoplankton genera that could be best predicted were mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (Figure \ref{fig:GF} $b$).
@@
 Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and the AMO index yielded a relatively low value around \qty{0.03}{}.
-This is expected given the noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the gradient forest algorithm \cite{pitcher_example_2012, roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}.
+This is expected given the noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the gradient forest algorithm \cite{roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}.
 It is important to consider that although predictive relationships between drivers and ecological variables are identified, this does not necessarily imply a mechanistic causal relationship.
@@
 \section{Conclusions}
-    We found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco Basin.
-Specifically, microscopy cell counts revealed the presence of two distinct clusters defined by high- and low-upwelling conditions that corresponded to anomalies in the AMO index.
-The community changes did not indicate a permanent regime shift, since a return to the early cluster was observed towards the end of the time series.
-We observed no significant changes in diversity metrics between the two clusters, except for lower turnover and higher community evenness under low-upwelling conditions.
-    Despite multi-annual trends in warming, reduced productivity, and drastic declines in micro-phytoplankton cell counts, the phytoplankton community returned to previous levels once the long-term climatic and local physical conditions were restored.
-    The strongest variable predicting changes in the phytoplankton community was the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local conditions through low-frequency natural cycles.
-We concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state for the phytoplankton community of the Cariaco Basin.
-    Notwithstanding a number of limitations, our analysis points to the incipient return of the Cariaco Basin to a pre-2004 regime, hinting at an oscillatory nature of phytoplankton communities under global change, an aspect that emphasizes the relevance of long-term time series and deserves further attention.
-    Future studies, ideally based on mathematical modeling, could focus on the relative impact of top-down vs. bottom-up drivers,  aspects remaining largely unexplored in the Cariaco Basin.
+    We found that decadal-scale climate variability, particularly linked to the AMO index, strongly influenced phytoplankton community composition in the tropical coastal ecosystem of the Cariaco Basin.
+Phytoplankton cell counts revealed two distinct community clusters associated with periods of high and low upwelling, correlating with negative and positive phases of the AMO index.
+While overall diversity remained similar between clusters, the low-upwelling period was characterized by reduced taxonomic turnover and increased evenness.
+These shifts did not reflect a permanent regime change.
+Rather, the community returned to a composition similar to the pre-2004 period in 2014, when physical conditions also reverted.
+This recovery occurred despite ongoing multi-year trends of surface warming, declining productivity, and steep reductions in microphytoplankton abundance, suggesting a degree of resilience in the phytoplankton community to climate-driven perturbations.
+
+    Our findings point to an oscillatory pattern in community structure, with transitions linked to low-frequency climate variability.
+This highlights the importance of long-term ecological time series for capturing non-linear ecosystem responses and underscores the potential of the AMO as a useful indicator of ecosystem state in this region, in line with the observations of \citeA{nye_ecosystem_2014}.
+Although our study is subject to some limitations, it provides evidence of a partial return to earlier ecosystem conditions and supports the notion that the phytoplankton community in the Cariaco Basin is shaped by recurring climate cycles.
+Future work, particularly using mathematical modeling, could help disentangle the relative impact of bottom-up versus top-down drivers, factors that remain largely unexplored in the Cariaco Basin.
 
@@
 This includes the Niskin bottle dataset for monthly in-situ measurements of chlorophyll \textit{a} and nutrient data \url{https://www.bco-dmo.org/dataset/3093}, the CTD dataset \url{https://www.bco-dmo.org/dataset/3092}, and the phytoplankton taxonomy dataset \url{https://www.bco-dmo.org/dataset/3095}.
-The climate variables, such as wind speed, precipitation and evaporation, were extracted from the "ERA5 monthly averaged data on single levels from 1940 to present" dataset \url{https://cds.climate.copernicus.eu/cdsapp#!/dataset/10.24381/cds.f17050d7}.
+The climate variables, such as wind speed, precipitation and evaporation, were extracted from the ``ERA5 monthly averaged data on single levels from 1940 to present'' dataset \url{https://cds.climate.copernicus.eu/cdsapp#!/dataset/10.24381/cds.f17050d7}.
 The AMO index was retrieved from the Climate Data Guide provided by the NSF National Center for Atmospheric Research at \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} and the MEI v.2 was accessed from NOAA Physical Sciences Laboratory (PSL) at \url{https://psl.noaa.gov/enso/mei/}.

### 2025-08-14 text-backup.tex
Final Text Revision

@@
+%OLDCONCLUSION
+We found a strong impact of decadal-scale climate variability on phytoplankton community shifts in the tropical coastal ecosystem of the Cariaco Basin.
+Specifically, phytoplankton cell counts revealed two distinct community clusters, corresponding to periods of high and low upwelling, which aligned with anomalies in the AMO index.
+We observed no significant differences in overall diversity between the two clusters, except for reduced taxonomic turnover and increased evenness during the low-upwelling phase.
+    These community changes did not constitute a permanent regime shift, as a return to the earlier cluster was observed toward the end of the time series (past 2014).
+Despite multi-annual trends in warming, reduced productivity, and drastic declines in micro-phytoplankton cell counts, the phytoplankton community returned to previous levels once the long-term climatic and local physical conditions were restored.
+    Notwithstanding a number of limitations, our analysis points to the incipient return of the Cariaco Basin to a pre-2004 regime, hinting at an oscillatory nature of phytoplankton communities under global change, an aspect that emphasizes the relevance of long-term time series and deserves further attention.
+
+    The strongest variable predicting changes in the phytoplankton community was the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local conditions through low-frequency natural cycles.
+We concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state, particularly for the phytoplankton community of the Cariaco Basin.
+    Future studies, ideally based on mathematical modeling, could focus on the relative impact of top-down vs. bottom-up drivers, aspects remaining largely unexplored in the Cariaco Basin.
+
 % TOP DOWN ASPECTS FROM INTRO:

### 2025-08-14 agujournaltemplate.tex
Final version submitted to JGR Biogeosciences

@@
 
-    All in situ data, including the Niskin bottle \cite{mullerkarger2019niskin}, CTD \cite{mullerkarger2019ctd}, and phytoplankton taxonomy \cite{troccoli2019phytoplankton}, were obtained from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) \url{https://www.bco-dmo.org/project/2047}.
+    All in situ data, including the Niskin bottle \cite{mullerkarger2019niskin}, CTD \cite{mullerkarger2019ctd}, and phytoplankton taxonomy \cite{troccoli2019phytoplankton}, were obtained from the Biological and Chemical Oceanography Data Management Office (BCO-DMO).
     From the CARIACO Time Series data, we used only those variables that cover the entire length of the time series without large gaps (no more than six months) in coverage.
@@
 Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}.
-We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels}.
+We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size).
 For simplicity, we refer to this variable as ``wind speed'' throughout the article.
@@
     In addition to the wind data, we examined two climatic indices, each of which provides information on the basin-scale climatic conditions and, consequently, the broad-scale forcing of marine ecosystems.
-The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies (\ang{0}N-\ang{60}N) \cite{trenberth2023amo}, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo}.
-
-    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}, were obtained from \url{https://psl.noaa.gov/enso/mei/}.
+The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies (\ang{0}N-\ang{60}N) \cite{trenberth2023amo}.
+
+    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}.
 In our analysis, the bi-monthly values were aligned with the tail month (December-January is treated as the value for January).
@@
     Data were processed and analyzed using the R programming language (version 4.3.3) \cite{r_core_team_r_2024} with data pipelines created using ``tidyverse'' packages \cite{wickham_welcome_2019}.
-All the codes we produced for data processing and data analysis are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
+All the codes we produced for data processing and data analysis are publicly available on GitHub \cite{benjamin_post_2025_16878343}.
     For all discrete depth measurements, the data were interpolated using the ``oce'' package \cite{kelley_oce_2023} with the adaptive ``unesco'' algorithm defined by the U.S.
@@
 We chose time lags that scored highest in the goodness-of-fit statistics for the final model run (Table \ref{tab:sup:GFlagtests} in Appendix).
-All the codes and data used in this analysis and used to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
+All the codes and data used in this analysis and used to create the plots are publicly available on GitHub at \url{https://github.com/ben1post/PhytoCariaco} \cite{benjamin_post_2025_16878343}.
 
@@
 The CARIACO time series data we used was accessed from the Biological and Chemical Oceanography Data Management Office (BCO-DMO) under the doi \url{10.1575/1912/bco-dmo.3095.1}.
-This includes the Niskin bottle dataset for monthly in-situ measurements of chlorophyll \textit{a} and nutrient data \url{https://www.bco-dmo.org/dataset/3093}, the CTD dataset \url{https://www.bco-dmo.org/dataset/3092}, and the phytoplankton taxonomy dataset \url{https://www.bco-dmo.org/dataset/3095}.
-The climate variables, such as wind speed, precipitation and evaporation, were extracted from the ``ERA5 monthly averaged data on single levels from 1940 to present'' dataset \url{https://cds.climate.copernicus.eu/cdsapp#!/dataset/10.24381/cds.f17050d7}.
+This includes the Niskin bottle dataset for monthly in-situ measurements of chlorophyll \textit{a} and nutrient data at \url{https://www.bco-dmo.org/dataset/3093}, the CTD dataset at \url{https://www.bco-dmo.org/dataset/3092}, and the phytoplankton taxonomy dataset at \url{https://www.bco-dmo.org/dataset/3095}.
+The climate variables, such as wind speed, precipitation and evaporation, were extracted from the ``ERA5 monthly averaged data on single levels from 1940 to present'' dataset at \url{https://cds.climate.copernicus.eu/cdsapp#!/dataset/10.24381/cds.f17050d7}.
 The AMO index was retrieved from the Climate Data Guide provided by the NSF National Center for Atmospheric Research at \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} and the MEI v.2 was accessed from NOAA Physical Sciences Laboratory (PSL) at \url{https://psl.noaa.gov/enso/mei/}.
-All the code and data used in this analysis and to create the plots are publicly available on GitHub \url{https://github.com/ben1post/PhytoCariaco}.
-%TODO add Zenodo link before submission
+All the code and data used in this analysis and to create the plots are publicly available on GitHub at \url{https://github.com/ben1post/PhytoCariaco} \cite{benjamin_post_2025_16878343}.
 

### 2025-12-15 text-backup.tex
Pre-Edits for Review Response

@@
+%OLDABSTRACT
+
+- The CARIACO phytoplankton community time series is characterized by two distinct temporal clusters separated by a shift in 2004 from high to low upwelling and by a return to high upwelling conditions in 2014.
+- No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low upwelling conditions.
+- Gradient forest analysis identifies the AMO climate index as the strongest predictor of community change, highlighting the influence of large-scale climatic oscillations on this tropical coastal ecosystem
+
+Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
+Long-term time series can capture this variability, allowing us to unravel the effects of environmental change on local communities.
+This study investigates changes in the phytoplankton community of the Cariaco Basin, off the coast of Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data.
+To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in local weather and upwelling in the Cariaco Basin.
+Cluster analysis identified two distinct community states separated by a shift from high to low upwelling conditions in 2004.
+In 2014, community and environmental conditions returned to a state similar to that before the shift in 2004.
+In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
+To quantify the effects of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2.
+Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a possible cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.
+
 %OLDCONCLUSION

### 2025-12-18 agujournaltemplate.tex
working on reviewer #1 response

@@
 \usepackage{siunitx}
+\sisetup{separate-uncertainty=true}
 \usepackage[section]{placeins}
@@
 \begin{keypoints}
-\item The CARIACO phytoplankton community time series is characterized by two distinct temporal clusters separated by a shift in 2004 from high to low upwelling and by a return to high upwelling conditions in 2014.
-\item No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low upwelling conditions.
-\item Gradient forest analysis identifies the AMO climate index as the strongest predictor of community change, highlighting the influence of large-scale climatic oscillations on this tropical coastal ecosystem.
+\item The CARIACO phytoplankton time series reveals two temporal clusters, linked to a 2004 shift from high to low upwelling and return in 2014.
+\item Diversity metrics show no significant difference except lower turnover and higher evenness during the cluster with low upwelling conditions.
+\item Gradient forest analysis finds AMO index as strongest predictor of community shifts, highlighting large-scale climate impacts on ecosystem.
 \end{keypoints}
@@
 Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
-Long-term time series can capture this variability, allowing us to unravel the effects of environmental change on local communities.
+Long-term time series capture this variability, allowing us to unravel the effects of environmental change on local communities.
 This study investigates changes in the phytoplankton community of the Cariaco Basin, off the coast of Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data.
-To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in local weather and upwelling in the Cariaco Basin.
+To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both linked to variations in local weather and upwelling in the Cariaco Basin.
 Cluster analysis identified two distinct community states separated by a shift from high to low upwelling conditions in 2004.
-In 2014, community and environmental conditions returned to a state similar to that before the shift in 2004.
-In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
-To quantify the effects of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2.
+In 2014, community and environmental conditions returned a state similar to that before the shift in 2004.
+In contrast, diversity declined between 1998 and 2009 independently of changes in the upwelling regime.
+To quantify effects of climate and bottom-up drivers on phytoplankton community changes, we employed gradient forest analysis and found the AMO index was the strongest predictor, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2.
 Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a possible cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.
@@
 The Cariaco Basin (Figure~\ref{fig:map}) is a large (approximately 160 km long and 70 km wide) and deep (approximately 1,400 m) basin situated on the northeastern Venezuelan continental shelf, with a sill to the north running almost parallel to the coastline at a mean depth of around \qty{100}{m}.
-This area is relatively productive ($\sim$ \qty{400}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the Bermuda Atlantic Time-series Study (BATS) and Hawaii Ocean Time-series (HOT) sites, resulting in the formation of anoxic conditions below a depth of  \qty{\sim 275}{m}.
+This area is relatively productive (mean integrated primary production across the entire time series of \qty{485 \pm 345}{g.C.m^{-2}.y^{-1}}), with primary production and vertical particulate organic matter fluxes exceeding those observed at the Bermuda Atlantic Time-series Study (BATS) and Hawaii Ocean Time-series (HOT) sites.
+This high productivity, combined with reduced ventilation across the sill, resulted in the formation of anoxic conditions below a depth of \qty{\sim 275}{m}.
 
@@
 Petersburg, USA).
-Samples from eight cruises were analyzed by both laboratories \cite{taylor_ecosystem_2012}.
-We used a merged dataset for all nutrients, in which coincidental monthly measurements are averaged.
+There is an overlap of eight nutrient samples analyzed by both laboratories \cite{taylor_ecosystem_2012}.
+We used a merged dataset for all nutrients, in which these eight coincidental monthly measurements are averaged.
 
@@
 Zonal wind (\textit{u}, easterly Trade Winds) across the southern Caribbean Sea drives nearshore upwelling of nutrient-rich waters \cite{rueda-roa_southern_2013}.
-We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the Cariaco time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size).
+We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the CARIACO time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size).
 For simplicity, we refer to this variable as ``wind speed'' throughout the article.
@@
 
-    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; sea surface temperature, SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}.
+    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}.
 In our analysis, the bi-monthly values were aligned with the tail month (December-January is treated as the value for January).
@@
     To provide an overview of the available data and highlight interannual trends and changes in the time series data, we calculated annual anomalies for relevant variables as normalized z scores, $z = \frac{(x – \mu)}{\sigma}$, where $x$ is the yearly mean, $\mu$ is the mean and $\sigma$ is the standard deviation of all yearly means.
-Data from 1995, 2016, and 2017 were not included because insufficient coverage (i.e.
+For the analysis of annual z-scores, data from 1995, 2016, and 2017 were not included because insufficient coverage (i.e.
 more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.
@@
     \subsubsection{Analysis of Phytoplankton Community Structure}
-    To examine temporal changes in the phytoplankton community structure in the Cariaco Basin, we applied complete linkage hierarchical clustering using the ``hclust'' function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix This analysis was based on presence-absence data of observed genera, over the upper 100 meters of the water column and aggregated annually.
+    To examine temporal changes in the phytoplankton community structure in the Cariaco Basin, we applied complete linkage hierarchical clustering using the ``hclust'' function in R \cite{r_core_team_r_2024} on the binary Jaccard distance matrix.
+This analysis was based on presence-absence data of observed genera, over the upper 100 meters of the water column and aggregated annually.
 By using binary data and temporal aggregation, we aimed to reduce the influence of extreme values in the cell counts and focus on broad compositional shifts in the community.
@@
 Silicate anomalies, in contrast, did not follow the same pattern as nitrate and phosphate.
-Instead, elevated concentrations were observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
+Instead, annual anomalies show an inverse relationship with precipitation and salinity, indicating effects of surface runoff.
+Elevated concentrations were observed between 2008 and 2013, potentially driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced consumption due to lower diatom abundance.
     Chlorophyll \textit{a} and functional group cell counts show positive anomalies throughout most of the first half of the time series, followed by a significant reduction after 2004, suggesting a shift in the community.
-This shift persists during the period between 2003 and 2013, when the AMO index presents mostly positive anomalies, while upwelling-related variables indicate a weak upwelling regime (low wind, high SST, deep isotherm).
+    This biological shift was preceded by a change in the AMO index anomalies that occurred after 2002, when the index transitioned to a period of mostly positive anomalies that persisted until 2013.
+This period was also characterized by upwelling-related variables indicating a weak upwelling regime (low wind, high SST, deep isotherm).
+Thus, while the AMO and physical parameters shifted after 2002, the corresponding change in chlorophyll a and cell counts occurred later, after 2004.
     Although data are sparse, nutrients and biological variables show a strong increase from 2014, coinciding with a return to negative anomalies in the AMO index and to more upwelling-favorable conditions.
     The annual means of diversity metrics do not follow this trend, but are reduced between 1998 and 2008.
-Shannon diversity specifically appears elevated past 2006 and the general pattern that emerges is that mean diversity is lower in years when mean chlorophyll \textit{a} is high.
+Shannon diversity z-scores are elevated between 2007 and 2013, when chlorophyll \textit{a} and cell counts were diminished.
 However, in 1996 and 1997 positive anomalies in cell counts, chlorophyll \textit{a} and diversity metrics coincide.
@@
 The yearly mean chlorophyll \textit{a} biomass showed the highest variability within the upper 25 meters.
-In some years, particularly in 2000, 2005, 2007, 2008, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deeper chlorophyll maximum in those years (Figure \ref{fig:divts} $a$).
+In some years, particularly in 1999, 2005, 2008, 2009, and 2013, the concentration in this layer was lower than in the 25–50 meter depth range, suggesting a deeper chlorophyll maximum in those years (Figure \ref{fig:divts} $a$).
 The year 2013 was an outlier, particularly for the upper 25 meters, as missing monthly sampling from the upwelling season (January and February) skewed the value downwards.
@@
 Chlorophyll \textit{a} biomass below 50 meters depth showed a relatively stable distribution throughout the year and across the time series.
-In fall, during the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters was markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
-
-    The calculated genus diversity from the microscopy cell counts generally followed the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons in winter.
+During the rainy season, mean chlorophyll \textit{a} biomass in the upper 25 meters was markedly reduced, while the 50–75 meter depth interval showed a slight increase, indicating a shift in the chlorophyll \textit{a} maximum to lower depths under weak upwelling conditions.
+
+    The calculated genus diversity from the microscopy cell counts generally followed the dynamics of chlorophyll \textit{a} biomass, with maximum values observed during the upwelling seasons.
 Genus richness was consistently highest in the top \qty{25}{meters} depth interval, also showing the highest variability over the time series.
@@
 Diversity remained high throughout 2000 and recovered more quickly but showed marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009, and 2016.
-The seasonal signal was lost (see Figure \ref{fig:divts} $f$).
+The seasonal signal that is apparent for chlorophyll \textit{a} and genus richness is not visible in the Shannon index (see Figure \ref{fig:divts} $f$).
 
     Community evenness, expressed by the Pielou index, shows no seasonal signal and a very large variance within seasons and through depth (see Figure \ref{fig:divts} $h$).
-The inter-annual dynamics of evenness are remarkably stable across all depth intervals, clustered around an intermediate 0.5 value.
-There were marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009, and 2016.
+The inter-annual dynamics of evenness were remarkably stable across all depth intervals, clustered around an intermediate 0.5 value.
+There were marked reductions in all depths towards 2000 and for all depths above 75 meters in 2007, 2009, and 2015.
 Throughout 2003 to 2013 evenness was almost consistently highest in the deepest interval below 75 meters, where the community appeared to have been more evenly distributed.
@@
     Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range.
-The black, dashed lines indicate cumulative importance across the entire community, and the solid, dashed lines indicate importance per phytoplankton group.
-For the AMO index, we see a clear pattern in which negative anomalies affect the community.
+The black lines indicate cumulative importance across the entire community, and the colored lines indicate importance per phytoplankton group.
+    Cumulative importance curves from gradient forest analysis show how the importance of a predictor accumulates across its range of values.
+The slope of the curve at any point indicates where the strongest compositional turnover occurs—steep sections represent threshold values where small changes in the predictor correspond to large changes in community composition, while flat sections indicate regions where the predictor has little influence on community structure.
+    For the AMO index, we see a clear pattern in which negative anomalies affect the community (slope is highest for negative values).
 This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes and cyanobacteria.
@@
     These changes are clearly visible in the yearly mean anomalies and the density distribution of monthly measurements.
-After 2004, we observed increased sea surface temperatures (SST), decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident with a shift in the Atlantic Multidecadal Oscillation (AMO) toward positive mean anomalies until 2014.
+After 2004, we observed increased SST, decreased wind speeds, reduced fluorometric chlorophyll \textit{a}, and drastically reduced cell counts of functional groups, all coincident with a shift in the AMO index toward positive mean anomalies until 2014.
     By considering the phytoplankton community time series up to the end of 2017, the temporally complete dataset, we found that the observed shift that occurred in 2004 produced a temporary regime that lasted only until 2014 (cluster 2), when the system reverted to the precedent regime of 1996-2004 (cluster 1), confirming the observations by \citeA{muller-karger_scientific_2019}.
@@
     Consistent with our results, independent spectrophotometric measurements of chlorophyll \textit{a} concentrations at \qty{20}{meters} depth, off the coast of Margarita Island, showed a reduction in chlorophyll concentration between 2004 and 2014 \cite{gomez_gaspar_variacion_2025}.
-    Additionally, despite decreased cell counts of all functional groupsing, the HPLC data indicated an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{\meter}.
+    Additionally, despite decreased cell counts of all functional grouping, the HPLC data indicated an increase in non-diatoms and a redistribution of biomass to depths below \qty{55}{\meter}.
 This provides a further strong indication that the phytoplankton community shifted to smaller cell sizes, which were not detected using light microscopy.
@@
     The temporal dynamics found in yearly mean genus richness, however, did not correspond directly to the observed clusters.
-Instead, there was a marked reduction from 1998 to 2010, after which richness recovered four years prior to 2014, when the community returned to levels characteristic of cluster 1.
+Instead, there was a marked reduction from 1998 to 2009, after which richness recovered four years prior to 2014, when the community returned to levels characteristic of cluster 1.
 The same trend was evident, albeit to a lesser degree, in the other diversity metrics.

### 2025-12-23 agujournaltemplate.tex
Start of Reviewer #2 responding

@@
 In 2014, community and environmental conditions returned a state similar to that before the shift in 2004.
-In contrast, diversity declined between 1998 and 2009 independently of changes in the upwelling regime.
-To quantify effects of climate and bottom-up drivers on phytoplankton community changes, we employed gradient forest analysis and found the AMO index was the strongest predictor, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2.
+No significant differences in diversity were observed between the two periods, however, the low-upwelling period was characterized by higher community evenness and lower species turnover.
+To quantify effects of climate and bottom-up drivers on phytoplankton community changes, we employed gradient forest analysis and found the AMO index was the strongest predictor of changes in the community, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2.
 Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a possible cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.
-
+% Reviewer two criticizes mention of cyclic behaviour, since missing data might skew results. Need to check!
 \end{abstract}
@@
 % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
-The shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensities from 2004 to 2013 did not appear to be a permanent change in regime.
-Data from 2014 to 2017, the final years of the CARIACO (CArbon Retention In A Colored Ocean) time series, indicated a return to strong upwelling intensities and high numbers of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
-The impact of large-scale climatic processes on the observed shifts in the phytoplankton community remains unclear.
-The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing the climate system in the region and the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}.
-The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean Sea \cite{stephenson_changes_2014}.
+The shift in phytoplankton community composition associated with increased temperature and reduced upwelling intensity from 2004 to 2013 did not appear to represent a permanent regime change.
+Hydrographic data from 2014 to 2017, the final years of the CARIACO (CArbon Retention In A Colored Ocean) time series, indicated a return to strong upwelling conditions, while aggregated microscopy cell counts from 2014 and 2015 showed increased abundances of diatoms and dinoflagellates \cite{muller-karger_scientific_2019}.
+Although the impact of large-scale climatic processes on the observed shifts in phytoplankton community composition remains unclear, we hypothesize that these changes were driven by decadal-scale changes in climatic factors.
+The Atlantic Multidecadal Oscillation (AMO) has emerged as an important factor influencing both the regional climate system and north Atlantic ecosystems \cite{nye_ecosystem_2014}.
+The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean Sea, where a negative phase is generally correlated with warmer and drier conditions and a positive phase with colder and wetter conditions \cite{stephenson_changes_2014}.
+A negative AMO phase has also been linked to a southward shift in the ITCZ \cite{knight_climate_2006}.
+Such a southward shift of the ITCZ is correlated with increased wind-driven upwelling in the Cariaco Basin \cite{taylor_ecosystem_2012}.
 The El Niño Southern Oscillation (ENSO) is teleconnected to the Caribbean Sea, given that during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures increase and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.
@@
 % What drives the changes??? Bottom-up/ Top-down ... CLIMATE 
-\citeA{taylor_ecosystem_2012} noted that the observed changes could not be clearly identified as ``unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation''.
+\citeA{taylor_ecosystem_2012} noted that the observed trends of increasing SST, reduced upwelling intensity, and reduced chlorophyll \textit{a} between 1996 and 2010 could not be clearly identified as ``unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation''.
 To our knowledge, all analyses of phytoplankton community diversity in the Cariaco Basin do not include time series data beyond 2012, and no detailed investigations have yet been conducted on changes in diversity based on microscopy cell counts.
@@
 We extracted the variable u10, the \qty{10}{\meter} eastward component of wind speed from the monthly averaged ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis \cite{hersbach2023era5} in the grid cell closest to the CARIACO time series sampling location (\ang{10.4}N, \ang{64.8}W, 31 km grid size).
-For simplicity, we refer to this variable as ``wind speed'' throughout the article.
+To signify the easterly wind, we used the negative of the u10 value in our analysis and, for simplicity, we refer to this variable as ``wind speed'' throughout the article.
 The data for precipitation and evaporation are also extracted from the ERA5 dataset for the aforementioned grid cell.
 Evaporation provides an estimate of the accumulated amount of water that has evaporated from Earth's surface, while precipitation is an estimate of total precipitation, comprising all water that falls on Earth's surface as rain or snow.
+These variables were all available at monthly resolution across the entire time series.
+We used these variables, in addition to the climate indices, in the analysis of annual z-scores, in the NMDS clustering analysis to explain variance in the monthly community data, and as predictors of community change in the gradient forest analysis.
 
     In addition to the wind data, we examined two climatic indices, each of which provides information on the basin-scale climatic conditions and, consequently, the broad-scale forcing of marine ecosystems.
-The monthly Atlantic Multidecadal Oscillation (AMO) index data (10-year low-pass), which represent area-averaged, detrended low-pass filtered North Atlantic SST anomalies (\ang{0}N-\ang{60}N) \cite{trenberth2023amo}.
-
-    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure, SLP; SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}.
-In our analysis, the bi-monthly values were aligned with the tail month (December-January is treated as the value for January).
+The monthly Atlantic Multidecadal Oscillation (AMO) index (10-year low-pass) is calculated using area-averaged, detrended low-pass filtered North Atlantic SST anomalies (\ang{0}N-\ang{60}N) \cite{trenberth2023amo}.
+    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) index (MEI v.2) is defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five variables (sea level pressure, SLP; SST; zonal and meridional components of surface wind; and outgoing long-wave radiation, OLR) over the tropical Pacific basin (\ang{30}S-\ang{30}N and \ang{100}E-\ang{70}W) \cite{noaa_psl_meiv2}.
+In our analysis, the bi-monthly values were aligned with the tail month (i.e., December–January was assigned to January).
 
@@
 National Oceanographic Data Center \cite{johnson2006world}.
+Depth discrete data was generally available at depths of 1, 7, 15, 25, 35, 55, 75, and 100 m.
 The interpolated values were depth integrated to \qty{100}{\meter} and in four discrete intervals between 0-25, 25-50, 50-75 and 75-100 meters for further analysis.
@@
 more than six missing monthly samples) in these years prevented us from producing reliable yearly estimates.
+    %TALK MORE ABOUT MISSING DATA
 
@@
     Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range.
-The black lines indicate cumulative importance across the entire community, and the colored lines indicate importance per phytoplankton group.
+The dashed black lines indicate cumulative importance across the entire community, and the solid colored lines indicate importance per phytoplankton group.
     Cumulative importance curves from gradient forest analysis show how the importance of a predictor accumulates across its range of values.
@@
 (c) Cumulative importance curves for the predictor variables.
-Black line shows cumulative importance across predictor range for the entire model and all genera.
-Dashed colored lines represent the responses of individual functional groups.}
+The dashed black lines show cumulative importance across predictor range for the entire model and all genera.
+Solid colored lines represent the responses of individual functional groups.}
     \label{fig:GF}
@@
 This increase in evenness, quantified by the Pielou index, is most likely explained by the lack of strong upwelling events and the resulting reduced disturbance of the community.
-The resulting decrease in compositional turnover was also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$).
+Under strong upwelling conditions, r-selected species are typically favored, leading to dominance by fewer taxa and consequently lower evenness.
+A decrease in compositional turnover was also observed in a decline in interannual species turnover during cluster 2 (see Figure \ref{fig:clustering} $b$).
 As the productivity and biomass of phytoplankton communities are driven primarily by nutrients supplied through disturbance events, this is in line with the inverse relationship between evenness and community performance observed in previous studies \cite{lehtinen_phytoplankton_2017, otero_phytoplankton_2020}.
@@
     The threshold responses indicate a shift in the abundances of dinoflagellates for negative MEI v.2 anomalies, which represent La Niña conditions.
-Findings from Bayesian modeling showed that dinoflagellates responding positively to warmer, stratified conditions in the Cariaco Basin \cite{mutshinda_environmental_2013}.
+Findings from Bayesian modeling showed that dinoflagellates responded positively to warmer, stratified conditions in the Cariaco Basin \cite{mutshinda_environmental_2013}.
 For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events.
@@
 This could be explained by previous findings, that strong El Niño events (as indicated by a positive MEI) increase the strength of currents in the Caribbean sea through modulating the easterly wind anomalies, thereby potentially increasing upwelling strength in the Cariaco Basin \cite{huang_enso_2023}.
+    In particular, this intensification of the Caribbean Current may enhance the secondary upwelling that typically occurs between June and August \cite{rueda-roa_description_2018}, rather than affecting the primary wind-driven upwelling between December and May.
+Additionally, enhanced eddy activity near the Cariaco Basin during strong El Niño events can stimulate productivity and alter community composition through ventilation of sub-oxic waters below the sill \cite{astor_seasonal_2003}.
 
@@
 Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, showed greater variance along NMDS 1.
+The fact that salinity and the MEI v.2 align is consistent with the influence of ENSO on regional hydrology, as El Niño events typically result in reduced precipitation and drier conditions in the southern Caribbean, whereas La Niña years are associated with higher precipitation, leading to elevated silicate concentrations and lower surface salinity \cite{giannini_interdecadal_2001}.
 Salinity, evaporation, and silicate concentration exhibited some explanatory power in the gradient forest analysis, although they were weaker predictors than SST, nitrate concentration, and climate indices (see Figure \ref{fig:GF}).
@@
 Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and the AMO index yielded a relatively low value around \qty{0.03}{}.
-This is expected given the noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the gradient forest algorithm \cite{roland_pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}.
+This is expected given the noisy nature of phytoplankton microscopy cell counts and is consistent with values reported in other studies utilizing the gradient forest algorithm \cite{pitcher_exploring_2012, roubeix_identification_2016, samhouri_defining_2017, fraker_temporal_2022}.
 It is important to consider that although predictive relationships between drivers and ecological variables are identified, this does not necessarily imply a mechanistic causal relationship.
@@
 
-    Our focus on bottom-up drivers of community change may have also restricted the explanatory power of our analysis, as top-down processes such as zooplankton grazing may have impacted the structure of the phytoplankton community \cite{frank_ups_2007, banas_adding_2011, acevedo-trejos_mechanisms_2015}.
-Due to the scarcity of data covering higher trophic levels and the absence of zooplankton data for the first six years of the time series, we suggest that mechanistic modeling is the ideal approach for investigating the relative impacts of top-down and bottom-up drivers.
+    Our focus on bottom-up drivers of community change may have restricted the explanatory power of our analysis, as top-down processes such as zooplankton grazing may have impacted the structure of the phytoplankton community \cite{frank_ups_2007, banas_adding_2011, acevedo-trejos_mechanisms_2015}.
+For instance, a trophic cascade resulting from the collapse of sardine fisheries and a subsequent increase in mesozooplankton abundance was proposed to have affected the phytoplankton community in the Cariaco Basin \cite{rueda-roa_spatial_2017, muller-karger_scientific_2019}.
+However, this predictor could not be included in our gradient forest analysis due to a lack of data at monthly resolution.
+Given the scarcity of data covering higher trophic levels and the absence of zooplankton data for the first six years of the time series, we suggest that mechanistic modeling would be the ideal approach for investigating the relative impacts of top-down and bottom-up drivers.
 
@@
 \noindent\includegraphics[width=\textwidth]{fig/FigureA3_GF_output_NOLAG_v1.pdf}
-\caption{Gradient forest analysis of monthly community data at a genus level.
+\caption{Gradient forest analysis of monthly community data at a genus level for all predictors without any time lags.
 (a) Predictor variables ranked according to overall weighted importance in predicting shifts in abundances.
@@
 (c) Cumulative importance curves for the predictor variables.
-Black line shows cumulative importance across predictor range for the entire model and all genera.
-Dashed colored lines represent the responses of individual functional groups.}
+The dashed black lines show cumulative importance across predictor range for the entire model and all genera.
+Solid colored lines represent the responses of individual functional groups.}
 \label{fig:sup:GFoutput_nolags}
@@
 \noindent\includegraphics[width=\textwidth]{fig/FigureA4_GF_output_Supplement_NOLAG_v1.pdf}
-\caption{Cumulative importance curves for all variables not shown in the previous figure.
-Dashed black line represents the cumulative importance across the entire community, colored lines represent functional groups.
+\caption{Cumulative importance curves for all variables not shown in the previous figure, showing the gradient forest analysis performed with all predictors without any time lags.
+Dashed black line represents the cumulative importance across the entire community.
+Solid colored lines represent functional groups.
 Gradient forest runs where done on Genus-level cell counts, integrated over the top 100 meters.}
